\section{Rate-Complexity Bounds}\label{sec:bounds}
%==============================================================================

We now prove the rate-complexity bounds that make DOF = 1 optimal. The key result: the gap between DOF-1-complete and DOF-1-incomplete architectures is \emph{unbounded}--it grows without limit as encoding systems scale (\LH{BND4}).

\subsection{Cost Model}\label{sec:cost-model}

\begin{definition}[Modification Cost Model]\label{def:cost-model}
Let $\delta_F$ be a modification to fact $F$ in encoding system $C$. The \emph{effective modification complexity} $M_{\text{effective}}(C, \delta_F)$ is the number of syntactically distinct edit operations that must be performed manually. Formally:
\[
M_{\text{effective}}(C, \delta_F) = |\{L \in \text{Locations}(C) : \text{requires\_manual\_edit}(L, \delta_F)\}|
\]
where $\text{requires\_manual\_edit}(L, \delta_F)$ holds iff location $L$ must be updated manually (not by automatic derivation) to maintain coherence after $\delta_F$.
\end{definition}

\textbf{Unit of cost:} One edit = one syntactic modification to one location. We count locations, not keystrokes or characters. This abstracts over edit complexity to focus on the scaling behavior.

\textbf{What we measure:} Manual edits only. Derived locations that update automatically have zero cost. This distinguishes DOF = 1 systems (where derivation handles propagation) from DOF $>$ 1 systems (where all updates are manual).

\textbf{Asymptotic parameter:} We measure scaling in the number of encoding locations for fact $F$. Let $n = |\{L \in C : \text{encodes}(L, F)\}|$ and $k = \text{DOF}(C, F)$. Bounds of $O(1)$ and $\Omega(n)$ are in this parameter; in particular, the lower bound uses $n = k$ independent locations.

\subsection{Upper Bound: DOF = 1 Achieves O(1)}\label{sec:upper-bound}

\begin{theorem}[DOF = 1 Upper Bound]\label{thm:upper-bound}
For an encoding system with DOF = 1 for fact $F$:
\[
M_{\text{effective}}(C, \delta_F) = O(1)
\]
Effective modification complexity is constant regardless of system size.
\end{theorem}
\leanmeta{\LH{BND2}}

\begin{proof}
Let $\text{DOF}(C, F) = 1$. By Definition~\ref{def:ssot}, $C$ has exactly one independent encoding location. Let $L_s$ be this single independent location.

When $F$ changes:
\begin{enumerate}
\tightlist
\item Update $L_s$ (1 manual edit)
\item All derived locations $L_1, \ldots, L_k$ are automatically updated by the derivation mechanism
\item Total manual edits: 1
\end{enumerate}

The number of derived locations $k$ can grow with system size, but the number of \emph{manual} edits remains 1. Therefore, $M_{\text{effective}}(C, \delta_F) = O(1)$.
\end{proof}

\textbf{Note on ``effective'' vs. ``total'' complexity:} Total modification complexity $M(C, \delta_F)$ counts all locations that change. Effective modification complexity counts only manual edits. With DOF = 1, total complexity can be $O(n)$ (many derived locations change), but effective complexity is $O(1)$ (one manual edit).

\subsection{Lower Bound: DOF $>$ 1 Requires \texorpdfstring{$\Omega(n)$}{Omega(n)}}\label{sec:lower-bound}

\begin{theorem}[DOF $>$ 1 Lower Bound]\label{thm:lower-bound}
For an encoding system with DOF $>$ 1 for fact $F$, if $F$ is encoded at $n$ independent locations:
\[
M_{\text{effective}}(C, \delta_F) = \Omega(n)
\]
\end{theorem}
\leanmeta{\LH{BND2}}

\begin{proof}
Let $\text{DOF}(C, F) = n$ where $n > 1$.

By Definition~\ref{def:independent}, the $n$ encoding locations are independent--updating one does not automatically update the others. When $F$ changes:
\begin{enumerate}
\tightlist
\item Each of the $n$ independent locations must be updated manually
\item No automatic propagation exists between independent locations
\item Total manual edits: $n$
\end{enumerate}

Therefore, $M_{\text{effective}}(C, \delta_F) = \Omega(n)$.
\end{proof}

\subsection{Information-theoretic converse (mutual information / Fano)}\label{sec:info-converse}

To make the converse argument explicit in information-theoretic terms we give a concise mutual-information / Fano-style bound showing that insufficient side information forces a non-vanishing error probability when attempting to recover structural facts from distributed encodings (\LH{CIA3}, \LH{COH1}).

\begin{remark}[Formalization status]
The results in this subsection apply Fano's inequality, a standard tool in information theory~\cite{cover2006elements}. We state and apply these results without separate Lean formalization; the proofs are direct applications of the classical inequality. The core capacity and complexity theorems (Sections~\ref{sec:capacity}--\ref{sec:requirements}) are fully machine-checked.
\end{remark}

\begin{theorem}[Fano-style Converse]\label{thm:fano-converse}
Let $F$ be a discrete fact taking values in a set of size $K$. Let $S$ denote the available side information (derived locations, observations). Let $\hat F(S)$ be any estimator of $F$ based on $S$, and let $P_e = \Pr(\hat F \neq F)$. Then
\[
I(F;S) \ge H(F) - H_b(P_e) - P_e\log(K-1),
\]
where $H_b(\cdot)$ is the binary entropy. In particular, if $P_e\to 0$ then $I(F;S)\to H(F)$, so $S$ must contain essentially $\log K$ bits about $F$.
\end{theorem}

\begin{proof}[Proof sketch]
This is the standard Fano inequality. Let $\hat F$ be any estimator formed from $S$. Then
\[
\begin{aligned}
H(F|S)
&\le H(F|\hat F) + I(F;\hat F|S) \\
&= H(F|\hat F) \\
&\le H_b(P_e) + P_e\log(K-1),
\end{aligned}
\]
and $I(F;S)=H(F)-H(F|S)$ yields the stated bound. The terms $H_b(P_e)$ and $P_e\log(K-1)$ vanish as $P_e\to0$, forcing $I(F;S)\approx H(F)$.
\end{proof}

\noindent This inequality formalizes the intuition used in the main text: unless the derived/side-information channels collectively convey essentially the full entropy of the fact, perfect (or vanishing-error) recovery is impossible. Translating this to modification complexity shows that when side-information capacity is limited, systems must retain multiple independent encodings (higher DOF), incurring larger manual update costs.

\subsection{Information-constrained DOF lower bound}
\label{sec:info-dof}

The following lemma makes the link explicit between mutual-information capacity of side information and the necessity of multiple independent encodings.

\begin{lemma}[Information-constrained DOF Lower Bound]\label{lem:info-dof}
Let $F$ be a fact uniformly distributed over $K$ values. Let $S$ be the collective side information provided by derived locations. If
\[I(F;S) < \log K - \delta\]
for some $\delta>0$, then any encoding system that guarantees vanishing-error recovery of $F$ from $S$ must include at least two independent encoding locations for some subset of values (i.e., DOF $>1$ for those values). Consequently, for facts encoded at $n$ independent locations the effective modification complexity satisfies $M_{\text{effective}}=\Omega(n)$.
\end{lemma}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:fano-converse}, if $I(F;S)<\log K-\delta$ then any estimator has error probability bounded away from zero for the $K$-ary hypothesis. To achieve correctness despite the insufficient side information, the system must retain additional independent encodings that act as auxiliary sources of information about $F$; these independent encodings increase DOF. Once DOF$>1$ over a set of $n$ independent locations, Theorem~\ref{thm:lower-bound} implies $M_{\text{effective}}=\Omega(n)$, completing the chain from information constraints to modification complexity.
\end{proof}

\subsection{Worked example: numeric mutual-information calculation}
\label{sec:worked-mi}

Consider a fact $F$ uniformly distributed over $K=4$ values, so $H(F)=\log_2 4 = 2$ bits. Suppose the collective derived side information $S$ conveys only $I(F;S)=1$ bit about $F$ (for example, a summary or hash with one bit of mutual information).

Apply the Fano-style bound (Theorem~\ref{thm:fano-converse}). Let $P_e$ denote the probability of incorrect recovery of $F$ from $S$. Then
\[
H(F|S) = H(F)-I(F;S) = 2 - 1 = 1\text{ bit.}
\]
Fano's inequality gives
\[
H_b(P_e) + P_e\log_2(K-1) \ge H(F|S) = 1.
\]
Plugging $K-1=3$ and solving numerically yields a lower bound $P_e \gtrsim 0.19$ (approximately 19\% error). Concretely: $H_b(0.19)\approx 0.695$ and $0.19\log_2 3\approx 0.301$, summing to $\approx 0.996\approx1$.
\[
H_b(0.19)\approx 0.695,\qquad
0.19\log_2 3\approx 0.301,\qquad
0.996\approx 1.
\]

Interpretation: with only 1 bit of side information for a 2-bit fact, no estimator can recover $F$ with error lower than roughly 19\%; therefore, to guarantee vanishing error one must supply more side information (increase $I(F;S)$) or retain extra independent encodings (increase DOF), which in turn raises manual modification costs per Theorem~\ref{thm:lower-bound}.

As a second illustration, if $I(F;S)=1.5$ bits then $H(F|S)=0.5$ and Fano's bound requires a much smaller $P_e$ (solve $H_b(P_e)+P_e\log_2 3\ge 0.5$ numerically to obtain $P_e\approx 0.08$), showing the smooth tradeoff between conveyed information and achievable error.



\textbf{Tightness (Achievability + Converse).} Theorems~\ref{thm:upper-bound} and~\ref{thm:lower-bound} form a tight information-theoretic bound: DOF = 1 achieves constant modification cost (achievability), while any encoding with more than one independent location incurs linear cost in the number of independent encodings (converse) (\LH{BND2}, \LH{BND2}). There is no intermediate regime with sublinear manual edits when $k > 1$ independent encodings are permitted.

\subsection{The Unbounded Gap}\label{sec:gap}

\begin{theorem}[Unbounded Gap]\label{thm:unbounded-gap}
The ratio of modification complexity between DOF-1-incomplete and DOF-1-complete architectures grows without bound:
\[
\lim_{n \to \infty} \frac{M_{\text{DOF}>1}(n)}{M_{\text{DOF}=1}} = \lim_{n \to \infty} \frac{n}{1} = \infty
\]
\end{theorem}
\leanmeta{\LH{BND4}}

\begin{proof}
By Theorem~\ref{thm:upper-bound}, $M_{\text{DOF}=1} = O(1)$. Specifically, $M_{\text{DOF}=1} = 1$ for any system size.

By Theorem~\ref{thm:lower-bound}, $M_{\text{DOF}>1}(n) = \Omega(n)$ where $n$ is the number of independent encoding locations.

The ratio is:
\[
\frac{M_{\text{DOF}>1}(n)}{M_{\text{DOF}=1}} = \frac{n}{1} = n
\]

As $n \to \infty$, the ratio $\to \infty$. The gap is unbounded.
\end{proof}

\begin{corollary}[Arbitrary Reduction Factor]\label{cor:arbitrary-reduction}
For any constant $k$, there exists a system size $n$ such that DOF = 1 provides at least $k\times$ reduction in modification complexity.
\end{corollary}
\leanmeta{\LH{BND4}}

\begin{proof}
Choose $n = k$. Then $M_{\text{DOF}>1}(n) = n = k$ and $M_{\text{DOF}=1} = 1$. The reduction factor is $k/1 = k$.
\end{proof}

\subsection{The (R, C, P) Tradeoff Space}\label{sec:rcp-tradeoff}

We now formalize the complete tradeoff space, analogous to rate-distortion theory in classical information theory (\LH{REG4}, \LH{REG4}, \LH{REG4}).

\begin{definition}[(R, C, P) Tradeoff]\label{def:rcp-tradeoff}
For an encoding system, define:
\begin{itemize}
\tightlist
\item $R$ = \emph{Rate} (DOF): Number of independent encoding locations
\item $C$ = \emph{Complexity}: Manual modification cost per change
\item $P$ = \emph{Coherence indicator}: $P = 1$ iff no incoherent state is reachable; otherwise $P = 0$
\end{itemize}
The \emph{(R, C, P) tradeoff space} is the set of achievable $(R, C, P)$ tuples.
\end{definition}

\begin{theorem}[Operating Regimes]\label{thm:operating-regimes}
The (R, C, P) space has three distinct operating regimes:
\begin{center}
\begin{tabularx}{\linewidth}{cccY}
\toprule
\textbf{Rate} & \textbf{Complexity} & \textbf{Coherence} & \textbf{Interpretation} \\
\midrule
$R = 0$ & $C = 0$ & $P = $ undefined & Fact not encoded \\
$R = 1$ & $C = O(1)$ & $P = 1$ & \textbf{Optimal (capacity-achieving)} \\
$R > 1$ & $C = \Omega(R)$ & $P = 0$ & Above capacity \\
\bottomrule
\end{tabularx}
\end{center}
\end{theorem}
\leanmeta{\LH{REG4}}

\begin{proof}
\textbf{$R = 0$:} No encoding exists. Complexity is zero (nothing to modify), but coherence is undefined (nothing to be coherent about).

\textbf{$R = 1$:} By Theorem~\ref{thm:upper-bound}, $C = O(1)$. By Theorem~\ref{thm:coherence-capacity}, $P = 1$ (coherence guaranteed). This is the capacity-achieving regime.

\textbf{$R > 1$:} By Theorem~\ref{thm:lower-bound}, $C = \Omega(R)$. By Theorem~\ref{thm:dof-gt-one-incoherence}, incoherent states are reachable, so $P = 0$.
\end{proof}

\begin{definition}[Pareto Frontier]\label{def:pareto-frontier}
A point $(R, C, P)$ is \emph{Pareto optimal} if no other achievable point dominates it (lower $R$, lower $C$, or higher $P$ without worsening another dimension).

The \emph{Pareto frontier} is the set of all Pareto optimal points.
\end{definition}

\begin{theorem}[Pareto Optimality of DOF = 1]\label{thm:pareto-optimal}
$(R=1, C=1, P=1)$ is the unique Pareto optimal point for encoding systems requiring coherence ($P = 1$).
\end{theorem}

\begin{proof}
We show $(1, 1, 1)$ is Pareto optimal and unique:

\textbf{Existence:} By Theorems~\ref{thm:upper-bound} and \ref{thm:coherence-capacity}, the point $(1, 1, 1)$ is achievable.

\textbf{Optimality:} Consider any other achievable point $(R', C', P')$ with $P' = 1$:
\begin{itemize}
\tightlist
\item If $R' = 0$: Fact is not encoded (excluded by requirement)
\item If $R' = 1$: Same as $(1, 1, 1)$ (by uniqueness of $C$ at $R=1$)
\item If $R' > 1$: By Theorem~\ref{thm:dof-gt-one-incoherence}, $P' < 1$, contradicting $P' = 1$
\end{itemize}

\textbf{Uniqueness:} No other point achieves $P = 1$ except $R = 1$.
\end{proof}

\textbf{Information-theoretic interpretation.} The Pareto frontier in rate-distortion theory is the curve $R(D)$ of minimum rate achieving distortion $D$. Here, the ``distortion'' is $1 - P$ (indicator of incoherence reachability), and the Pareto frontier collapses to a single point: $R = 1$ is the unique rate achieving $D = 0$.

\begin{corollary}[No Tradeoff at $P = 1$]\label{cor:no-tradeoff}
Unlike rate-distortion where you can trade rate for distortion, there is no tradeoff at $P = 1$ (perfect coherence). The only option is $R = 1$.
\end{corollary}

\begin{proof}
Direct consequence of Theorem~\ref{thm:coherence-capacity}.
\end{proof}

\textbf{Comparison to rate-distortion.} In rate-distortion theory:
\begin{itemize}
\tightlist
\item You can achieve lower distortion with higher rate (more bits)
\item The rate-distortion function $R(D)$ is monotonically decreasing
\item $D = 0$ (lossless) requires $R = H(X)$ (source entropy)
\end{itemize}

In our framework:
\begin{itemize}
\tightlist
\item You \emph{cannot} achieve higher coherence ($P$) with more independent locations
\item Higher rate ($R > 1$) \emph{eliminates} coherence guarantees ($P = 0$)
\item $P = 1$ (perfect coherence) requires $R = 1$ exactly
\end{itemize}

The key difference: redundancy (higher $R$) \emph{hurts} rather than helps coherence (without coordination). This inverts the intuition from error-correcting codes, where redundancy enables error detection/correction. Here, redundancy without derivation enables errors (incoherence).

\subsection{Practical Implications}\label{sec:practical-implications}

The unbounded gap has practical implications:

\textbf{1. DOF = 1 matters more at scale.} For small systems ($n = 3$), the difference between 3 edits and 1 edit is minor. For large systems ($n = 50$), the difference between 50 edits and 1 edit is significant.

\textbf{2. The gap compounds over time.} Each modification to fact $F$ incurs the complexity cost. If $F$ changes $m$ times over the system lifetime, total cost is $O(mn)$ with DOF $>$ 1 vs. $O(m)$ with DOF = 1.

\textbf{3. The gap affects error rates.} Each manual edit is an opportunity for error. With $n$ edits, the probability of at least one error is $1 - (1-p)^n$ where $p$ is the per-edit error probability. As $n$ grows, this approaches 1.

\begin{example}[Error Rate Calculation]\label{ex:error-rate}
Assume a 1\% error rate per edit ($p = 0.01$).

\begin{center}
\begin{tabular}{ccc}
\toprule
\textbf{Edits ($n$)} & \textbf{P(at least one error)} & \textbf{Architecture} \\
\midrule
1 & 1.0\% & DOF = 1 \\
10 & 9.6\% & DOF = 10 \\
50 & 39.5\% & DOF = 50 \\
100 & 63.4\% & DOF = 100 \\
\bottomrule
\end{tabular}
\end{center}

With 50 independent encoding locations (DOF = 50), there is a 39.5\% chance of introducing an error when modifying fact $F$. With DOF = 1, the chance is 1\%.
\end{example}

\subsection{Amortized Analysis}\label{sec:amortized}

The complexity bounds assume a single modification. Over the lifetime of an encoding system, facts are modified many times (\LH{REG4}).

\begin{theorem}[Amortized Complexity]\label{thm:amortized}
Let fact $F$ be modified $m$ times over the system lifetime. Let $n$ be the number of independent encoding locations. Total modification cost is:
\begin{itemize}
\tightlist
\item DOF = 1: $O(m)$
\item DOF = $n > 1$: $O(mn)$
\end{itemize}
\end{theorem}

\begin{proof}
Each modification costs $O(1)$ with DOF = 1 and $O(n)$ with DOF = $n$. Over $m$ modifications, total cost is $m \cdot O(1) = O(m)$ with DOF = 1 and $m \cdot O(n) = O(mn)$ with DOF = $n$.
\end{proof}

For a fact modified 100 times with 50 independent encoding locations:
\begin{itemize}
\tightlist
\item DOF = 1: 100 edits total
\item DOF = 50: 5,000 edits total
\end{itemize}

The 50$\times$ reduction factor applies to every modification, compounding over the system lifetime.

%==============================================================================
