\subsection{The Identification Problem}

Consider an encoder-decoder pair communicating about entities from a large universe $\mathcal{V}$. The decoder must \emph{identify} each entity, determining which of $k$ classes it belongs to, using only:
\begin{itemize}
\item A \emph{tag} of $L$ bits stored with the entity, and/or
\item \emph{Queries} to a binary oracle: ``does entity $v$ satisfy attribute $I$?''
\end{itemize}

This is not reconstruction (the decoder need not recover $v$), but \emph{identification} in the sense of Ahlswede and Dueck \cite{ahlswede1989identification}: the decoder must answer ``which class?'' with zero or bounded error. Our work extends this framework to consider the tradeoff between tag storage, query complexity, and identification accuracy.

We prove three results:
\begin{enumerate}
\item \textbf{Information barrier (identifiability limit).} When the attribute profile $\pi: \mathcal{V} \to \{0,1\}^n$ is not injective on classes, zero-error identification via queries alone is impossible: any decoder produces identical output on colliding classes, so cannot be correct for both.
\item \textbf{Optimal tagging (achievability).} A tag of $L = \lceil \log_2 k \rceil$ bits achieves zero-error identification with $W = O(1)$ query cost. For maximal-barrier domains ($A_\pi = k$), this is the unique Pareto-optimal point in the $(L, W, D)$ tradeoff space at $D=0$; in general domains, the converse depends on $A_\pi := \max_u |\{c : \pi(c)=u\}|$.
\item \textbf{Matroid structure (query complexity).} Minimal sufficient query sets form the bases of a matroid. The \emph{distinguishing dimension} (the common cardinality of all minimal sets) is well-defined and lower-bounds the query cost $W$ for any tag-free scheme.
\end{enumerate}

These results are universal: the theory applies to type systems, databases, biological taxonomy, and knowledge graphs. We develop the mathematics in full generality, then exhibit concrete instantiations.

\subsection{The Observation Model}

We formalize the observational constraint as a family of binary predicates. The terminology is deliberately abstract; concrete instantiations follow in Section~\ref{sec:applications}.

\begin{definition}[Entity space and attribute family]
Let $\mathcal{V}$ be a set of entities (program objects, database records, biological specimens, library items). Let $\mathcal{I}$ be a finite set of binary \emph{attributes}. Each $I \in \mathcal{I}$ induces a bipartition of $\mathcal{V}$ via $q_I$, and the full family induces the observational equivalence partition.
\end{definition}

\begin{remark}[Terminology]
We use ``attribute'' for the abstract concept. In type systems, attributes are \emph{interfaces} or \emph{method signatures}. In databases, they are \emph{columns}. In taxonomy, they are \emph{phenotypic characters}. In library science, they are \emph{facets}. The mathematics is identical.
\end{remark}

\begin{definition}[Attribute observation family]
For each $I \in \mathcal{I}$, define the attribute-membership observation $q_I: \mathcal{V} \to \{0,1\}$:
\[
q_I(v) = \begin{cases} 1 & \text{if } v \text{ satisfies attribute } I \\ 0 & \text{otherwise} \end{cases}
\]
Let $\Phi_{\mathcal{I}} = \{q_I : I \in \mathcal{I}\}$ denote the attribute observation family.
\end{definition}

\begin{remark}[Notation for size parameters]
We write $n := |\mathcal{I}|$ for the ambient number of available attributes. We write $d$ for the distinguishing dimension (the common size of all minimal distinguishing query sets; Definition~\ref{def:distinguishing-dimension}), so $d \le n$ and there exist worst-case families with $d = n$. We write $m$ for the number of \emph{query sites} (call sites) that perform attribute checks in a program or protocol (used only in the complexity-of-maintenance discussion).
When discussing a particular identification/verification task, we may write $s$ for the number of attributes actually queried/traversed by the procedure (e.g., members/fields checked in a structural type test, phenotypic characters checked in taxonomy), with $s \le n$. The maintenance-only parameter $m$ appears only in Section~\ref{sec:complexity}.
\end{remark}

\begin{definition}[Attribute profile]
The attribute profile function $\pi: \mathcal{V} \to \{0,1\}^{|\mathcal{I}|}$ maps each value to its complete attribute signature:
\[
\pi(v) = (q_I(v))_{I \in \mathcal{I}}
\]
\end{definition}

\begin{definition}[Attribute indistinguishability]
Values $v, w \in \mathcal{V}$ are \emph{attribute-indistinguishable}, written $v \sim w$, iff $\pi(v) = \pi(w)$.
\end{definition}

The relation $\sim$ is an equivalence relation. We write $[v]_\sim$ for the equivalence class of $v$.

\begin{definition}[Attribute-only observer]
An \emph{attribute-only observer} is any procedure whose interaction with a value $v \in \mathcal{V}$ is limited to queries in $\Phi_{\mathcal{I}}$. Formally, the observer interacts with $v$ only via primitive attribute queries $q_I \in \Phi_{\mathcal{I}}$; hence any transcript (and output) factors through $\pi(v)$.
\end{definition}

\subsection{The Central Question}

The central question is: \textbf{what semantic properties can an attribute-only observer compute?}

A semantic property is a function $P: \mathcal{V} \to \{0,1\}$ (or more generally, $P: \mathcal{V} \to Y$ for some codomain $Y$). We say $P$ is \emph{attribute-computable} if there exists a function $f: \{0,1\}^{|\mathcal{I}|} \to Y$ such that $P(v) = f(\pi(v))$ for all $v$.

\subsection{The Information Barrier}

\begin{theorem}[Information barrier]\label{thm:information-barrier}
Let $P: \mathcal{V} \to Y$ be any function. If $P$ is attribute-computable, then $P$ is constant on $\sim$-equivalence classes:
\[
v \sim w \implies P(v) = P(w)
\]
Equivalently: no attribute-only observer can compute any property that varies within an equivalence class.
\end{theorem}
\leanmeta{ \LH{ACS8}, \LH{IMP1}, \LH{COH1}}

\begin{proof}
Suppose $P$ is attribute-computable via $f$, i.e., $P(v) = f(\pi(v))$ for all $v$. Let $v \sim w$, so $\pi(v) = \pi(w)$. Then:
\[
P(v) = f(\pi(v)) = f(\pi(w)) = P(w)
\]
\end{proof}

\begin{remark}[Information-theoretic nature]
The barrier is \emph{informational}, not computational. Given unlimited time, memory, and computational power, an attribute-only observer still cannot distinguish $v$ from $w$ when $\pi(v) = \pi(w)$. The constraint is on the evidence itself.
\end{remark}

\begin{remark}[Role in the paper]
Theorem~\ref{thm:information-barrier} is the foundational invariance statement. The technical contribution is the downstream structure built on top of it: the ambiguity-based converse (Theorem~\ref{thm:converse}), the Pareto characterization (Theorem~\ref{thm:lwd-optimal}), and the matroid/equicardinality results (Section~\ref{sec:matroid}).
\end{remark}

\begin{corollary}[Class identity is not attribute-computable]\label{cor:provenance-barrier}
Let $C: \mathcal{V} \to \{1,\ldots,k\}$ be the class assignment function. If there exist values $v, w$ with $\pi(v) = \pi(w)$ but $C(v) \neq C(w)$, then class identity is not attribute-computable.
\end{corollary}
\leanmeta{ \LH{ACS9}}

\begin{proof}
Direct application of Theorem~\ref{thm:information-barrier} to $P = C$.
\end{proof}

\subsection{The Positive Result: Nominal Tagging}

We now show that augmenting attribute observations with a single primitive, nominal-tag access, achieves constant witness cost.

\begin{definition}[Nominal-tag access]
A \emph{nominal tag} is a value $\tau(v) \in \mathcal{T}$ associated with each $v \in \mathcal{V}$, representing the class identity of $v$. The \emph{nominal-tag access} operation returns $\tau(v)$ in $O(1)$ time.
\end{definition}

\begin{definition}[Primitive query set]
The extended primitive query set is $\Phi_{\mathcal{I}}^+ = \Phi_{\mathcal{I}} \cup \{\tau\}$, where $\tau$ denotes nominal-tag access.
\end{definition}

\begin{definition}[Witness cost]
Let $W(P)$ denote the minimum number of primitive queries from $\Phi_{\mathcal{I}}^+$ required to compute property $P$. We distinguish two tasks:
\begin{itemize}
\item $W_{\text{id}}$: Cost to identify the class of a single entity.
\item $W_{\text{eq}}$: Cost to determine if two entities have the same class.
\end{itemize}
Unless specified, $W$ refers to $W_{\text{eq}}$.
\end{definition}

\begin{theorem}[Constant witness for class identity]\label{thm:constant-witness}
Under nominal-tag access, class identity checking has constant witness cost:
\[
W(\text{class-identity}) = O(1)
\]
Specifically, the witness procedure is: return $\tau(v_1) = \tau(v_2)$.
\end{theorem}
\leanmeta{ \LH{ACS7}, \LH{FORC2}, \LH{DOM1}}

\begin{proof}
The procedure makes exactly 2 primitive queries (one $\tau$ access per value) and one comparison. This is $O(1)$ regardless of the number of available attributes $|\mathcal{I}|$.
\end{proof}

\begin{theorem}[Attribute-only lower bound]\label{thm:interface-lower-bound}
For attribute-only observers, class identity checking requires:
\[
W(\text{class-identity}) = \Omega(d)
\]
in the worst case, where $d$ is the distinguishing dimension (Definition~\ref{def:distinguishing-dimension}).
\end{theorem}
\leanmeta{ \LH{ACS1}}

\begin{proof}
Assume a zero-error attribute-only procedure halts after fewer than $d$ queries on every execution path. Fix any execution path and let $Q \subseteq \mathcal{I}$ be the set of queried attributes on that path, so $|Q|<d$. Since $d$ is the cardinality of every minimal distinguishing set, no set of size $<d$ is distinguishing; hence there exist values $v,w$ from different classes with identical answers on all attributes in $Q$.

An adversary can answer the procedure's queries consistently with both $v$ and $w$ along this path. Therefore the resulting transcript (and output) is identical on $v$ and $w$, contradicting zero-error class identification. So some execution path must use at least $d$ queries, giving worst-case cost $\Omega(d)$.
\end{proof}

\subsection{Main Contributions}

This paper establishes the following results:

\begin{enumerate}
\item \textbf{Information Barrier Theorem} (Theorem~\ref{thm:information-barrier}): Attribute-only observers cannot compute any property that varies within $\sim$-equivalence classes. This is an information-theoretic impossibility, not a computational limitation.

\item \textbf{Constant-Witness Theorem} (Theorem~\ref{thm:constant-witness}): Nominal-tag access achieves $W(\text{class-identity}) = O(1)$, with matching lower bound $\Omega(d)$ for attribute-only observers (Theorem~\ref{thm:interface-lower-bound}), where $d$ is the distinguishing dimension (Definition~\ref{def:distinguishing-dimension}).

\item \textbf{Complexity Separation} (Section~\ref{sec:complexity}): We establish O(1) vs O(k) vs $\Omega(d)$ complexity bounds for error localization under different observation regimes (where $d$ is the distinguishing dimension).

\item \textbf{Matroid Structure} (Section~\ref{sec:matroid}): Minimal distinguishing query sets form the bases of a matroid. All such sets have equal cardinality, establishing a well-defined ``distinguishing dimension.''

\item \textbf{$(L, W, D)$ Optimality} (Section~\ref{sec:lwd}): We characterize the zero-error converse via collision multiplicity $A_\pi$ and prove uniqueness of the nominal point in the maximal-barrier regime ($A_\pi=k$).

\item \textbf{Machine-Checked Proofs}: All results formalized in Lean 4 (\LeanTotalLines\ lines, \LeanTotalTheorems\ theorem/lemma statements, \LeanTotalSorry\ \texttt{sorry} placeholders).
\end{enumerate}

\subsection{Related Work and Positioning}

\textbf{Identification via channels.} Our work extends the identification paradigm introduced by Ahlswede and Dueck \cite{ahlswede1989identification, ahlswede1989identification2}. In their framework, a decoder need not reconstruct a message but only answer ``is the message $m$?'' for a given hypothesis. This yields dramatically different capacity: double-exponential codebook sizes become achievable. Our setting differs in three ways: (1) we consider zero-error identification rather than vanishing error, (2) queries are adaptive rather than block codes, and (3) we allow auxiliary tagging (rate $L$) to reduce query cost. The $(L, W, D)$ tradeoff generalizes Ahlswede-Dueck to a multi-dimensional operating regime.

\textbf{Rate-distortion theory.} The $(L, W, D)$ framework connects to Shannon's rate-distortion theory \cite{shannon1959coding, berger1971rate} with an important twist: the ``distortion'' $D$ is semantic (class misidentification), and there is a second resource $W$ (query cost) alongside rate $L$. Classical rate-distortion asks: what is the minimum rate to achieve distortion $D$? We ask: given rate $L$, what is the minimum query cost $W$ to achieve distortion $D = 0$? Theorem~\ref{thm:lwd-optimal} gives the converse in terms of collision multiplicity and identifies the unique nominal point in the maximal-barrier regime.

\textbf{Rate-distortion-perception tradeoffs.} Blau and Michaeli \cite{blau2019rethinking} extended rate-distortion theory by adding a perception constraint, creating a three-way tradeoff. Our query cost $W$ plays an analogous role: it measures the interactive cost of achieving low distortion rather than a distributional constraint. This parallel suggests that $(L, W, D)$ tradeoffs may admit similar geometric characterizations. Section~\ref{sec:extensions} develops this connection further.

\textbf{Zero-error information theory.} The matroid structure (Section~\ref{sec:matroid}) connects to zero-error capacity and graph entropy. K\"orner \cite{korner1973coding} and Witsenhausen \cite{witsenhausen1976zero} studied zero-error source coding where confusable symbols must be distinguished. Our distinguishing dimension (Definition~\ref{def:distinguishing-dimension}) is the minimum number of binary queries to separate all classes, which is precisely the zero-error identification cost when $L = 0$.

\textbf{Query complexity and communication complexity.} The $\Omega(d)$ lower bound for attribute-only identification relates to decision tree complexity \cite{buhrman2002complexity} and interactive communication \cite{orlitsky1991worst}. The key distinction is that our queries are constrained to a fixed observable family $\mathcal{I}$, not arbitrary predicates.

\textbf{Compression in classification systems.} The framework applies uniformly to databases, knowledge graphs, taxonomy, and typed software systems: for zero-error identification, ambiguity induces a minimum metadata requirement $L \ge \log_2 A_\pi$ (Theorem~\ref{thm:converse}), with maximal-barrier specialization $L \ge \log_2 k$.

\textbf{Programming-language corollary (secondary).} In nominal-vs-structural typing settings \cite{Cardelli1985, cook1990inheritance}, the model yields a concrete cost statement: under attribute collisions, purely structural identification has worst-case $\Omega(d)$ witness cost, while nominal tags achieve $O(1)$ identification using $O(\log A_\pi)$ bits. This is the paper's PL-facing corollary; the main contribution remains the information-theoretic characterization.

\subsection{Paper Organization}

Section~\ref{sec:framework} formalizes the compression framework and defines the $(L, W, D)$ tradeoff. Section~\ref{sec:complexity} establishes complexity bounds for error localization. Section~\ref{sec:matroid} proves the matroid structure of distinguishing query families. Section~\ref{sec:witness} analyzes witness cost in detail. Section~\ref{sec:lwd} proves the ambiguity-based converse and Pareto characterization. Section~\ref{sec:applications} provides cross-domain instantiations as secondary illustrations. Section~\ref{sec:conclusion} concludes. Appendix~\ref{sec:lean} describes the Lean 4 formalization.
