\section{Encoding Systems and Coherence}\label{sec:foundations}
%==============================================================================

We formalize exact consistency for multi-location encodings with modification constraints. The goal of this section is to isolate the abstract IT model before any software-specific interpretation: facts take values in finite alphabets, system states expose multiple representations of those facts, edits move the system among reachable states, and the observer must determine whether the represented fact is uniquely recoverable.

\subsection{Model assumptions and notation}\label{sec:assumptions}

For rigour we state the modeling assumptions used throughout the paper. These are intentionally minimal and are made explicit so reviewers can judge applicability.

\begin{enumerate}
\item \textbf{Fact value model (A1):} Each fact $F$ takes values in a finite set $\mathcal{V}_F$; $H(F)=\log_2|\mathcal{V}_F|$ denotes its entropy under a chosen prior. When we write ``zero probability of incoherence'' we mean $\Pr(\text{incoherent})=0$ under the model for edits and randomness specified below.
\item \textbf{Update model (A2):} Modifications occur as discrete events (rounds). An edit $\delta_F$ changes the source value for $F$; derived locations update deterministically when causal propagation is present. We do not require a blocklength parameter; our asymptotic statements are in the number of encoding locations $n$ or in ensemble scaling of the codebase.
\item \textbf{Adversary / randomness (A3):} When randomness is needed we assume a benign stochastic model for edits (e.g., uniformly sampled new values for illustration). For impossibility/converse statements we make no cooperative assumptions of the editor: adversarial sequences of edits that produce incoherence are allowed when DOF $>1$.
\item \textbf{Side-information channel (A4):} Derived locations and runtime queries collectively form side information $S$ available to any verifier. In the finite zero-error layer, the relevant quantity is the number of distinguishable tags that $S$ can expose.
\end{enumerate}

References to these assumptions use labels (A1)--(A4). Full formal versions of the operational model are given in the mechanized supplement (Supplement~A).

\subsection{The Encoding Model}\label{sec:epistemic}

We begin with the abstract encoding model: locations, values, reachable states, and ambiguity.

\begin{definition}[Encoding System]\label{def:encoding-system}
An \emph{encoding system} for a fact $F$ is a collection of locations $\{L_1, \ldots, L_n\}$, each capable of holding a value for $F$.
\end{definition}

\begin{definition}[Coherence]\label{def:coherence}
An encoding system is \emph{coherent} iff all locations hold the same value:
\[
\forall i, j: \text{value}(L_i) = \text{value}(L_j)
\]
\end{definition}

\begin{definition}[Incoherence]\label{def:incoherence}
An encoding system is \emph{incoherent} iff some locations disagree:
\[
\exists i, j: \text{value}(L_i) \neq \text{value}(L_j)
\]
\end{definition}

\begin{definition}[Ambiguity Class]\label{def:ambiguity-class}
For a system state $x$, the \emph{ambiguity class} of $x$ is the set
\[
\mathcal A(x) = \{v : \exists L \text{ with } \text{value}(L)=v\}.
\]
When $|\mathcal A(x)|>1$, the state presents multiple internally supported candidate values for the same fact.
\end{definition}

\textbf{The zero-error resolution problem.} Given a state $x$, a resolver must output the correct value of $F$ with zero error. When $|\mathcal A(x)|>1$, the internal state alone may fail to identify a unique answer.

\begin{theorem}[Oracle Arbitrariness]\label{thm:oracle-arbitrary}
For any incoherent encoding system state $x$ and any oracle $O$ that resolves $x$ to a value $v \in \mathcal A(x)$, there exists a value $v' \in \mathcal A(x)$ such that $v' \neq v$.
\end{theorem}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
By incoherence, $|\mathcal A(x)|>1$, so there exist $v_1,v_2\in\mathcal A(x)$ with $v_1\neq v_2$. Either $O$ picks $v_1$ (then $v_2$ disagrees) or it does not pick $v_1$ (then $v_1$ disagrees).
\end{proof}

\textbf{Interpretation.} This is the first zero-error obstruction in the paper: once an ambiguity class contains more than one value, the state itself does not identify which value is authoritative. Any exact resolver must therefore use auxiliary information not contained in the state representation alone.

\begin{definition}[Degrees of Freedom]\label{def:dof-epistemic}
The \emph{degrees of freedom} (DOF) of an encoding system is the number of locations that can be modified independently.
\end{definition}

\begin{theorem}[DOF = 1 Guarantees Coherence]\label{thm:dof-one-coherence}
If DOF = 1, then the encoding system is coherent in all reachable states.
\end{theorem}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
With DOF = 1, exactly one location is independent. All other locations are derived (automatically updated when the source changes). Derived locations cannot diverge from their source. Therefore, all locations hold the value determined by the single independent source. Disagreement is impossible.
\end{proof}

\begin{theorem}[DOF $>$ 1 Permits Incoherence]\label{thm:dof-gt-one-incoherence}
If DOF $> 1$, then incoherent states are reachable.
\end{theorem}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
With DOF $> 1$, at least two locations are independent. Independent locations can be modified separately. A sequence of edits can set $L_1 = v_1$ and $L_2 = v_2$ where $v_1 \neq v_2$. This is an incoherent state.
\end{proof}

\begin{corollary}[Coherence Forces DOF = 1]\label{cor:coherence-forces-ssot}
If coherence must be guaranteed (no incoherent states reachable), then DOF = 1 is necessary and sufficient.
\end{corollary}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

This is the basic exact-consistency criterion in the model (\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}).

The following subsections refine this model and then connect it back to concrete systems only as examples.

% Section divider

\subsection{Computational Realizations}\label{sec:edit-space}

The abstract encoding model (Definitions~\ref{def:encoding-system}--\ref{def:dof-epistemic}) applies to any system where:
\begin{enumerate}
\tightlist
\item Facts are encoded at multiple locations
\item Locations can be modified
\item Correctness requires coherence across modifications
\end{enumerate}

\textbf{Domains satisfying these constraints:}
\begin{itemize}
\tightlist
\item \textbf{Software codebases:} Type definitions, registries, configurations
\item \textbf{Distributed databases:} Replica consistency under updates
\item \textbf{Configuration systems:} Multi-file settings (e.g., infrastructure-as-code)
\item \textbf{Version control:} Merge resolution under concurrent modifications
\end{itemize}

We focus on \emph{computational realizations} only to anchor the model. Software codebases are a convenient example, but the theorems in this section are stated independently of any language, runtime, or toolchain.

\begin{definition}[Codebase (Software Realization)]
A \emph{codebase} $C$ is a finite collection of source files, each containing syntactic constructs (classes, functions, statements, expressions). This is the canonical computational encoding system.
\end{definition}

\begin{definition}[Location]
A \emph{location} $L \in C$ is a syntactically identifiable region: a class definition, function body, configuration value, type annotation, database field, or configuration entry.
\end{definition}

\begin{definition}[Modification Space]
For encoding system $C$, the \emph{modification space} $E(C)$ is the set of all valid modifications. Each edit $\delta \in E(C)$ transforms $C$ into $C' = \delta(C)$.
\end{definition}

The modification space is large (exponential in system size). But we focus on modifications that \emph{change a specific fact}.

\subsection{Facts: Atomic Units of Specification}\label{sec:facts}

\begin{definition}[Fact]\label{def:fact}
A \emph{fact} $F$ is an atomic unit of program specification: a single piece of knowledge that can be independently modified. Facts are the indivisible units of meaning in a specification.
\end{definition}

The granularity of facts is determined by the specification, not the implementation. If two pieces of information must always change together, they constitute a single fact. If they can change independently, they are separate facts.

\noindent\textbf{Examples of facts:}

\begin{center}
\small
\setlength{\tabcolsep}{3pt}
\begin{tabularx}{\linewidth}{@{}>{\RaggedRight\arraybackslash\hsize=0.95\hsize}X>{\RaggedRight\arraybackslash\hsize=1.05\hsize}X@{}}
\toprule
\textbf{Fact} & \textbf{Description} \\
\midrule
$F_1$: ``threshold = 0.5'' & A configuration value \\
$F_2$: ``\texttt{PNGLoader} handles \texttt{.png}'' & A type-to-handler mapping \\
$F_3$: ``\texttt{validate()} returns \texttt{bool}'' & A method signature \\
$F_4$: ``\texttt{Detector} is a subclass of \texttt{Processor}'' & An inheritance relationship \\
$F_5$: ``\texttt{Config} has field \texttt{name: str}'' & A dataclass field \\
\bottomrule
\end{tabularx}
\end{center}

\begin{definition}[Structural Fact]\label{def:structural-fact}
A fact $F$ is \emph{structural} with respect to encoding system $C$ iff the locations encoding $F$ are fixed at definition time:
\[
\begin{aligned}
\text{structural}(F, C) \Longleftrightarrow\;
&\forall L:\; \text{encodes}(L, F) \\
&\rightarrow\; L \in \text{DefinitionSyntax}(C)
\end{aligned}
\]
where $\text{DefinitionSyntax}(C)$ comprises declarative constructs that cannot change post-definition without recreation.
\end{definition}

\textbf{Examples across domains:}
\begin{itemize}
\tightlist
\item \textbf{Software:} Class declarations, method signatures, inheritance clauses, attribute definitions
\item \textbf{Databases:} Schema definitions, table structures, foreign key constraints
\item \textbf{Configuration:} Infrastructure topology, service dependencies
\item \textbf{Version control:} Branch structure, merge policies
\end{itemize}

\textbf{Key property:} Structural facts are fixed at \emph{definition time}. Once defined, their structure cannot change without recreation. This is why structural coherence requires definition-time computation: the encoding locations are only mutable during creation.

\textbf{Non-structural facts} (runtime values, mutable state) have encoding locations modifiable post-definition. Achieving DOF = 1 for non-structural facts requires different mechanisms (reactive bindings, event systems) and is outside this paper's scope. We focus on structural facts because they demonstrate the impossibility results most clearly.

\subsection{Encoding: The Correctness Relationship}\label{sec:encoding}

\begin{definition}[Encodes]\label{def:encodes}
Location $L$ \emph{encodes} fact $F$, written $\text{encodes}(L, F)$, iff correctness requires updating $L$ when $F$ changes.

Formally:
\[
\text{encodes}(L, F) \Longleftrightarrow \forall \delta_F: \neg\text{updated}(L, \delta_F) \rightarrow \text{incorrect}(\delta_F(C))
\]

where $\delta_F$ is an edit targeting fact $F$.
\end{definition}

\textbf{Key insight:} This definition is induced by correctness rather than by annotation. If failing to update location $L$ when fact $F$ changes produces an incorrect system, then $L$ participates in the representation of $F$.

\begin{example}[Encoding in Practice]\label{ex:encoding}
Consider a type registry:

\begin{code}
# Location L1: Class definition
class PNGLoader(ImageLoader):
    format = "png"

# Location L2: Registry entry
LOADERS = {"png": PNGLoader, "jpg": JPGLoader}

# Location L3: Documentation
# Supported formats: png, jpg
\end{code}

The fact $F$ = ``\texttt{PNGLoader} handles \texttt{png}'' is encoded at:
\begin{itemize}
\tightlist
\item $L_1$: The class definition (primary encoding)
\item $L_2$: The registry dictionary (secondary encoding)
\item $L_3$: The documentation comment (tertiary encoding)
\end{itemize}

If $F$ changes (e.g., to ``\texttt{PNGLoader} handles \texttt{png} and \texttt{apng}''), all three locations must be updated for correctness. The program is incorrect if $L_2$ still says \texttt{\{"png": PNGLoader\}} when the class now handles both formats.
\end{example}

\subsection{Modification Complexity}\label{sec:mod-complexity}

\begin{definition}[Modification Complexity]\label{def:mod-complexity}
\[
M(C, \delta_F) = |\{L \in C : \text{encodes}(L, F)\}|
\]
The number of locations that must be updated when fact $F$ changes.
\end{definition}

Modification complexity is the central metric of this paper. It measures the \emph{cost} of changing a fact. A codebase with $M(C, \delta_F) = 47$ requires 47 edits to correctly implement a change to fact $F$. A codebase with $M(C, \delta_F) = 1$ requires only 1 edit.

\begin{theorem}[Correctness Forcing]\label{thm:correctness-forcing}
$M(C, \delta_F)$ is the \textbf{minimum} number of edits required for correctness. Fewer edits imply an incorrect program.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Suppose $M(C, \delta_F) = k$, meaning $k$ locations encode $F$. By Definition~\ref{def:encodes}, each encoding location must be updated when $F$ changes. If only $j < k$ locations are updated, then $k - j$ locations still reflect the old value of $F$. These locations create inconsistencies:

\begin{enumerate}
\tightlist
\item The specification says $F$ has value $v'$ (new)
\item Locations $L_1, \ldots, L_j$ reflect $v'$
\item Locations $L_{j+1}, \ldots, L_k$ reflect $v$ (old)
\end{enumerate}

By Definition~\ref{def:encodes}, the program is incorrect. Therefore, all $k$ locations must be updated, and $k$ is the minimum.
\end{proof}

\subsection{Independence and Degrees of Freedom}\label{sec:dof}

Not all encoding locations are created equal. Some are \emph{derived} from others.

\begin{definition}[Independent Locations]\label{def:independent}
Locations $L_1, L_2$ are \emph{independent} for fact $F$ iff they can diverge. Updating $L_1$ does not automatically update $L_2$, and vice versa.

Formally: $L_1$ and $L_2$ are independent iff there exists a sequence of edits that makes $L_1$ and $L_2$ encode different values for $F$.
\end{definition}

\begin{definition}[Derived Location]\label{def:derived}
Location $L_{\text{derived}}$ is \emph{derived from} $L_{\text{source}}$ iff updating $L_{\text{source}}$ automatically updates $L_{\text{derived}}$. Derived locations are not independent of their sources.
\end{definition}

\begin{example}[Independent vs. Derived]\label{ex:independence}
Consider two architectures for the type registry:

\textbf{Architecture A (independent locations):}
\begin{code}
# L1: Class definition
class PNGLoader(ImageLoader): ...

# L2: Manual registry (independent of L1)
LOADERS = {"png": PNGLoader}
\end{code}

Here $L_1$ and $L_2$ are independent. A developer can change $L_1$ without updating $L_2$, causing inconsistency.

\textbf{Architecture B (derived location):}
\begin{code}
# L1: Class definition with registration
class PNGLoader(ImageLoader):
    format = "png"

# L2: Derived registry (computed from L1)
LOADERS = {cls.format: cls for cls in ImageLoader.__subclasses__()}
\end{code}

Here $L_2$ is derived from $L_1$. Updating the class definition automatically updates the registry. They cannot diverge.
\end{example}

\begin{definition}[Degrees of Freedom]\label{def:dof}
\[
\text{DOF}(C, F) = |\{L \in C : \text{encodes}(L, F) \land \text{independent}(L)\}|
\]
The number of \emph{independent} locations encoding fact $F$.
\end{definition}

DOF is the key metric. Modification complexity $M$ counts all encoding locations. DOF counts only the independent ones. If all but one encoding location is derived, DOF = 1 even though $M$ can be large.

\begin{theorem}[DOF = Incoherence Potential]\label{thm:dof-inconsistency}
$\text{DOF}(C, F) = k$ implies that up to $k$ distinct values for $F$ can coexist simultaneously. In particular, $k>1$ makes ambiguity reachable.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Each independent location can hold a different value. By Definition~\ref{def:independent}, no constraint forces agreement between independent locations. Therefore, $k$ independent locations can simultaneously realize up to $k$ distinct values.
\end{proof}

\begin{corollary}[DOF $>$ 1 Implies Incoherence Risk]\label{cor:dof-risk}
$\text{DOF}(C, F) > 1$ implies incoherent states are reachable.
\end{corollary}
\leanmeta{\LH{UNQ1}}

\subsection{The DOF Lattice}\label{sec:dof-lattice}

DOF values partition the model into three regimes:

\begin{center}
\begin{tabularx}{\linewidth}{cX}
\toprule
\textbf{DOF} & \textbf{Encoding Status} \\
\midrule
0 & Fact $F$ is not encoded (no representation) \\
1 & Coherence guaranteed (optimal rate under coherence constraint) \\
$k > 1$ & Incoherence possible (redundant independent encodings) \\
\bottomrule
\end{tabularx}
\end{center}

\begin{theorem}[DOF = 1 is Uniquely Coherent]\label{thm:dof-optimal}
For any fact $F$ that must be encoded, $\text{DOF}(C, F) = 1$ is the unique value guaranteeing coherence:
\begin{enumerate}
\tightlist
\item DOF = 0: Fact is not represented
\item DOF = 1: Coherence guaranteed (by Theorem~\ref{thm:dof-one-coherence})
\item DOF $>$ 1: Incoherence reachable (by Theorem~\ref{thm:dof-gt-one-incoherence})
\end{enumerate}
This theorem is deliberately simple: it records the exact-consistency trichotomy of the model before the later side-information and complexity consequences are added. The same trichotomy is recast as a capacity statement in Section~\ref{sec:capacity}.
\end{theorem}
\leanmeta{\LH{UNQ1}}

%==============================================================================
\subsection{Encoding-Theoretic CAP and FLP}\label{sec:cap-flp}

We record two structural analogies to CAP and FLP inside the encoding model (\LH{UNQ1}, \LH{UNQ1}). These are secondary to the main IT contribution.

\begin{definition}[Local Availability]\label{def:local-availability}
An encoding system for fact $F$ is \emph{locally available} iff for every encoding location $L$ of $F$ and every value $v$, there exists a valid edit $\delta \in E(C)$ such that $\text{updated}(L, \delta)$ and for every other encoding location $L'$, $\neg \text{updated}(L', \delta)$. Informally: each encoding location can be updated without coordinating with others.
\end{definition}

\begin{definition}[Partition Tolerance]\label{def:partition-tolerance}
An encoding system for fact $F$ is \emph{partition-tolerant} iff $F$ is encoded at two or more locations:
\[
|\{L \in C : \text{encodes}(L, F)\}| \ge 2.
\]
This is the minimal formal notion of ``replication'' in our model; without it, partitions are vacuous.
\end{definition}

\begin{theorem}[CAP in the Encoding Model]\label{thm:cap-encoding}
No encoding system can simultaneously guarantee coherence (Definition~\ref{def:coherence}), local availability (Definition~\ref{def:local-availability}), and partition tolerance (Definition~\ref{def:partition-tolerance}) for the same fact $F$.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Partition tolerance gives at least two encoding locations. Local availability allows each to be updated without updating any other encoding location, so by Definition~\ref{def:independent} there exist two independent locations and thus $\text{DOF}(C, F) > 1$. By Theorem~\ref{thm:dof-gt-one-incoherence}, incoherent states are reachable, contradicting coherence.
\end{proof}

\begin{definition}[Resolution Procedure]\label{def:resolution-procedure}
A \emph{resolution procedure} is a deterministic function $R$ that maps an encoding system state to a value present in that state.
\end{definition}

\begin{theorem}[Static FLP in the Encoding Model]\label{thm:static-flp}
For any incoherent encoding system state and any resolution procedure $R$, the returned value is arbitrary relative to the other values present; no deterministic $R$ can be justified by internal information alone.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Immediate from Theorem~\ref{thm:oracle-arbitrary}: in an incoherent state, at least two distinct values are present, and any choice leaves another value disagreeing.
\end{proof}

These theorems are the encoding-theoretic counterparts of CAP~\cite{brewer2000cap,gilbert2002cap} and FLP~\cite{flp1985impossibility}: CAP corresponds to the impossibility of coherence when replicated encodings remain independently updatable; FLP corresponds to the impossibility of truth-preserving resolution in an incoherent state without side information.

%==============================================================================
\subsection{Information-Theoretic Quantities}\label{sec:it-quantities}

Before establishing the main threshold result, we define the information-theoretic quantities used to express ambiguity, redundancy, and side-information requirements.

\begin{definition}[Encoding Rate]\label{def:encoding-rate}
The \emph{encoding rate} of system $C$ for fact $F$ is $R(C, F) = \text{DOF}(C, F)$, the number of independent encoding locations. This counts locations that can hold distinct values simultaneously.
\end{definition}

\begin{definition}[Value Entropy]\label{def:value-entropy}
For a fact $F$ with value space $\mathcal{V}_F$, the \emph{value entropy} is:
\[
H(F) = \log_2 |\mathcal{V}_F|
\]
This is the information content of specifying $F$'s value.
\end{definition}

\begin{definition}[Encoding Redundancy]\label{def:redundancy}
The \emph{encoding redundancy} of system $C$ for fact $F$ is:
\[
\rho(C, F) = \text{DOF}(C, F) - 1
\]
Redundancy measures excess independent encodings beyond the minimum needed to represent $F$.
\end{definition}

\begin{theorem}[Redundancy-Incoherence Equivalence]\label{thm:redundancy-incoherence}
Encoding redundancy $\rho > 0$ is necessary and sufficient for incoherence reachability:
\[
\rho(C, F) > 0 \Longleftrightarrow \text{incoherent states reachable}
\]
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
$(\Rightarrow)$ If $\rho > 0$, then DOF $> 1$. By Theorem~\ref{thm:dof-gt-one-incoherence}, incoherent states are reachable.

$(\Leftarrow)$ If incoherent states are reachable, then by contrapositive of Theorem~\ref{thm:dof-one-coherence}, DOF $\neq 1$. Since the fact is encoded, DOF $\geq 1$, so DOF $> 1$, hence $\rho > 0$.
\end{proof}

\begin{definition}[Incoherence Entropy]\label{def:incoherence-entropy}
For a state whose ambiguity class has size $k$, the \emph{incoherence entropy} is:
\[
H_{\text{inc}} = \log_2 k
\]
This quantifies the uncertainty about which value is authoritative.
\end{definition}

%==============================================================================
\subsection{Zero-Incoherence Capacity Theorem}\label{sec:capacity}

We now establish the central exact-consistency characterization. The argument is elementary at the combinatorial level, but it serves as the base layer for the later side-information and rate-complexity consequences. In that sense the section plays the role that an achievability/converse theorem plays in classical IT.

\textbf{Background: Zero-Error Capacity.} Shannon~\cite{shannon1956zero} introduced zero-error capacity: the maximum rate at which information can be transmitted with \emph{zero} probability of error. Körner~\cite{korner1973graphs} and Lovász~\cite{lovasz1979shannon} linked such questions to confusability structure. Our zero-incoherence capacity is the corresponding exact-consistency threshold for modifiable encodings.

\begin{definition}[Zero-Incoherence Capacity]\label{def:coherence-capacity}
The \emph{zero-incoherence capacity} of an encoding system is:
\[
C_0 = \sup\{R : \text{encoding rate } R \Rightarrow \text{incoherence probability} = 0\}
\]
where ``incoherence probability = 0'' means no incoherent state is reachable under any modification sequence.
\end{definition}

\begin{theorem}[Zero-Incoherence Capacity]\label{thm:coherence-capacity}
The zero-incoherence capacity of any encoding system under independent modification is exactly 1:
\[
C_0 = 1
\]
\end{theorem}
\leanmeta{\LH{UNQ1}}

We state the result in achievability/converse form because later sections attach nontrivial side-information and complexity consequences to the same threshold.

\begin{theorem}[Achievability]\label{thm:capacity-achievability}
Encoding rate DOF $= 1$ achieves zero incoherence. Therefore $C_0 \geq 1$.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Let DOF$(C, F) = 1$. By Definition~\ref{def:dof}, exactly one location $L_s$ is independent; all other locations $\{L_i\}$ are derived from $L_s$.

By Definition~\ref{def:derived}, derived locations satisfy: $\text{update}(L_s) \Rightarrow \text{automatically\_updated}(L_i)$. Therefore, for any reachable state:
\[
\forall i: \text{value}(L_i) = f_i(\text{value}(L_s))
\]
where $f_i$ is the derivation function. All values are determined by $L_s$. Disagreement requires two locations with different determining sources, but only $L_s$ exists. Therefore, no incoherent state is reachable.
\end{proof}

\begin{theorem}[Converse]\label{thm:capacity-converse}
Encoding rate DOF $> 1$ does not achieve zero incoherence. Therefore $C_0 < R$ for all $R > 1$.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Let DOF$(C, F) = k > 1$. By Definition~\ref{def:independent}, there exist locations $L_1, L_2$ that can be modified independently.

\textbf{Construction of incoherent state:} Consider the modification sequence:
\begin{enumerate}
\tightlist
\item $\delta_1$: Set $L_1 = v_1$ for some $v_1 \in \mathcal{V}_F$
\item $\delta_2$: Set $L_2 = v_2$ for some $v_2 \neq v_1$
\end{enumerate}

Both modifications are valid (each location accepts values from $\mathcal{V}_F$). By independence, $\delta_2$ does not affect $L_1$. The resulting state has:
\[
\text{value}(L_1) = v_1 \neq v_2 = \text{value}(L_2)
\]

By Definition~\ref{def:incoherence}, this state is incoherent. Since it is reachable, zero incoherence is not achieved.

The converse is intentionally combinatorial: once two independently writable locations exist, one can construct a reachable disagreement state. Later sections refine the same obstruction using finite zero-error side-information counting arguments.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:coherence-capacity}]
By Theorem~\ref{thm:capacity-achievability}, $C_0 \geq 1$.

By Theorem~\ref{thm:capacity-converse}, $C_0 < R$ for all $R > 1$.

Therefore $C_0 = 1$ exactly. The bound is tight: achieved at DOF $= 1$, not achieved at any DOF $> 1$.
\end{proof}

\textbf{Comparison to Shannon capacity.} The analogy to channel coding is structural rather than literal:

\begin{center}
\footnotesize
\setlength{\tabcolsep}{3pt}
\begin{tabularx}{\linewidth}{@{}>{\RaggedRight\arraybackslash}p{0.19\linewidth}YY@{}}
\toprule
\textbf{Aspect} & \textbf{Shannon (Channel)} & \textbf{This Work (Encoding)} \\
\midrule
Rate & Bits per channel use & Independent encoding locations \\
Constraint & Error probability $\to 0$ & Incoherence probability $= 0$ \\
Achievability & $R < C$ achieves & DOF $= 1$ achieves \\
Converse & $R > C$ fails & DOF $> 1$ fails \\
Capacity & $C = \max I(X;Y)$ & $C_0 = 1$ \\
\bottomrule
\end{tabularx}
\end{center}

The key difference is that Shannon capacity concerns communication over noisy channels, while the present threshold concerns exact consistency of stored representations under allowed edits.

\begin{corollary}[Capacity-Achieving Rate is Unique]\label{cor:capacity-unique}
DOF $= 1$ is the unique capacity-achieving encoding rate. No alternative achieves zero incoherence at higher rate.
\end{corollary}
\leanmeta{\LH{L1}}

\begin{corollary}[Redundancy Above Capacity]\label{cor:redundancy-above}
Any encoding with $\rho > 0$ (redundancy) operates above capacity and cannot guarantee coherence.
\end{corollary}
\leanmeta{\LH{UNQ1}}

\subsubsection{Zero-Distortion Boundary}

The model induces a degenerate zero-distortion boundary: the only rate achieving perfect coherence is rate $1$.

\begin{definition}[Rate-Incoherence Function]\label{def:rate-incoherence}
For $\epsilon\in[0,1]$, define
\[
R(\epsilon) = \inf\{R : \exists \text{ encoding achieving } P(\text{incoherence}) \leq \epsilon\}
\]
where $\epsilon \in [0, 1]$ is the tolerable incoherence probability.
\end{definition}

\begin{theorem}[Rate-Incoherence has a Sharp Zero-Error Threshold]\label{thm:rate-incoherence-step}
At the zero-error point, the minimum achievable rate is exactly one:
\[
R(0)=1.
\]
For every $\epsilon>0$, the model imposes no nontrivial upper bound on admissible independent rate.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
The zero-error statement is exactly Theorem~\ref{thm:coherence-capacity}. Once a positive incoherence budget is allowed, the model no longer forbids higher independent rate, so no analogous finite cutoff remains.
\end{proof}

\textbf{Interpretation.} Unlike classical rate-distortion curves, the interesting point here is the sharp zero-error boundary: perfect coherence isolates a single admissible rate.

\subsubsection{Design Necessity}

The threshold result also yields a simple design corollary:

\begin{theorem}[Design Necessity]\label{thm:design-necessity}
If an encoding system requires zero incoherence for correctness, then the encoding rate must satisfy DOF $\leq 1$. This is a necessary condition; no design choice can circumvent it.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Suppose a system requires $P(\text{incoherence}) = 0$ and has DOF $> 1$. By Theorem~\ref{thm:capacity-converse}, incoherent states are reachable. This contradicts the requirement. Therefore DOF $\leq 1$ is necessary.

The bound is not an implementation artifact: with DOF $> 1$, independent locations can diverge by construction.
\end{proof}

\begin{corollary}[Architectural Forcing]\label{cor:architectural-forcing}
For any fact $F$ requiring coherence guarantees, system architecture must either:
\begin{enumerate}
\tightlist
\item Achieve DOF$(F) = 1$ through derivation mechanisms, or
\item Accept that coherence cannot be guaranteed and implement resolution protocols with $\geq \log_2(\text{DOF})$ bits of side information.
\end{enumerate}
There is no third option.
\end{corollary}
\leanmeta{\LH{L1}}

\subsubsection{Coding Interpretation of Derivation}

The coding-theoretic interpretation is compact: derivation is the capacity-achieving dependence structure. One location carries the independent value and the remaining locations are deterministic views. Those views improve observability without increasing independent rate.

\subsubsection{Connection to Confusability Graphs}

Körner~\cite{korner1973graphs} and Lovász~\cite{lovasz1979shannon} characterized zero-error capacity via \emph{confusability graphs}: vertices are channel inputs, edges connect inputs that can produce the same output. Zero-error capacity is determined by the graph's independence number.

We can define an analogous \emph{incoherence graph} for encoding systems:

\begin{definition}[Incoherence Graph]\label{def:incoherence-graph}
For encoding system $C$ and fact $F$, the incoherence graph $G = (V, E)$ has:
\begin{itemize}
\tightlist
\item Vertices $V = $ independent encoding locations
\item Edge $(L_i, L_j) \in E$ iff $L_i$ and $L_j$ can hold different values (are truly independent)
\end{itemize}
\end{definition}

\begin{theorem}[Incoherence Graph Characterization]\label{thm:incoherence-graph}
Zero incoherence is achievable iff the incoherence graph has at most one vertex: $|V| \leq 1$.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
If $|V| = 0$, no locations encode $F$. If $|V| = 1$, DOF $= 1$ and zero incoherence is achieved by Theorem~\ref{thm:capacity-achievability}. If $|V| \geq 2$, DOF $\geq 2$ and incoherence is reachable by Theorem~\ref{thm:capacity-converse}.
\end{proof}

This graph view is intentionally simple. Its value is not a deep graph invariant, but a compact bridge to zero-error language before the richer confusability bounds used later.

\subsection{Side Information for Resolution}\label{sec:side-information}

When an encoding system is incoherent, exact resolution requires external side information. This is the point where the paper becomes genuinely information-theoretic: ambiguity size induces a quantitative lower bound on the information a resolver must obtain.

\subsubsection{The Multi-Terminal View}

We can view an encoding system as a \emph{multi-terminal source coding} problem~\cite{slepian1973noiseless,cover2006elements}:
\begin{itemize}
\tightlist
\item Each encoding location is a \emph{terminal} that can transmit (store) a value
\item The fact $F$ is the \emph{source} to be encoded
\item \emph{Derivation} introduces correlation: a derived terminal's output is a deterministic function of its source terminal
\item The \emph{decoder} (any observer querying the system) must reconstruct $F$'s value
\end{itemize}

In Slepian-Wolf coding, correlated sources $(X, Y)$ can be encoded at rates $(R_X, R_Y)$ satisfying $R_X \geq H(X|Y)$, $R_Y \geq H(Y|X)$, $R_X + R_Y \geq H(X,Y)$. With perfect correlation ($Y = f(X)$), we have $H(X|Y) = 0$--the correlated source needs zero rate.

\textbf{Derivation as perfect correlation.} In our model, derivation creates perfect correlation: if $L_d$ is derived from $L_s$, then $\text{value}(L_d) = f(\text{value}(L_s))$ deterministically. The ``rate'' of $L_d$ is effectively zero--it contributes no independent information. This is why derived locations do not contribute to DOF.

\textbf{Independence as zero correlation.} Independent locations have no correlation. Each can hold any value in $\mathcal{V}_F$. The total ``rate'' is DOF $\cdot H(F)$, but with DOF $> 1$, the locations can encode \emph{different} values--incoherence.

\subsubsection{Side Information Bound}

\begin{theorem}[Side Information Requirement]\label{thm:side-info}
Given a state whose ambiguity class has size $k$, resolving to the correct value requires at least $\log_2 k$ bits of side information.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
The ambiguity class presents $k$ internally supported alternatives. By Theorem~\ref{thm:oracle-arbitrary}, no internal rule distinguishes them.

Let $S \in \{1, \ldots, k\}$ denote the index of the correct source. The decoder must determine $S$ to resolve correctly. Since $S$ is uniformly distributed over $k$ alternatives (by symmetry of the incoherent state):
\[
H(S) = \log_2 k
\]

Any resolution procedure is a function $R: \text{State} \to \mathcal{V}_F$. Without side information $Y$:
\[
H(S | R(\text{State})) = H(S) = \log_2 k
\]
because $R$ can be computed from the state, which contains no information about which source is correct.

With side information $Y$ that identifies the correct source:
\[
H(S | Y) = 0 \Rightarrow I(S; Y) = H(S) = \log_2 k
\]

Therefore, side information must expose at least $\log_2 k$ bits' worth of distinguishable tags.
\end{proof}

\begin{corollary}[DOF = 1 Requires Zero Side Information]\label{cor:dof1-zero-side}
With DOF = 1, resolution requires 0 bits of side information.
\end{corollary}
\leanmeta{\LH{UNQ1}}

\begin{proof}
With $k = 1$, there is one independent location. $H(S) = \log_2 1 = 0$. No uncertainty exists about the authoritative source.
\end{proof}

\begin{corollary}[Side Information Scales with Redundancy]\label{cor:side-info-redundancy}
The required side information equals $\log_2(1 + \rho)$ where $\rho$ is encoding redundancy:
\[
\text{Side information required} = \log_2(\text{DOF}) = \log_2(1 + \rho)
\]
\end{corollary}
\leanmeta{\LH{UNQ1}}

\textbf{Connection to Slepian-Wolf.} In Slepian-Wolf coding, the decoder has side information $Y$ and decodes $X$ at rate $H(X|Y)$. Our setting inverts this: the decoder needs side information $S$ (source identity) to ``decode'' (resolve) which value is correct. The required side information is exactly the entropy of the source-selection variable.

\textbf{Connection to zero-error decoding.} In zero-error channel coding, the decoder must identify the transmitted message with certainty. Without sufficient side information (channel structure), this is impossible. Similarly, without $\log_2 k$ bits identifying the authoritative source, zero-error resolution is impossible.

\begin{example}[Side Information in Practice]\label{ex:side-info-practice}
Consider a configuration system with DOF = 3:
\begin{itemize}
\tightlist
\item \texttt{config.yaml}: \texttt{threshold: 0.5}
\item \texttt{settings.json}: \texttt{"threshold": 0.7}
\item \texttt{params.toml}: \texttt{threshold = 0.6}
\end{itemize}

Resolution requires $\log_2 3 \approx 1.58$ bits of side information:
\begin{itemize}
\tightlist
\item A priority ordering: ``YAML $>$ JSON $>$ TOML'' (encodes permutation $\to$ identifies source)
\item A timestamp comparison: ``most recent wins'' (encodes ordering $\to$ identifies source)
\item An explicit declaration: ``params.toml is authoritative'' (directly encodes source identity)
\end{itemize}

With DOF = 1, zero bits suffice--the single source is self-evidently authoritative.
\end{example}

\subsubsection{Multi-Terminal Capacity Interpretation}

The exact-consistency threshold (Theorem~\ref{thm:coherence-capacity}) can be restated in multi-terminal language:

\begin{corollary}[Multi-Terminal Interpretation]\label{cor:multi-terminal}
An encoding system achieves zero incoherence iff all terminals are perfectly correlated (every terminal's output is a deterministic function of one source terminal). Equivalently: the correlation structure must be a tree with one root.
\end{corollary}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
Perfect correlation means DOF = 1 (only the root is independent). By Theorem~\ref{thm:capacity-achievability}, this achieves zero incoherence. If any two terminals are independent (not perfectly correlated through the root), DOF $\geq 2$, and by Theorem~\ref{thm:capacity-converse}, incoherence is reachable.
\end{proof}

This connects to network information theory: in a multi-terminal network, achieving coherence requires complete dependence on a single source--no ``multicast'' of independent copies.

\subsection{Structure Theorems: The Derivation Lattice}\label{sec:derivation-structure}

The set of derivation relations on an encoding system has algebraic structure. We characterize this structure and its computational implications (\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}).

\begin{definition}[Derivation Relation]\label{def:derivation-relation}
A \emph{derivation relation} $D \subseteq L \times L$ on locations $L$ is a directed relation where $(L_s, L_d) \in D$ means $L_d$ is derived from $L_s$. We require $D$ be acyclic (no location derives from itself through any chain).
\end{definition}

\begin{definition}[DOF under Derivation]\label{def:dof-derivation}
Given derivation relation $D$, the degrees of freedom is:
\[
\text{DOF}(D) = |\{L : \nexists L'. (L', L) \in D\}|
\]
The count of locations with no incoming derivation edges (source locations).
\end{definition}

\begin{theorem}[Derivation Lattice]\label{thm:derivation-lattice}
The set of derivation relations on a fixed set of locations $L$, ordered by inclusion, forms a bounded lattice:
\begin{enumerate}
\tightlist
\item \textbf{Bottom ($\bot$):} $D = \emptyset$ (no derivations, DOF = $|L|$)
\item \textbf{Top ($\top$):} Maximal acyclic $D$ with DOF = 1 (all but one location derived)
\item \textbf{Meet ($\land$):} $D_1 \land D_2 = D_1 \cap D_2$
\item \textbf{Join ($\lor$):} $D_1 \lor D_2 = $ transitive closure of $D_1 \cup D_2$ (if acyclic)
\end{enumerate}
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
\textbf{Bottom:} $\emptyset$ is trivially a derivation relation with all locations independent.

\textbf{Top:} For $n$ locations, a maximal acyclic relation has one source (root) and $n-1$ derived locations forming a tree or DAG. DOF = 1.

\textbf{Meet:} Intersection of acyclic relations is acyclic. The intersection preserves only derivations present in both.

\textbf{Join:} If $D_1 \cup D_2$ is acyclic, its transitive closure is the smallest relation containing both. If cyclic, join is undefined (partial lattice).

Bounded: $\emptyset \subseteq D \subseteq \top$ for all valid $D$.
\end{proof}

\begin{theorem}[DOF is Anti-Monotonic]\label{thm:dof-antimonotone}
DOF is anti-monotonic in the derivation lattice:
\[
D_1 \subseteq D_2 \Rightarrow \text{DOF}(D_1) \geq \text{DOF}(D_2)
\]
More derivations imply fewer independent locations.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Adding a derivation edge $(L_s, L_d)$ to $D$ can only decrease DOF: if $L_d$ was previously a source (no incoming edges), it now has an incoming edge and is no longer a source. Sources can only decrease or stay constant as derivations are added.
\end{proof}

\begin{corollary}[Minimal DOF = 1 Derivations]\label{cor:minimal-dof1}
A derivation relation $D$ with DOF($D$) = 1 is \emph{minimal} iff removing any edge increases DOF.
\end{corollary}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}}

\textbf{Computational implication:} Given an encoding system, there can be multiple DOF-1-achieving derivation structures. The minimal ones use the fewest derivation edges--the most economical way to achieve coherence.

\textbf{Representation model for complexity.} For the algorithmic results below, we assume the derivation relation $D$ is given explicitly as a DAG over the location set $L$. The input size is $|L| + |D|$, and all complexity bounds are measured in this explicit representation.

\begin{theorem}[DOF Computation Complexity]\label{thm:dof-complexity}
Given an encoding system with explicit derivation relation $D$:
\begin{enumerate}
\tightlist
\item Computing DOF($D$) is $O(|L| + |D|)$ (linear in locations plus edges)
\item Deciding if DOF($D$) = 1 is $O(|L| + |D|)$
\item Finding a minimal DOF-1 extension of $D$ is $O(|L|^2)$ in the worst case
\end{enumerate}
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
\textbf{(1) DOF computation:} Count locations with in-degree 0 in the DAG. Single pass over edges: $O(|D|)$ to compute in-degrees, $O(|L|)$ to count zeros.

\textbf{(2) DOF = 1 decision:} Compute DOF, compare to 1. Same complexity.

\textbf{(3) Minimal extension:} Must connect $k-1$ source locations to reduce DOF from $k$ to 1. Finding which connections preserve acyclicity requires reachability queries. Naive: $O(|L|^2)$. With better data structures (e.g., dynamic reachability): $O(|L| \cdot |D|)$ amortized.
\end{proof}

%==============================================================================
