\section{Derivation as the Capacity-Achieving Structure}\label{sec:ssot}
%==============================================================================

Having established the encoding model (Section~\ref{sec:epistemic}), we now identify the structural mechanism that attains the exact-consistency regime: derivation induces perfect dependence, so derived locations add visibility without adding independent rate.

\subsection{Rate 1 as the Exact-Consistency Regime}\label{sec:ssot-def}

Rate $1$ is not merely a design preference. In the present model it is the exact-consistency threshold for facts encoded in systems with modification constraints (\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}).

\begin{definition}[Optimal Encoding (DOF = 1)]\label{def:ssot}
Encoding system $C$ achieves \emph{optimal encoding rate} for fact $F$ iff:
\[
\text{DOF}(C, F) = 1
\]
Equivalently: exactly one independent encoding location exists for $F$. All other encodings are derived.
\end{definition}

\textbf{Interpretation.} The language of ``single authoritative source'' is an application-level reading of a more basic IT statement: one independent source plus deterministic derivations is the unique zero-incoherence regime.

Equivalently, rate 1 means one independent encoding location and only deterministic views elsewhere.

\begin{corollary}[DOF = 1 Guarantees Determinacy]\label{thm:ssot-determinate}
If $\text{DOF}(C, F) = 1$, then for all reachable states of $C$, the value of $F$ is determinate: all encodings agree. \emph{(This restates Theorem~\ref{thm:dof-one-coherence} in terms of determinacy.)}
\end{corollary}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

As an application-level interpretation, this recovers Hunt \& Thomas's ``single, unambiguous, authoritative representation''~\cite{hunt1999pragmatic}.

Our contribution here is to show that the capacity-achieving structure is exactly the one in which all non-source locations are deterministic functions of a single source (\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}).

\begin{corollary}[DOF = 1 Achieves $O(1)$ Update]\label{thm:ssot-optimal}
If $\text{DOF}(C, F) = 1$, then coherence restoration requires $O(1)$ updates. \emph{(See Theorem~\ref{thm:upper-bound}.)}
\end{corollary}
\leanmeta{\LH{UNQ1}}

\begin{corollary}[Uniqueness of Optimal Rate]\label{thm:ssot-unique}
DOF = 1 is the unique rate guaranteeing coherence. \emph{(See Theorem~\ref{thm:coherence-capacity}.)}
\end{corollary}
\leanmeta{\LH{L1}, \LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

\begin{corollary}[Incoherence Under Redundancy]\label{cor:no-redundancy}
Multiple independent sources encoding the same fact permit incoherent states. DOF $> 1 \Rightarrow$ incoherence reachable.
\end{corollary}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
Direct application of Theorem~\ref{thm:dof-gt-one-incoherence}. With DOF $> 1$, independent locations can be modified separately, reaching states where they disagree.
\end{proof}
See the formal Zero-Incoherence Capacity statement in Section~\ref{sec:capacity} (Theorem~\ref{thm:coherence-capacity}, \LH{UNQ1}). The remainder of this section focuses on why derivation is the relevant dependence structure.

\subsection{Information-theoretic proof sketches}
\label{sec:info-sketches}

We give concise proof sketches connecting the combinatorial DOF arguments to finite zero-error information-theoretic tools:

\begin{itemize}
\item \textbf{Counting converse:} To show that limited side information cannot enable exact recovery of $F$, apply the finite counting converse (Theorem~\ref{thm:fano-converse}, \LH{UNQ1}). If $K$ ambiguous states must be distinguished using an $L$-bit side-information tag, exact recovery requires $K\le 2^L$, equivalently $L\ge \log_2 K$.

\item \textbf{Confusability graph:} Lemma~\ref{lem:confusability-clique} converts combinatorial indistinguishability (cliques) into information lower bounds: in the finite zero-error model, a $k$-clique forces at least $k$ tag values (\LH{UNQ1}, \LH{UNQ1}); globally, the required tag alphabet is lower-bounded by both worst-case fiber size (\LH{UNQ1}) and the observation-alphabet occupancy floor (\LH{UNQ1}), equivalently $|\mathcal V|\le |\mathcal O|\cdot |\mathcal T|$ for any zero-error scheme (\LH{UNQ1}). This gives a language-independent certificate that side information is insufficient in environments with limited introspection.

\item \textbf{DOF implication:} If side information cannot supply enough distinguishable tags for exact recovery, the system must retain multiple independent encodings (higher DOF) to achieve correctness under modifications. Higher DOF incurs $\Omega(n)$ manual edits (Theorem~\ref{thm:lower-bound}, \LH{UNQ1}), completing the converse chain from information constraints to modification complexity.
\end{itemize}


\subsection{Rate-Complexity Tradeoff}\label{sec:ssot-vs-m}

The DOF metric creates a direct tradeoff between independent rate and manual synchronization cost.

\textbf{Question:} When fact $F$ changes, how many manual updates are required to restore coherence?

\begin{itemize}
\tightlist
\item \textbf{DOF = 1:} $O(1)$ updates. The single source determines all derived locations automatically.
\item \textbf{DOF = $n > 1$:} $\Omega(n)$ updates. Each independent location must be synchronized manually.
\end{itemize}

This is a rate-complexity consequence of the exact-consistency threshold: higher independent rate incurs higher synchronization cost, while DOF = 1 minimizes that cost.

\textbf{Key insight:} Many locations can encode $F$ (high total encoding locations), but if DOF = 1, coherence restoration requires only 1 manual update. The derivation mechanism handles propagation automatically.

\begin{example}[Encoding Rate vs. Modification Complexity]\label{ex:ssot-large-m}
Consider an encoding system where a fact $F$ = ``all processors must implement operation $P$'' is encoded at 51 locations:
\begin{itemize}
\tightlist
\item 1 abstract specification location
\item 50 concrete implementation locations
\end{itemize}

\textbf{Architecture A (DOF = 51):} All 51 locations are independent.
\begin{itemize}
\tightlist
\item Modification complexity: Changing $F$ requires 51 manual updates
\item Coherence risk: After $k < 51$ updates, system is incoherent (partial updates)
\item Only after all 51 updates is coherence restored
\end{itemize}

\textbf{Architecture B (DOF = 1):} The abstract specification is the single source; implementations are derived.
\begin{itemize}
\tightlist
\item Modification complexity: Changing $F$ requires 1 update (the specification)
\item Coherence guarantee: Derived locations update automatically via enforcement mechanism
\item The \emph{specification} has a single authoritative source
\end{itemize}

\textbf{Realization:} Abstract base classes with enforcement (type checkers, runtime validation) achieve DOF = 1 for contract specifications. Changing the abstract method signature updates the contract; type checkers flag non-compliant implementations.

Note: Implementations are separate facts. DOF = 1 for the contract specification does not eliminate implementation updates--it ensures the specification itself is determinate.
\end{example}

\subsection{Derivation: The Coherence Mechanism}\label{sec:derivation}

Derivation is the mechanism by which DOF is reduced without losing encodings. A derived location cannot diverge from its source, eliminating it as a source of incoherence.

\begin{definition}[Derivation]\label{def:derivation}
Location $L_{\text{derived}}$ is \emph{derived from} $L_{\text{source}}$ for fact $F$ iff:
\[
\text{updated}(L_{\text{source}}) \rightarrow \text{automatically\_updated}(L_{\text{derived}})
\]
No manual intervention is required. Coherence is maintained automatically.
\end{definition}

Derivation can occur at different times depending on the encoding system:

\begin{center}
\begin{tabularx}{\linewidth}{lY}
\toprule
\textbf{Derivation Time} & \textbf{Examples Across Domains} \\
\midrule
Compile/Build time & C++ templates, Rust macros, database schema generation, infrastructure-as-code compilation \\
Definition time & Python metaclasses, ORM model registration, dynamic schema creation \\
Query/Access time & Database views, computed columns, lazy evaluation \\
\bottomrule
\end{tabularx}
\end{center}

\textbf{Structural facts require definition-time derivation.} Structural facts (class existence, schema structure, service topology) are fixed when defined. Compile-time derivation that runs before the definition is fixed is too early (the declarative source is not yet fixed). Runtime is too late (structure already immutable). Definition-time is the unique opportunity for structural derivation---a consequence of timing trichotomy (\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}).

\begin{theorem}[Derivation Preserves Coherence]\label{thm:derivation-excludes}
If $L_{\text{derived}}$ is derived from $L_{\text{source}}$, then $L_{\text{derived}}$ cannot diverge from $L_{\text{source}}$ and does not contribute to DOF.
\end{theorem}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
By Definition~\ref{def:derivation}, derived locations are automatically updated when the source changes. Let $L_d$ be derived from $L_s$. If $L_s$ encodes value $v$, then $L_d$ encodes $f(v)$ for some function $f$. When $L_s$ changes to $v'$, $L_d$ automatically changes to $f(v')$.

There is no reachable state where $L_s = v'$ and $L_d = f(v)$ with $v' \neq v$. Divergence is impossible. Therefore, $L_d$ does not contribute to DOF.
\end{proof}

\begin{corollary}[Derivation Achieves Coherence]\label{cor:metaprogramming}
If all encodings of $F$ except one are derived from that one, then $\text{DOF}(C, F) = 1$ and coherence is guaranteed.
\end{corollary}
\leanmeta{\LH{UNQ1}, \LH{UNQ1}}

\begin{proof}
Let $L_s$ be the non-derived encoding. All other encodings $L_1, \ldots, L_k$ are derived from $L_s$. By Theorem~\ref{thm:derivation-excludes}, none can diverge. Only $L_s$ is independent. Therefore, $\text{DOF}(C, F) = 1$, and by Theorem~\ref{thm:dof-one-coherence}, coherence is guaranteed.
\end{proof}

\subsection{Illustrative Realizations}\label{sec:ssot-patterns}

This subsection is illustrative only. It shows how the dependence structure appears in concrete systems, but the theorem itself is independent of these realizations.

\textbf{Software: Subclass Registration (Python)}

\begin{code}
class Registry:
    _registry = {}
    def __init_subclass__(cls, **kwargs):
        Registry._registry[cls.__name__] = cls

class PNGHandler(Registry):  # Automatically registered
    pass
\end{code}

\textbf{Encoding structure:}
\begin{itemize}
\tightlist
\item Source: Class definition (declared once)
\item Derived: Registry dictionary entry (computed at definition time via \texttt{\_\_init\_subclass\_\_})
\item DOF = 1: Registry cannot diverge from class hierarchy
\end{itemize}

\textbf{Databases: Materialized Views}

\begin{code}
CREATE TABLE users (id INT, name TEXT, created_at TIMESTAMP);
CREATE MATERIALIZED VIEW user_count AS
  SELECT COUNT(*) FROM users;
\end{code}

\textbf{Encoding structure:}
\begin{itemize}
\tightlist
\item Source: Base table \texttt{users}
\item Derived: Materialized view \texttt{user\_count} (updated on refresh)
\item DOF = 1: View cannot diverge from base table (consistency guaranteed by DBMS)
\end{itemize}

\textbf{Configuration: Infrastructure as Code (Terraform)}

\begin{code}
resource "aws_instance" "app" {
  ami = "ami-12345"
  instance_type = "t2.micro"
}

output "instance_ip" {
  value = aws_instance.app.public_ip
}
\end{code}

\textbf{Encoding structure:}
\begin{itemize}
\tightlist
\item Source: Resource declaration (authoritative configuration)
\item Derived: Output value (computed from resource state)
\item DOF = 1: Output cannot diverge from actual resource (computed at apply time)
\end{itemize}

\textbf{Common pattern:} In all cases, one location carries the independent value and the others are deterministic views computed at definition/build/query time. Manual synchronization disappears because the dependence structure, not human process, enforces consistency.

%==============================================================================
