\section{Information-Theoretic Realizability Requirements}\label{sec:requirements}
%==============================================================================

This section identifies, at an abstract information-theoretic level, the two properties an encoding system must provide to realize DOF = 1. Concrete programming-language and system-level instantiations are presented in the supplementary material; here we present the abstract requirements and their necessity/sufficiency proofs.

Informally, an encoding system provides a set of locations holding values and a space of allowed modifications. DOF = 1 is realizable when exactly one independent location exists (the source) and all other encodings are deterministic functions (derivations) of that source so divergence is impossible.

We show that two encoder properties are necessary and sufficient (\LH{SOT1}):
\begin{enumerate}
\item \textbf{Causal update propagation:} updates to the source must automatically propagate to derived locations (eliminating transient incoherence windows);
\item \textbf{Provenance observability:} the system must expose, to verification procedures, which locations are sources and which are derived (providing the side information necessary to verify DOF = 1).
\end{enumerate}

The remainder of this section gives formal statements and proofs of necessity and sufficiency for these properties in the abstract model.

\subsection{Confusability and Side-Information}
\label{sec:confusability}

To connect realizability to the information available at runtime, define the \emph{confusability graph} for a fact $F$: vertices are distinct values of $F$, and an undirected edge joins two values $x,x'$ iff they are indistinguishable given the available side information (i.e., no observation of derived locations can reliably separate $x$ from $x'$).

\begin{lemma}[Confusability Clique Bound]\label{lem:confusability-clique}
If the confusability graph for $F$ contains a clique of size $k$, then any side-information channel that yields vanishing error probability for distinguishing all $k$ values must convey at least $\log k$ bits about $F$. Equivalently, if the mutual information between $F$ and side information $S$ satisfies $I(F;S)<\log k - \epsilon$, then the error probability for distinguishing the $k$-clique is bounded away from zero.
\end{lemma}

\begin{proof}[Proof sketch]
Restrict $F$ to the $k$ values in the clique; this induces a uniform $k$-ary hypothesis testing problem. Apply the Fano-style bound (Theorem~\ref{thm:fano-converse} in Section~\ref{sec:info-converse}) to conclude that vanishing error requires $I(F;S)\ge\log k - o(1)$. If $I(F;S)$ is strictly smaller than $\log k$, Fano's inequality implies a non-zero lower bound on error probability.
\end{proof}

\noindent The confusability graph provides a compact combinatorial certificate of necessary side-information: large cliques force large information requirements and hence constrain realizability in languages/environments with limited introspection or definition-time hooks (\LH{CIA3}).

\subsection{The Structural Timing Constraint}\label{sec:timing}

For certain classes of facts--\emph{structural facts}--there is a fundamental timing constraint that shapes realizability.

\begin{definition}[Structural Fact]\label{def:structural-fact-req}
A fact $F$ is \emph{structural} if its encoding locations are fixed at the moment of definition. After definition, the structure cannot be retroactively modified--only new structures can be created.
\end{definition}

\textbf{Examples across domains:}
\begin{itemize}
\tightlist
\item \textbf{Programming languages}: Class definitions, method signatures, inheritance relationships
\item \textbf{Databases}: Schema definitions, table structures, foreign key constraints
\item \textbf{Configuration systems}: Resource declarations, dependency specifications
\item \textbf{Version control}: Branch structures, commit ancestry
\end{itemize}

The key property: structural facts have a \emph{definition moment} after which their encoding is immutable. This creates a timing constraint for derivation.

\begin{theorem}[Timing Constraint for Structural Derivation]\label{thm:timing-forces}
For structural facts, derivation must occur at or before the moment the structure is fixed.
\end{theorem}

\begin{proof}
Let $F$ be a structural fact. Let $t_{\text{fix}}$ be the moment $F$'s encoding is fixed. Any derivation $D$ that depends on $F$ must execute at some time $t_D$.

\textbf{Case 1}: $t_D < t_{\text{fix}}$. Derivation executes before $F$ is fixed. $D$ cannot derive from $F$ because $F$ does not yet exist.

\textbf{Case 2}: $t_D > t_{\text{fix}}$. Derivation executes after $F$ is fixed. $D$ can read $F$ but cannot modify structures derived from $F$--they are already fixed.

\textbf{Case 3}: $t_D = t_{\text{fix}}$. Derivation executes at the moment $F$ is fixed. $D$ can both read $F$ and create derived structures before they are fixed.

Therefore, structural derivation requires $t_D = t_{\text{fix}}$.
\end{proof}

This timing constraint is the information-theoretic reason why derivation must be \emph{causal}--triggered by the act of defining the source, not by later access (\LH{REQ1}, \LH{REQ2}).

%==============================================================================
\subsection{Requirement 1: Causal Update Propagation}\label{sec:causal-propagation}
%==============================================================================

\begin{definition}[Causal Update Propagation]\label{def:causal-propagation}
An encoding system has \emph{causal update propagation} if changes to a source location automatically trigger updates to all derived locations, without requiring explicit synchronization commands.

Formally: let $L_s$ be a source location and $L_d$ a derived location. The system has causal propagation iff:
\[
\text{update}(L_s, v) \Rightarrow \text{automatically\_updated}(L_d, f(v))
\]
where $f$ is the derivation function. No separate ``propagate'' or ``sync'' operation is required.
\end{definition}

\textbf{Information-theoretic interpretation:} Causal propagation is analogous to \emph{channel coding with feedback}. In classical channel coding, the encoder sends a message and waits for acknowledgment. With feedback, the encoder can immediately react to channel state. Causal propagation provides ``feedback'' from the definition event to the derivation mechanism--the encoder (source) and decoder (derived locations) are coupled in real-time.

\textbf{Connection to multi-version coding:} Rashmi et al.~\cite{rashmi2015multiversion} formalize consistent distributed storage where updates to a source must propagate to replicas while maintaining consistency. Their ``multi-version code'' requires that any $c$ servers can decode the latest common version--a consistency guarantee analogous to our coherence requirement. Causal propagation is the mechanism by which this consistency is maintained under updates.

\textbf{Why causal propagation is necessary:}

Without causal propagation, there exists a temporal window between source modification and derived location update. During this window, the system is incoherent--the source and derived locations encode different values.

\begin{theorem}[Causal Propagation is Necessary for DOF = 1]\label{thm:causal-necessary}
Achieving DOF = 1 for structural facts requires causal update propagation.
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:timing-forces}, structural derivation must occur at definition time. Without causal propagation, derived locations are not updated when the source is defined. This means:
\begin{enumerate}
\tightlist
\item The source exists with value $v$
\item Derived locations have not been updated; they either do not exist yet or hold stale values
\item The system is temporarily incoherent
\end{enumerate}

For DOF = 1, incoherence must be \emph{impossible}, not merely transient. Causal propagation eliminates the temporal window: derived locations are updated \emph{as part of} the source definition, not after.

Contrapositive: If an encoding system lacks causal propagation, DOF = 1 for structural facts is unrealizable.
\end{proof}

\textbf{Realizations across domains:}

\begin{center}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Domain} & \textbf{Causal Propagation Mechanism} \\
\midrule
Python & \texttt{\_\_init\_subclass\_\_}, metaclass \texttt{\_\_new\_\_} \\
CLOS & \texttt{:after} methods on class initialization \\
Smalltalk & Class creation protocol, \texttt{subclass:} method \\
Databases & Triggers on schema operations (PostgreSQL event triggers) \\
Distributed systems & Consensus protocols (Paxos, Raft) \\
Configuration & Terraform dependency graph, reactive bindings \\
\bottomrule
\end{tabularx}
\end{center}

\textbf{Systems lacking causal propagation:}
\begin{itemize}
\tightlist
\item \textbf{Java}: Annotations are metadata, not executable. No code runs at class definition.
\item \textbf{C++}: Templates expand at compile time but don't execute arbitrary user code.
\item \textbf{Go}: No hook mechanism. Interface satisfaction is implicit.
\item \textbf{Rust}: Proc macros run at compile time but generate static code, not runtime derivation.
\end{itemize}

%==============================================================================
\subsection{Requirement 2: Provenance Observability}\label{sec:provenance-observability}
%==============================================================================

\begin{definition}[Provenance Observability]\label{def:provenance-observability}
An encoding system has \emph{provenance observability} if the system supports queries about derivation structure:
\begin{enumerate}
\tightlist
\item What locations exist encoding a given fact?
\item Which locations are sources vs. derived?
\item What is the derivation relationship (which derived from which)?
\end{enumerate}
\end{definition}

\textbf{Information-theoretic interpretation:} Provenance observability is the encoding-system analog of \emph{side information at the decoder}. In Slepian-Wolf coding~\cite{slepian1973noiseless}, the decoder has access to correlated side information that enables decoding at rates below the source entropy. Provenance observability provides ``side information'' about the encoding structure itself--enabling verification that DOF = 1 holds.

Without provenance observability, the encoding system is a ``black box''--you can read locations but cannot determine which are sources and which are derived. This makes DOF uncomputable from within the system.

\begin{theorem}[Provenance Observability is Necessary for Verifiable DOF = 1]\label{thm:provenance-necessary}
Verifying that DOF = 1 holds requires provenance observability.
\end{theorem}

\begin{proof}
Verification of DOF = 1 requires confirming:
\begin{enumerate}
\tightlist
\item All locations encoding fact $F$ are enumerable
\item Exactly one location is independent (the source)
\item All other locations are derived from that source
\end{enumerate}

Step (1) requires querying what structures exist. Step (2) requires distinguishing sources from derived locations. Step (3) requires querying the derivation relationship.

Without provenance observability, none of these queries are answerable from within the system. DOF = 1 can hold but cannot be verified. Bugs in derivation logic go undetected until coherence violations manifest.
\end{proof}

\textbf{Connection to coding theory:} In coding theory, a code's structure (generator matrix, parity-check matrix) must be known to the decoder. Provenance observability is analogous: the derivation structure must be queryable for verification.

\textbf{Realizations across domains:}

\begin{center}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Domain} & \textbf{Provenance Observability Mechanism} \\
\midrule
Python & \texttt{\_\_subclasses\_\_()}, \texttt{\_\_mro\_\_}, \texttt{dir()}, \texttt{vars()} \\
CLOS & \texttt{class-direct-subclasses}, MOP introspection \\
Smalltalk & \texttt{subclasses}, \texttt{allSubclasses} \\
Databases & System catalogs (\texttt{pg\_depend}), query plan introspection \\
Distributed systems & Vector clocks, provenance tracking, \texttt{etcd} watch \\
Configuration & Terraform \texttt{graph}, Kubernetes API server \\
\bottomrule
\end{tabularx}
\end{center}

\textbf{Systems lacking provenance observability:}
\begin{itemize}
\tightlist
\item \textbf{C++}: Cannot query ``what types instantiated template \texttt{Foo<T>}?''
\item \textbf{Rust}: Proc macro expansion is opaque at runtime.
\item \textbf{TypeScript}: Types are erased. Runtime cannot query type relationships.
\item \textbf{Go}: No type registry. Cannot enumerate interface implementations.
\end{itemize}

%==============================================================================
\subsection{Independence of Requirements}\label{sec:independence}
%==============================================================================

The two requirements--causal propagation and provenance observability--are independent. Neither implies the other (\LH{IND1}, \LH{IND2}).

\begin{theorem}[Requirements are Independent]\label{thm:independence}
\begin{enumerate}
\tightlist
\item An encoding system can have causal propagation without provenance observability
\item An encoding system can have provenance observability without causal propagation
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{(1) Causal without provenance:} Rust proc macros execute at compile time (causal propagation: definition triggers code generation). But the generated code is opaque at runtime--the program cannot query what was generated (no provenance observability).

\textbf{(2) Provenance without causal:} Java provides reflection (\texttt{Class.getMethods()}, \texttt{Class.getInterfaces()})--provenance observability. But no code executes when a class is defined--no causal propagation.
\end{proof}

This independence means both requirements must be satisfied for DOF = 1 realizability (\LH{SOT1}).

%==============================================================================
\subsection{The Realizability Theorem}\label{sec:realizability-theorem}
%==============================================================================

\begin{theorem}[Necessary and Sufficient Realizability Conditions]\label{thm:ssot-iff}
An encoding system $S$ can achieve verifiable DOF = 1 for structural facts if and only if:
\begin{enumerate}
\tightlist
\item $S$ provides causal update propagation, AND
\item $S$ provides provenance observability
\end{enumerate}
\end{theorem}

\begin{proof}
$(\Rightarrow)$ \textbf{Necessity:} Suppose $S$ achieves verifiable DOF = 1 for structural facts.
\begin{itemize}
\tightlist
\item By Theorem~\ref{thm:causal-necessary}, $S$ must provide causal propagation
\item By Theorem~\ref{thm:provenance-necessary}, $S$ must provide provenance observability
\end{itemize}

$(\Leftarrow)$ \textbf{Sufficiency:} Suppose $S$ provides both capabilities.
\begin{itemize}
\tightlist
\item Causal propagation enables derivation at the right moment (when structure is fixed)
\item Provenance observability enables verification that all secondary encodings are derived
\item Therefore, DOF = 1 is achievable: create one source, derive all others causally, verify completeness via provenance queries
\end{itemize}
\end{proof}

\begin{definition}[DOF-1-Complete Encoding System]\label{def:dof-complete}
An encoding system is \emph{DOF-1-complete} if it satisfies both causal propagation and provenance observability. Otherwise it is \emph{DOF-1-incomplete}.
\end{definition}

\textbf{Information-theoretic interpretation:} DOF-1-completeness is analogous to \emph{channel capacity achievability}. A channel achieves capacity if there exist codes that approach the Shannon limit. An encoding system is DOF-1-complete if there exist derivation mechanisms that achieve the coherence-optimal rate (DOF = 1). The two requirements (causal propagation, provenance observability) are the ``channel properties'' that enable capacity achievement.

%==============================================================================
\subsection{Connection to Write-Once Memory Codes}\label{sec:wom-connection}
%==============================================================================

Our realizability requirements connect to \emph{write-once memory (WOM) codes}~\cite{rivest1982wom,wolf1984wom}, an established area of coding theory.

A WOM is a storage medium where bits can only transition in one direction (typically $0 \to 1$). Rivest and Shamir~\cite{rivest1982wom} showed that WOMs can store more information than their apparent capacity by encoding multiple ``writes'' cleverly--the capacity for $t$ writes is $\log_2(t+1)$ bits per cell.

The connection to our framework:
\begin{itemize}
\tightlist
\item \textbf{WOM constraint}: Bits can only increase (irreversible state change)
\item \textbf{Structural fact constraint}: Structure is fixed at definition (irreversible encoding)
\item \textbf{WOM coding}: Clever encoding enables multiple logical writes despite physical constraints
\item \textbf{DOF = 1 derivation}: Clever derivation enables multiple logical locations from one physical source
\end{itemize}

Both settings involve achieving optimal encoding under irreversibility constraints. WOM codes achieve capacity via coding schemes; DOF-1-complete systems achieve coherence via derivation mechanisms.

%==============================================================================
\subsection{The Logical Chain (Summary)}\label{sec:chain}
%==============================================================================

\begin{center}
\fbox{\parbox{0.95\columnwidth}{
\textbf{Observation:} Structural facts are fixed at definition time (irreversible encoding).

$\downarrow$ (timing analysis)

\textbf{Theorem~\ref{thm:timing-forces}:} Derivation for structural facts must occur at definition time.

$\downarrow$ (requirement derivation)

\textbf{Theorem~\ref{thm:causal-necessary}:} Causal update propagation is necessary for DOF = 1.

\textbf{Theorem~\ref{thm:provenance-necessary}:} Provenance observability is necessary for verifiable DOF = 1.

$\downarrow$ (conjunction)

\textbf{Theorem~\ref{thm:ssot-iff}:} An encoding system achieves DOF = 1 iff it has both properties.

$\downarrow$ (evaluation)

\textbf{Classification:} Python, CLOS, Smalltalk are DOF-1-complete. Java, C++, Rust, Go are DOF-1-incomplete.
}}
\end{center}

\textbf{Mechanization status.} The core requirement chain is machine-checked in Lean 4; current build statistics are \LeanTotalLines\ lines, \LeanTotalTheorems\ theorem/lemma statements, and \LeanTotalSorry\ \texttt{sorry} placeholders.

%==============================================================================
\subsection{Concrete Impossibility Demonstration}\label{sec:impossibility}
%==============================================================================

We demonstrate exactly why DOF-1-incomplete systems cannot achieve DOF = 1 for structural facts.

\textbf{The structural fact:} ``\texttt{PNGHandler} handles \texttt{.png} files.''

This fact must be encoded in two places:
\begin{enumerate}
\tightlist
\item The handler definition (where the handler is defined)
\item The registry/dispatcher (where format$\to$handler mapping lives)
\end{enumerate}

\textbf{Python (DOF-1-complete) achieves DOF = 1:}

\begin{code}
class ImageHandler:
    _registry = {}

    def __init_subclass__(cls, format=None, **kwargs):
        super().__init_subclass__(**kwargs)
        if format:
            ImageHandler._registry[format] = cls  # DERIVED (causal)

class PNGHandler(ImageHandler, format="png"):  # SOURCE
    def load(self, path): ...
\end{code}

\textbf{Causal propagation:} When \texttt{class PNGHandler} executes, \texttt{\_\_init\_subclass\_\_} fires immediately, adding the registry entry. No temporal gap.

\textbf{Provenance observability:} \texttt{ImageHandler.\_\_subclasses\_\_()} returns all handlers. The derivation structure is queryable.

\textbf{DOF = 1}: The \texttt{format="png"} in the class definition is the single source. The registry entry is derived causally. Adding a new handler requires changing exactly one location.

\textbf{Java (DOF-1-incomplete) cannot achieve DOF = 1:}

\begin{code}
// File 1: PNGHandler.java
@Handler(format = "png")  // Metadata, not executable
public class PNGHandler implements ImageHandler { ... }

// File 2: HandlerRegistry.java (SEPARATE SOURCE)
public class HandlerRegistry {
    static { register("png", PNGHandler.class); }  // Manual
}
\end{code}

\textbf{No causal propagation:} The \texttt{@Handler} annotation is data, not code. Nothing executes when the class is defined.

\textbf{Provenance partially present:} Java has reflection, but cannot enumerate ``all classes with \texttt{@Handler}'' without classpath scanning.

\textbf{DOF = 2}: The annotation and the registry are independent encodings. Either can be modified without the other. Incoherence is reachable.

\begin{theorem}[Generated Files Are Independent Sources]\label{thm:generated-second}
A generated source file constitutes an independent encoding, not a derivation. Code generation does not achieve DOF = 1.
\end{theorem}

\begin{proof}
Let $E_1$ be the annotation on \texttt{PNGHandler.java}. Let $E_2$ be the generated \texttt{HandlerRegistry.java}.

Test: If $E_2$ is deleted or modified, does system behavior change? \textbf{Yes}--the handler is not registered.

Test: Can $E_2$ diverge from $E_1$? \textbf{Yes}--$E_2$ is a separate file that can be edited, fail to generate, or be stale.

Therefore, $E_1$ and $E_2$ are independent encodings. The fact that $E_2$ was \emph{generated from} $E_1$ does not make it derived in the DOF sense, because:
\begin{enumerate}
\tightlist
\item $E_2$ exists as a separate artifact that can diverge
\item The generation process is external to the runtime and can be bypassed
\item There is no causal coupling--modification of $E_1$ does not automatically update $E_2$
\end{enumerate}

Contrast with Python: the registry entry exists only in memory, created causally by the class statement. There is no second file. DOF = 1.
\end{proof}

%==============================================================================
\subsection{Summary: The Information-Theoretic Requirements}\label{sec:req-summary}
%==============================================================================

\begin{center}
\begin{tabularx}{\linewidth}{lYY}
\toprule
\textbf{Requirement} & \textbf{IT Interpretation} & \textbf{Why Necessary} \\
\midrule
Causal propagation & Channel with feedback; encoder-decoder coupling & Eliminates temporal incoherence window \\
Provenance observability & Side information at decoder; codebook visibility & Enables DOF verification \\
\bottomrule
\end{tabularx}
\end{center}

Both requirements are necessary. Neither is sufficient alone. Together they enable DOF-1-complete encoding systems that achieve the coherence-optimal rate.

%==============================================================================
