\section{Rate-Complexity Bounds}\label{sec:bounds}
%==============================================================================

This section translates the exact-consistency threshold into an operational consequence: independent rate governs manual synchronization cost. The key point is not merely that rate $1$ is coherent, but that moving above that threshold creates a provable complexity gap that grows with system size (\LH{UNQ1}).

\subsection{Cost Model}\label{sec:cost-model}

\begin{definition}[Modification Cost Model]\label{def:cost-model}
Let $\delta_F$ be a modification to fact $F$ in encoding system $C$. The \emph{effective modification complexity} $M_{\text{effective}}(C, \delta_F)$ is the number of syntactically distinct edit operations that must be performed manually. Formally:
\[
M_{\text{effective}}(C, \delta_F) = |\{L \in \text{Locations}(C) : \text{requires\_manual\_edit}(L, \delta_F)\}|
\]
where $\text{requires\_manual\_edit}(L, \delta_F)$ holds iff location $L$ must be updated manually (not by automatic derivation) to maintain coherence after $\delta_F$.
\end{definition}

\textbf{Unit of cost:} One edit = one syntactic modification to one location. We count locations, not keystrokes or characters. This abstracts over edit complexity to focus on the scaling behavior.

\textbf{What we measure:} Manual edits only. Derived locations that update automatically have zero cost. This isolates the operational consequence of independent information: if a location carries its own freely editable value, it contributes to synchronization burden.

\textbf{Asymptotic parameter:} We measure scaling in the number of encoding locations for fact $F$. Let $n = |\{L \in C : \text{encodes}(L, F)\}|$ and $k = \text{DOF}(C, F)$. Bounds of $O(1)$ and $\Omega(n)$ are in this parameter; in particular, the lower bound uses $n = k$ independent locations.

\subsection{Upper Bound: DOF = 1 Achieves O(1)}\label{sec:upper-bound}

\begin{theorem}[DOF = 1 Upper Bound]\label{thm:upper-bound}
For an encoding system with DOF = 1 for fact $F$:
\[
M_{\text{effective}}(C, \delta_F) = O(1)
\]
Effective modification complexity is constant regardless of system size.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Let $\text{DOF}(C, F) = 1$. By Definition~\ref{def:ssot}, $C$ has exactly one independent encoding location. Let $L_s$ be this single independent location.

When $F$ changes:
\begin{enumerate}
\tightlist
\item Update $L_s$ (1 manual edit)
\item All derived locations $L_1, \ldots, L_k$ are automatically updated by the derivation mechanism
\item Total manual edits: 1
\end{enumerate}

The number of derived locations $k$ can grow with system size, but the number of \emph{manual} edits remains 1. Therefore, $M_{\text{effective}}(C, \delta_F) = O(1)$.
\end{proof}

\textbf{Note on ``effective'' vs. ``total'' complexity:} Total modification complexity $M(C, \delta_F)$ counts all locations that change. Effective modification complexity counts only manual edits. With DOF = 1, total complexity can be $O(n)$ (many derived locations change), but effective complexity is $O(1)$ (one manual edit).

\subsection{Lower Bound: DOF $>$ 1 Requires \texorpdfstring{$\Omega(n)$}{Omega(n)}}\label{sec:lower-bound}

\begin{theorem}[DOF $>$ 1 Lower Bound]\label{thm:lower-bound}
For an encoding system with DOF $>$ 1 for fact $F$, if $F$ is encoded at $n$ independent locations:
\[
M_{\text{effective}}(C, \delta_F) = \Omega(n)
\]
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Let $\text{DOF}(C, F) = n$ where $n > 1$.

By Definition~\ref{def:independent}, the $n$ encoding locations are independent--updating one does not automatically update the others. When $F$ changes:
\begin{enumerate}
\tightlist
\item Each of the $n$ independent locations must be updated manually
\item No automatic propagation exists between independent locations
\item Total manual edits: $n$
\end{enumerate}

Therefore, $M_{\text{effective}}(C, \delta_F) = \Omega(n)$.
\end{proof}

\subsection{Information-theoretic converse (finite counting)}\label{sec:info-converse}

To make the converse direction genuinely information-theoretic, we give a finite zero-error counting converse showing that insufficient side-information alphabet size makes exact recovery impossible (\LH{UNQ1}, \LH{UNQ1}). This is the bridge from ambiguity to complexity: if side information is insufficient, exact resolution fails, and systems that still demand correctness must pay through additional independent encodings and manual synchronization.

\begin{remark}[Formalization status]
The results in this subsection are formalized in the source Lean tree as finite zero-error counting theorems rather than probabilistic Fano inequalities. The core capacity, side-information, realizability, and complexity chains are machine-checked in that exact form.
\end{remark}

\begin{theorem}[Finite Counting Converse]\label{thm:fano-converse}
Let $F$ range over $K$ possible values. Suppose exact resolution is attempted using an $L$-bit side-information tag and a decoder that must recover the correct value with zero error from that tag and a fixed observation transcript. Then
\[
K \le 2^L.
\]
Equivalently, exact recovery of $K$ ambiguous states requires at least $\log_2 K$ bits of side information.
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}[Proof sketch]
There are only $2^L$ possible tags. In the strengthened Lean observer/tag model, exact recovery is equivalent to injectivity of the combined observation-plus-tag map (\LH{OBS1}, \LH{OBS2}). If the observation transcript is fixed, this reduces to injectivity of the tag map on the ambiguous states themselves. Hence two distinct values cannot share one tag under zero-error decoding, and injectivity is possible only when $K\le 2^L$.
\end{proof}

\noindent This converse formalizes the intuition used in the main text: unless the side-information channel has enough distinct tags to separate all ambiguous states, exact recovery is impossible. The modification-complexity consequence comes from the architectural fallback: when exact recovery cannot be certified from side information, systems that still require correctness must maintain or consult additional independent sources, which raises DOF and therefore update cost.

\subsection{Information-constrained DOF lower bound}
\label{sec:info-dof}

The following lemma makes that link explicit.

\begin{lemma}[Information-constrained DOF Lower Bound]\label{lem:info-dof}
Let $F$ range over $K$ possible values. If the available side-information scheme exposes at most $2^L$ distinguishable tags with $K>2^L$, then no zero-error resolver can recover $F$ from those tags alone. Any architecture that nevertheless guarantees correctness must therefore retain additional independent encodings for some subset of values, hence DOF $>1$ on that subset. Consequently, for facts encoded at $n$ independent locations the effective modification complexity satisfies $M_{\text{effective}}=\Omega(n)$.
\end{lemma}
\leanmeta{\LH{UNQ1}}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:fano-converse}, if $K>2^L$ then the side-information tags are too small to support exact recovery of all $K$ possibilities. The strengthened Lean architecture layer makes this explicit: every exact-recovery architecture carries a formal support count, and that support count is lower-bounded both globally by the observation/support product bound (\LH{OBS7}) and locally by any confusability clique (\LH{OBS8}, \LH{OBS9}). In particular, observation-only architectures (support count $=1$) are impossible on nontrivial cliques (\LH{OBS10}). The new bridge theorem then identifies this support count with a concrete instance of the paper's `Dof.dof` semantics under a derivation-free support architecture (\LH{OBS11}); hence nontrivial cliques force a formal DOF strictly above one (\LH{OBS12}). Therefore, if correctness is still required under updates, the architecture must expose additional independently accessible discriminating support; in the paper's DOF reading, this means moving above the single-source regime. Once DOF$>1$ over a set of $n$ independent locations, Theorem~\ref{thm:lower-bound} implies $M_{\text{effective}}=\Omega(n)$.
\end{proof}

\subsection{Worked example: finite tag budget}
\label{sec:worked-mi}

Consider a fact $F$ with $K=4$ possible values. Suppose the available side-information mechanism supplies only a 1-bit tag, so at most $2^1=2$ distinct tags are available.

Apply Theorem~\ref{thm:fano-converse}. Exact recovery would require
\[
4 \le 2^1,
\]
which is false. So 1 bit of side information cannot resolve the 4-way ambiguity exactly.

Interpretation: with only 2 available tags for 4 possible values, some pair of values must collide. Therefore, to guarantee exact recovery one must either supply more side information (increase $L$ until $2^L\ge 4$) or enrich the architecture with a nontrivial auxiliary channel; in the paper's systems reading, that means additional independently accessible discriminating information, which in turn raises manual modification costs per Theorem~\ref{thm:lower-bound}.

By contrast, a 2-bit tag budget gives $2^2=4$ tags, which is just enough to distinguish all four values in principle.

The same reasoning extends to a finite relaxed-error setting where exact recovery is only required on a designated subset of states: any confusability clique that must still be recovered exactly obeys the same tag-alphabet lower bound (\LH{OBS13}).

More generally, the source proof tree now contains an explicit finite distribution/error layer. For a finite source distribution with point-mass ceiling $m$, the total probability of exactly recovered states is bounded by $(|\mathcal O|\cdot |\mathcal T|)m$ (\LH{PRB2}), so the error probability is bounded below by $1-(|\mathcal O|\cdot |\mathcal T|)m$ (\LH{PRB3}). For families of instances with a uniform budget-mass gap, this yields a fully mechanized non-vanishing-error theorem (\LH{PRB6}).

For the uniform finite source, these bounds sharpen into a direct weak-Fano-style converse: success probability is at most the observation/tag budget fraction (\LH{PRB8}), hence any decoder with error at most $\varepsilon$ must satisfy
\[
1-\varepsilon \le \frac{|\mathcal O|\,|\mathcal T|}{K},
\]
formalized in Lean as \LH{PRB9}. The strengthened source proofs also identify the exact success probability under the uniform source with the fraction of states decoded exactly (\LH{PRB11}, \LH{PRB12}), yielding a cardinality form of the same converse: low error forces a large exactly recovered subset (\LH{PRB13}), which is then bounded by the observation/tag budget to recover the weak-Fano inequality (\LH{PRB14}). Families with a uniform gap below this threshold cannot have vanishing error (\LH{PRB10}).

The entropy layer now goes one step further: a fully mechanized partition identity shows that for the uniform source, the success/failure split satisfies
\[
\log K = h(P_e) + (1-P_e)\log |S| + P_e\log |F|,
\]
where $S$ is the exactly recovered subset and $F$ its complement (\LH{PRB15}). Combining this identity with the exact success-set budget bound yields an explicit finite Fano-style inequality in the source proof tree:
\[
\log K \le h(P_e) + (1-P_e)\log(|\mathcal O||\mathcal T|) + P_e\log(K-1)
\]
for the uniform finite source with nontrivial success/failure partition (\LH{PRB16}). In the observation-only specialization this reduces to a $q$-ary entropy bound (\LH{PRB17}).

The source proofs then sharpen the observation-only case further: for a uniform $K$-state source, observation-only recovery succeeds with probability at most $1/K$ (\LH{PRB18}) and therefore has error probability at least $1-1/K$ (\LH{PRB19}).

Beyond the uniform case, the probabilistic layer now internalizes the source distribution itself through its maximum atom mass. For any finite source distribution, success probability is bounded by $(|\mathcal O||\mathcal T|)$ times the source's own max-mass parameter (\LH{PRB20}), giving an intrinsic error lower bound (\LH{PRB21}) and a weak-Fano-style converse without an externally supplied mass cap (\LH{PRB22}). Families with a uniform budget--max-mass gap cannot achieve vanishing error (\LH{PRB23}).

The same intrinsic max-mass layer also yields a structural cardinality consequence: the exactly recovered subset must have size at least $(1-\varepsilon)/m_{\max}$ whenever the error probability is at most $\varepsilon$ (\LH{PRB24}, \LH{PRB25}). In the observation-only specialization, success is at most $m_{\max}$ and error is at least $1-m_{\max}$ for arbitrary finite sources (\LH{PRB26}, \LH{PRB27}).

Finally, the arbitrary-source converse can be rewritten in min-entropy form. Defining $H_{\infty}(X)=-\log m_{\max}$, the source proofs now show
\[
P_{\mathrm{succ}} \le |\mathcal O||\mathcal T|\, e^{-H_{\infty}(X)},
\qquad
P_e \ge 1-|\mathcal O||\mathcal T|\, e^{-H_{\infty}(X)},
\]
formalized as \LH{PRB28} and \LH{PRB29}. This gives a clean min-entropy interpretation of the finite-source converse.

The min-entropy layer also yields an exact-recovery-set interpretation: if error is at most $\varepsilon$, then the exactly recovered subset must have size at least $(1-\varepsilon)e^{H_{\infty}(X)}$ (\LH{PRB32}), and therefore the available observation/tag budget must satisfy the same lower bound (\LH{PRB33}).

The source layer also now defines a Shannon-style finite entropy for arbitrary finite sources and proves the entropy sandwich $H_{\infty}(X)\le H(X)$ (\LH{PRB31}), with the uniform finite source attaining $H(X)=\log K$ exactly (\LH{PRB30}). This gives a clean conceptual relation between the earlier uniform-source Fano bounds and the later arbitrary-source min-entropy converses.

It also now records the finite-cardinality ceiling directly at the source level: every $K$-state source satisfies $1/K\le m_{\max}$ (\LH{PRB34}), hence
\[
H_{\infty}(X) \le \log K,
\]
formalized as \LH{PRB35}. In the observation-only specialization, the min-entropy converse collapses to the intrinsic one-state budget bound
\[
P_{\mathrm{succ}} \le e^{-H_{\infty}(X)},
\qquad
P_e \ge 1-e^{-H_{\infty}(X)},
\]
formalized as \LH{PRB36} and \LH{PRB37}. This isolates the best possible observation-only performance directly in terms of source concentration, without introducing any auxiliary alphabet-size parameter.

The Shannon layer now closes the other side of the finite-source picture as well. The source entropy is always nonnegative (\LH{PRB38}) and is upper-bounded by the ambient alphabet entropy,
\[
H(X) \le \log K,
\]
formalized as \LH{PRB39}. Combined with the previously mechanized min-entropy lower comparison (\LH{PRB31}), the source proof tree now contains the full finite entropy chain
\[
H_{\infty}(X) \le H(X) \le \log K,
\]
so the arbitrary-source converses are no longer expressed only through max-mass and min-entropy, but also situated inside the standard Shannon entropy envelope of the finite alphabet.

\subsection{Unified Converse Viewpoint}\label{sec:unified-converse}

The point of this section is that the many finite lemmas are not separate claims but one converse family. Let $X$ be the latent source, let $Y=(O,T)$ denote the deterministic observation/tag transcript, and let $\hat X$ denote any deterministic decoder output. Then the source proof tree supports the following synthesis:
\begin{theorem}[Unified finite converse, informal synthesis]
If the observation transcript does not already isolate the true state, then the same finite budget simultaneously constrains
\begin{enumerate}
\tightlist
\item zero-error resolution of ambiguity classes and confusability cliques,
\item the conditional entropy $H(X\mid Y)$,
\item the decoder-output entropy $H(\hat X)$,
\item the deterministic mutual information $I(X;Y)=H(Y)$,
\item and the entropy deficits $\log |\mathcal Y|-H(Y)$ and $\log |\hat{\mathcal X}|-H(\hat X)$ via KL divergence from the corresponding uniform output laws.
\end{enumerate}
Equivalently, the paper's counting, clique, Fano-style, decoder-output, and PMF/KL statements are different normal forms of one finite converse theorem family.
\end{theorem}

Read the theorem stack in that order: exact zero-error separation first, then entropy converses, then PMF/KL saturation and gap witnesses. The same obstruction simply reappears in several classical information-theoretic languages.

The unified converse can be read in four compressed layers.

\textbf{Min-entropy and Shannon envelope.} The budget first appears as a min-entropy bound:
\[
H_{\infty}(X) \le \log(|\mathcal O||\mathcal T|) - \log(1-\varepsilon),
\]
formalized as \LH{PRB40}, with the observation-only specialization \LH{PRB41}. Together with the finite-source envelope
\[
H_{\infty}(X) \le H(X) \le \log K
\]
(\LH{PRB31}, \LH{PRB38}, \LH{PRB39}), this places the later Fano-style converses inside a standard entropy hierarchy.

\textbf{Source-side Fano and conditional entropy.} For arbitrary finite sources, the source entropy obeys the exact success/failure partition bound \LH{PRB42}, the budgeted Fano inequality
\[
H(X) \le h(P_e) + (1-P_e)\log(|\mathcal O||\mathcal T|) + P_e\log(K-1),
\]
(\LH{PRB43}), and its observation-only specialization \LH{PRB44}. The same family admits a conditional-entropy form: the model defines a genuine deterministic $H(X\mid Y)$ with chain rule \LH{PRB47}, yields
\[
H(X\mid Y) \le h(P_e) + (1-P_e)\log(|\mathcal O||\mathcal T|) + P_e\log(K-1)
\]
(\LH{PRB50}), and specializes to observation-only as \LH{PRB51}. The graph layer meets this Shannon layer through clique-sensitive entropy bounds \LH{PRB54} and \LH{PRB55}.

\textbf{Observation-side mutual information and KL gaps.} The same model identifies deterministic mutual information with output entropy,
\[
I(X;Y)=H(Y)=H(X)-H(X\mid Y),
\]
formalized by \LH{PRB49}, \LH{PRB57}, and \LH{PRB58}. The same layer gives the budget and source ceilings \LH{PRB52}, \LH{PRB53}, \LH{PRB60}, \LH{PRB62}, and \LH{PRB63}, deterministic data processing under coarsenings \LH{PRB64}--\LH{PRB68}, and a PMF/KL saturation story: zero KL to the uniform output law forces $H(Y)=I(X;Y)=\log(|\mathcal O||\mathcal T|)$ (\LH{PRB69}--\LH{PRB76}, \LH{ENT10}--\LH{ENT12}).

\textbf{Decoder-output and success/failure forms.} The success/failure indicator is itself a finite pushforward source and Bernoulli PMF with entropy exactly $h(P_e)$ (\LH{PRB56}, \LH{PRB77}, \LH{PRB79}, \LH{ENT9}). The decoder output obeys the same converse program:
\[
H(\hat X) \le I(X;Y) \le H(X),
\]
with explicit source and alphabet gaps \LH{PRB81}--\LH{PRB83}, a decoder-native Fano bound
\[
H(\hat X) \le h(P_e) + (1-P_e)\log(|\mathcal O||\mathcal T|) + P_e\log(K-1)
\]
(\LH{PRB84}, \LH{PRB85}, \LH{PRB87}), with its own PMF/KL saturation story on `$\mathrm{Fin}(K+1)$` (\LH{PRB91}--\LH{PRB95}, \LH{ENT13}). The observable/random-variable wrappers simply package these deterministic pushforwards in a reusable interface.



\textbf{Tightness (Achievability + Converse).} Theorems~\ref{thm:upper-bound} and~\ref{thm:lower-bound} together show that the operational cost transition at rate $1$ is sharp: rate $1$ achieves constant manual cost, while any architecture with more than one independent location incurs linear cost in the number of independent encodings (\LH{UNQ1}, \LH{UNQ1}).

\subsection{The Unbounded Gap}\label{sec:gap}

\begin{theorem}[Unbounded Gap]\label{thm:unbounded-gap}
The ratio of modification complexity between DOF-1-incomplete and DOF-1-complete architectures grows without bound:
\[
\lim_{n \to \infty} \frac{M_{\text{DOF}>1}(n)}{M_{\text{DOF}=1}} = \lim_{n \to \infty} \frac{n}{1} = \infty
\]
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
By Theorem~\ref{thm:upper-bound}, $M_{\text{DOF}=1} = O(1)$. Specifically, $M_{\text{DOF}=1} = 1$ for any system size.

By Theorem~\ref{thm:lower-bound}, $M_{\text{DOF}>1}(n) = \Omega(n)$ where $n$ is the number of independent encoding locations.

The ratio is:
\[
\frac{M_{\text{DOF}>1}(n)}{M_{\text{DOF}=1}} = \frac{n}{1} = n
\]

As $n \to \infty$, the ratio $\to \infty$. The gap is unbounded.
\end{proof}

\begin{corollary}[Arbitrary Reduction Factor]\label{cor:arbitrary-reduction}
For any constant $k$, there exists a system size $n$ such that DOF = 1 provides at least $k\times$ reduction in modification complexity.
\end{corollary}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Choose $n = k$. Then $M_{\text{DOF}>1}(n) = n = k$ and $M_{\text{DOF}=1} = 1$. The reduction factor is $k/1 = k$.
\end{proof}

\subsection{The (R, C, P) Tradeoff Space}\label{sec:rcp-tradeoff}

We now summarize the tradeoff space induced by the model.

\begin{definition}[(R, C, P) Tradeoff]\label{def:rcp-tradeoff}
For an encoding system, define:
\begin{itemize}
\tightlist
\item $R$ = \emph{Rate} (DOF): Number of independent encoding locations
\item $C$ = \emph{Complexity}: Manual modification cost per change
\item $P$ = \emph{Coherence indicator}: $P = 1$ iff no incoherent state is reachable; otherwise $P = 0$
\end{itemize}
The \emph{(R, C, P) tradeoff space} is the set of achievable $(R, C, P)$ tuples.
\end{definition}

\begin{theorem}[Operating Regimes]\label{thm:operating-regimes}
The (R, C, P) space has three distinct operating regimes:
\begin{center}
\begin{tabularx}{\linewidth}{cccY}
\toprule
\textbf{Rate} & \textbf{Complexity} & \textbf{Coherence} & \textbf{Interpretation} \\
\midrule
$R = 0$ & $C = 0$ & $P = $ undefined & Fact not encoded \\
$R = 1$ & $C = O(1)$ & $P = 1$ & \textbf{Optimal (capacity-achieving)} \\
$R > 1$ & $C = \Omega(R)$ & $P = 0$ & Above capacity \\
\bottomrule
\end{tabularx}
\end{center}
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
\textbf{$R = 0$:} No encoding exists. Complexity is zero (nothing to modify), but coherence is undefined (nothing to be coherent about).

\textbf{$R = 1$:} By Theorem~\ref{thm:upper-bound}, $C = O(1)$. By Theorem~\ref{thm:coherence-capacity}, $P = 1$ (coherence guaranteed). This is the capacity-achieving regime.

\textbf{$R > 1$:} By Theorem~\ref{thm:lower-bound}, $C = \Omega(R)$. By Theorem~\ref{thm:dof-gt-one-incoherence}, incoherent states are reachable, so $P = 0$.
\end{proof}

\begin{definition}[Pareto Frontier]\label{def:pareto-frontier}
A point $(R, C, P)$ is \emph{Pareto optimal} if no other achievable point dominates it (lower $R$, lower $C$, or higher $P$ without worsening another dimension).

The \emph{Pareto frontier} is the set of all Pareto optimal points.
\end{definition}

\begin{theorem}[Pareto Optimality of DOF = 1]\label{thm:pareto-optimal}
$(R=1, C=1, P=1)$ is the unique Pareto optimal point for encoding systems requiring coherence ($P = 1$).
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
We show $(1, 1, 1)$ is Pareto optimal and unique:

\textbf{Existence:} By Theorems~\ref{thm:upper-bound} and \ref{thm:coherence-capacity}, the point $(1, 1, 1)$ is achievable.

\textbf{Optimality:} Consider any other achievable point $(R', C', P')$ with $P' = 1$:
\begin{itemize}
\tightlist
\item If $R' = 0$: Fact is not encoded (excluded by requirement)
\item If $R' = 1$: Same as $(1, 1, 1)$ (by uniqueness of $C$ at $R=1$)
\item If $R' > 1$: By Theorem~\ref{thm:dof-gt-one-incoherence}, $P' < 1$, contradicting $P' = 1$
\end{itemize}

\textbf{Uniqueness:} No other point achieves $P = 1$ except $R = 1$.
\end{proof}

\textbf{Information-theoretic interpretation.} If one uses $1-P$ as a coarse distortion surrogate, then the zero-distortion boundary collapses to a single point: $R=1$ is the only rate achieving perfect coherence.

\begin{corollary}[No Tradeoff at $P = 1$]\label{cor:no-tradeoff}
Unlike rate-distortion where you can trade rate for distortion, there is no tradeoff at $P = 1$ (perfect coherence). The only option is $R = 1$.
\end{corollary}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Direct consequence of Theorem~\ref{thm:coherence-capacity}.
\end{proof}

\textbf{Comparison to rate-distortion.} In rate-distortion theory:
\begin{itemize}
\tightlist
\item You can achieve lower distortion with higher rate (more bits)
\item The rate-distortion function $R(D)$ is monotonically decreasing
\item $D = 0$ (lossless) requires $R = H(X)$ (source entropy)
\end{itemize}

In our framework:
\begin{itemize}
\tightlist
\item You \emph{cannot} achieve higher coherence ($P$) with more independent locations
\item Higher rate ($R > 1$) \emph{eliminates} coherence guarantees ($P = 0$)
\item $P = 1$ (perfect coherence) requires $R = 1$ exactly
\end{itemize}

The key difference is that extra independent representations increase ambiguity unless they are coordinated by derivation. In this setting, uncontrolled redundancy hurts exact consistency.

\subsection{Practical Implications}\label{sec:practical-implications}

The unbounded gap has practical implications, but these should be read as consequences of the abstract rate/computation result rather than as the main theorem of the paper:

\textbf{1. DOF = 1 matters more at scale.} For small systems ($n = 3$), the difference between 3 edits and 1 edit is minor. For large systems ($n = 50$), the difference between 50 edits and 1 edit is significant.

\textbf{2. The gap compounds over time.} Each modification to fact $F$ incurs the complexity cost. If $F$ changes $m$ times over the system lifetime, total cost is $O(mn)$ with DOF $>$ 1 vs. $O(m)$ with DOF = 1.

\textbf{3. The gap affects error rates.} Each manual edit is an opportunity for error. With $n$ edits, the probability of at least one error is $1 - (1-p)^n$ where $p$ is the per-edit error probability. As $n$ grows, this approaches 1.

\begin{example}[Error Rate Calculation]\label{ex:error-rate}
Assume a 1\% error rate per edit ($p = 0.01$).

\begin{center}
\begin{tabular}{ccc}
\toprule
\textbf{Edits ($n$)} & \textbf{P(at least one error)} & \textbf{Architecture} \\
\midrule
1 & 1.0\% & DOF = 1 \\
10 & 9.6\% & DOF = 10 \\
50 & 39.5\% & DOF = 50 \\
100 & 63.4\% & DOF = 100 \\
\bottomrule
\end{tabular}
\end{center}

With 50 independent encoding locations (DOF = 50), there is a 39.5\% chance of introducing an error when modifying fact $F$. With DOF = 1, the chance is 1\%.
\end{example}

\subsection{Amortized Analysis}\label{sec:amortized}

The complexity bounds assume a single modification. Over the lifetime of an encoding system, facts are modified many times (\LH{UNQ1}).

\begin{theorem}[Amortized Complexity]\label{thm:amortized}
Let fact $F$ be modified $m$ times over the system lifetime. Let $n$ be the number of independent encoding locations. Total modification cost is:
\begin{itemize}
\tightlist
\item DOF = 1: $O(m)$
\item DOF = $n > 1$: $O(mn)$
\end{itemize}
\end{theorem}
\leanmeta{\LH{UNQ1}}

\begin{proof}
Each modification costs $O(1)$ with DOF = 1 and $O(n)$ with DOF = $n$. Over $m$ modifications, total cost is $m \cdot O(1) = O(m)$ with DOF = 1 and $m \cdot O(n) = O(mn)$ with DOF = $n$.
\end{proof}

For a fact modified 100 times with 50 independent encoding locations:
\begin{itemize}
\tightlist
\item DOF = 1: 100 edits total
\item DOF = 50: 5,000 edits total
\end{itemize}

The 50$\times$ reduction factor applies to every modification, compounding over the system lifetime.

%==============================================================================
