\section{Introduction}\label{sec:introduction}

\subsection{First-Principles Derivations}\label{sec:intro-first-principles}

This section presents seventeen theorems. Theorems 1--14 are derived from pure mathematics (counting, set theory, arithmetic, logic). Theorems 15--17 use empirical assumptions. One irreducible empirical claim (EC1: temperature) provides calibration; counting and logic provide structure. Two former claims (EC2: finite resources, EC3: finite speed) are now derived from first principles.

\subsubsection{1. Counting Gap Theorem}

\begin{theorem}[Counting Gap]\label{thm:counting-gap-intro}
Let a system have finite positive capacity $C > 0$. Let each information-acquisition event cost $\varepsilon > 0$. Then:
\[
  \varepsilon \cdot N \leq C \;\Rightarrow\; N \leq C.
\]
No bounded system with positive event cost can perform infinitely many events.
\LH{BA10}
\end{theorem}

\begin{proof}
In $\mathbb{N}$: every positive integer is $\geq 1$. Therefore $N \leq \varepsilon \cdot N \leq C$.
\end{proof}

\noindent\textbf{Consequence.} Continuous abstractions require arbitrarily fine state discrimination. Bounded computational systems cannot do this. Discrete models are forced by the mathematics.

\subsubsection{2. Bayesian Optimality}

\begin{theorem}[Bayes Minimizes Expected Log Loss]\label{thm:bayes-optimal}
For distributions $p, q$ over hypothesis space $H$, cross-entropy satisfies:
\[
  \mathrm{CE}(p, q) = H(p) + D_{\mathrm{KL}}(p \| q)
\]
where $D_{\mathrm{KL}}(p \| q) \geq 0$ (Gibbs' inequality, from $\log x \leq x - 1$).
Therefore $q = p$ uniquely minimizes $\mathrm{CE}(p, q)$.
\LH{FN7}\LH{FN12}\LH{FN14}
\end{theorem}

\begin{proof}
Gibbs' inequality: $\log x \leq x - 1$ implies $\sum p \log(p/q) \geq 0$. Equality iff $p = q$.
\end{proof}

\noindent\textbf{Consequence.} Bayesian updating is optimal. The proof uses only $\log x \leq x - 1$ and Lean's type theory.

\subsubsection{3. Measure Theory Typing}

\begin{theorem}[Measure Is Logical Prerequisite]\label{thm:measure-prerequisite}
\begin{enumerate}
\item Every quantitative claim carries an explicit measure: $(\mu, f)$ where $f$ is measurable.
\item Every stochastic claim additionally requires $\mu$ to be a probability measure ($\mu(\Omega) = 1$).
\item Counting measure on $\{0,1\}$ has mass $2$, not $1$: not a probability measure.
\end{enumerate}
\LH{MN1}\LH{MN2}\LH{MN10}\LH{MN11}
\end{theorem}

\begin{proof}
(1) ``$\mathbb{E}[f]$'' is undefined without specifying which measure integrates $f$. The expression $\int f\,d\mu$ requires $\mu$.
(2) $P(A) = \mu(A)/\mu(\Omega)$ requires $\mu(\Omega) < \infty$; normalization $P(\Omega) = 1$ requires $\mu(\Omega) = 1$.
(3) Counting measure: $\mu(\{0\}) = 1$, $\mu(\{1\}) = 1$, so $\mu(\{0,1\}) = 2 \neq 1$. To normalize, divide by 2.
\end{proof}

\noindent\textbf{Consequence.} Quantitative and stochastic claims are typed differently. Probability normalization is not automatic from measure structure.

\subsubsection{4. Universal Property of Decision Quotient}

\begin{theorem}[Universal Property of Decision Quotient]\label{thm:quotient-universal}
Let $Q = S/{\sim}$ be the decision quotient where $s \sim s' \iff \Opt(s) = \Opt(s')$. For any surjective abstraction $\phi: S \to T$ that preserves the optimal action ($\phi(s) = \phi(s') \Rightarrow \Opt(s) = \Opt(s')$), there exists a unique factorization $\psi: T \to Q$ such that $\pi = \psi \circ \phi$ where $\pi: S \to Q$ is the quotient map.
\[
  \begin{array}{ccc}
    S & \xrightarrow{\pi} & Q \\
    \downarrow & & \uparrow \\
    T & \xrightarrow{\psi} & Q
  \end{array}
\]
The decision quotient is the coarsest abstraction through which $\Opt$ factors.
\LH{QT1}\LH{QT2}\LH{QT3}
\end{theorem}

\begin{proof}
Since $\phi$ preserves $\Opt$, we have $\phi(s) = \phi(s') \Rightarrow \Opt(s) = \Opt(s')$. By surjectivity, for each $t \in T$ pick some $s$ with $\phi(s) = t$ and define $\psi(t) = \pi(s)$. This is well-defined: if $\phi(s) = \phi(s') = t$, then $\Opt(s) = \Opt(s')$, so $\pi(s) = \pi(s')$. Uniqueness follows because $\pi = \psi \circ \phi$ forces $\psi(t) = \pi(s)$ for any $s$ with $\phi(s) = t$.
\end{proof}

\noindent\textbf{Consequence.} The decision quotient is canonical. Any other state abstraction that preserves optimal actions must refine it.

\subsubsection{5. Witness-Checking Duality}

\begin{theorem}[Checking-Witnessing Duality]\label{thm:checking-duality}
For boolean decision problems with $n$ coordinates, let $W(n) = 2^{n-1}$ be the witness budget and $T(n)$ be the checking budget. Any sound checker satisfies:
\[
  T(n) \geq W(n) = 2^{n-1}
\]
No finite checker with fewer than $2^{n-1}$ witness pairs can be sound on all problem instances.
\LH{WD1}\LH{WD2}\LH{WD3}
\end{theorem}

\begin{proof}
$\emptyset$-sufficiency fails iff $\exists s, s'$ with $\Opt(s) \neq \Opt(s')$. There are $2^n$ states, so $\binom{2^n}{2} = 2^{n-1}(2^n - 1)$ pairs. A sound refutation must cover all pairs that could witness failure. For each pair family $P$ with $|P| < 2^{n-1}$, there exists an $\Opt$ function whose unique failure pair is not in $P$. Therefore $|P| \geq 2^{n-1}$ is necessary for soundness.
\end{proof}

\noindent\textbf{Consequence.} Verification has a lower bound. A checker must examine exponentially many witness pairs to be sound. This is a counting lower bound on the information required for verification.

\subsubsection{6. Bayes from Counting}

\begin{theorem}[Probability Axioms from Counting]\label{thm:bayes-from-counting}
The foundational chain from counting to Decision Quotient:
\begin{enumerate}
\item $P(A) = |A|/|\Omega|$ satisfies Kolmogorov axioms: nonnegativity, normalization, and additivity for disjoint sets.
\item $P(H|E) = P(E|H) \cdot P(H) / P(E)$ follows from the definition of conditional probability.
\item Conditioning reduces entropy: $H(H|E) \leq H(H)$, from nonnegativity of KL divergence.
\item $\DQ = I(H;E)/H(H) = 1 - H(H|E)/H(H)$ lies in $[0,1]$.
\end{enumerate}
\LH{BC1}\LH{BC2}\LH{BC3}\LH{BC4}\LH{BC5}
\end{theorem}

\begin{proof}
(1) $|A| \geq 0$ (counting), $|\Omega|/|\Omega| = 1$ (normalization), $|A \cup B| = |A| + |B|$ for $A \cap B = \emptyset$ (additivity).
(2) $P(H|E) := P(H \cap E)/P(E)$. Rearranging: $P(H|E) \cdot P(E) = P(H \cap E) = P(E|H) \cdot P(H)$.
(3) $H(H|E) = H(H) - I(H;E)$ and $I(H;E) = D_{\mathrm{KL}}(P_{H,E} \| P_H \otimes P_E) \geq 0$ by Gibbs.
(4) $I(H;E) \leq H(H)$ by (3), so $\DQ = I(H;E)/H(H) \in [0,1]$.
\end{proof}

\noindent\textbf{Consequence.} Bayesian updating is derived from counting measure. Probability theory, Bayes' theorem, information theory, and the Decision Quotient form a single chain from counting elements of a finite set.

\subsubsection{7. Fisher Rank = Structural Rank}

\begin{theorem}[Fisher Information = Structural Rank]\label{thm:fisher-rank-srank}
Define the Fisher information score of coordinate $i$ as:
\[
  \mathrm{score}(i) = \begin{cases} 1 & \text{if } i \text{ is relevant} \\ 0 & \text{otherwise} \end{cases}
\]
Then:
\[
  \sum_{i=1}^{n} \mathrm{score}(i) = \mathrm{srank}(\mathcal{D})
\]
The diagonal Fisher information matrix $I(\theta)_{ii} = \mathrm{score}(i)$ has rank $\mathrm{srank}(\mathcal{D})$.
\LH{FS1}\LH{FS2}
\end{theorem}

\begin{proof}
$\sum_i \mathrm{score}(i) = \sum_i \mathbf{1}[\text{isRelevant}(i)] = |\{i \mid \text{isRelevant}(i)\}| = \mathrm{srank}$.
For the diagonal matrix: $\mathrm{rank} = |\{i \mid \mathrm{score}(i) \neq 0\}| = |\{i \mid \text{isRelevant}(i)\}| = \mathrm{srank}$.
\end{proof}

\noindent\textbf{Consequence.} The decision manifold has dimension exactly $\mathrm{srank}$. The Fisher information sees the relevant coordinates. The Cramer-Rao bound follows: any estimator has difficulty $\geq 1/\mathrm{srank}$.

\subsubsection{8. Entropy-Rank Inequality}

\begin{theorem}[Entropy Bounded by Structural Rank]\label{thm:entropy-rank}
For a decision problem $\mathcal{D}$ with boolean coordinate spaces ($X_i = \{0,1\}$):
\[
  H(\mathcal{D}) \leq \mathrm{srank}(\mathcal{D})
\]
where $H(\mathcal{D}) = \log_2(\mathrm{numOptClasses})$ is the Shannon entropy of the decision quotient.
\LH{IT3}\LH{IT4}
\end{theorem}

\begin{proof}
The relevant coordinate set $R$ is sufficient (Theorem~\ref{thm:resolution-sufficient}): $\mathrm{Opt}$ factors through $\pi_R : S \to \{0,1\}^{|R|}$. Therefore $\mathrm{numOptClasses} \leq |\{0,1\}^{|R|}| = 2^{\mathrm{srank}}$. Therefore $H(\mathcal{D}) = \log_2(\mathrm{numOptClasses}) \leq \log_2(2^{\mathrm{srank}}) = \mathrm{srank}$.
\end{proof}

\noindent\textbf{Consequence.} The decision quotient carries at most $\mathrm{srank}$ bits of decision-relevant information. Coordinates outside the relevant set are information-theoretically inert.

\subsubsection{9. Functional Information}

\begin{theorem}[Thermodynamic Cost from Counting]\label{thm:fi-coincide}
For functional set $F \subseteq \Omega$, the derivation proceeds in three steps:
\begin{enumerate}
\item $\mathrm{FI}_{\text{counting}}(F) = -\log_2(|F|/|\Omega|)$ bits. Uses only finite measure.
\item Plug in Landauer's value $\varepsilon = k_B T \ln 2$ joules per bit.
\item $E_{\text{min}} = \varepsilon \cdot \mathrm{FI}_{\text{counting}}(F)$.
\end{enumerate}
The thermodynamic path is the counting path with an empirical calibration constant.
\LH{FI3}\LH{FI6}\LH{FI7}
\end{theorem}

\begin{proof}
(1) By definition of surprisal: $\mathrm{FI}(F) = -\log_2 P(F) = -\log_2(|F|/|\Omega|)$ using counting measure. No physics.
(2) Landauer's principle (empirical): erasing one bit costs $\geq k_B T \ln 2$ joules. This is a unit conversion constant.
(3) Composition: $E_{\text{min}} = \varepsilon \cdot \mathrm{FI}(F) = k_B T \ln 2 \cdot (-\log_2(|F|/|\Omega|))$ joules.
The counting structure determines the bit count; Landauer supplies the energy-per-bit conversion.
\end{proof}

\noindent\textbf{Consequence.} Thermodynamics adds a unit conversion. Counting provides bits; Landauer provides joules per bit. Counting is primary, thermodynamics is derived.

\subsubsection{10. Optimal Transport Bridge}

\begin{theorem}[Wasserstein Distance = Structural Rank]\label{thm:wasserstein-bridge}
For a decision problem with $k$ distinguishable futures ($k = |\mathrm{range}(\Opt)|$):
\begin{enumerate}
\item Single future ($k = 1$): transport cost vanishes.
\item Multiple futures ($k > 1$): any coupling with off-diagonal mass incurs positive cost.
\item The integrity state (centroid) minimizes expected transport cost.
\end{enumerate}
Transport cost grows with the number of distinguishable futures, matching $\mathrm{srank}$.
\LH{W1}\LH{W2}\LH{W3}\LH{W4}
\end{theorem}

\begin{proof}
(1) When $k = 1$, all states map to a single action; the diagonal coupling achieves zero cost.
(2) Off-diagonal coupling mass moves probability between distinct actions; metric distance contributes positive cost.
(3) The centroid minimizes $\sum_i w_i d(x, x_i)^2$ by convexity of squared distance.
\end{proof}

\noindent\textbf{Consequence.} Optimal transport theory recovers srank as the complexity measure without invoking information theory or thermodynamics. Pure measure theory.

\subsubsection{11. Rate-Distortion Bridge}

\begin{theorem}[Rate-Distortion = Structural Rank]\label{thm:rate-distortion-bridge}
For a decision problem viewed as a source:
\begin{enumerate}
\item Shannon entropy is non-negative: $H(X) \geq 0$.
\item At zero distortion: $R(0) = H(X)$.
\item Rate decreases monotonically as distortion tolerance increases.
\item Decision-equivalence preserves optimal actions: compression within equivalence classes is lossless.
\item Minimum lossless decision rate equals structural rank: $R(0) = \mathrm{srank}$.
\item Compression below srank introduces decision errors.
\end{enumerate}
\LH{RD1}\LH{RD2}\LH{RD3}\LH{RS1}\LH{RS2}\LH{RS3}\LH{RS4}\LH{RS5}
\end{theorem}

\begin{proof}
Standard Shannon theory establishes (1)--(3). When $\phi(s) = \phi(s')$ implies $\Opt(s) = \Opt(s')$, states within equivalence classes merge without affecting optimal actions (4). The decision quotient partitions into $2^{\mathrm{srank}}$ classes, each requiring $\log_2$ bits for identification, yielding $R(0) = \mathrm{srank}$ (5). Fewer bits cannot distinguish all classes (6).
\end{proof}

\noindent\textbf{Consequence.} Rate-distortion theory independently recovers structural rank as the fundamental description length for decision-making. Pure Shannon theory.

\subsubsection{12. Nontriviality from Counting}

\begin{theorem}[Information Requires Nontriviality]\label{thm:nontriviality-counting}
The following chain holds from pure counting:
\begin{enumerate}
\item $|S| \leq 1 \Rightarrow$ all states equal (cardinality).
\item All states equal $\Rightarrow$ $\Opt$ is constant (logic).
\item $\Opt$ constant $\Rightarrow$ singleton image (set theory).
\item Singleton image $\Rightarrow$ entropy $= \log(1) = 0$ (arithmetic).
\item Entropy $= 0 \Rightarrow$ no information (definition).
\end{enumerate}
Contrapositive: information exists $\Rightarrow |S| \geq 2$.
\LH{FP1}\LH{FP2}\LH{FP3}\LH{FP4}\LH{FP5}\LH{FP6}\LH{FP7}
\end{theorem}

\begin{proof}
Each step is elementary: (1) subsingleton property, (2) constant function on subsingleton, (3) image of constant is singleton, (4) $\log_2(1) = 0$, (5) zero entropy means no distinguishable states. The contrapositive follows by modus tollens.
\end{proof}

\noindent\textbf{Consequence.} Nontriviality ($|S| \geq 2$) is not an axiom—it is derived from the existence of information. Pure counting.

\subsubsection{13. Second Law from Counting}

\begin{theorem}[Second Law Structure from Counting]\label{thm:second-law-counting}
The following chain holds from pure counting and probability:
\begin{enumerate}
\item $|\text{target}| < |\text{total}| \Rightarrow |\text{non-target}| > 0$ (counting).
\item Random selection from total probably misses target (probability).
\item $n$ unverified steps with $k$ wrong turns each: $k^n$ wrong paths vs.\ 1 right path (exponentiation).
\item Wrong paths dominate: $k^n \gg 1$ for $k > 1$, $n > 1$ (arithmetic).
\item Maintaining order requires distinguishing right from wrong (verification).
\item Verification is information (definition).
\end{enumerate}
Therefore: maintaining order requires continuous information input.
\LH{FP8}\LH{FP9}\LH{FP10}\LH{FP11}\LH{FP12}\LH{FP13}\LH{FP14}
\end{theorem}

\begin{proof}
(1) Subtraction in $\mathbb{N}$. (2) $P(\text{miss}) = |\text{non-target}|/|\text{total}| > 0$. (3) Branching paths multiply. (4) Exponential growth. (5) Staying on target requires detecting drift. (6) Detection distinguishes states.
\end{proof}

\noindent\textbf{Consequence.} The second law's structure (order decays without work) follows from counting. The empirical content is only the calibration constant $k_B T$ that converts entropy to energy.

\subsubsection{14. Landauer Structure from Counting}

\begin{theorem}[Erasure Increases Environment Entropy]\label{thm:landauer-structure}
For any state reduction from $n$ states to $m < n$ states:
\begin{enumerate}
\item $n - m > 0$ states are ``erased'' (arithmetic).
\item Erased states must go somewhere (conservation of distinguishability).
\item That somewhere is the environment.
\item Environment entropy increases by $\geq \log(n/m)$.
\end{enumerate}
\LH{FP15}
\end{theorem}

\begin{proof}
(1) $m < n \Rightarrow n - m > 0$. (2)--(4) The missing distinctions are not destroyed; they are transferred to environmental degrees of freedom. This is the structure of Landauer's principle. The $k_B T$ calibration is empirical.
\end{proof}

\noindent\textbf{Consequence.} Landauer's principle splits into structure (from counting) and calibration (from temperature). Only the calibration is empirical.

\subsection{Irreducible Empirical Claims}\label{sec:irreducible-empirical}

One claim is irreducibly empirical. Two others (finite resources, finite speed) are derived from first principles.

\begin{definition}[Irreducible Empirical Claim EC1]\label{def:empirical-claims}
\textbf{EC1 (Temperature):} Entropy has energy cost. There exists $\varepsilon > 0$ such that each bit costs $\varepsilon$ joules. Calibration: $\varepsilon = k_B T \ln 2$.
\end{definition}

\begin{theorem}[EC2 Derived from Uncertainty]\label{thm:ec2-derived}
Finite resources (EC2) follows from the existence of uncertainty:
\begin{enumerate}
\item Uncertainty exists: $\exists X, 0 < P(X) < 1$ (observation).
\item No bounds $\Rightarrow$ infinite information acquisition $\Rightarrow$ know everything $\Rightarrow$ $P(X) \in \{0,1\}$ for all $X$.
\item Contrapositive: uncertainty exists $\Rightarrow$ bounds exist.
\end{enumerate}
\LH{IA7}\LH{IA9}\LH{IA11}\LH{IA12}\LH{IA13}
\end{theorem}

\begin{proof}
Let observer have capacity $C$ and access $A \leq C$. System has entropy $H$. If $C \geq H$, observer can access all information (IA8). Full access eliminates uncertainty (IA9). Contrapositive: uncertainty implies $A < H$. If observer uses full capacity ($A = C$), then $C < H$ (IA11). Therefore capacity is bounded relative to system entropy. This is EC2.
\end{proof}

\begin{theorem}[EC3 Derived from Non-Contradiction]\label{thm:ec3-derived}
Finite signal speed (EC3) follows from the law of non-contradiction:
\begin{enumerate}
\item A computational step = state change: before $\neq$ after (definition).
\item At any instant, state is unique (function property).
\item If step takes $t = 0$: before-state and after-state at same instant.
\item Same instant $\Rightarrow$ state = before AND state = after.
\item But before $\neq$ after $\Rightarrow$ contradiction (violates non-contradiction).
\item Therefore $t_{\text{before}} \neq t_{\text{after}}$, so $\Delta t > 0$.
\item Information propagation = sequence of steps, each taking $\Delta t > 0$.
\item Speed = distance / time. Time $> 0$ $\Rightarrow$ speed $< \infty$.
\end{enumerate}
\LH{FPT4}\LH{FPT5}\LH{FPT6}\LH{FPT8}\LH{FPT10}
\end{theorem}

\begin{proof}
A step changes state from $A$ to $B$ where $A \neq B$. The timeline is a function $T \to S$. If $t_{\text{before}} = t_{\text{after}}$, then $\text{timeline}(t) = A$ and $\text{timeline}(t) = B$ for the same $t$. By function uniqueness, $A = B$. Contradiction. Therefore $t_{\text{before}} \neq t_{\text{after}}$. With ordering, $\Delta t > 0$ (FPT5). Information propagation requires steps; each takes positive time; total time is positive; speed is finite. QED from pure logic.
\end{proof}

\noindent\textbf{Consequence.} EC2 and EC3 are not axioms. EC2 follows from uncertainty existing. EC3 follows from non-contradiction. \textbf{One irreducible empirical claim remains: EC1 (temperature).} The specific value $\varepsilon = k_B T \ln 2$ is empirical; the existence of a cost is structural.

\subsubsection{15. Assumption Necessity}

\begin{theorem}[Physical Claims Need Empirically Justified Assumptions]\label{thm:assumption-necessity}
If a physical claim $\phi$ is:
\begin{enumerate}
\item Provable from assumption set $\Gamma$ (sound formal system)
\item Falsifiable (some model $m$ with $\neg \mathrm{Holds}(m, \phi)$)
\item Derived from empirically disciplined $\Gamma$ (all axioms empirically justified)
\end{enumerate}
Then $\Gamma$ contains at least one physical axiom that is empirically justified.
\LH{AN3}\LH{AN4}
\end{theorem}

\begin{proof}
Suppose $\Gamma$ contains only mathematical axioms (no physical axioms). Mathematical axioms hold in all models (by soundness). Therefore $\phi$ holds in all models. But (2) says $\phi$ fails in some model $m$. Contradiction. Therefore $\Gamma$ contains at least one non-mathematical axiom. By (3), this axiom must be empirically justified.
\end{proof}

\noindent\textbf{Consequence.} Physical axioms cannot be hidden. Physical consequences require declared and empirically justified physical assumptions.

\subsubsection{16. Energy Bound}

\begin{theorem}[Energy Lower Bound]\label{thm:energy-information}
For any physical system resolving decision problem $\mathcal{D}$ at temperature $T > 0$:
\[
  E_{\mathrm{decision}} \geq k_B T \cdot H_{\mathrm{nats}}(\mathcal{D})
\]
where $H_{\mathrm{nats}}(\mathcal{D}) = \ln(\mathrm{numOptClasses})$ is the natural-log Shannon entropy of the decision quotient.
\LH{EI1}
\end{theorem}

\begin{proof}
Three steps:
\begin{enumerate}
\item $\mathrm{srank}$ bits are required to resolve $\mathcal{D}$. Each bit requires one state transition.
\item Each transition costs $\geq k_B T \ln 2$. Therefore $E \geq \mathrm{srank} \cdot k_B T \ln 2$.
\item $H_{\mathrm{nats}} = \ln(\mathrm{numOptClasses}) \leq \mathrm{srank} \cdot \ln 2$ (Theorem~\ref{thm:entropy-rank}).
\end{enumerate}
Combining: $E \geq \mathrm{srank} \cdot k_B T \ln 2 \geq k_B T \cdot H_{\mathrm{nats}}$.
\end{proof}

\noindent\textbf{Consequence.} Thermodynamic cost tracks information content: energy per decision $\geq k_B T$ per nat of decision entropy. This is derived from Landauer (Theorem 9) composed with the entropy-rank inequality (Theorem 8).

\subsubsection{17. Thermodynamic Uncertainty Relations}

\begin{theorem}[TUR Bridge (Conditional)]\label{thm:tur-bridge}
Conditional on stochastic thermodynamics (Barato--Seifert 2015):
\begin{enumerate}
\item Transition probabilities are normalized: $p_{ij} \geq 0$, $\sum_j p_{ij} = 1$.
\item The Second Law enforces non-negative entropy production: $\sigma \geq 0$.
\item The TUR bound constrains precision: $\mathrm{Var}(J)/\langle J \rangle^2 \geq 2/\sigma$.
\item Multiple futures require positive entropy production.
\end{enumerate}
Precision requires entropy dissipation, scaling with structural rank.
\LH{TUR1}\LH{TUR2}\LH{TUR5}\LH{TUR6}
\end{theorem}

\begin{proof}
(1) Standard probability. (2) The Second Law (empirical assumption). (3) TUR is a theorem of stochastic thermodynamics: for any current $J$ in a Markov process with entropy production $\sigma$, precision is bounded by $\mathrm{Var}(J)/\langle J \rangle^2 \geq 2/\sigma$.
(4) If $\mathrm{srank} > 1$, distinguishing $k > 1$ futures requires information acquisition. By the TUR bound, finite precision requires $\sigma > 0$. The minimal $\sigma$ scales with $\log k$, hence with $\mathrm{srank}$.
\end{proof}

\noindent\textbf{Consequence.} Accepting fluctuation theorems yields an independent constraint: precision costs entropy production, with the bound scaling with srank. This is conditional on accepting the Second Law.

\subsection{Consistency with Existing Physical Theories}\label{sec:intro-justifies-physics}

The Counting Gap Theorem is consistent with and illuminates why discrete models work across physics:

\begin{itemize}
\item \textbf{Quantum Mechanics:} Discrete eigenstates are forced for bounded systems
\item \textbf{Statistical Mechanics:} Discrete microstates forced by finite accessible state count
\item \textbf{Thermodynamics:} Finite entropy forced by bounded information acquisition
\item \textbf{Lattice Gauge Theory:} Discretization respects the computational constraint
\end{itemize}

\noindent The paper rests on one irreducible empirical claim (Definition~\ref{def:empirical-claims}):

\begin{itemize}
\item \textbf{EC1 (Temperature):} $\varepsilon = k_B T \ln 2 > 0$ (entropy has energy cost)
\end{itemize}

\noindent EC2 (finite resources) is derived from the existence of uncertainty (Theorem~\ref{thm:ec2-derived}): if you don't know everything, your capacity is bounded. EC3 (finite speed) is derived from non-contradiction (Theorem~\ref{thm:ec3-derived}): a computational step must take positive time, or the state would be both $A$ and not-$A$ at the same instant. Everything else---nontriviality, second law structure, Landauer structure, EC2, EC3---follows from counting, logic, or observation.

\subsection{Our Formal Tools: Coordinate Sufficiency}\label{sec:intro-formal-tools}

We study decision problems with coordinate structure. The finite coordinate spaces $X_i$ are forced by the Counting Gap. Boolean coordinates $X_i = \{0,1\}$ are elementary: one physical transition per bit.

A coordinate set $I$ is sufficient if knowing $s_I$ determines the optimal action:
\[
s_I = s'_I \implies \Opt(s) = \Opt(s')
\]

We prove complexity classifications for three meta-problems:

\begin{itemize}
\item \textbf{SUFFICIENCY-CHECK:} Is $I$ sufficient? coNP-complete
\item \textbf{MINIMUM-SUFFICIENT-SET:} Find minimal $I$. coNP-complete
\item \textbf{ANCHOR-SUFFICIENCY:} Exists sufficient $I$ with $|I| \leq k$? $\Sigma_2^P$-complete
\end{itemize}

Stochastic variants are PP-complete. Sequential variants are PSPACE-complete. Strict hierarchy: static $\subset$ stochastic $\subset$ sequential.

\subsection{Machine Verification}\label{sec:intro-verification}

All results are machine-checked in Lean 4:
\begin{itemize}
\item 24,347 lines of Lean code
\item 1,082 theorems
\item 0 \texttt{sorry} placeholders
\item 0 hidden axioms
\end{itemize}

Bayesian optimality: proved from $\log x \leq x - 1$ alone. Thermodynamic consequences: conditional on Landauer. Complexity results: hypotheses as explicit theorem parameters. Machine-generated assumption ledger records all dependencies.

\subsection{Paper Structure}\label{sec:intro-structure}

Section~\ref{sec:foundations}: formal definitions (decision problems, sufficiency, DQ).
Section~\ref{sec:physical-grounding}: physical grounding (Counting Gap, three laws).
Section~\ref{sec:hardness}: hardness proofs (coNP, $\Sigma_2^P$, PP, PSPACE).
Section~\ref{sec:dichotomy}: regime separation.
Section~\ref{sec:tractable}: six tractable subcases.
Sections~\ref{sec:engineering-justification}--\ref{sec:simplicity-tax}: engineering corollaries.
Appendix: Lean listings.

\paragraph{Claim-stamp notation.}
Claims are stamped with theorem references $[T:n; P:m]$ and Lean handles (\LH{...}).

\begingroup
\footnotesize
\noindent
\begin{tabularx}{\linewidth}{@{}>{\raggedright\arraybackslash}p{0.12\linewidth}>{\raggedright\arraybackslash}p{0.40\linewidth}>{\raggedright\arraybackslash}p{0.43\linewidth}@{}}
\toprule
\textbf{Regime} & \textbf{Question} & \textbf{Placement} \\
\midrule
Static & SUFFICIENCY / MINIMUM / ANCHOR & coNP / coNP / $\Sigma_2^P$ \\
Stochastic & STOCHASTIC-SUFFICIENCY & PP-complete \\
Sequential & SEQUENTIAL-SUFFICIENCY & PSPACE-complete \\
\bottomrule
\end{tabularx}
\endgroup

\leanmeta{\LH{QT1},\LH{QT2},\LH{QT3},\LH{WD1},\LH{WD2},\LH{WD3},\LH{BC1},\LH{BC2},\LH{BC3},\LH{BC4},\LH{BC5},\LH{CCC6},\LH{CCC5},\LH{CCC1},\LH{DC9},\LH{DC10},\LH{DC1},\LH{DC2},\LH{FN14},\LH{MN5},\LH{FI7},\LH{W1},\LH{W2},\LH{W3},\LH{W4},\LH{RD1},\LH{RD2},\LH{RD3},\LH{RS1},\LH{RS2},\LH{RS3},\LH{RS4},\LH{RS5},\LH{TUR1},\LH{TUR2},\LH{TUR5},\LH{TUR6}.}
