\section{Formal Foundations}\label{sec:foundations}

We formalize decision problems with coordinate structure, sufficiency of coordinate sets, and the decision quotient, drawing on classical decision theory \cite{savage1954foundations, raiffa1961applied}.

\subsection{Decision Problems with Coordinate Structure}

\begin{definition}[Decision Problem]\label{def:decision-problem}
A \emph{decision problem with coordinate structure} is a tuple $\mathcal{D} = (A, X_1, \ldots, X_n, U)$ where:
\begin{itemize}
\item $A$ is a finite set of \emph{actions} (alternatives)
\item $X_1, \ldots, X_n$ are finite \emph{coordinate spaces}
\item $S = X_1 \times \cdots \times X_n$ is the \emph{state space}
\item $U : A \times S \to \mathbb{Q}$ is the \emph{utility function}
\end{itemize}
\end{definition}

\begin{definition}[Projection]\label{def:projection}
For state $s = (s_1, \ldots, s_n) \in S$ and coordinate set $I \subseteq \{1, \ldots, n\}$:
\[
s_I := (s_i)_{i \in I}
\]
is the \emph{projection} of $s$ onto coordinates in $I$.
\end{definition}

\begin{definition}[Optimizer Map]\label{def:optimizer}
For state $s \in S$, the \emph{optimal action set} is:
\[
\Opt(s) := \arg\max_{a \in A} U(a, s) = \{a \in A : U(a,s) = \max_{a' \in A} U(a', s)\}
\]
\end{definition}

\subsection{Sufficiency and Relevance}

\begin{definition}[Sufficient Coordinate Set]\label{def:sufficient}
A coordinate set $I \subseteq \{1, \ldots, n\}$ is \emph{sufficient} for decision problem $\mathcal{D}$ if:
\[
\forall s, s' \in S: \quad s_I = s'_I \implies \Opt(s) = \Opt(s')
\]
Equivalently, the optimal action depends only on coordinates in $I$.
\end{definition}

\begin{definition}[Minimal Sufficient Set]\label{def:minimal-sufficient}
A sufficient set $I$ is \emph{minimal} if no proper subset $I' \subsetneq I$ is sufficient.
\end{definition}

\begin{definition}[Relevant Coordinate]\label{def:relevant}
Coordinate $i$ is \emph{relevant} if it belongs to some minimal sufficient set.
\end{definition}

\begin{example}[Weather Decision]
Consider deciding whether to carry an umbrella:
\begin{itemize}
\item Actions: $A = \{\text{carry}, \text{don't carry}\}$
\item Coordinates: $X_1 = \{\text{rain}, \text{no rain}\}$, $X_2 = \{\text{hot}, \text{cold}\}$, $X_3 = \{\text{Monday}, \ldots, \text{Sunday}\}$
\item Utility: $U(\text{carry}, s) = -1 + 3 \cdot \mathbf{1}[s_1 = \text{rain}]$, $U(\text{don't carry}, s) = -2 \cdot \mathbf{1}[s_1 = \text{rain}]$
\end{itemize}

The minimal sufficient set is $I = \{1\}$ (only rain forecast matters). Coordinates 2 and 3 (temperature, day of week) are irrelevant.
\end{example}

\subsection{The Decision Quotient}

\begin{definition}[Decision Equivalence]\label{def:decision-equiv}
For coordinate set $I$, states $s, s'$ are \emph{$I$-equivalent} (written $s \sim_I s'$) if $s_I = s'_I$.
\end{definition}

\begin{definition}[Decision Quotient]\label{def:decision-quotient}
The \emph{decision quotient} for state $s$ under coordinate set $I$ is:
\[
\text{DQ}_I(s) = \frac{|\{a \in A : a \in \Opt(s') \text{ for some } s' \sim_I s\}|}{|A|}
\]
This measures the fraction of actions that are optimal for at least one state consistent with $I$.
\end{definition}

\begin{proposition}[Sufficiency Characterization]\label{prop:sufficiency-char}
Coordinate set $I$ is sufficient if and only if $\text{DQ}_I(s) = |\Opt(s)|/|A|$ for all $s \in S$.
\end{proposition}

\begin{proof}
If $I$ is sufficient, then $s \sim_I s' \implies \Opt(s) = \Opt(s')$, so the set of actions optimal for some $s' \sim_I s$ is exactly $\Opt(s)$.

Conversely, if the condition holds, then for any $s \sim_I s'$, the optimal actions form the same set (since $\text{DQ}_I(s) = \text{DQ}_I(s')$ and both equal the relative size of the common optimal set).
\end{proof}

\subsection{Computational Model and Input Encoding}\label{sec:encoding}

We fix the computational model used by the complexity claims.

\paragraph{Succinct encoding (primary for hardness).}
This succinct circuit encoding is the standard representation for decision problems in complexity theory; hardness is stated with respect to the input length of the circuit description \cite{arora2009computational}.
An instance is encoded as:
\begin{itemize}
\item a finite action set $A$ given explicitly,
\item coordinate domains $X_1,\ldots,X_n$ given by their sizes in binary,
\item a Boolean or arithmetic circuit $C_U$ that on input $(a,s)$ outputs $U(a,s)$.
\end{itemize}
The input length is $L = |A| + \sum_i \log |X_i| + |C_U|$. Polynomial time and all complexity classes (\coNP, $\Sigma_2^P$, ETH, W[2]) are measured in $L$. All hardness results in Section~\ref{sec:hardness} use this encoding.

\paragraph{Explicit-state encoding (used for enumeration algorithms and experiments).}
The utility is given as a full table over $A \times S$. The input length is $L_{\text{exp}} = \Theta(|A||S|)$ (up to the bitlength of utilities). Polynomial time is measured in $L_{\text{exp}}$. Results stated in terms of $|S|$ use this encoding.

Unless explicitly stated otherwise, ``polynomial time'' refers to the succinct encoding.

\subsection{Structural Complexity vs Representational Hardness}

\begin{definition}[Structural Complexity]\label{def:structural-complexity}
For a fixed formal decision relation (e.g., ``$I$ is sufficient for $\mathcal{D}$''), \emph{structural complexity} means its placement in standard complexity classes within the formal model (coNP, $\Sigma_2^P$, etc.), as established by class-membership arguments and reductions.
\end{definition}

\begin{definition}[Representational Hardness]\label{def:representational-hardness}
For a fixed decision relation and an encoding regime $E$ (Section~\ref{sec:encoding}), \emph{representational hardness} is the worst-case computational cost incurred by solvers whose input access is restricted to $E$.
\end{definition}

\begin{remark}[Interpretation Contract]
This paper keeps the decision relation fixed and varies the encoding regime explicitly. Thus, later separations are read as changes in representational hardness under fixed structural complexity, not as changes to the underlying sufficiency semantics.
\end{remark}


% Include the hardness proofs (already developed)
\input{hardness_proofs.tex}
