\section{Substrate Cost}\label{sec:substrate-cost}

We extend the integrity-competence matrix from Paper 4 with substrate cost $\kappa$.

\subsection{Definition}

\begin{definition}[Substrate Cost Function]\label{def:substrate-cost}
$\kappa : (\text{MatrixCell} \times \text{SubstrateType}) \to (\mathbb{R}_{\ge 0}, \leq)$

Where:
\begin{itemize}
\item MatrixCell = (integrity, attempted, competenceAvailable) triple from Paper 4
\item SubstrateType = opaque type representing substrate (silicon, carbon, biochemical, etc.)
\item The codomain is a totally ordered cost space
\end{itemize}
\end{definition}

\begin{claim}{Verdict is substrate-independent}{th:substrate-independence-verdict}
For all cells $c$ and substrate types $\tau_1, \tau_2$:
\[
\text{verdict}(c; \tau_1) = \text{verdict}(c; \tau_2).
\]
\claimstamp{T\ref{th:substrate-independence-verdict}}{RG}
\end{claim}

\leanmeta{\LH{CC64}}

\begin{claim}{Trajectory is substrate-dependent}{th:substrate-dependence-trajectory}
There exist substrate types $\tau_1, \tau_2$ and problem sequences $\sigma$ such that:
\[
\text{trajectory}(\sigma; \tau_1) \neq \text{trajectory}(\sigma; \tau_2)
\]
even though $\text{verdict}(\sigma_t; \tau_1) = \text{verdict}(\sigma_t; \tau_2)$ for all $t$.
\claimstamp{T\ref{th:substrate-dependence-trajectory}}{RG}
\end{claim}

\leanmeta{\LH{CC63}}

\subsection{What $\kappa$ Captures}

$\kappa$ is a placeholder for everything the $(A, S, U)$ tuple doesn't model:
\begin{itemize}
\item interface latency / switching overhead
\item thermal dissipation limits
\item noise floor and correction burden
\item material degradation rate
\item energy-density constraints
\end{itemize}

We do NOT decompose $\kappa$. We acknowledge it exists, show where it enters the dynamics, and prove that the matrix verdict is $\kappa$-independent while the trajectory is $\kappa$-dependent.

\begin{claim}{Minimal formal hook}{th:substrate-kappa-minimal}
$\kappa(\text{cell}, \text{substrateType})$ is the sufficient statistic of substrate for the trajectory problem.
\claimstamp{T\ref{th:substrate-kappa-minimal}}{RG}
\end{claim}

\leanmeta{\LH{CC65}}

This is the clean separation: the model captures exactly the behaviorally relevant component of substrate, and nothing more.

We do NOT claim that $\kappa$ captures every microscopic property of substrate physics. We claim that $\kappa$ captures the \emph{decision-relevant} substrate component that affects trajectory through the problem sequence.

\subsection{Where Substrate Enters}

The matrix verdict is computed from (integrity, attempted, competenceAvailable). None of these depend on substrate. The verdict is universal.

Substrate enters at exactly one point: the COST of being in a cell. This cost determines:
\begin{itemize}
\item Whether the system attempts the next instance or pauses
\item How the system allocates resources across a sequence
\item The effective adaptation rate under the substrate
\item The abstention threshold under uncertainty and budget limits
\end{itemize}

This connects to Paper 5 (Credibility): $\kappa = \kappa_{\text{compute}} + \kappa_{\text{communicate}}$, where $\kappa_{\text{communicate}}$ is bounded by costly/cheap signaling constraints.
