\section{Conclusion}

\subsection*{Methodology and Disclosure}

\textbf{Role of LLMs in this work.} This paper was developed through human-AI collaboration. The author provided the core intuitions: the connection between decision-relevance and computational complexity, the conjecture that SUFFICIENCY-CHECK is \coNP-complete, and the insight that the $\SigmaP{2}$ structure collapses for MINIMUM-SUFFICIENT-SET. Large language models (Claude, GPT-4) served as implementation partners for proof drafting, Lean formalization, and \LaTeX{} generation.

The Lean 4 proofs were iteratively refined: the author specified the target statements, the LLM proposed proof strategies, and the Lean compiler served as the arbiter of correctness. The complexity-theoretic reductions required careful human oversight to ensure the polynomial bounds were correctly established.

\textbf{What the author contributed:} The problem formulations (SUFFICIENCY-CHECK, MINIMUM-SUFFICIENT-SET, ANCHOR-SUFFICIENCY), the hardness conjectures, the tractability conditions, and the connection to over-modeling in engineering practice.

\textbf{What LLMs contributed:} \LaTeX{} drafting, Lean tactic exploration, reduction construction assistance, and prose refinement.

The proofs are machine-checked; their validity is independent of generation method. This methodology is disclosed for academic transparency.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection*{Summary of Results}

This paper establishes the computational complexity of coordinate sufficiency problems:

\begin{itemize}
\item \textbf{SUFFICIENCY-CHECK} is \coNP-complete (Theorem~\ref{thm:sufficiency-conp})
\item \textbf{MINIMUM-SUFFICIENT-SET} is \coNP-complete (Theorem~\ref{thm:minsuff-conp})
\item \textbf{ANCHOR-SUFFICIENCY} is $\SigmaP{2}$-complete (Theorem~\ref{thm:anchor-sigma2p})
\item An encoding-regime separation contrasts explicit-state polynomial-time (polynomial in $|S|$) with a succinct worst-case ETH lower bound witnessed by a hard family with $k^*=n$ (Theorem~\ref{thm:dichotomy})
\item Full intermediate query-access lower bounds are formalized as a finite-state Opt-oracle core ($\Omega(|S|)$, Boolean instantiation $\Omega(2^n)$) plus Boolean-interface refinements for value-entry and state-batch access, with explicit subproblem-to-full transfer, weighted randomized robustness, and oracle-lattice transfer/strictness closures (Propositions~\ref{prop:query-regime-obstruction}--\ref{prop:oracle-lattice-strict})
\item Tractable subcases exist for explicit-state encoding, separable utility, and tree-structured utility with explicit local factors (Theorem~\ref{thm:tractable})
\item Discrete-time interface semantics are formalized: time is $\mathbb{N}$-valued, decision events are exactly one-tick transitions, run length equals elapsed time, and substrate-consistent realizations preserve the same one-unit event law (Propositions~\ref{prop:time-discrete}--\ref{prop:substrate-unit-time})
\end{itemize}

These results place the problem of identifying decision-relevant coordinates at the first and second levels of the polynomial hierarchy.

Beyond classification, the paper contributes a formal claim-typing framework (Section~\ref{sec:interpretive-foundations}): structural complexity is a property of the fixed decision question, while representational hardness is regime-conditional access cost. This is why encoding-regime changes can move practical hardness without changing the underlying semantics.

The reduction constructions and key equivalence theorems are machine-checked in Lean 4 (see Appendix~\ref{app:lean} for proof listings). The formalization verifies that the TAUTOLOGY reduction correctly maps tautologies to sufficient coordinate sets; complexity classifications (coNP-completeness, $\SigmaP{2}$-completeness) follow by composition with standard complexity-theoretic results (TAUTOLOGY is \coNP-complete, $\exists\forall$-SAT is $\SigmaP{2}$-complete). The strengthened gadget showing that non-tautologies yield instances with \emph{all coordinates relevant} is also formalized.

\subsection*{Complexity Characterization}

The results provide precise complexity characterizations within the formal model:

\begin{enumerate}
\item \textbf{Exact bounds.} SUFFICIENCY-CHECK is \coNP-complete, both \coNP-hard and in \coNP.

\item \textbf{Constructive reductions.} The reductions from TAUTOLOGY and $\exists\forall$-SAT are explicit and machine-checked.

\item \textbf{Encoding-regime separation.} Under [E], SUFFICIENCY-CHECK is polynomial in $|S|$. Under [S+ETH], there exist succinct worst-case instances (with $k^*=n$) requiring $2^{\Omega(n)}$ time. Under [Q\_fin], the Opt-oracle core has $\Omega(|S|)$ worst-case query complexity (Boolean instantiation $\Omega(2^n)$), and under [Q\_bool] the value-entry/state-batch refinements preserve the obstruction with weighted-cost transfer closures.
\end{enumerate}

\subsection*{The Complexity Redistribution Corollary}

Section~\ref{sec:simplicity-tax} develops a quantitative consequence: when a problem requires $k$ dimensions and a model handles only $j < k$ natively, the remaining $k - j$ dimensions must be handled externally at each decision site. For $n$ sites, total external work is $(k-j) \times n$.\claimstamp{T\ref{thm:tax-grows}}{LC}

The set identity is elementary; its operational content comes from composition with the hardness results on exact relevance minimization. This redistribution corollary is formalized in Lean 4 (\texttt{HardnessDistribution.lean}), proving:
\begin{itemize}
\item Redistribution identity: complexity burden cannot be eliminated by omission, only moved between native handling and external handling
\item Dominance: complete models have lower total work than incomplete models
\item Amortization: there exists a threshold $n^*$ beyond which higher-dimensional models have lower total cost
\end{itemize}

\subsection*{Open Questions}

The landscape above is complete for the static sufficiency class under C1--C4 and the declared regimes; the items below are adjacent-class extensions or secondary refinements.
Several questions remain for future work:
\begin{itemize}
\item \textbf{Fixed-parameter tractability (primary):} Is SUFFICIENCY-CHECK FPT when parameterized by the minimal sufficient-set size $k^*$, or is it W[2]-hard under this parameterization?
\item \textbf{Sequential/stochastic bridge extension:} Characterize the exact frontier where adjacent sequential objectives reduce to the static class via Proposition~\ref{prop:one-step-bridge}, and where genuinely new complexity objects (e.g., horizon/sample/regret complexity) must replace the present coNP/\(\Sigma_2^P\) analysis.
\item \textbf{Average-case complexity:} What is the complexity under natural distributions on decision problems?
\item \textbf{Typed-confidence calibration:} For signaled reports $(r,g,p_{\mathrm{self}},p_{\mathrm{cert}})$, characterize proper scoring and calibration objectives that preserve signal consistency and typed admissibility.
\item \textbf{Learning cost formalization:} Can central cost $H_{\text{central}}$ be formalized as the rank of a concept matroid, making the amortization threshold precisely computable?
\end{itemize}

\subsection*{Practical Corollaries}

The practical corollaries are regime-indexed and theorem-indexed:
\begin{itemize}
\item \textbf{[E] and structured regimes:} polynomial-time exact procedures exist (Theorem~\ref{thm:tractable}).
\item \textbf{[Q\_fin]/[Q\_bool] query-access lower bounds:} worst-case Opt-oracle complexity is $\Omega(|S|)$ in the finite-state core (Boolean instantiation $\Omega(2^n)$), and value-entry/state-batch interfaces satisfy the same obstruction in the mechanized Boolean refinement (Propositions~\ref{prop:query-regime-obstruction}--\ref{prop:oracle-lattice-strict}), with randomized weighted robustness and oracle-lattice closures.
\item \textbf{[S+ETH] hard families:} exact minimization inherits exponential worst-case cost (Theorem~\ref{thm:dichotomy} together with Theorem~\ref{thm:sufficiency-conp}).
\item \textbf{[S\_bool] mechanized criterion:} minimization reduces to relevance-cardinality constraints (Corollary~\ref{cor:practice-diagnostic}).
\item \textbf{Redistribution consequences:} omitted native coverage externalizes work with explicit growth/amortization laws (Theorems~\ref{thm:tax-conservation}--\ref{thm:amortization}).
\end{itemize}

Hence the design choice is typed: enforce a tractable regime, or adopt weakened guarantees with explicit verification boundaries.\claimstamp{T\ref{thm:tractable},\ref{thm:dichotomy}}{RG}
Equivalently, under the attempted-competence matrix, over-specification is rational only in attempted-competence-failure cells; once exact competence is available in the active regime (or no attempted exact competence was made), persistent over-specification is irrational for the same verified-cost objective.\claimstamp{P\ref{prop:attempted-competence-matrix}}{AR}
By Proposition~\ref{prop:attempted-competence-matrix}, the integrity-preserving matrix contains 3 irrational cells and 1 rational cell.\claimstamp{P\ref{prop:attempted-competence-matrix}}{AR}

\subsection*{Epistemic Alignment Implications for Agent Training}

Within this paper's formalism, epistemic alignment is a claim-discipline property: outputs are asserted only when certifiable, and otherwise emitted as abstain reports (optionally carrying tentative guesses and self-reflected confidence under zero certified confidence).
Formally, this is the integrity/competence split of Section~\ref{sec:interpretive-foundations}, not a full value-alignment claim.\claimstamp{D\ref{def:solver-integrity},\ref{def:competence-regime};P\ref{prop:integrity-competence-separation}}{TR}

\paragraph{Truth-Objective Abstention Requirement.}
If a solver-agent declares a truth-preserving objective for claims about relation $\mathcal{R}$ in regime $\Gamma$, uncertified outputs are admissible only as $\mathsf{ABSTAIN}/\mathsf{UNKNOWN}$ rather than asserted answers.
Under the signal discipline, this includes abstain-with-guess/self-confidence reports while keeping certification evidence-gated.\claimstamp{D\ref{def:signaled-typed-report},\ref{def:signal-consistency};P\ref{prop:certified-confidence-gate},\ref{prop:abstain-guess-self-signal};C\ref{cor:exact-no-competence-zero-certified}}{AR}
Equivalently, training policies that suppress abstention in uncertified regions violate the typed claim discipline under the declared task model.\claimstamp{D\ref{def:solver-integrity};P\ref{prop:integrity-competence-separation},\ref{prop:integrity-resource-bound},\ref{prop:attempted-competence-matrix}}{AR}

This yields a training doctrine for solver-agents under declared tasks/regimes:
\begin{enumerate}
\item Treat $\mathsf{ABSTAIN}/\mathsf{UNKNOWN}$ as first-class admissible outputs, not errors.\claimstamp{D\ref{def:solver-integrity}}{TR}
\item For signaled reports, maintain two channels: self-reflected confidence may be emitted, but certified confidence is positive only with evidence.\claimstamp{D\ref{def:signal-consistency};P\ref{prop:certified-confidence-gate},\ref{prop:self-confidence-not-certification}}{AR}
\item Treat runtime disclosure as scalar-witnessed: always emit exact step-count, and emit completion percentage only when a declared bound exists.\claimstamp{P\ref{prop:steps-run-scalar},\ref{prop:no-fraction-without-bound},\ref{prop:fraction-defined-under-bound}}{AR}
\item Optimize competent coverage subject to integrity constraints, rather than optimizing answer-rate alone.\claimstamp{D\ref{def:competence-regime};P\ref{prop:integrity-competence-separation}}{AR}
\item In hardness-blocked regimes, abstention or explicit guarantee-weakening is structurally required unless regime assumptions are changed.\claimstamp{T\ref{thm:dichotomy};P\ref{prop:integrity-resource-bound},\ref{prop:attempted-competence-matrix}}{S,S+ETH}
\end{enumerate}

Operationally, a reporting schema consistent with this framework tracks at least five quantities: integrity violations, competent non-abstaining coverage on the declared scope, abstention quality under regime shift, the calibration gap between self-reflected and certified confidence channels, and scalar runtime witness distributions (with bound-qualified fraction semantics when bounds are declared). This is consistent with bounded-rationality feasibility discipline under declared resource models \cite{sep_bounded_rationality,arora2009computational}.


\appendix
