We study \emph{physical information sufficiency} as a decision-theoretic meta-problem. For \(\mathcal{D}=(A,S,U)\) with factored state space \(S=X_1\times\cdots\times X_n\), a coordinate set \(I\) is sufficient when
\[
s_I=s'_I \implies \Opt(s)=\Opt(s').
\]
This asks whether projected information preserves the full optimal-action correspondence.

\textbf{Complexity landscape.}
\begin{itemize}
\item SUFFICIENCY-CHECK is \coNP-complete; MINIMUM-SUFFICIENT-SET is \coNP-complete; ANCHOR-SUFFICIENCY is \(\SigmaP{2}\)-complete. Informally, this gap separates verification (checking whether a given coordinate set suffices) from discovery (finding a grounding under which the optimal action is determined), and suggests the latter is structurally harder unless the polynomial hierarchy collapses.
\item Under explicit-state encoding, polynomial-time algorithms hold for bounded actions, separable utility, and tree-structured utility.
\item Under succinct encoding, hardness is regime-dependent: with ETH, there are worst-case families with \(k^*=n\) requiring \(2^{\Omega(n)}\) time.
\item Under query access, the finite-state core has worst-case Opt-oracle complexity \(\Omega(|S|)\), with Boolean value-entry and state-batch refinements preserving the obstruction.
\end{itemize}

\textbf{Physical grounding.}
The paper formalizes a physical-to-core encoding \(E:\mathcal{P}\to\mathcal{D}\) and a transport rule: declared physical assumptions transfer to core assumptions, and core claims lift back to encoded physical instances. Encoded physical counterexamples induce core failures on the encoded slice. Discrete-time interface semantics (decision event = one tick) and budgeted thermodynamic lifts (bit lower bounds to energy/carbon lower bounds under declared constants) are formalized in the same assumption-typed framework.
\claimstamp{P\ref{prop:physical-claim-transport},\ref{prop:law-instance-objective-bridge},\ref{prop:time-discrete},\ref{prop:decision-unit-time},\ref{prop:run-time-accounting},\ref{prop:substrate-unit-time},\ref{prop:budgeted-crossover},\ref{prop:thermo-lift};C\ref{cor:physical-counterexample-core-failure},\ref{cor:finite-budget-threshold-impossibility};T\ref{thm:physical-bridge-bundle}}{AR}
\leanmeta{\LH{CT6},\LH{CT8},\LH{CT4},\LH{CT1},\LH{DQ548},\LH{DQ537},\LH{DQ539},\LH{DQ541},\LH{PBC1},\LH{CC65},\LH{DQ569}.}

All theorem-level claims are machine-checked in Lean 4 (\LeanTotalLines\ lines, \LeanTotalTheorems\ theorem/lemma statements). Complexity-class completeness follows by composition with standard complexity results; regime-dependent and physical-transport consequences are proved as assumption-explicit closures.

\textbf{Keywords:} computational complexity, physical information, decision theory, polynomial hierarchy, formal verification
