\section{Introduction}\label{sec:introduction}

The decision-relevance problem asks which parts of a state description matter for optimal action. This raises two questions: how hard relevance is to certify, and what structure emerges once the relevant-coordinate support has been identified. This paper addresses the second.

The unifying object is the optimizer quotient. States are identified when they induce the same optimal-action set, and the resulting quotient is the coarsest abstraction that preserves the decision boundary. In \textbf{Set}, this is the coimage/image factorization of the optimizer map. The question is then which quantitative summaries of that quotient recover the same structural content.

Several mathematical frameworks can be compared against the same invariant. The support size of the relevant coordinates, written $\mathrm{srank}$, appears directly in Fisher-style support counting, bounds quotient entropy, and constrains zero-distortion summaries through the optimizer quotient. The Bayesian chain is parallel rather than another recovery theorem: it shows that the same finite counting setup also supports the standard Bayes-optimality identities. Convergence therefore means that these frameworks are organized around the same notion of decision-relevant structure, not that they all yield the same formula.

The novelty is not in any individual ingredient. Entropy, Bayes, quotient objects, and Fisher information are classical. The contribution is the theorem-level synthesis: once the optimizer quotient is fixed as the canonical decision-preserving abstraction, these frameworks can be compared in a single formal setting and related to the same structural notion of decision-relevant dimension.

The argument is built from shared definitions, quotient structure, and information-theoretic inequalities.

\subsection{Contributions}

\begin{enumerate}
\item \textbf{Counting-gap preliminaries.} We isolate a discrete counting bound and a measure-before-probability theorem that place normalization and Bayesian identities on explicit finite setup.

\item \textbf{Optimizer quotient as canonical structure.} We restate the shared setup needed for independent readability, including an explicit structural-rank support proposition, and identify the optimizer quotient as the canonical decision-preserving abstraction.

\item \textbf{Structural-rank comparison across frameworks.} We show that Fisher-style support counting recovers $\mathrm{srank}$ exactly, quotient entropy is bounded by $\mathrm{srank}$, lossless zero-distortion summaries are constrained by the same optimizer-quotient structure, and the transport bridge tracks the same quotient branching regimes.

\item \textbf{Bayesian optimality from minimal assumptions.} We derive probability normalization, Bayes' rule, cross-entropy decomposition, and Bayes optimality under log loss from finite counting and the inequality $\log x \le x - 1$.

\item \textbf{Machine-checked support.} The Lean development records the quotient, Bayesian, and information-theoretic lemmas used in the paper.
\end{enumerate}

\subsection{Paper Structure}

Section~\ref{sec:counting-preliminaries} states the counting and measure preliminaries. Section~\ref{sec:formal-setup} restates the shared quotient setup needed for independence. Section~\ref{sec:convergence-frameworks} develops the structural-rank convergence story. Section~\ref{sec:bayesian-optimality} gives the Bayesian optimality chain. Section~\ref{sec:related} situates the paper relative to information theory, sufficient statistics, and formalized mathematics. Appendix~\ref{app:lean} summarizes the corresponding Lean support.

\subsection{Proof Strategy at a Glance}

Each theorem in the body is proved by a short explicit chain:
\begin{enumerate}
\item \textbf{Counting layer (Section~\ref{sec:counting-preliminaries}):} finite counting bounds and normalization identities.
\item \textbf{Quotient layer (Section~\ref{sec:formal-setup}):} sufficiency/relevance definitions and universal property of the optimizer quotient.
\item \textbf{Bridge layer (Section~\ref{sec:convergence-frameworks}):} entropy bounds, support counting, zero-distortion summaries, and transport regimes all tied back to the same optimizer-quotient core.
\item \textbf{Statistical consequences (Section~\ref{sec:bayesian-optimality}):} Bayes optimality on the normalized finite-probability side.
\end{enumerate}

This removes ambiguity about where each claim comes from: no hidden assumptions, no skipped quantifier steps, and no dependence on informal analogies.
