\section{Related Work}\label{sec:related}

\subsection{Sufficient Statistics and Information Reduction}

The closest classical analogue to coordinate sufficiency is the theory of sufficient statistics \cite{cover2006elements,csiszar2011information,savage1954foundations}. In both settings the goal is to replace a large object by a smaller one without losing decision-relevant structure. The difference here is one of object and emphasis: the retained information is indexed by optimal-action preservation rather than parameter inference, and the central summary is the optimizer quotient rather than a statistic defined relative to a parametric family.

\subsection{Information Geometry and Bayesian Decision Theory}

The Bayesian chain used here is standard: cross-entropy decomposes into entropy plus KL divergence, and Gibbs' inequality yields Bayes optimality under log loss \cite{cover2006elements,csiszar2011information,raiffa1961applied}. What is distinctive here is the way those classical identities are attached to the optimizer quotient and to structural rank, so that Bayesian optimality appears as one part of a larger convergence story rather than as an isolated fact.

\subsection{Rate-Distortion and Lossless Compression}

The link between lossless compression and decision structure is conceptually close to the minimum-description-length viewpoint \cite{shannon1959coding,blahut1972computation,rissanen1978modeling,grunwald2007minimum}. We focus on the zero-distortion boundary relevant to exact decision preservation: once only decision-equivalent states may be merged, the quotient structure fixes the admissible compression target. This places rate-distortion language in direct contact with the relevant-coordinate support.

\subsection{Categorical and Quotient Viewpoints}

The optimizer quotient is a familiar object in \textbf{Set}: it is the coimage/image factorization of the optimizer map \cite{maclane1998categories}. The contribution here is not novelty of construction but novelty of use. The quotient is the organizing object through which the Bayesian, entropy, and structural-rank viewpoints can be compared within a single framework.

\subsection{Formalized Mathematics and Verified Complexity}

Formal verification has made major progress in algebra, analysis, and basic computability \cite{forster2019verified,kunze2019formal,nipkow2002isabelle,mathlib2020,demoura2021lean4}. The present work belongs to the growing body of research that uses mechanization to stabilize arguments spanning more than one mathematical viewpoint. Here the mechanized layer supports quotient-theoretic, information-theoretic, and Bayesian statements within a single development.
