\section{Lean 4 Proof Listings}\label{app:lean}

The Lean development contains the shared definitions and the quotient-, Bayesian-, and information-theoretic lemmas used by this paper. This appendix lists the support most relevant to the main results.

\subsection{What Is Mechanized}

The Lean development records:
\begin{itemize}
\item core definitions of decision problems, sufficiency, relevance, and optimizer quotients,
\item quotient/image-factorization lemmas for the optimizer map,
\item finite probability and Bayesian optimality lemmas,
\item structural-rank support lemmas and related information-theoretic bridge statements.
\end{itemize}

\subsection{Lean Handle ID Map}\label{sec:lean-handle-id-map}
This appendix lists the handle families most relevant to the results of the paper:
\begin{itemize}
\item \texttt{QT1}, \texttt{QT2}, \texttt{QT3}, \texttt{QT7} for quotient/coimage factorization;
\item \texttt{SK1}, \texttt{SK2}, \texttt{SK3} for structural-rank support facts;
\item \texttt{IT3}, \texttt{IT4} for quotient-entropy bounds;
\item \texttt{FS1}, \texttt{FS2} for Fisher-style structural-rank recovery;
\item \texttt{RS1}, \texttt{RS2}, \texttt{RS3}, \texttt{RS4}, \texttt{RD1}, \texttt{RD2}, \texttt{RD3} for rate-distortion and lossless-summary statements;
\item \texttt{BC1}--\texttt{BC5}, \texttt{FN7}, \texttt{FN8}, \texttt{FN12}, \texttt{FN14} for the finite Bayesian chain.
\end{itemize}

These are the handles most relevant to the claims developed in the manuscript. They provide a focused map from the text to the corresponding mechanized support.

\subsection{Representative Modules}

\begin{itemize}
\item \texttt{DecisionQuotient/Quotient.lean} -- quotient/coimage structure for the optimizer map
\item \texttt{DecisionQuotient/BayesFoundations.lean} -- finite probability and cross-entropy identities
\item \texttt{DecisionQuotient/BayesOptimalityProof.lean} -- Bayes optimality under log loss
\item \texttt{DecisionQuotient/Information.lean} and \texttt{DecisionQuotient/Information/RateDistortion.lean} -- quotient entropy and compression-facing lemmas
\item \texttt{DecisionQuotient/Statistics/FisherInformation.lean} -- support-count style structural-rank lemmas
\end{itemize}

\subsection{Scope}

For readability, this appendix provides a focused manual summary rather than repository-level omnibus tables. A paper-specific handle table should include only the statements proved or cited here, together with the Lean handles that support them.
