\section{Axis Inclusion Requirements and Fixed-Axis Incompleteness}\label{universal-dominance}

\textbf{Framing note:} This section establishes dominance relations \emph{within} fixed-axis systems. Section~\ref{axis-parametric-theory} proves the stronger result: any fixed-axis system is incomplete for some domain. The correct approach is axis-parametric (deriving the minimal axis set from requirements).

\textbf{Thought experiment:} What if \texttt{type()} only took
namespace?

Given that the semantic core is (bases, namespace), what if we further
reduce to just namespace?

\begin{verbatim}
# Hypothetical minimal class constructor
def type_minimal(namespace: dict) -> type:
    """Create a class from namespace only."""
    return type("", (), namespace)
\end{verbatim}

\textbf{Definition 3.1 (Namespace-Only System).} A namespace-only class
system is one where: - Classes are characterized entirely by their
namespace (attributes/methods) - No explicit inheritance mechanism
exists (bases axis absent)

\textbf{Theorem 3.1 (Structural Typing Is Correct for Namespace-Only
Systems).}

In a namespace-only system, structural typing is the unique correct
typing discipline.

\emph{Proof.} 1. Let A and B be classes in a namespace-only system 2. A
\(\equiv\) B iff namespace(A) = namespace(B) (by definition of
namespace-only) 3. Structural typing checks: namespace(x) \(\supseteq\)
signature(T) 4. This is the only information available for type checking
5. Therefore structural typing is correct and complete. \qed

\textbf{Corollary 3.2 (Go's Design Is Consistent).} Go has no
inheritance. Interfaces are method sets. Structural typing is correct
for Go.

\textbf{Corollary 3.3 (TypeScript's Static Type System).} TypeScript's
\emph{static} type system is structural. Class compatibility is
determined by shape, not inheritance. However, at runtime, JavaScript's
prototype chain provides nominal identity (\texttt{instanceof} checks
the chain) \cite{mdnPrototypeChain}. This creates a coherence tension
discussed in Section 8.7.

\textbf{The Critical Observation (Semantic Axes):}

\small
\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
System & Semantic Axes & Correct Discipline \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Namespace-only & \texttt{(namespace)} & Structural \\
Full Python & \texttt{(bases,\ namespace)} & Nominal \\
\end{longtable}

The syntactic name (\texttt{\_\_name\_\_}) is metadata in both cases---it doesn't affect
which typing discipline is correct. See Definition 2.7 for the distinction between syntactic names and semantic identity.

\textbf{Binder vs.~Observable Identity.} In \emph{pure} structural
typing, ``name'' is only a binder/alias for a shape; it is not an
observable discriminator. Conformance depends solely on namespace
(structure). Any observable discriminator (brand/tag/nominal identity)
is an added axis: once it is observable to the conformance relation, the
discipline is no longer purely structural.

\textbf{Lineage axis = ordered identities.} The Bases axis \(B\) can be
viewed as the ordered lineage \(\text{MRO}(T)\) (C3 linearization). The
``identity'' capability is a projection of this lineage:
\(\text{head}(\text{MRO}(T)) = T\) (exact type), and instance-of is
membership \(U \in \text{MRO}(T)\). Provenance and conflict resolution
are the other projections. There is no separate ``I axis''; identity is
one of the queries made available by \(B\). A discipline that can only
test \(\text{head}(\text{MRO}(T))\) has tag identity but not inheritance
capabilities; it is a strict subset of full \(B\).

\textbf{Theorem 3.4 (Axis Inclusion Dominance).} When a
\texttt{bases} axis exists in the class system ($B \neq \emptyset$), systems including $\{B, S\}$
Pareto-dominate all systems using only $\{S\}$ (provides strictly more
capabilities with zero additional cost). This dominance is
universal, not limited to greenfield development.

\textbf{Remark:} In type system terminology, $\{B, S\}$-inclusive systems are called ``nominal typing''; $\{S\}$-only systems are called ``structural typing.'' The theorem establishes that axis inclusion is determined by domain requirements, not preference.

\emph{Proof.} We prove this in two steps: (1) strict dominance holds
unconditionally, (2) retrofit constraints do not constitute an
exception.

\textbf{Step 1: Strict Dominance is Unconditional.}

Let \(D_{\{S\}}\) be any system using only the $\{S\}$ axis (shape-based). Let \(D_{\{B,S\}}\) be a system using both axes.

By Theorem 2.15 (Axis Lattice Dominance):
\[\text{capabilities}(D_{\{S\}}) \subseteq \text{capabilities}(D_{\{B,S\}})\]

By Theorem 2.17 (Capability Completeness), \(D_{\{B,S\}}\)
provides four capabilities that \(D_{\{S\}}\) cannot: provenance,
identity, enumeration, conflict resolution.

Therefore:
\(\text{capabilities}(D_{\{S\}}) \subset \text{capabilities}(D_{\{B,S\}})\)
(strict subset).

This dominance holds \textbf{regardless of whether the system currently
uses these capabilities}. The capability gap exists by the structure of
axis subsets, not by application requirements.

\textbf{Step 2: Retrofit Constraints Do Not Constitute an Exception.}

One might object: ``In retrofit contexts, external types cannot be made
to inherit from my ABCs, so nominal typing is unavailable.''

This objection was addressed in Theorem 2.10j (Protocol Dominated by
Adapters): when \(B \neq \emptyset\), nominal typing with adapters
provides all capabilities of Protocol plus four additional capabilities.
The ``retrofit exception'' is not an exception. Adapters are the
mechanism that makes nominal typing universally available.

\begin{itemize}
\tightlist
\item
  External type cannot inherit from your ABC? Wrap it in an adapter that
  does.
\item
  Protocol avoids the adapter? Yes, but avoiding adapters is a
  convenience, not a capability (Corollary 2.10k).
\end{itemize}

\textbf{Conclusion: Systems Excluding Required Axes Have Negative Expected Value.}

Given two available options \(A\) and \(B\) where
\(\text{capabilities}(A) \subset \text{capabilities}(B)\) and
\(\text{cost}(A) = \text{cost}(B)\), choosing \(A\) is
\textbf{Pareto-dominated}: there exists no rational justification for
\(A\) over \(B\) under expected utility maximization.

When \(B \neq \emptyset\):
\begin{itemize}
\item Systems using only $\{S\}$ are Pareto-dominated by systems using $\{B, S\}$
\item Same declaration cost: one class definition per interface for both (Definition 2.14c)
\item Foreclosure is permanent: Missing capabilities cannot be added later (Theorem 3.67)
\item Ignorant choice has expected cost: \(E[\text{gap}] > 0\) (Theorem 3.68)
\item Retrofit cost dominates analysis cost: \(\text{cost}_{\text{retrofit}} > \text{cost}_{\text{analysis}}\) (Theorem 3.69)
\item Analysis has positive expected value: \(E[V_{\text{analysis}}] > 0\) (Theorem 3.70)
\end{itemize}

Therefore: \textbf{Excluding the $B$ axis when inheritance exists ($B \neq \emptyset$)
has negative expected value under capability-based utility.}

\textbf{Note:} In type system terminology, ``nominal typing'' denotes $\{B, S\}$-inclusive systems; ``structural typing'' denotes $\{S\}$-only systems.

\textbf{Note on ``what if I don't need the extra capabilities?''}

This objection misunderstands option value. A Pareto-dominated choice
has negative expected value \textbf{even if the additional capabilities
are never exercised}:

\begin{enumerate}
\item Capability availability has zero marginal cost (identical \texttt{isinstance()} syntax)
\item Future requirements are uncertain; capability foreclosure has negative option value (Theorem 3.70)
\item Capability gaps are irreversible: cannot transition from shape to nominal without discipline rewrite (Theorem 3.67)
\item Architecture choices have persistent effects; one-time decisions determine long-term capability sets
\end{enumerate}

The \texttt{bases} axis creates an information asymmetry: nominal typing
can access inheritance structure; shape-based typing cannot. Adapters
ensure nominal typing is universally available.

\textbf{Theorem 3.71 (Unique Optimum):} Under capability-based utility
maximization, nominal typing is the unique optimal choice when \(B \neq
\emptyset\). Choosing shape-based typing while maximizing capabilities
contradicts the stated objective (Theorem 3.72: proven incoherence).
\qed

\textbf{Theorem 3.5 (Axis Inclusion Dominance, Concrete).} Systems including $\{B, S\}$
strictly dominate systems using only $\{S\}$ whenever \(B \neq \emptyset\):
$\{B, S\}$ systems provide all capabilities of $\{S\}$-only systems plus additional
capabilities, at equal or lower cost.

\textbf{Remark:} This theorem is traditionally stated as ``nominal typing strictly dominates structural typing.'' We state it in axis terms to emphasize that the result follows from axis inclusion, not from advocacy for a particular discipline name.

\emph{Proof.} Consider Python's concrete implementations: - $\{S\}$-only (coherent):
\texttt{typing.Protocol} (structural typing) - $\{B, S\}$:Abstract Base
Classes (ABCs, nominal typing)

Let S = capabilities provided by Protocol, N = capabilities provided by
ABCs.

\textbf{What Protocols ($\{S\}$-only) provide:} 1. Interface enforcement via method
signature matching 2. Type checking at static analysis time (mypy,
pyright) 3. No runtime isinstance() check (by default)

\textbf{What ABCs ($\{B, S\}$) provide:} 1. Interface enforcement via
\texttt{@abstractmethod} (equivalent to Protocol) 2. Type checking at
static analysis time (equivalent to Protocol) 3. \textbf{Type identity
via isinstance()} (Protocol cannot provide this) 4. \textbf{Provenance
tracking via MRO position} (Protocol cannot provide this) 5.
\textbf{Exhaustive enumeration via \texttt{\_\_subclasses\_\_()}}
(Protocol cannot provide this) 6. \textbf{Type-as-dictionary-key via
type() identity} (Protocol cannot provide this) 7. \textbf{Runtime
enforcement at instantiation} (Protocol only checks statically)

Therefore S \(\subset\) N (strict subset). Both require explicit type
declarations. The declaration cost is equivalent: one class definition
per interface. Therefore, $\{B, S\}$ systems provide strictly more
capabilities at equal or lower cost (earlier failure). \qed

\textbf{Corollary 3.6 (Axis Exclusion Incorrectness).} When
\(B \neq \emptyset\), excluding $B$ from the axis set is not suboptimal: it is incorrect.

\emph{Proof.} By Theorem 3.5, systems including $\{B, S\}$ strictly dominate
systems using only $\{S\}$. By Theorem 2.10j, adapters make $\{B, S\}$ systems
universally available. Choosing a strictly dominated option when the
superior option is available is definitionally incorrect.
\qed

\textbf{Remark:} In type system terminology, this means structural typing is incorrect when inheritance exists; nominal typing is correct.

\subsection{The Absolute Claim}\label{the-absolute-claim}

\textbf{Claim (Axis Inclusion Correctness).} In any programming
language with explicit inheritance syntax (\(B \neq \emptyset\)),
systems excluding the $B$ axis (structural typing, Protocol-based typing) are \textbf{incorrect}.
Duck typing is \textbf{incoherent} (Theorem 2.10d). It is not even a valid typing discipline.
Systems including $\{B, S\}$ (nominal typing) are \textbf{correct}. This
is not a preference, recommendation, or tradeoff. It is a mathematical
fact derivable from the structure of class systems.

\emph{Proof.} By Theorem 2.18 (Strict Dominance), systems using $\{B, S\}$
provide all capabilities of $\{S\}$-only systems plus additional
capabilities (provenance, type identity, subtype enumeration,
type-as-key). By Theorem 2.10j, adapters eliminate the retrofit
exception. Therefore, choosing $\{S\}$-only when
\(B \neq \emptyset\) is choosing the strictly dominated option.
\qed

\textbf{What ``incorrect'' means:} 1. \textbf{Information-theoretic}:
Excluding the $B$ axis discards available information. Discarding available
information without compensating benefit is suboptimal by definition. 2.
\textbf{Capability-theoretic}: Excluding $B$ forecloses
capabilities that including it provides. Foreclosing capabilities for
zero benefit is incorrect. 3. \textbf{Decision-theoretic}: Given the
choice between two options where one strictly dominates, choosing the
dominated option is irrational.

\subsection{Information-Theoretic
Foundations}\label{information-theoretic-foundations}

This section establishes the formal foundation of our
results. We prove three theorems that make claims about all possible
typing disciplines, not just our particular model.

\paragraph{3.8.1 The Impossibility
Theorem}\label{the-impossibility-theorem}

\textbf{Definition 3.10 (Typing Discipline).} A \emph{typing discipline}
\(\mathcal{D}\) over axis set \(A \subseteq \{B, S\}\) is a
collection of computable functions that take as input only the
projections of types onto axes in \(A\).

\textbf{Definition 3.11 (Axis-Exclusive Discipline).}
A \emph{$\{S\}$-only discipline} is a typing discipline over \(\{S\}\). It
has access to namespaces (shape), but not to the Bases axis.

\textbf{Remark:} In type system terminology, $\{S\}$-only disciplines include structural typing (with declared interfaces) and duck typing (without declared interfaces). Both exclude the $B$ axis.

\textbf{Remark (Syntactic Names and Type Identity).} A discipline that also inspects type identity (e.g., \texttt{type(x) is T}) is using information from $B$, not a third axis. The type identity is $\text{head}(\text{MRO}(T)) = T$, a projection of $B$ that discards all ancestors. Such a discipline has access to $\{S\}$ plus the \emph{leaf} of $B$, but not the lineage. Since provenance requires ancestor information (which ancestor provides attribute $a$?), projecting $B$ to its leaf loses exactly the information needed. This is why the impossibility results hold even for disciplines that inspect type identity. Note: the syntactic name (\texttt{\_\_name\_\_}) is distinct from type identity (see Definition 2.7).

\textbf{Definition 3.12 (Provenance Function).} The \emph{provenance
function} is:
\[\text{prov} : \text{Type} \times \text{Attr} \to \text{Type}\] where
\(\text{prov}(T, a)\) returns the type in \(T\)'s MRO that provides
attribute \(a\).

\textbf{Theorem 3.13 (Provenance Impossibility, Universal).}\label{thm:provenance-impossibility}
\leanmeta{\LH{IMP1}}
Let
\(\mathcal{D}\) be ANY $\{S\}$-only discipline (typing discipline over
\(\{S\}\), even with the leaf of \(B\)). Then \(\mathcal{D}\) cannot compute \(\text{prov}\).

\textbf{Remark:} In type system terminology, this means neither structural typing nor duck typing can provide provenance. Only systems including $B$ (nominal typing) can answer provenance queries.

\emph{Proof.} We prove this by showing that \(\text{prov}\) requires
information that is information-theoretically absent from \((S)\).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Information content of \((S)\).} A $\{S\}$-only discipline
  receives: the type name \(N(T)\) and the namespace
  \(S(T) = \{a_1, a_2, \ldots, a_k\}\) (the set of attributes \(T\)
  declares or inherits).
\item
  \textbf{Information content required by \(\text{prov}\).} The function
  \(\text{prov}(T, a)\) must return \emph{which ancestor type}
  originally declared \(a\). This requires knowing the MRO of \(T\) and
  which position in the MRO declares \(a\).
\item
  \textbf{MRO is defined exclusively by \(B\).} By Definition 2.11,
  \(\text{MRO}(T) = \text{C3}(T, B(T))\). The C3 linearization of
  \(T\)'s base classes. The function
  \(B : \text{Type} \to \text{List}[\text{Type}]\) is the Bases axis.
\item
  \textbf{\((S)\) contains no information about \(B\).} The namespace
  \(S(T)\) is the \emph{union} of attributes from all ancestors. It
  does not record \emph{which} ancestor contributed each attribute. Two
  types with identical \(S\) can have completely different \(B\) (and
  therefore different MROs and different provenance answers).
\item
  \textbf{Concrete counterexample.} Let:

  \begin{itemize}
  \tightlist
  \item
    \(A = \text{type}(\text{"A"}, (), \{\text{"x"}: 1\})\)
  \item
    \(B_1 = \text{type}(\text{"B1"}, (A,), \{\})\)
  \item
    \(B_2 = \text{type}(\text{"B2"}, (), \{\text{"x"}: 1\})\)
  \end{itemize}

  Then \(S(B_1) = S(B_2) = \{\text{"x"}\}\) (both have attribute ``x''),
  but:

  \begin{itemize}
  \tightlist
  \item
    \(\text{prov}(B_1, \text{"x"}) = A\) (inherited from parent)
  \item
    \(\text{prov}(B_2, \text{"x"}) = B_2\) (declared locally)
  \end{itemize}

  A $\{S\}$-only discipline cannot distinguish \(B_1\) from \(B_2\), therefore
  cannot compute \(\text{prov}\). \qed
\end{enumerate}

\textbf{Corollary 3.14 (No Algorithm Exists).} There exists no
algorithm, heuristic, or approximation that allows a $\{S\}$-only discipline to
compute provenance. This is not a limitation of current
implementations: it is information-theoretically impossible.

\emph{Proof.} The proof of Theorem 3.13 shows that the input \((S)\)
contains strictly less information than required to determine
\(\text{prov}\). No computation can extract information that is not
present in its input. \qed

\textbf{Significance:} This is not ``our model doesn't have
provenance.'' It is ``NO model over \((S)\) can have provenance.''
The impossibility is mathematical, not implementational.

\textbf{Theorem 3.13a (Semantic Non-Embeddability).}\label{thm:non-embeddability}
\leanmeta{\LH{EMB1}}
Nominal typing cannot be embedded into structural typing. Formally: there exists no function \(f : \mathcal{D}_{\{B,S\}} \to \mathcal{D}_{\{S\}}\) that preserves type query answers.

\emph{Proof.} By Theorem 3.13, provenance is answerable in \(\mathcal{D}_{\{B,S\}}\) but not in \(\mathcal{D}_{\{S\}}\). Any embedding must preserve query answers. Since no function in \(\mathcal{D}_{\{S\}}\) computes provenance, no such embedding exists. \qed

\textbf{Corollary 3.13b (Strict Information Gap).} The gap between nominal and structural typing is not a matter of convenience or expressiveness---it is an information-theoretic impossibility. Structural typing has strictly less information content than nominal typing. No encoding, simulation, or workaround can bridge this gap.

This result generalizes: for any axis sets \(\mathcal{A}_1 \subset \mathcal{A}_2\), the discipline using \(\mathcal{A}_1\) cannot embed the discipline using \(\mathcal{A}_2\).

\paragraph{3.8.2 The Derived Characterization
Theorem}\label{the-derived-characterization-theorem}

A potential objection is that our capability enumeration
\(\mathcal{C}_B = \{\text{provenance, identity, enumeration, conflict resolution}\}\)
is arbitrary. We now prove it is \textbf{derived from information
structure}, not chosen.

\textbf{Definition 3.15 (Query).} A \emph{query} is a computable
function \(q : \text{Type}^k \to \text{Result}\) that a typing
discipline evaluates.

\textbf{Definition 3.16 (Shape-Respecting Query).} A query \(q\) is
\emph{shape-respecting} if for all types with \(S(A) = S(B)\):
\[q(\ldots, A, \ldots) = q(\ldots, B, \ldots)\]

That is, shape-equivalent types produce identical query results.

\textbf{Definition 3.17 (B-Dependent Query).} A query \(q\) is
\emph{B-dependent} if there exist types \(A, B\) with \(S(A) = S(B)\)
but \(q(A) \neq q(B)\).

\textbf{Theorem 3.18 (Query Space Partition).} Every query is either
shape-respecting or B-dependent. These categories are mutually exclusive
and exhaustive.

\emph{Proof.} - \emph{Mutual exclusion:} If \(q\) is shape-respecting,
then \(S(A) = S(B) \Rightarrow q(A) = q(B)\). If \(q\) is B-dependent,
then \(\exists A, B: S(A) = S(B) \land q(A) \neq q(B)\). These are
logical negations. - \emph{Exhaustiveness:} For any query \(q\), either
\(\forall A, B: S(A) = S(B) \Rightarrow q(A) = q(B)\) (shape-respecting)
or \(\exists A, B: S(A) = S(B) \land q(A) \neq q(B)\) (B-dependent).
Tertium non datur. \qed

\textbf{Theorem 3.19 (Capability Gap = B-Dependent Queries).} The
capability gap between systems using only $\{S\}$ and systems using $\{B, S\}$ is EXACTLY the set of
B-dependent queries:
\[\text{Capabilities}_{\{B,S\}} \setminus \text{Capabilities}_{\{S\}} = \{q : q \text{ is B-dependent}\}\]

\textbf{Remark:} In type system terminology, this characterizes the gap between structural/duck typing ($\{S\}$-only) and nominal typing ($\{B, S\}$).

\emph{Proof.} - (\(\supseteq\)) If \(q\) is B-dependent, then
\(\exists A, B\) with \(S(A) = S(B)\) but \(q(A) \neq q(B)\). Shape
disciplines cannot distinguish \(A\) from \(B\), so cannot compute
\(q\). Nominal disciplines have access to \(B\), so can distinguish
\(A\) from \(B\) via MRO. Therefore \(q\) is in the gap. -
(\(\subseteq\)) If \(q\) is in the gap, then nominal can compute it but
shape cannot. If \(q\) were shape-respecting, shape could compute it
(contradiction). Therefore \(q\) is B-dependent. \qed

\textbf{Theorem 3.20 (Four Capabilities Are Complete).} The set
\(\mathcal{C}_B = \{\text{provenance, identity, enumeration, conflict resolution}\}\)
is the complete set of B-dependent query classes.

\emph{Proof.} We show that every B-dependent query reduces to one of
these four:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Provenance queries} (``which type provided \(a\)?''): Any
  query requiring ancestor attribution.
\item
  \textbf{Identity queries} (``is \(x\) an instance of \(T\)?''): Any
  query requiring MRO membership.
\item
  \textbf{Enumeration queries} (``what are all subtypes of \(T\)?''):
  Any query requiring inverse MRO.
\item
  \textbf{Conflict resolution queries} (``which definition wins?''): Any
  query requiring MRO ordering.
\end{enumerate}

\textbf{Completeness argument:} A B-dependent query must use information
from \(B\). The only information in \(B\) is: - Which types are
ancestors (enables identity, provenance) - The order of ancestors
(enables conflict resolution) - The inverse relation (enables
enumeration)

These three pieces of information (ancestor set, ancestor order, inverse
relation) generate exactly four query classes. No other information
exists in \(B\). \qed

\textbf{Corollary 3.21 (Capability Set Is Minimal).}
\(|\mathcal{C}_B| = 4\) and no element is redundant.

\emph{Proof.} Each capability addresses a distinct aspect of \(B\): -
Provenance: forward lookup by attribute - Identity: forward lookup by
type - Enumeration: inverse lookup - Conflict resolution: ordering

Removing any one leaves queries that the remaining three cannot answer.
\qed

\paragraph{3.8.3 The Axis Packing Dichotomy}\label{the-axis-packing-dichotomy}

A common objection to the capability gap is ``packing'': encoding $B$ into $S$ to recover $B$-dependent queries without adding an axis. We prove this is impossible.

\textbf{Definition 3.22a (Full Semantics).} The full semantics of a type is:
\[\text{Full} := S \times B\]
where $S$ is the namespace (shape) and $B$ is the bases (inheritance).

\textbf{Definition 3.22b (Shape Observation).} The shape observation function is the projection:
\[\text{obs}_S : \text{Full} \to S, \quad \text{obs}_S(s, b) = s\]

This formalizes what structural/shape-based typing can see: only the namespace.

\textbf{Definition 3.22c (Shape-Only Definability).} A query $q : \text{Full} \to Y$ is \emph{shape-only} iff it factors through $\text{obs}_S$:
\[\exists f : S \to Y.\; q = f \circ \text{obs}_S\]

That is, $q(s, b) = f(s)$ for all $(s, b) \in \text{Full}$.

\textbf{Definition 3.22d (B-Dependence).} A query $q : \text{Full} \to Y$ \emph{depends on $B$} iff:
\[\exists s, b_1, b_2.\; b_1 \neq b_2 \land q(s, b_1) \neq q(s, b_2)\]

That is, $q$ distinguishes types with identical shape but different bases.

\textbf{Theorem 3.22e (No Free Lunch for Axis Packing).} If $q$ depends on $B$, then $q$ is not shape-only.

\emph{Proof.} Suppose $q = f \circ \text{obs}_S$ for some $f$. Let $s, b_1, b_2$ witness $B$-dependence. Then:
\[q(s, b_1) = f(\text{obs}_S(s, b_1)) = f(s) = f(\text{obs}_S(s, b_2)) = q(s, b_2)\]
But $q(s, b_1) \neq q(s, b_2)$ by $B$-dependence. Contradiction. \qed

This is the core impossibility: if the observation interface is only $S$, then no query that distinguishes on $B$ can be defined.

\textbf{Definition 3.22f (Packing).} A \emph{packing} of $B$ into an extended shape $S'$ is an encoding:
\[\text{enc} : \text{Full} \to S'\]
plus an extended observation:
\[\text{obs}_{S'} : S' \to S\]
such that the shape interface is preserved:
\[\text{obs}_{S'}(\text{enc}(s, b)) = s \quad \text{for all } (s, b)\]

This formalizes ``packing $B$ into $S$ without changing what structural typing observes.''

\textbf{Theorem 3.22g (Packing Cannot Recover B-Queries).} Let $\text{enc}$, $\text{obs}_{S'}$ be a packing that preserves the $S$-interface. Let $q_B : \text{Full} \to Y$ be any $B$-dependent query. Then there is no $q' : S' \to Y$ that is shape-only w.r.t.\ $\text{obs}_{S'}$ and satisfies:
\[q'(\text{enc}(s, b)) = q_B(s, b) \quad \text{for all } (s, b)\]

\emph{Proof.} Suppose $q' = f \circ \text{obs}_{S'}$ for some $f : S \to Y$. Let $s, b_1, b_2$ witness $B$-dependence of $q_B$. Then:
\begin{align*}
q_B(s, b_1) &= q'(\text{enc}(s, b_1)) = f(\text{obs}_{S'}(\text{enc}(s, b_1))) = f(s) \\
q_B(s, b_2) &= q'(\text{enc}(s, b_2)) = f(\text{obs}_{S'}(\text{enc}(s, b_2))) = f(s)
\end{align*}
Therefore $q_B(s, b_1) = q_B(s, b_2)$, contradicting $B$-dependence. \qed

\textbf{Theorem 3.22h (The Axis Packing Dichotomy).} To make $B$-dependent queries answerable after packing, \textbf{exactly one} of the following must happen:
\begin{enumerate}
\item \textbf{Extend the observable interface.} Introduce an operation $\pi_B : S' \to B$ (or equivalent) that exposes $B$ information. This is \emph{adding an axis}, so $B$ becomes observable.
\item \textbf{Fail to answer required queries.} Remain shape-only w.r.t.\ $\text{obs}_{S'}$ and lose provenance, identity, enumeration, and conflict resolution.
\end{enumerate}

\emph{Proof.} By Theorem 3.22g, if the $S$-interface is preserved, $B$-dependent queries are unanswerable. To answer them, the interface must be extended. Adding $\pi_B$ makes $B$ observable. This is precisely what ``adding an axis'' means: making new information available to queries. \qed

\textbf{Corollary 3.22i (Packing Maneuver Is Futile).} The ``packing'' objection to the capability gap fails in all cases:
\begin{itemize}
\item If packing preserves the $S$-interface: $B$-queries remain impossible (Theorem 3.22g).
\item If packing extends the interface with $\pi_B$: this is adding an axis, not avoiding one.
\end{itemize}

There is no third option. ``Packing $B$ into $S$'' is revealed as either useless (information hidden) or equivalent to adding an axis (information exposed).

\textbf{Remark 3.22j (Categorical Interpretation).} The dichotomy has a clean categorical form: given the product $\text{Full} = S \times B$ with canonical projections $\pi_S, \pi_B$, any query depending on $\pi_B$ must factor through $\pi_B$. Packing replaces the product with an encoding $S'$ that generally lacks canonical projections. To recover $B$-queries, one must add $\pi_B : S' \to B$ as an explicit operation---i.e., restore the axis. The product structure is not an accident; it is the minimal representation preserving all queryable information.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\paragraph{3.8.4 The Complexity Lower Bound
Theorem}\label{the-complexity-lower-bound-theorem}

Our O(1) vs $\Omega(n)$ complexity claim requires proving that $\Omega(n)$ is a
\textbf{lower bound}, not merely an upper bound. We must show that NO
algorithm can do better.

\textbf{Definition 3.22 (Computational Model).} We formalize error
localization as a decision problem in the following model:

\begin{itemize}
\tightlist
\item
  \textbf{Input:} A program \(P\) with \(n\) call sites
  \(c_1, \ldots, c_n\), each potentially accessing attribute \(a\) on
  objects of type \(T\).
\item
  \textbf{Oracle:} The algorithm may query an oracle
  \(\mathcal{O}(c_i) \in \{\text{uses } a, \text{does not use } a\}\)
  for each call site.
\item
  \textbf{Output:} The set \(V \subseteq \{c_1, \ldots, c_n\}\) of call
  sites that access \(a\) on objects lacking \(a\).
\item
  \textbf{Correctness:} The algorithm must output the exact set \(V\)
  for all valid inputs.
\end{itemize}

This model captures duck typing's fundamental constraint: type
compatibility is checked at each call site, not at declaration.

\textbf{Definition 3.23 (Inspection Cost).} The \emph{cost} of an
algorithm is the number of oracle queries in the worst case over all
inputs.

\textbf{Theorem 3.24 (Duck Typing Lower Bound).} Any algorithm that
correctly solves error localization in the above model requires
\(\Omega(n)\) oracle queries in the worst case.

\emph{Proof.} By adversary argument and information-theoretic counting.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Adversary construction.} Fix any deterministic algorithm
  \(\mathcal{A}\). We construct an adversary that forces \(\mathcal{A}\)
  to query at least \(n-1\) call sites.
\item
  \textbf{Adversary strategy.} The adversary maintains a set \(S\) of
  ``candidate violators'', call sites that could be the unique
  violating site. Initially \(S = \{c_1, \ldots, c_n\}\). When
  \(\mathcal{A}\) queries \(\mathcal{O}(c_i)\):

  \begin{itemize}
  \tightlist
  \item
    If \(|S| > 1\): Answer ``does not use \(a\)'' and set
    \(S \leftarrow S \setminus \{c_i\}\)
  \item
    If \(|S| = 1\): Answer consistently with \(c_i \in S\) or
    \(c_i \notin S\)
  \end{itemize}
\item
  \textbf{Lower bound derivation.} The algorithm must distinguish
  between \(n\) possible inputs (exactly one of \(c_1, \ldots, c_n\)
  violates). Each query eliminates at most one candidate. After
  \(k < n-1\) queries, \(|S| \geq 2\), so the algorithm cannot determine
  the unique violator. Therefore \(\mathcal{A}\) requires at least
  \(n-1 \in \Omega(n)\) queries.
\item
  \textbf{Generalization.} For the case where multiple call sites may
  violate: there are \(2^n\) possible subsets. Each binary query
  provides at most 1 bit. Therefore \(\log_2(2^n) = n\) queries are
  necessary to identify the exact subset. \qed
\end{enumerate}

\textbf{Remark (Static Analysis).} Static analyzers precompute call site
information via control-flow analysis over the program text. This shifts
the \(\Omega(n)\) cost to analysis time rather than eliminating it. The
bound characterizes the inherent information content required, \(n\)
bits to identify \(n\) potential violation sites, regardless of when
that information is gathered.

\textbf{Theorem 3.25 (Nominal Typing Upper Bound).} Nominal error
localization requires exactly 1 inspection.

\emph{Proof.} In nominal typing, constraints are declared at the class
definition. The constraint ``type \(T\) must have attribute \(a\)'' is
checked at the single location where \(T\) is defined. If the constraint
is violated, the error is at that location. No call site inspection is
required. \qed

\textbf{Corollary 3.26 (Complexity Gap Is Unbounded).} The ratio
\(\frac{\text{DuckCost}(n)}{\text{NominalCost}}\) grows without bound:
\[\lim_{n \to \infty} \frac{\Omega(n)}{O(1)} = \infty\]

\emph{Proof.} Immediate from Theorems 3.24 and 3.25. \qed

\textbf{Corollary 3.27 (Lower Bound Is Tight).} The $\Omega(n)$ lower bound for
duck typing is achieved by naive inspection; no algorithm can do
better, and simple algorithms achieve this bound.

\emph{Proof.} Theorem 3.24 proves \(\Omega(n)\) is necessary. Linear
scan of call sites achieves \(O(n)\). Therefore the bound is tight.
\qed

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Summary: Core Formal
Results}\label{summary-core-formal-results}

We have established four theorems with universal scope:

\small
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.2195}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.2683}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.5122}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Theorem
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Statement
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Proof Technique
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{3.13 (Impossibility)} & No shape discipline can compute
provenance & Information-theoretic: input lacks required data \\
\textbf{3.19 (Derived Characterization)} & Capability gap = B-dependent
queries & Mathematical: query space partitions exactly \\
\textbf{3.22h (Packing Dichotomy)} & Packing $B$ into $S$ either fails or adds an axis &
Categorical: observation preservation blocks recovery \\
\textbf{3.24 (Lower Bound)} & Duck typing requires $\Omega$(n) inspections &
Adversary argument: any algorithm can be forced \\
\end{longtable}

These are not claims about our model; they are claims about \textbf{the
universe of possible typing systems}. The theorems establish:
\begin{itemize}
\tightlist
\item Theorem 3.13 proves no model over $(S)$ can provide provenance.
\item Theorem 3.19 proves the capability enumeration is derived from information structure.
\item Theorem 3.22h proves ``packing'' cannot circumvent the capability gap; it either fails or is equivalent to adding an axis.
\item Theorem 3.24 proves no algorithm can overcome the information-theoretic limitation.
\end{itemize}

These results follow from the structure of the problem, not from our particular formalization.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Information-Theoretic
Completeness}\label{information-theoretic-completeness}

For completeness, we restate the original characterization in the
context of the new foundations.

\textbf{Definition 3.28 (Query).} A \emph{query} is a predicate
\(q : \text{Type} \to \text{Bool}\) that a typing discipline can
evaluate.

\textbf{Definition 3.29 (Axis-Respecting Query).} A query \(q\) is
\emph{$\{S\}$-respecting} if for all types \(A, B\) with \(S(A) = S(B)\):
\[q(A) = q(B)\]

That is, shape-equivalent types cannot be distinguished by \(q\).

\textbf{Theorem 3.30 (Capability Gap Characterization).} Let
$\{S\}$-Queries be the set of all $\{S\}$-respecting queries, and let
AllQueries be the set of all queries. If there exist types \(A \neq B\)
with \(S(A) = S(B)\), then:
\[\text{$\{S\}$-Queries} \subsetneq \text{AllQueries}\]

\textbf{Remark:} In type system terminology, "$\{S\}$-Queries" corresponds to structural/shape/duck typing capabilities.

\emph{Proof.} The identity query \(\text{isA}(T) := (T = A)\) is in
AllQueries but not $\{S\}$-Queries, because isA(A) = true but isA(B) =
false despite \(S(A) = S(B)\). \qed

\textbf{Corollary 3.31 (Derived Capability Set).} The capability gap
between $\{S\}$-only and $\{B,S\}$ systems is \textbf{exactly} the set of
queries that depend on the Bases axis:
\[\text{Capability Gap} = \{ q \mid \exists A, B.\ S(A) = S(B) \land q(A) \neq q(B) \}\]

\textbf{Remark:} This characterizes the gap between structural/duck typing ($\{S\}$) and nominal typing ($\{B,S\}$).

This is not an enumeration; it's a \textbf{characterization}. Our
listed capabilities (provenance, identity, enumeration, conflict
resolution) are instances of this set, not arbitrary choices.

\textbf{Information-Theoretic Interpretation:} Information theory tells
us that discarding information removes the ability to answer queries that depend on that
information. The Bases axis contains information about inheritance
relationships. Shape-based typing discards this axis. Therefore, any
query that depends on inheritance (provenance, identity, enumeration,
conflict resolution) cannot be answered. This follows from the structure
of the information available.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Completeness and Robustness Theorems}\label{completeness-robustness-theorems}

This section presents additional theorems that establish the completeness
and robustness of our results. Each theorem addresses a specific aspect
of the model's coverage.

\paragraph{3.11.1 Model Completeness}\label{model-completeness}

\textbf{Theorem 3.32 (Model Completeness).}\label{thm:model-completeness}
\leanmeta{\LH{MDL1}}
The \((B, S)\) model
captures all information constitutive of a type. Any computable function
over types is expressible as a function of \((B, S)\).

\textbf{Remark (Two Senses of Completeness).} ``Model completeness'' (this theorem) means \((B, S)\) captures all constitutive information of a type object. ``Axis set completeness for domain \(D\)'' (Definition 2.22) means the axis set can answer all queries in \(D\). These are distinct: \((B, S)\) is model-complete (captures all type information) but may be domain-incomplete (cannot answer queries requiring axes beyond \(B\) and \(S\), such as scope-chain queries requiring \(H\)).

\emph{Proof.} The proof proceeds by constitutive definition, not empirical
enumeration.

In Python, \texttt{type(name, bases, namespace)} is the universal type
constructor. Every type \(T\) is created by some invocation
\texttt{type(n, b, s)}, either explicitly or via the \texttt{class}
statement (which is syntactic sugar for \texttt{type()}). A type does not
merely \emph{have} \((B, S)\); a type \emph{is} \((B, S)\). There is
no other information constitutive of a type object.

Therefore, for any computable function \(g : \text{Type} \to \alpha\):
\[g(T) = g(\texttt{type}(n, b, s)) = h(n, b, s)\]
for some computable \(h\). Any function of a type is definitionally a
function of the triple that constitutes it.

\textbf{Remark (Derived vs.\ Constitutive).} Properties like
\texttt{\_\_mro\_\_} (method resolution order) or \texttt{\_\_module\_\_}
are not counterexamples: MRO is computed from \(B\) by C3 linearization;
\texttt{\_\_module\_\_} is stored in the namespace \(S\). These are
\emph{derived from} or \emph{contained in} \((B, S)\), not independent
of it.

This is a definitional closure: a critic cannot exhibit a third axis
because any proposed axis is either (a) stored in \(S\) (including \texttt{\_\_name\_\_}),
(b) computable from \((B, S)\), or (c) not part of the type's semantic identity
(e.g., memory address). \qed

\textbf{Corollary 3.33 (No Hidden Information).} There exists no
third axis that shape-based typing could use to recover provenance.
The information is structurally absent, not because we failed to model
it, but because types \emph{are} \((B, S)\) by construction.

\paragraph{3.11.2 Capability Comparison}\label{capability-comparison}

\textbf{Theorem 3.34 (Capability Superset).} Let
\(\mathcal{C}_{\text{duck}}\) be the capabilities available under duck
typing. Let \(\mathcal{C}_{\text{nom}}\) be the capabilities under
nominal typing. Then:
\[\mathcal{C}_{\text{duck}} \subseteq \mathcal{C}_{\text{nom}}\]

\emph{Proof.} Duck typing operations are: 1. Attribute access:
\texttt{getattr(obj,\ "name")} 2. Attribute existence:
\texttt{hasattr(obj,\ "name")} 3. Method invocation:
\texttt{obj.method()}

All three operations are available in nominal systems. Nominal typing
adds type identity operations; it does not remove duck typing
operations. \qed

\textbf{Theorem 3.35 (Strict Superset).} The inclusion is strict:
\[\mathcal{C}_{\text{duck}} \subsetneq \mathcal{C}_{\text{nom}}\]

\emph{Proof.} Nominal typing provides provenance, identity, enumeration,
and conflict resolution (Theorem 2.17). Duck typing cannot provide these
(Theorem 3.13). Therefore:
\[\mathcal{C}_{\text{nom}} = \mathcal{C}_{\text{duck}} \cup \mathcal{C}_B\]
where \(\mathcal{C}_B \neq \emptyset\). \qed

\textbf{Corollary 3.36 (No Capability Tradeoff).} Choosing nominal
typing over duck typing: - Forecloses \textbf{zero} capabilities - Gains
\textbf{four} capabilities

There is no capability tradeoff. Nominal typing strictly dominates.

\textbf{Remark (Capability vs.~Code Compatibility).} The capability
superset does not mean ``all duck-typed code runs unchanged under
nominal typing.'' It means ``every operation expressible in duck typing
is expressible in nominal typing.'' The critical distinction:

\begin{itemize}
\item
  \textbf{False equivalence} (duck typing): \texttt{WellFilterConfig}
  and \texttt{StepWellFilterConfig} are structurally identical but
  semantically distinct (different MRO positions, different scopes).
  Duck typing conflates them; it literally cannot answer ``which type
  is this?'' This is not flexibility; it is \textbf{information
  destruction}.
\item
  \textbf{Type distinction} ($\{B,S\}$ systems):
  \texttt{isinstance(config,\ StepWellFilterConfig)} distinguishes them
  in O(1). The distinction is expressible because $\{B,S\}$ systems
  preserve type identity.
\end{itemize}

The ``acceptance'' of structurally-equivalent types in $\{S\}$-only systems is not a
capability; it is the \emph{absence} of the capability to distinguish
them. $\{B,S\}$ systems add this capability without removing any structural
typing operation. See Case Study 1 (\S{}5.2, Theorem 5.1) for the complete
production example demonstrating that structural identity $\neq$ semantic
identity.

\textbf{Remark:} The traditional terms "duck typing" and "nominal typing" correspond to $\{S\}$-only and $\{B,S\}$ systems respectively.

\paragraph{3.11.3 Axiom Justification}\label{axiom-justification}

\textbf{Lemma 3.37 (Axis-Exclusion is Definitional).}\label{lem:shape-axiom-definitional}
\leanmeta{\LH{AXI1}}
The axiom
``$\{S\}$-only typing treats same-namespace types identically'' is not an
assumption; it is the \textbf{definition} of $\{S\}$-only typing.

\emph{Proof.} $\{S\}$-only typing is defined as a typing discipline over
\(\{S\}\) only (Definition 2.10). If a discipline uses information from
\(B\) (the Bases axis) to distinguish types, it is, by definition, not
$\{S\}$-only.

The axiom is not: ``We assume $\{S\}$-only typing can't distinguish same-shape
types.'' The axiom is: ``$\{S\}$-only typing means treating same-shape types
identically.''

Any system that distinguishes same-shape types is using \(B\)
(explicitly or implicitly). \qed

\textbf{Remark:} In type system terminology, "$\{S\}$-only typing" corresponds to shape-based/duck/structural typing.

\textbf{Corollary 3.38 (No Axis-Extension Without Including Axis).} There exists no
``clever'' $\{S\}$-only system that can distinguish types \(A\) and \(B\)
with \(S(A) = S(B)\). Such a system would, by definition, not be
$\{S\}$-only.

\paragraph{3.11.4 Extension
Impossibility}\label{extension-impossibility}

\textbf{Theorem 3.39 (Extension Impossibility).}\label{thm:extension-impossibility}
\leanmeta{\LH{EXT1}}
Let \(\mathcal{D}\) be
any $\{S\}$-only typing system. Let \(\mathcal{D}'\) be \(\mathcal{D}\) extended
with any computable function \(f : \text{Namespace} \to \alpha\). Then
\(\mathcal{D}'\) still cannot compute provenance.

\emph{Proof.} Provenance requires distinguishing types \(A\) and \(B\)
where \(S(A) = S(B)\) but \(\text{prov}(A, a) \neq \text{prov}(B, a)\)
for some attribute \(a\).

Any function \(f : \text{Namespace} \to \alpha\) maps \(A\) and \(B\) to
the same value, since \(S(A) = S(B)\) implies \(f\) receives identical
input for both.

Therefore, \(f\) provides no distinguishing information. The only way to
distinguish \(A\) from \(B\) is to use information not in
\(\text{Namespace}\), i.e., the Bases axis \(B\).

No computable extension over \(\{S\}\) (even with the leaf of \(B\)) can recover provenance.
\qed

\textbf{Corollary 3.40 (No Future Fix).} No future language feature,
library, or tool operating within the duck typing paradigm can provide
provenance. The limitation is structural, not technical.

\paragraph{3.11.5 Capability
Exhaustiveness}\label{capability-exhaustiveness}

\textbf{Theorem 3.43a (Capability Exhaustiveness).} The four
capabilities (provenance, identity, enumeration, conflict resolution)
are \textbf{exhaustive}; they are the only capabilities derivable from
the Bases axis.

\emph{Proof.} (Machine-checked in \texttt{nominal\_resolution.lean},
Section 6: CapabilityExhaustiveness)

The Bases axis provides MRO, a \textbf{list of types}. A list has
exactly three queryable properties: 1. \textbf{Ordering}: Which element
precedes which? $\rightarrow$ \emph{Conflict resolution} (C3 linearization selects
based on MRO order) 2. \textbf{Membership}: Is element X in the list? $\rightarrow$
\emph{Enumeration} (subtype iff in some type's MRO) 3. \textbf{Element
identity}: Which specific element? $\rightarrow$ \emph{Provenance} and \emph{type
identity} (distinguish structurally-equivalent types by MRO position)

These are exhaustive by the structure of lists; there are no other
operations on a list that do not reduce to ordering, membership, or
element identity. Therefore, the four capabilities are derived from MRO
structure, not enumerated by inspection. \qed

\textbf{Corollary 3.43b (No Missing Capability).} Any capability claimed
to require \(B\) reduces to one of the four. There is no ``fifth
capability'' that \(B\) provides.

\emph{Proof.} Any operation on \(B\) is an operation on MRO. Any
operation on MRO is an operation on a list. List operations are
exhaustively \{ordering, membership, identity\}. \qed

\textbf{Theorem 3.43b-bis (Capability Reducibility).} Every B-dependent
query reduces to a composition of the four primitive capabilities.

\emph{Proof.} Let \(q : \text{Type} \to \alpha\) be any B-dependent
query (per Definition 3.17). By Definition 3.17, \(q\) distinguishes
types with identical structure:
\(\exists A, B: S(A) = S(B) \land q(A) \neq q(B)\).

The only information distinguishing \(A\) from \(B\) is \(B(A) \neq B(B)\) (different bases), which distinguishes via:
\begin{itemize}
  \item Ancestor membership: is \(T \in \text{ancestors}(A)\)? $\rightarrow$ covered by \textbf{provenance}
  \item Subtype enumeration: what are all \(T : T <: A\)? $\rightarrow$ covered by \textbf{enumeration}
  \item MRO position: which type wins for attribute \(a\)? $\rightarrow$ covered by \textbf{conflict\_resolution}
  \item Type identity: is \(\text{head}(\text{MRO}(x)) = T\)? $\rightarrow$ covered by \textbf{type\_identity}
\end{itemize}
Note: Syntactic names (\texttt{\_\_name\_\_}) are metadata external to this model (see Definition 2.7).

No other distinguishing information exists (Theorem 3.32: \((B, S)\)
is complete).

Therefore any B-dependent query \(q\) can be computed by composing:
\[q(T) = f(\text{provenance}(T), \text{identity}(T), \text{enumeration}(T), \text{conflict\_resolution}(T))\]
for some computable \(f\). \qed

\paragraph{3.11.6a Adapter Cost Analysis}\label{a-adapter-cost-analysis}

\textbf{Theorem 3.43c (Adapter Declaration is Information-Preserving).}
An adapter declares information that is \textbf{already true}, that a
type conforms to an interface. Declaration does not create the
conformance; it makes it explicit.

\emph{Proof.} If \texttt{TheirType} does not satisfy \texttt{YourABC}'s
interface, the adapter fails at definition time (missing method error).
If \texttt{TheirType} does satisfy the interface, the conformance
existed before the adapter. The adapter is not implementation; it is
documentation of pre-existing fact. \qed

\textbf{Theorem 3.43d (Adapter Amortization).} Adapter cost is O(1).
Manual capability implementation is O(N) where N is the number of use
sites.

\emph{Proof.} (Machine-checked in \texttt{nominal\_resolution.lean},
Section 7: AdapterAmortization)

Under nominal typing (with adapter): - Provenance: Automatic via
\texttt{type(obj).\_\_mro\_\_} (0 additional code per use) - Identity:
Automatic via \texttt{isinstance()} (0 additional code per use) -
Enumeration: Automatic via \texttt{\_\_subclasses\_\_()} (0 additional
code per use) - Conflict resolution: Automatic via C3 (0 additional code
per use)

Under structural typing (without adapter), to recover any capability
manually: - Provenance: Must thread source information through call
sites (1 additional parameter $\times$ N calls) - Identity: Must maintain
external type registry (1 registry + N registration calls) -
Enumeration: Must maintain external subtype set (1 set + N insertions) -
Conflict resolution: Must implement manual dispatch (1 dispatcher + N
cases)

The adapter is 2 lines. Manual implementation is \(\Omega(N)\). For
\(N \geq 1\), adapter dominates. \qed

\textbf{Corollary 3.43e (Negative Adapter Cost).} Adapter ``cost'' is
negative, a net benefit.

\emph{Proof.} The adapter enables automatic capabilities that would
otherwise require O(N) manual implementation. The adapter costs O(1).
For any system requiring the capabilities, adapter provides net savings
of \(\Omega(N) - O(1) = \Omega(N)\). The ``cost'' is negative.
\qed

\textbf{Corollary 3.43f (Adapter Cost Objection is Invalid).} Objecting
to adapter cost is objecting to O(1) overhead while accepting O(N)
overhead. This is mathematically incoherent.

\paragraph{3.11.6b Methodological
Independence}\label{b-methodological-independence}

\textbf{Theorem 3.43g (Methodological Independence).}\label{thm:methodological-independence}
\leanmeta{\LH{IMP1}, \LH{DOM1}, \LH{ALT1}}
The dominance
theorems are derived from the structure of \((B, S)\), not from any
implementation. OpenHCS is an existential witness, not a premise.

\emph{Proof.} We distinguish two logical roles:
\begin{itemize}
\tightlist
\item \textbf{Premise:} A proposition the conclusion depends on. If false,
  the conclusion may not follow.
\item \textbf{Existential witness:} A concrete example demonstrating
  satisfiability. Removing it does not affect the theorem's validity.
\end{itemize}

Examine the proof of Theorem 3.13 (Provenance Impossibility): it shows
that \((S)\) contains insufficient information to compute provenance.
This is an information-theoretic argument referencing no codebase. The
proof could be written before any codebase existed.

\textbf{Proof chain (no OpenHCS references):}
\begin{enumerate}
\tightlist
\item Theorem 2.17 (Capability Gap): Proved from the definition of
  $\{S\}$-only typing
\item Theorem 3.5 (Strict Dominance): Proved from Theorem 2.17 + Theorem 2.18
\item Theorem 2.10j (Adapters): Proved from capability comparison
\end{enumerate}

OpenHCS appears only to demonstrate that the four capabilities are
\emph{achievable}, that real systems use provenance, identity,
enumeration, and conflict resolution. This is an existence proof
(``such systems exist''), not a premise (``if OpenHCS works, then
the theorems hold'').

\textbf{Analogy:} Proving ``comparison-based sorting requires
\(\Omega(n \log n)\)'' does not require testing on multiple arrays.
Exhibiting quicksort demonstrates achievability, not theorem validity.
\qed

\textbf{Corollary 3.43h (Cross-Codebase Validity).} The theorems apply
to any codebase in any language where \(B \neq \emptyset\). OpenHCS is a
sufficient example, not a necessary one.

\paragraph{3.11.6c Inheritance Ubiquity}\label{c-inheritance-ubiquity}

\textbf{Theorem 3.43i (Inheritance Ubiquity).} In Python,
\(B = \emptyset\) requires actively avoiding all standard tooling. Any
project using \(\geq 1\) of the following has \(B \neq \emptyset\) by
construction:

\small
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.2273}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.2273}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.5455}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Examples
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Why \(B \neq \emptyset\)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Exceptions & \texttt{raise\ MyError()} & Must subclass
\texttt{Exception} \\
Web frameworks & Django, Flask, FastAPI & Views/models inherit framework
bases \\
Testing & pytest classes, unittest & Test classes inherit
\texttt{TestCase} or use class fixtures \\
ORM & SQLAlchemy, Django ORM & Models inherit declarative
\texttt{Base} \\
Data validation & Pydantic, attrs & Models inherit \texttt{BaseModel} \\
Enumerations & \texttt{class\ Color(Enum)} & Must subclass
\texttt{Enum} \\
Abstract interfaces & ABC, Protocol with inheritance & Defines
inheritance hierarchy \\
Dataclasses & \texttt{@dataclass} with inheritance & Parent class in
\texttt{\_\_bases\_\_} \\
Context managers & Class-based \texttt{\_\_enter\_\_/\_\_exit\_\_} &
Often inherit helper bases \\
Type extensions & \texttt{typing.NamedTuple}, \texttt{TypedDict} &
Inherit from typing constructs \\
\end{longtable}

\emph{Proof.} Each listed feature requires defining or inheriting from a
class with non-trivial bases. In Python, even an ``empty'' class
\texttt{class\ X:\ pass} has \texttt{X.\_\_bases\_\_\ ==\ (object,)}, so
\(B \supseteq \{\texttt{object}\}\). For \(B = \emptyset\) to hold, a
project must use:

\begin{itemize}
\tightlist
\item
  No user-defined exceptions (use only built-in exceptions)
\item
  No web frameworks (no Django, Flask, FastAPI, Starlette, etc.)
\item
  No ORM (no SQLAlchemy, Django ORM, Peewee, etc.)
\item
  No Pydantic, attrs, or dataclass inheritance
\item
  No Enum
\item
  No ABC or Protocol inheritance
\item
  No pytest/unittest class-based tests
\item
  No class-based context managers
\item
  Pure functional style with only module-level functions and built-in
  types
\end{itemize}

This describes a pathologically constrained subset of Python, not
``most code'' but ``no OOP at all.'' \qed

\textbf{Corollary 3.43j (B=$\emptyset$ Is Exceptional).} The \(B = \emptyset\)
case applies only to: 1. Languages without inheritance by design (Go) 2.
Pure data serialization boundaries (JSON parsing before domain modeling)
3. FFI boundaries (ctypes, CFFI) before wrapping in domain types 4.
Purely functional codebases with no class definitions

In all other cases---which constitute the overwhelming majority of
production Python, Java, C\#, TypeScript, Kotlin, Swift, Scala, and C++
code---\(B \neq \emptyset\) and nominal typing strictly dominates.

\textbf{Corollary 3.43k (Inheritance Prevalence).} A claim that
``\(B = \emptyset\) is the common case'' would require exhibiting a non-trivial
production codebase using none of the tooling in Theorem 3.43i. No such
codebase is known to exist in the Python ecosystem.

\paragraph{3.11.7 Generics and Parametric
Polymorphism}\label{generics-and-parametric-polymorphism}

\textbf{Theorem 3.43 (Generics Preserve Axis Structure).} Parametric
polymorphism does not introduce a third axis. Type parameters refine
$(B, S)$, not add orthogonal information.

\emph{Proof.} A parameterized type \(G\langle T \rangle\) (e.g.,
\texttt{List\textless{}Dog\textgreater{}}) has:
\begin{itemize}
  \item \(B(G\langle T \rangle) = B(G)[T/\tau]\). Bases with parameter substituted
  \item \(S(G\langle T \rangle) = S(G)[T/\tau]\). Namespace with parameter in signatures
\end{itemize}
The type parameter affects both $B$ and $S$ after substitution. No additional axis is required; generics are fully captured by $(B, S)$.
\qed

\textbf{Theorem 3.44 (Generic Shape Indistinguishability).} Under
shape-based typing, \texttt{List\textless{}Dog\textgreater{}} and
\texttt{Set\textless{}Cat\textgreater{}} are indistinguishable if
\(S(\text{List}\langle\text{Dog}\rangle) = S(\text{Set}\langle\text{Cat}\rangle)\).

\emph{Proof.} Shape typing uses only \(S\). If two parameterized types
have the same method signatures (after parameter substitution), shape
typing treats them identically. It cannot distinguish: - The base
generic type (\texttt{List} vs \texttt{Set}) - The type parameter
(\texttt{Dog} vs \texttt{Cat}) - The generic inheritance hierarchy

These require \(N\) (for parameter identity) and \(B\) (for hierarchy).
\qed

\textbf{Theorem 3.45 (Generic Capability Gap Extends).} The four
capabilities from \(\mathcal{C}_B\) (provenance, identity, enumeration,
conflict resolution) apply to generic types. Generics do not reduce the
capability gap; they \textbf{increase the type space} where it applies.

\emph{Proof.} For generic types, the four capabilities manifest as: 1.
\textbf{Provenance:} ``Which generic type provided this method?'' ---
requires \(B\) 2. \textbf{Identity:} ``Is this
\texttt{List\textless{}Dog\textgreater{}} or
\texttt{Set\textless{}Cat\textgreater{}}?'' --- requires parameterized
\(N\) 3. \textbf{Enumeration:} ``What are the subtypes of
\texttt{Collection\textless{}T\textgreater{}}?'' --- requires \(B\) 4.
\textbf{Conflict resolution:} ``Which
\texttt{Comparable\textless{}T\textgreater{}} implementation wins?'' ---
requires \(B\)

Additionally, generics introduce \textbf{variance} (covariant,
contravariant, invariant), which requires \(B\) to track inheritance
direction. Shape typing discards \(B\) and the parameter component of
\(N\), losing all four capabilities plus variance. \qed

\textbf{Corollary 3.45.1 (Same Four, Larger Space).} Generics do not
create new capabilities---they apply the same four capabilities to a
larger type space. The capability gap is preserved, not reduced.

\textbf{Theorem 3.46 (Erasure Does Not Save Shape Typing).} In languages
with type erasure (Java), the capability gap still exists
\cite{jlsErasure}.

\emph{Proof.} Type checking occurs at compile time, where full
parameterized types are available. Erasure only affects runtime
representations. Our theorems about typing disciplines apply to the type
system (compile time), not runtime behavior.

At compile time: - The type checker has access to
\texttt{List\textless{}Dog\textgreater{}} vs
\texttt{List\textless{}Cat\textgreater{}} - Shape typing cannot
distinguish them if method signatures match - Nominal typing can
distinguish them

At runtime (erased): - Both become \texttt{List} (erased) - Shape typing
cannot distinguish \texttt{ArrayList} from \texttt{LinkedList} - Nominal
typing can (via \texttt{instanceof})

The capability gap exists at both levels. \qed

\textbf{Theorem 3.47 (Universal Extension).} All capability gap theorems
(3.13, 3.19, 3.24) extend to generic type systems. The formal results
apply to:

\begin{itemize}
\tightlist
\item
  \textbf{Erased generics:} Java, Scala, Kotlin
\item
  \textbf{Reified generics:} C\#, Kotlin (inline reified)
\item
  \textbf{Monomorphized generics:} Rust, C++ (templates)
\item
  \textbf{Compile-time only:} TypeScript, Swift
\end{itemize}

\emph{Proof.} Each language encodes generics as parameterized \(N\) (see
Table 2.2). The \((B, S)\) model applies uniformly. Type checking
occurs at compile time where full parameterized types are available.
Runtime representation (erased, reified, or monomorphized) is irrelevant
to typing discipline. \qed

\textbf{Corollary 3.48 (No Generic Escape).} Generics do not provide an
escape from the capability gap. No major language invented a fourth
axis.

\textbf{Remark 3.49 (Exotic Type Features).} Intersection types, union
types, row polymorphism, higher-kinded types, and multiple dispatch do
not escape the \((B, S)\) model:

\begin{itemize}
\tightlist
\item
  \textbf{Intersection/union types} (TypeScript \texttt{A\ \&\ B},
  \texttt{A\ \textbar{}\ B}): Refine \(N\), combine \(B\) and \(S\).
  Still two axes (N derivable from B).
\item
  \textbf{Row polymorphism} (OCaml
  \texttt{\textless{}\ x:\ int;\ ..\ \textgreater{}}): Pure structural
  typing using \(S\) only, but with a \emph{declared} interface (unlike
  duck typing). OCaml row types are coherent (Theorem 2.10d does not
  apply) because the object types and row variables are declared
  explicitly \cite{ocamlObjectsRowVars}; they still lose the four
  \(B\)-dependent capabilities
  (provenance, identity, enumeration, conflict resolution) and cannot
  provide metaprogramming hooks (Theorem 2.10p).
\item
  \textbf{Higher-kinded types} (Haskell \texttt{Functor},
  \texttt{Monad}): Parameterized \(N\) at the type-constructor level.
  Typeclass hierarchies provide \(B\).
\item
  \textbf{Multiple dispatch} (Julia): Type hierarchies exist
  (\texttt{AbstractArray\ \textless{}:\ Any}). \(B\) axis present.
  Dispatch semantics are orthogonal to type structure.
\item
  \textbf{Prototype-based inheritance} (JavaScript): Prototype chain IS
  the \(B\) axis at object level. \texttt{Object.getPrototypeOf()}
  traverses MRO.
\end{itemize}

No mainstream type system feature introduces a third axis orthogonal to
\((B, S)\).

\paragraph{3.11.7 Scope Expansion: From Greenfield to
Universal}\label{scope-expansion-from-greenfield-to-universal}

\textbf{Theorem 3.50 (Universal Optimality).} Wherever inheritance
hierarchies exist and are accessible, nominal typing provides strictly
more capabilities than shape-based typing. This is not limited to
greenfield development.

\emph{Proof.} The capability gap (Theorem 3.19) is
information-theoretic: shape typing discards \(B\), losing four
capabilities. This holds regardless of: - Whether code is new or legacy
- Whether the language is compiled or interpreted - Whether types are
manifest or inferred - Whether the system uses classes, traits,
protocols, or typeclasses

The gap exists wherever \(B\) exists. \qed

\textbf{Corollary 3.51 (Scope of Shape Typing).} Shape-based typing is
capability-equivalent to nominal typing only when:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{No hierarchy exists:} \(B = \emptyset\) (e.g., Go interfaces,
  JSON objects)
\item
  \textbf{Hierarchy is inaccessible:} True FFI boundaries where type
  metadata is lost
\end{enumerate}

When \(B \neq \emptyset\), shape-based typing is \textbf{always
dominated} by nominal typing with adapters (Theorem 2.10j).
``Deliberately ignored'' is not a valid justification---it is an
admission of choosing the dominated option.

\textbf{Claim 3.52 (Universal).} For ALL object-oriented systems where
inheritance hierarchies exist and are accessible---including legacy
codebases, dynamic languages, and functional languages with
typeclasses---nominal typing is strictly optimal. Shape-based typing is
a \textbf{capability sacrifice}, not an alternative with tradeoffs.

\paragraph{3.11.8 Discipline Optimality vs Migration
Optimality}\label{discipline-optimality-vs-migration-optimality}

A critical distinction:
\textbf{discipline optimality} (which typing paradigm has more
capabilities) is independent of \textbf{migration optimality} (whether
migrating an existing codebase is beneficial).

\textbf{Definition 3.53 (Pareto Dominance).} Discipline \(A\) Pareto
dominates discipline \(B\) if: 1. \(A\) provides all capabilities of
\(B\) 2. \(A\) provides at least one capability \(B\) lacks 3. The
declaration cost of \(A\) is at most the declaration cost of \(B\)

\textbf{Theorem 3.54 (Nominal Pareto Dominates Shape).} Nominal typing
Pareto dominates shape-based typing.

\emph{Proof.} (Machine-checked in \texttt{discipline\_migration.lean})
1. Shape capabilities = \{attributeCheck\} 2. Nominal capabilities =
\{provenance, identity, enumeration, conflictResolution,
attributeCheck\} 3. Shape $\subset$ Nominal (strict subset) 4. Declaration cost:
both require one class definition per interface 5. Therefore nominal
Pareto dominates shape. \qed

\textbf{Theorem 3.55 (Dominance Does Not Imply Migration).}\label{thm:dominance-not-migration}
\leanmeta{\LH{CMP1}}
Pareto
dominance of discipline \(A\) over \(B\) does NOT imply that migrating
from \(B\) to \(A\) is beneficial for all codebases.

\emph{Proof.} (Machine-checked in \texttt{discipline\_migration.lean})

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Dominance is codebase-independent.} \(D(A, B)\) (``\(A\)
  dominates \(B\)'') is a relation on typing disciplines. It depends
  only on capability sets:
  \(\text{Capabilities}(A) \supset \text{Capabilities}(B)\). This is a
  property of the disciplines themselves, not of any codebase.
\item
  \textbf{Migration cost is codebase-dependent.} Let \(C(ctx)\) be the
  cost of migrating codebase \(ctx\) from \(B\) to \(A\). Migration
  requires modifying: type annotations using \(B\)-specific constructs,
  call sites relying on \(B\)-specific semantics, and external API
  boundaries (which may be immutable). Each of these quantities is
  unbounded: there exist codebases with arbitrarily many annotations,
  call sites, and external dependencies.
\item
  \textbf{Benefit is bounded.} The benefit of migration is the
  capability gap:
  \(|\text{Capabilities}(A) \setminus \text{Capabilities}(B)|\). For
  nominal vs structural, this is 4 (provenance, identity, enumeration,
  conflict resolution). This is a constant, independent of codebase
  size.
\item
  \textbf{Unbounded cost vs bounded benefit.} For any fixed benefit
  \(B\), there exists a codebase \(ctx\) such that \(C(ctx) > B\). This
  follows from (2) and (3): cost grows without bound, benefit does not.
\item
  \textbf{Existence of both cases.} For small \(ctx\): \(C(ctx) < B\)
  (migration beneficial). For large \(ctx\): \(C(ctx) > B\) (migration
  not beneficial).
\end{enumerate}

Therefore dominance does not determine migration benefit.
\qed

\textbf{Corollary 3.55a (Category Error).} Conflating ``discipline \(A\)
is better'' with ``migrate to \(A\)'' is a category error: the former is
a property of disciplines (universal), the latter is a property of
(discipline, codebase) pairs (context-dependent).

\textbf{Corollary 3.56 (Discipline vs Migration Independence).} The
question ``which discipline is better?'' (answered by Theorem 3.54) is
independent of ``should I migrate?'' (answered by cost-benefit
analysis).

This distinguishes ``nominal provides more capabilities'' from ``rewrite everything
in nominal.'' The theorems are:
\begin{itemize}
\tightlist
\item \textbf{Discipline comparison}: Universal, always true (Theorem 3.54)
\item \textbf{Migration decision}: Context-dependent, requires cost-benefit
analysis (Theorem 3.55)
\end{itemize}

\paragraph{3.11.9 Context Formalization: Greenfield and Retrofit
(Historical)}\label{context-formalization-greenfield-and-retrofit-historical}

\textbf{Note.} The following definitions were used in earlier versions
of this paper to distinguish contexts where nominal typing was
``available'' from those where it was not. Theorem 2.10j (Adapters)
eliminates this distinction: adapters make nominal typing available in
all retrofit contexts. We retain these definitions for completeness and
because the Lean formalization verifies them.

\textbf{Definition 3.57 (Greenfield Context).} A development context is
\emph{greenfield} if: 1. All modules are internal (architect can modify
type hierarchies) 2. No constraints require structural typing (e.g.,
JSON API compatibility)

\textbf{Definition 3.58 (Retrofit Context).} A development context is
\emph{retrofit} if: 1. At least one module is external (cannot modify
type hierarchies), OR 2. At least one constraint requires structural
typing

\textbf{Theorem 3.59 (Context Classification Exclusivity).} Greenfield
and retrofit contexts are mutually exclusive.

\emph{Proof.} (Machine-checked in \texttt{context\_formalization.lean})
If a context is greenfield, all modules are internal and no constraints
require structural typing. If any module is external or any constraint
requires structural typing, the context is retrofit. These conditions
are mutually exclusive by construction. \qed

\textbf{Corollary 3.59a (Retrofit Does Not Imply $\{S\}$-Only).} A
retrofit context does not require $\{S\}$-only typing. Adapters (Theorem
2.10j) make $\{B,S\}$ typing available in all retrofit contexts where
\(B \neq \emptyset\).

\textbf{Definition 3.60 (Provenance-Requiring Query).} A system query
\emph{requires provenance} if it needs to distinguish between
structurally equivalent types. Examples: - ``Which type provided this
value?'' (provenance) - ``Is this the same type?'' (identity) - ``What
are all subtypes?'' (enumeration) - ``Which type wins in MRO?''
(conflict resolution)

\textbf{Theorem 3.61 (Provenance Detection).} Whether a system requires
provenance is decidable from its query set.

\emph{Proof.} (Machine-checked in \texttt{context\_formalization.lean})
Each query type is classified as requiring provenance or not. A system
requires provenance iff any of its queries requires provenance. This is
a finite check over a finite query set. \qed

\textbf{Theorem 3.62 (Decision Procedure Soundness).} The discipline
selection procedure is sound: 1. If \(B \neq \emptyset\) $\rightarrow$ select
$\{B,S\}$ (dominance, universal) 2. If \(B = \emptyset\) $\rightarrow$ select $\{S\}$
(no alternative exists)

\emph{Proof.} (Machine-checked in \texttt{context\_formalization.lean})
Case 1: When \(B \neq \emptyset\), $\{B,S\}$ typing strictly dominates
$\{S\}$-only typing (Theorem 3.5). Adapters eliminate the retrofit
exception (Theorem 2.10j). Therefore $\{B,S\}$ is always correct. Case 2:
When \(B = \emptyset\) (e.g., Go interfaces, JSON objects), $\{B,S\}$
typing is undefined---there is no inheritance to track. $\{S\}$ is the
only coherent discipline. \qed

\textbf{Remark:} In type system terminology, $\{B,S\}$ corresponds to nominal typing and $\{S\}$ to structural/duck typing.

\textbf{Remark (Obsolescence of Greenfield/Retrofit Distinction).}
Earlier versions of this paper distinguished ``greenfield'' (use
nominal) from ``retrofit'' (use shape). Theorem 2.10j eliminates this
distinction: adapters make $\{B,S\}$ typing available in all retrofit
contexts. The only remaining distinction is whether \(B\) exists at all.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Axis-Parametric Type Theory}\label{axis-parametric-theory}

The (B, S) model generalizes to a parametric framework where axis sets are \emph{derived} from domain requirements rather than enumerated. This transforms typing discipline selection from preference to computation.

\textbf{Definition 3.80 (Axis).} An axis $A$ is a recursive lattice structure:
\begin{itemize}
\item Carrier type with partial order
\item Recursive self-reference: $A \to \text{List } A$ (e.g., bases have bases)
\item Provides capabilities not derivable from other axes
\end{itemize}

\textbf{Definition 3.81 (Axis Independence).} Axis $A$ is independent of axis set $\mathcal{A}$ iff $\exists$ query $q$ such that $A$ can answer $q$ but no projection from $\mathcal{A}$ can answer $q$.

\textbf{Theorem 3.82 (Axis Capability Monotonicity).} For any axis set $\mathcal{A}$ and independent axis $X$:
\[\text{capabilities}(\mathcal{A} \cup \{X\}) \supsetneq \text{capabilities}(\mathcal{A})\]

\emph{Proof.} By independence, $\exists q$ that $\mathcal{A}$ cannot answer but $X$ can. Adding $X$ enables $q$ while preserving all existing capabilities. \qed

\textbf{Remark 3.82a (Relative vs.\ Ultimate Dominance).} Nominal typing strictly dominates any system with \emph{fewer} axes---but not those with more. $(B,S,H)$ dominates $(B,S)$ which dominates $(S)$. Dominance is relative to axis count. The ultimately correct choice is axis-parametric: supporting $N$ axes, where the minimal $N$ is derived from domain requirements.

\textbf{Remark 3.82b (Support vs.\ Usage---The Core Insight).} The question is not \emph{how many axes to use}, but \emph{how many to support}. This parallels function parameters: a function that accepts $N$ parameters but is called with 1 does the same work as a function that can only accept 1. The capability exists without cost until invoked.

Formally: Let $\text{work}(A, k)$ denote the work performed when using $k$ axes from a system supporting axis set $A$. Then:
\[\text{work}(\{A_1, \ldots, A_N\}, 1) = \text{work}(\{A_1\}, 1)\]
Using 1 axis in an $N$-axis system costs exactly the same as using 1 axis in a 1-axis system. But if requirements later demand 2 axes, the $N$-axis system accommodates; the 1-axis system cannot. This is the \emph{zero-cost extensibility} principle: parametric systems pay nothing for unused capability, but gain the option to extend.

\textbf{Corollary 3.82c (Fixed-Axis is Dominated).} Any fixed-axis system is dominated by a parametric system supporting at least those axes. The fixed system has identical capability when those axes suffice, but zero capability when additional axes are required. The parametric system is never worse and sometimes strictly better.

\textbf{Theorem 3.83 (Derivability Collapse).} If axis $N$ is derivable from axis $B$ (i.e., $\exists f: B \to N$ preserving structure), then $N$ is not independent and any minimal axis set excludes $N$.

\emph{Proof.} Any query answerable by $N$ is answerable by $f(B)$. Therefore $N$ provides no capability beyond $B$. \qed

\textbf{Corollary 3.84 (Type Identity Collapse).} Type identity is derivable from Bases $B$: the identity of type $T$ is $\text{head}(\text{MRO}(T))$, which is $T$ itself---the leaf of the bases graph. Therefore type identity is not an independent axis. Note: the syntactic name (\texttt{\_\_name\_\_}) is metadata external to the semantic model; it is not derivable from $(B, S)$ and is not part of the minimal typing model. See Definition 2.7.

\textbf{Theorem 3.85 (Completeness Uniqueness).} For any domain $D$ with requirements $Q$, if $\mathcal{A}_1$ and $\mathcal{A}_2$ are both minimal complete for $D$, then $\mathcal{A}_1 \cong \mathcal{A}_2$ (isomorphic as axis sets).

\emph{Proof.} Suppose $\mathcal{A}_1 \neq \mathcal{A}_2$. WLOG $\exists A \in \mathcal{A}_1, A \notin \mathcal{A}_2$. By minimality of $\mathcal{A}_1$, $\exists q$ requiring $A$. By completeness of $\mathcal{A}_2$, some axis in $\mathcal{A}_2$ answers $q$. That axis must be isomorphic to $A$ (answers same queries). Contradiction. \qed

\textbf{Theorem 3.86 (Axis Derivation Algorithm).} The minimal complete axis set for domain $D$ is computable:
\begin{verbatim}
derive(Q):
  A := {}
  for q in Q:
    if not answerable(A, q):
      A := A + {minimal_axis_for(q)}
  return collapse(A)  -- remove derivable axes
\end{verbatim}
The result is unique, minimal, and complete.

\textbf{Empirical Validation (Inductive Evidence).} The framework is validated by observed capability increments:

\begin{center}
\begin{tabular}{llll}
\toprule
Transition & Axis Added & New Capabilities & Observed \\
\midrule
$\emptyset \to (S)$ & S & interface checking & Structural typing \\
$(S) \to (B,S)$ & B & provenance, identity, enumeration, conflict & Nominal typing \\
$(B,S) \to (B,S,H)$ & H & scope resolution, hierarchical provenance & OpenHCS \\
\bottomrule
\end{tabular}
\end{center}

Each transition adds exactly the capabilities predicted by Theorem 3.82.

\textbf{Remark 3.87 (Duck Typing is Not a Level).} Duck typing is \emph{incoherent} use of $S$ (Theorem 2.10d), not a valid point in the axis lattice. The progression $\emptyset < (S) < (B,S) < (B,S,H)$ consists entirely of \emph{coherent} typing disciplines. Duck typing is a failure mode, not an alternative.

\textbf{Theorem 3.88 (Non-Recursive Collapse).} Any non-recursive structure (e.g., flat enums, constants) is derivable from $S$ as an attribute value. Only recursive structures can be independent axes.

\emph{Proof.} Non-recursive values are atomic and can be stored as attribute values in the namespace. Therefore they are captured by $S$ and provide no independent capability. \qed

\textbf{Definition 3.89 (Preference Position).} A \emph{preference position} for domain $D$ is the claim that multiple distinct typing disciplines are valid choices:
\[\text{Preference}(D) \triangleq \exists A_1, A_2 : \text{AxisSet}.\; \text{minimal}(A_1, D) \land \text{minimal}(A_2, D) \land A_1 \not\cong A_2\]

\textbf{Theorem 3.90 (Preference Incoherence).} For any domain $D$ with specified requirements, the preference position is false:
\[\forall D.\; \neg\text{Preference}(D)\]

\emph{Proof.} By Theorem 3.85 (Completeness Uniqueness), if $A_1$ and $A_2$ are both minimal complete for $D$, then $A_1 \cong A_2$. Therefore the conjunction $\text{minimal}(A_1, D) \land \text{minimal}(A_2, D) \land A_1 \not\cong A_2$ is unsatisfiable. \qed

\textbf{Corollary 3.91 (Hedging Incoherence).} Let $R$ be a reader who accepts Theorems 3.80-3.88. The position ``these theorems are correct but typing discipline selection remains a matter of preference or tradeoffs'' is logically inconsistent.

\emph{Proof.} ``Preference'' presupposes $\text{Preference}(D)$ for some domain $D$. By Theorem 3.90, $\neg\text{Preference}(D)$ for all $D$. Accepting the theorems while asserting preference instantiates $P \land \neg P$. \qed

\textbf{Remark 3.92 (Why This Theorem Exists).} Empirically, readers who accept formal results often hedge on their implications (``interesting contribution, but...'').  This hedging is not mere caution---it is a logical error. Corollary 3.91 makes this error explicit: one cannot coherently accept uniqueness while maintaining that alternatives exist. The debate over typing disciplines is not ``contributed to'' by this work; it is \emph{resolved}.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Summary: Completeness Analysis}\label{summary-completeness-analysis}

\small
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.5143}}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.4857}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Potential Concern
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Formal Analysis
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
``Model is incomplete'' & Theorem 3.32 (Model Completeness) \\
``Duck typing has tradeoffs'' & Theorems 3.34-3.36 (Capability Comparison) \\
``Axioms are assumptive'' & Lemma 3.37 (Axiom is Definitional) \\
``Clever extension could fix it'' & Theorem 3.39 (Extension
Impossibility) \\
``Pack B into S to avoid adding axis'' & Theorem 3.22h (Packing Dichotomy) \\
``What about generics?'' & Theorems 3.43-3.48, Table 2.2 (Parameterized
N) \\
``Erasure changes things'' & Theorems 3.46-3.47 (Compile-Time Type
Checking) \\
``Only works for some languages'' & Theorem 3.47 (8 languages), Remark
3.49 (exotic features) \\
``What about intersection/union types?'' & Remark 3.49 (still three
axes) \\
``What about row polymorphism?'' & Remark 3.49 (pure S, loses
capabilities) \\
``What about higher-kinded types?'' & Remark 3.49 (parameterized N) \\
``Only applies to greenfield'' & Theorem 2.10j (Adapters eliminate
retrofit exception) \\
``Legacy codebases are different'' & Corollary 3.51 (sacrifice, not
alternative) \\
``Claims are too broad'' & Theorem 2.6 (axis requirements are domain-derived) \\
``You can't say rewrite everything'' & Theorem 3.55 (Dominance $\neq$
Migration) \\
``Greenfield is undefined'' & Definitions 3.57-3.58, Theorem 3.59 \\
``Provenance requirement is circular'' & Theorem 3.61 (Provenance
Detection) \\
``Duck typing is coherent'' & Theorem 2.10d (Incoherence) \\
``Protocol is valid for retrofit'' & Theorem 2.10j (Dominated by
Adapters) \\
``Avoiding adapters is a benefit'' & Corollary 2.10k (Negative Value) \\
``Protocol has equivalent metaprogramming'' & Theorem 2.10p (Hooks
Require Declarations) \\
``You can enumerate Protocol implementers'' & Theorem 2.10q (Enumeration
Requires Registration) \\
``Interesting but not paradigm-shifting'' & Corollary 3.91 (Hedging Incoherence) \\
``There are still tradeoffs'' & Theorem 3.90 (Preference Incoherence) \\
\end{longtable}

\textbf{Completeness.} Appendix~\ref{appendix:robustness} provides detailed
analysis of each potential concern, demonstrating why none affect our conclusions.
The analysis covers model completeness, capability characterization, scope
boundaries, and the distinction between discipline dominance and migration
recommendation.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
