\section{Cross-Regime Transfer}\label{sec:cross-regime-transfer}

When does a sufficiency result in regime $R_1$ help in regime $R_2$?

\subsection{Transfer Conditions}

\begin{enumerate}
\item \textbf{Static $\to$ Stochastic}: Transfers iff distribution $P$ concentrates on a tractable static subcase. Formally: if $\text{supp}(P)$ lies within a structure-detectable class $C$, then static sufficiency for $C$ implies stochastic sufficiency under $P$. Otherwise: bridge failure (already mechanized).

\item \textbf{Static $\to$ Sequential}: Transfers iff horizon $T = 1$ and transitions are deterministic. This is the one-step bridge (already proved in Paper 4: \LH{CC30}). For $T > 1$: bridge failure (\LH{CC18}).

\item \textbf{Stochastic $\to$ Sequential}: Transfers iff transitions are memoryless AND observation model is regime-compatible. Formally: if $T(s'|a,s)$ depends only on the stochastically sufficient coordinates of $s$, then stochastic sufficiency lifts to sequential sufficiency. Otherwise: new bridge-failure witness needed.

\item \textbf{Sequential $\to$ Static}: Always degenerates. A sequential problem with $T=1$ and deterministic transitions IS a static problem.
\end{enumerate}

\begin{claim}{Transfer static to stochastic}{th:transfer-static-stochastic}
Static sufficiency transfers to stochastic iff $P$ concentrates on tractable subcase.
\claimstamp{T\ref{th:transfer-static-stochastic}}{RG}
\end{claim}

\leanmeta{\LH{CC53}}

\begin{claim}{Transfer stochastic to sequential}{th:transfer-stochastic-sequential}
Stochastic sufficiency transfers to sequential iff memoryless transitions and regime-compatible observations.
\claimstamp{T\ref{th:transfer-stochastic-sequential}}{RG}
\end{claim}

\leanmeta{\LH{CC57}}

\subsection{Heuristic Transfer Across Regimes}

Paper 4's heuristic reusability bridge generalizes: if structure detection is polynomial in regime $R_1$, and the detected structure is preserved under regime extension to $R_2$, then the heuristic transfers.

Key question: when is structure preserved?
\begin{itemize}
\item Product distributions preserve separable utility
\item Deterministic transitions preserve tree structure
\end{itemize}
