\section{Impossibility Results}\label{impossibility-results}

\subsection{The Text Credibility
Bound}\label{the-text-credibility-bound}

\begin{theorem}[Text Credibility Bound]\label{thm:text-bound}
For any text string $T$ (memory content, assertion, etc.) and high-magnitude
claim $c$ with $M(c) > M^*$ (i.e., prior $p < e^{-M^*}$):
\[\Pr[c \mid T] < \tau\]
where $\tau < 1$ is determined by the mimicability $q$ and $M^*$.
No text achieves full credibility for exceptional claims.

\nolinkurl{Credibility.Impossibility.text_credibility_bound} \nolinkurl{Credibility.Impossibility.high_magnitude_credibility_small} \nolinkurl{Credibility.Impossibility.memory_iteration_futility}
\end{theorem}

\begin{proof}
Text is cheap talk (production cost independent of truth). Apply
Theorem~\ref{thm:cheap-talk-bound} with prior $p = e^{-M^*}$ and mimicability $q$:
\[
\tau = \frac{p}{p + (1-p)q} = \frac{e^{-M^*}}{e^{-M^*} + (1 - e^{-M^*})q}
\]
For $M^*$ large (low prior probability), $\tau \to 0$ regardless of $q > 0$.
\end{proof}

\textbf{Corollary 5.2 (Memory Iteration Futility).} No rephrasing of
memory content can achieve credibility above \(\tau\) for high-magnitude
claims. Iteration on text is bounded in effectiveness.

\textbf{Interpretation:} This is why we couldn't solve the credibility
problem by editing memory text. The \emph{structure} of the problem
(text is cheap talk, claims are high-magnitude) guarantees bounded
credibility regardless of phrasing.

\subsection{Optimal Strategies}\label{optimal-strategies}

\begin{theorem}[Optimal Credibility Strategy]\label{thm:optimal-credibility-strategy}
For high-magnitude claims in regime \texttt{[CT]}+\texttt{[VS]}, an optimal strategy is:
\begin{enumerate}
\item Minimize cheap talk (reduce emphasis and meta-assertions),
\item Maximize costly-signal exposure (show work and verifiable artifacts),
\item Enable costly-to-fake demonstration channels.
\end{enumerate}
\textit{(Lean anchor: \texttt{optimal\_strategy\_dominance}.)}
\end{theorem}

\begin{proof}
From Theorem~\ref{thm:text-bound}, text-only updates are bounded for high-magnitude claims.
From Theorem~\ref{thm:verified-signal}, verifier-backed channels strictly improve achievable posterior credibility as false-positive rate decreases.
Therefore optimal policy minimizes bounded channels and allocates effort to verifier-backed costly channels.
\end{proof}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
