\section{Leverage Integration}\label{leverage-integration}

\subsection{Credibility as DOF
Minimization}\label{credibility-as-dof-minimization}

Applying the leverage framework (Paper 3) \cite{paper3_leverage}:

\textbf{Signal DOF:} Words in an assertion are degrees of freedom. Each
word can be independently modified.

\textbf{Signal Leverage:} \(L_S = \frac{\Delta C}{\text{Words}}\)

\begin{theorem}[Credibility Leverage]\label{thm:credibility-leverage}
For cheap-talk signals with nonnegative credibility impact, leverage is maximized by minimizing word count:
\[
\arg\max_s L_S(s) = \arg\min_s |s|
\]
subject to conveying the claim.

\LH{C_c960ce0f} \LH{C_c49e8c99} \LH{C_55861ab0}
\end{theorem}

\begin{proof}
With impact fixed and nonnegative, \(L_S=\Delta C/\text{Words}\) is inverse-monotone in word count.
Hence shorter valid signals weakly dominate longer ones in leverage.
\end{proof}

\textbf{Interpretation:} Shorter, terser memory entries achieve higher
credibility leverage than verbose explanations. ``70k lines, deployed in
3 labs'' beats lengthy justification.

\subsection{Energy Budget Interpretation}\label{energy-budget-interpretation}

The cost differential $\Delta$ (in joules) induces a natural budget interpretation.
Let $B$ be an agent's total energy budget for signal production. Then:

\begin{itemize}
\item A cheap-talk strategy exhausts $B$ producing signals with $\Delta = 0$, yielding
  credibility bounded by Theorem~\ref{thm:cheap-talk-bound} regardless of $B$.
\item A costly-signal strategy concentrates $B$ on signals with large $\Delta$, yielding
  credibility approaching $1$ as $\Delta \to \infty$ (Theorem~\ref{thm:costly-signal}).
\item Optimal credibility leverage (Theorem~\ref{thm:credibility-leverage}) therefore
  allocates budget entirely to high-$\Delta$ channels: \emph{demonstrations, working code,
  machine-checked proofs} rather than assertions.
\end{itemize}

\textbf{Corollary (Minimality of honest signaling).} An honest agent with true claim $c$
achieves maximal $L_C$ by producing the minimal proof of $c$: one correctly compiling Lean
file, zero additional assertions. Additional words add DOF (increase $|s|$) without increasing $\Delta$,
strictly decreasing leverage.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
