\section{Discussion}\label{sec:discussion}

We address anticipated questions about the scope and implications of the main results.

\subsection{On Physical Grounding of Computational Hardness}\label{sec:locality-physics}

The complexity results rest on $\P \neq \coNP$ (equivalently, $\P \neq \NP$). We make precise what rejecting this assumption requires.

Consider a bounded spacetime region $R$ with finite energy capacity $E$. Any observation requires energy (Landauer's principle: $kT\ln 2$ per bit~\cite{landauer1961irreversibility}). If the state space has $|S|$ configurations and each query costs $\varepsilon > 0$ energy, then $E < |S| \cdot \varepsilon$ implies the region cannot observe all configurations (\LH{LP32}, \LH{LP33}).

This forces \emph{locality}: bounded regions cannot have complete knowledge of the universe (\LH{LP34}). But locality implies \emph{independent local configurations}: spacelike-separated regions maintain configurations that need not agree (\LH{LP35}, \LH{LP10c}). When such regions interact (``board merge''), their configurations may contradict (\LH{LP36}, \LH{LP18}).

\begin{theorem}[P $\neq$ NP Necessary for Physics, informal]\label{thm:pne-np-physics}
Under:
\begin{enumerate}
\item Landauer's principle (queries cost energy) --- \LH{LP41}
\item Nontrivial state space ($|S| > 1$) --- \LH{LP42}
\item Finite signal speed (special relativity) --- \LH{LP43}
\item Finite universe energy --- \LH{LP44}
\end{enumerate}
spacelike-separated regions can hold locally valid configurations that contradict on merge (\LH{LP38}).
\end{theorem}

The formal statement and proof are in \texttt{LocalityPhysics.lean} (\LH{LP38}). If $\P = \NP$, bounded-energy observers could determine global consistency in polynomial time, eliminating the locality that makes spatially distributed matter possible (\LH{LP39}, \LH{LP40}).

To reject this theorem, one must deny:
\begin{itemize}
\item Landauer's principle (empirically verified~\cite{berut2012experimental}), or
\item that physics has multiple states (we observe them), or
\item special relativity (signal speed is finite), or
\item finite universe energy (thermodynamics)
\end{itemize}

Each denial contradicts established physics. The ``integrity trap'' (\LH{LP45}) is that rejecting the mathematical claim requires rejecting physical law.

\paragraph{Explicit Regime Specification.}
The theorem holds under the following physical regime. The spacetime region $R$ must satisfy four empirically verified conditions. First, bounded energy capacity $E < \infty$ at temperature $T > 0$ (thermodynamic regime, \LH{EP2}). Second, nonzero energy cost per bit query $\varepsilon \geq kT\ln 2$ (Landauer's principle, \LH{EP1}). Third, finite signal speed $c < \infty$ (special relativity regime, \LH{EP3}). Fourth, state space with $|S| > 1$ configurations (nontrivial physics regime, \LH{EP4}).

\paragraph{Step-by-Step Deduction.}
The proof proceeds by necessary implication through seven steps. Step 1: Finite energy $E$ and nonzero query cost $\varepsilon$ imply $E < |S| \cdot \varepsilon$ for sufficiently large $|S|$ (\LH{LP32}). Step 2: Energy constraint implies incomplete observation of the state space (\LH{LP33}). Step 3: Incomplete observation forces locality; bounded regions cannot have complete knowledge of the universe (\LH{LP34}). Step 4: Locality implies independent regions can exist; spacelike separated regions maintain configurations that need not agree (\LH{LP35}). Step 5: Independent regions can disagree; when such regions interact their configurations may contradict (\LH{LP36}). Step 6: If $\P = \NP$, bounded energy observers could resolve global consistency in polynomial time, eliminating the locality requirement (\LH{LP39}). Step 7: Eliminating locality contradicts the finite energy constraint. Therefore P $\neq$ NP is necessary for physical locality as observed (\LH{LP38}).

\paragraph{The Integrity Trap Structure.}
To reject this theorem, one must deny at least one of four empirically justified axioms. Denying Landauer's principle (\LH{LP41}) contradicts experiments by Berut et al. (2012) that verified the $kT\ln 2$ energy cost. Denying nontrivial state space (\LH{LP42}) contradicts direct observation of multiple physical configurations. Denying finite signal speed (\LH{LP43}) contradicts special relativity. Denying finite universe energy (\LH{LP44}) contradicts thermodynamics. The trap is complete; the mathematical claim is shielded by empirical axioms.

\subsection{On Tractability Despite Hardness}

\textbf{Question:} ``coNP-complete problems have good heuristics or approximations in practice. Does the hardness result preclude practical solutions?''

\textbf{Response:} This observation actually \emph{strengthens} our thesis. The point is not that practitioners cannot find useful approximations; they clearly do (feature selection heuristics in ML, sensitivity analysis in economics, configuration defaults in software). The point is that \emph{optimal} dimension selection is provably hard.

The prevalence of heuristics across domains is itself evidence of the computational barrier. If optimal selection were tractable, we would see optimal algorithms, not heuristics. The widespread adoption of ``include more than necessary'' strategies is a rational response to coNP-completeness.

\subsection{On Coordinate Structure Assumptions}

\textbf{Question:} ``Real decision problems are messier than the clean product-space model. Is the coordinate structure assumption too restrictive?''

\textbf{Response:} The assumption is weak. Any finite state space admits a binary-coordinate encoding; our hardness results apply to this encoding. More structured representations make the problem \emph{easier}, not harder; so hardness for structured problems implies hardness for unstructured ones.

The coordinate structure abstracts common patterns: independent sensors, orthogonal configuration parameters, factored state spaces. These are ubiquitous in practice precisely because they enable tractable reasoning in special cases (Theorem~\ref{thm:tractable}).

\subsection{On the SAT Reduction}

\textbf{Question:} ``The reduction from SAT/TAUT is an artifact of complexity theory. Do real decision problems encode Boolean formulas?''

\textbf{Response:} All coNP-completeness proofs use reductions. The reduction demonstrates that TAUT instances admit an encoding as sufficiency-checking problems while preserving computational structure. This is standard methodology \cite{cook1971complexity, karp1972reducibility}.

The claim is not that practitioners encounter SAT problems in disguise, but that sufficiency checking is \emph{at least as hard as} TAUT. If sufficiency checking is tractable, then TAUT is solvable in polynomial time, contradicting the widely-believed $\P \neq \NP$ conjecture.

\subsection{On Tractable Subcase Scope}

\textbf{Question:} ``The tractable subcases (bounded actions, separable utility, tree structure) seem restrictive. Do they cover real problems?''

\textbf{Response:} These subcases characterize \emph{when} dimension selection becomes feasible:

\begin{itemize}
\item \textbf{Bounded actions:} Many real decisions have few options (buy/sell/hold, accept/reject, left/right/straight)
\item \textbf{Separable utility:} Additive decomposition is common in economics and operations research
\item \textbf{Tree structure:} Hierarchical dependencies appear in configuration, organizational decisions, and causal models
\end{itemize}

The dichotomy theorem (Theorem~\ref{thm:dichotomy}) precisely identifies the boundary in the stated encoding regimes. The contribution is not that all problems are hard, but that hardness is the \emph{default} unless special structure exists.

\subsection{On the Value of Formalization}

\textbf{Question:} ``Feature selection is known to be hard. Does this paper just add mathematical notation to folklore?''

\textbf{Response:} The contribution is unification. Prior work established hardness for specific domains (feature selection in ML \cite{guyon2003introduction}, factor identification in economics, variable selection in statistics). We prove a general result that applies to decision problems with coordinate structure within the formal model.

This generality explains why the same ``over-modeling'' pattern recurs across unrelated domains. It's not that each domain independently discovered the same heuristic; it's that each domain independently hit the same computational barrier.

\subsection{On the Lean Formalization Scope}

\textbf{Question:} ``The Lean formalization models an idealized version of the problem. Real coNP-completeness proofs involve Turing machines.''

\textbf{Response:} The Lean formalization captures the mathematical structure of the reduction, not the Turing machine details. We prove:

\begin{enumerate}
\item The sufficiency-checking problem is in coNP (verifiable counterexample)
\item TAUT reduces to sufficiency checking (polynomial-time construction)
\item The reduction preserves yes/no answers (correctness)
\end{enumerate}

These are the mathematical claims that establish coNP-completeness. The Turing machine encoding is implicit in Lean's computational semantics. The formalization provides machine-checked verification that the reduction is correct.

\subsection{On the Dichotomy Gap}

\textbf{Question:} ``The dichotomy between $O(\log n)$ and $\Omega(n)$ minimal sufficient sets leaves a gap. What about $O(\sqrt{n})$?''

\textbf{Response:} The dichotomy is tight in the stated regimes under ETH. The gap corresponds to problems reducible to a polynomial number of SAT instances; these sit in the polynomial hierarchy between P and coNP.

In practice, the dichotomy captures the relevant cases: either the problem has logarithmic dimension (tractable) or linear dimension (intractable). Intermediate cases exist theoretically; we do not rule them out.

\subsection{On Practical Guidance}

\textbf{Question:} ``Proving hardness doesn't help engineers solve their problems. Does this paper offer constructive guidance?''

\textbf{Response:} Understanding limits is constructive. The paper provides:

\begin{enumerate}
\item \textbf{Tractable subcases} (Theorem~\ref{thm:tractable}): Check if your problem has bounded actions, separable utility, or tree structure
\item \textbf{Justification for heuristics}: Over-modeling is not laziness; it's computationally rational
\item \textbf{Focus for optimization}: Don't waste effort on optimal dimension selection; invest in good defaults and local search
\end{enumerate}

Knowing that optimal selection is coNP-complete frees practitioners to use heuristics without guilt. This is actionable guidance.

\subsection{On Learning Costs}

\textbf{Question:} ``The Simplicity Tax analysis ignores learning costs. Simple tools have lower barrier to entry, which matters for team adoption.''

\textbf{Response:} This conflates $H_{\text{central}}$ (learning cost) with total cost. Yes, simple tools have lower learning cost. But for $n$ use sites, the total cost is:
\[H_{\text{total}} = H_{\text{central}} + n \times H_{\text{distributed}}\]

The learning cost is paid once; the per-site cost is paid $n$ times. For $n > H_{\text{central}} / H_{\text{distributed}}$, the ``complex'' tool with higher learning cost has lower total cost.

\subsection{On Organizational Constraints}

\textbf{Question:} ``In practice, teams use what they know. Advocating for 'complex' tools ignores organizational reality.''

\textbf{Response:} The Simplicity Tax is paid regardless of whether your team recognizes it. If your team writes boilerplate at 50 locations because they don't know metaclasses, they pay the tax; in time, bugs, and maintenance.

Organizational reality is a constraint on \emph{implementation}, not on \emph{what is optimal}. The Simplicity Tax Theorem identifies the optimal; the practitioner's task is to approach it within organizational constraints.

\subsection{On Deferred Refactoring}

\textbf{Question:} ``Start simple, refactor when needed. Technical debt is manageable.''

\textbf{Response:} Refactoring from distributed to centralized is $O(n)$ work; you pay the accumulated Simplicity Tax all at once. If you have $n$ sites each paying tax $k$, refactoring costs at least $nk$ effort.

Moreover, distributed implementations create dependencies. Each workaround becomes a local assumption that must be preserved during refactoring. Under dependency models where local assumptions interact, refactoring cost is \emph{superlinear} in $n$.

\subsection{On Weighted Importance}

\textbf{Question:} ``Real problems have axes of varying importance. A tool that covers the important axes is good enough.''

\textbf{Response:} The theorem is conservative: it counts axes uniformly. Weighted versions strengthen the result.

If axis $a$ has importance $w_a$, define weighted tax:
\[\text{WeightedTax}(T, P) = \sum_{a \in R(P) \setminus A(T)} w_a\]

The incomplete tool pays $\sum w_a \times n$ while the complete tool pays 0. The qualitative result is unchanged.

The ``cover important axes'' heuristic only works if you \emph{correctly identify} which axes are important. By Theorem~\ref{thm:sufficiency-conp}, this identification is coNP-complete; returning us to the original hardness result.

\subsection{On Tool-Relative Collapse Claims}

\textbf{Question:} ``Can we state `tools collapse complexity' rigorously without claiming universal complexity-class collapse?''

\textbf{Response:} Yes. We formalize a baseline work model and a tool-augmented model with a pointwise no-worse condition, then define an effective-collapse predicate: never worse and eventually strictly better. Lean proves no-worse monotonicity and collapse construction from eventual strict improvement (`DecisionQuotient.ToolCollapse.tool\_never\_worse`, `DecisionQuotient.ToolCollapse.effective\_collapse\_of\_eventual\_strict`). A concrete witness family is fully mechanized: baseline work $2^n$ versus tool work $n$, yielding eventual strict improvement and effective collapse (`DecisionQuotient.ToolCollapse.linear\_tool\_eventual\_strict`, `DecisionQuotient.ToolCollapse.linear\_tool\_effective\_collapse`). This is explicitly model-relative and does not assert a universal class collapse.
