\section{Cheap Talk Theorems}\label{cheap-talk-theorems}

\subsection{The Cheap Talk Bound}\label{the-cheap-talk-bound}

\begin{theorem}[Cheap-talk credibility is a likelihood-ratio bound]\label{thm:cheap-talk-bound}
Let $C\in\{0,1\}$ denote the truth of a claim ($C=1$ true), with prior
$p := \Pr[C=1]\in(0,1)$. Let $S$ be the event that the receiver observes a
particular message-pattern (signal) $s$.

\LH{C_c6e273f0} \LH{C_123d1f33}

Define the emission rates
\[
\alpha := \Pr[S \mid C=1],\qquad \beta := \Pr[S \mid C=0].
\]
Then the posterior credibility of the claim given observation of $s$ is
\[
\Pr[C=1 \mid S] \;=\; \frac{p\,\alpha}{p\,\alpha + (1-p)\,\beta}.
\]
Equivalently, in odds form,
\[
\frac{\Pr[C=1 \mid S]}{\Pr[C=0 \mid S]}
\;=\;
\frac{p}{1-p}\cdot \frac{\alpha}{\beta}.
\]

In particular, if $s$ is a \emph{cheap-talk} pattern in the sense that:
\begin{enumerate}
\item[(i)] truthful senders emit $s$ with certainty ($\alpha=1$), and
\item[(ii)] deceptive senders can mimic $s$ with probability at least $q$
(i.e.\ $\beta \ge q$),
\end{enumerate}
then credibility obeys the tight upper bound
\[
\Pr[C=1 \mid S] \;\le\; \frac{p}{p+(1-p)q}.
\]
Moreover this bound is \emph{tight}: equality holds whenever $\alpha=1$ and $\beta=q$.
\end{theorem}

\begin{proof}
By Bayes' rule,
\[
\Pr[C=1 \mid S]
= \frac{\Pr[S\mid C=1]\Pr[C=1]}{\Pr[S\mid C=1]\Pr[C=1] + \Pr[S\mid C=0]\Pr[C=0]}
= \frac{p\alpha}{p\alpha+(1-p)\beta}.
\]
If $\alpha=1$ and $\beta \ge q$, the denominator is minimized by setting $\beta=q$,
yielding
\[
\Pr[C=1 \mid S]\le \frac{p}{p+(1-p)q}.
\]
Tightness is immediate when $\beta=q$.
\end{proof}

\textbf{Remark (Notation reconciliation).} In this paper we use $q$ to denote the
\emph{mimicability} of a cheap-talk signal: the probability that a deceptive sender
successfully produces the same message pattern as a truthful sender. If one prefers
to work with detection probability $\pi_d$ (the probability deception is detected),
then $q = 1 - \pi_d$ and the bound becomes
$\Pr[C=1 \mid S] \le p / (p + (1-p)(1-\pi_d))$.

\textbf{Interpretation:} No matter how emphatically you assert
something, cheap talk credibility is capped. The cap depends on how
likely deception is in the population.

\subsection{The Magnitude Penalty}\label{the-magnitude-penalty}

\begin{theorem}[Magnitude Penalty]\label{thm:magnitude-penalty}
For claims $c_1, c_2$ with $M(c_1) < M(c_2)$ (i.e., $p_1 := P(c_1) > p_2 := P(c_2)$)
and identical cheap talk signals $s$ with mimicability $q$:
\[\Pr[c_1 \mid S] > \Pr[c_2 \mid S]\]
Higher-magnitude claims receive less credibility from identical signals.

\LH{C_6c29d508} \LH{C_a54e07fe}
\end{theorem}

\begin{proof}
From Theorem~\ref{thm:cheap-talk-bound}, the bound $p/(p+(1-p)q)$ is strictly
increasing in $p$ for fixed $q \in (0,1)$. Since $p_1 > p_2$, we have
$\Pr[c_1 \mid S] > \Pr[c_2 \mid S]$.
\end{proof}

\textbf{Interpretation:} Claiming you wrote one good paper gets more
credibility than claiming you wrote four. The signal (your assertion) is
identical; the prior probability differs.

\subsection{The Emphasis Penalty}\label{the-emphasis-penalty}

\begin{theorem}[Emphasis Penalty]\label{thm:emphasis-penalty}
Let \(s_1, s_2, \ldots, s_n\) be cheap talk signals all asserting claim \(c\) with the
same mimicability $q$ and prior $p = \Pr[C=1]$. Assume the signaling equilibrium condition:
\begin{itemize}
\item[(EP)] Deceptive agents produce repetitions at least as readily as honest agents:
$\Pr[s_k \mid C{=}0,\, s_{1..k-1}] \;\geq\; \Pr[s_k \mid C{=}1,\, s_{1..k-1}]$ for $k > k^*$.
\end{itemize}
Then for $n > k^*$: $\frac{\partial C(c, s_{1..n})}{\partial n} \leq 0$, with strict decrease whenever
the inequality in (EP) is strict.

\LH{C_10ce8e6e} \LH{C_5efdc080} \LH{C_cd94ef32}
\end{theorem}

\begin{proof}
Let $p_k := \Pr[C=1 \mid s_1,\ldots,s_k]$ denote the posterior after $k$ assertions.
By Bayes' rule applied to the $k$-th update:
\[
p_k \;=\; \frac{p_{k-1}\,\alpha_k}{p_{k-1}\,\alpha_k + (1-p_{k-1})\,\beta_k}
\]
where $\alpha_k := \Pr[s_k \mid C{=}1, s_{1..k-1}]$ and $\beta_k := \Pr[s_k \mid C{=}0, s_{1..k-1}]$.

The odds form is:
\[
\frac{p_k}{1-p_k} \;=\; \frac{p_{k-1}}{1-p_{k-1}} \cdot \frac{\alpha_k}{\beta_k}.
\]
Under condition (EP), $\alpha_k / \beta_k \leq 1$ for $k > k^*$.
Therefore the odds ratio is non-increasing, i.e., $p_k \leq p_{k-1}$ for $k > k^*$.
When (EP) holds with strict inequality, $p_k < p_{k-1}$, giving $\partial C / \partial n < 0$.

\textit{Physical interpretation.} Since all assertions are cheap talk ($\Delta = 0$ in joules),
additional assertions carry zero energy cost differential. A rational receiver therefore
assigns the repetition count $n$ the same analysis as any cheap-talk signal:
it is informative only about the sender's strategy. Under common knowledge of rationality,
deceptive senders choose large $n$ to mimic honest senders; receivers anticipate this, making
(EP) the equilibrium condition.
\end{proof}

\textbf{Interpretation:} ``Trust me, I'm serious, this is absolutely
true, I swear'' is \emph{less} credible than just stating the claim. The
emphasis signals desperation.

\subsection{The Meta-Assertion Trap}\label{the-meta-assertion-trap}

\begin{theorem}[Meta-Assertion Trap]\label{thm:meta-assertion-trap}
Let \(a\) be a cheap talk assertion and \(m\) be a meta-assertion ``assertion \(a\) is credible.''
Then: \[C(c, a \cup m) \leq C(c, a) + \epsilon\]
where \(\epsilon \to 0\) as common knowledge of rationality increases.

\LH{C_95cbc13f} \LH{C_a0cb6371} \LH{C_d80a8806}
\end{theorem}

\begin{proof}
Meta-assertion \(m\) is itself cheap talk (costs nothing to produce
regardless of truth). Therefore \(m\) is subject to the Cheap Talk Bound
(Theorem~\ref{thm:cheap-talk-bound}).

Under common knowledge of rationality, agents anticipate that deceptive
agents will produce meta-assertions. Therefore:
\[P(m | \text{deceptive}) \approx P(m | \text{honest})\]

The signal provides negligible information; \(\epsilon \to 0\).
\end{proof}

\textbf{Interpretation:} ``My claims are verified'' is cheap talk about
cheap talk. It doesn't escape the bound---it's \emph{subject to} the
bound recursively. Adding ``really verified, I promise'' makes it worse.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

