\section{References}\label{references}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Barrett, K., et al.~(1996). ``A Monotonic Superclass Linearization for
  Dylan.'' OOPSLA.
\item
  Van Rossum, G. (2002). ``Unifying types and classes in Python 2.2.''
  PEP 253.
\item
  The Python Language Reference, §3.3.3: ``Customizing class creation.''
\item
  Malayeri, D. \& Aldrich, J. (2008). ``Integrating Nominal and
  Structural Subtyping.'' ECOOP.
\item
  Malayeri, D. \& Aldrich, J. (2009). ``Is Structural Subtyping Useful?
  An Empirical Study.'' ESOP.
\item
  Abdelgawad, M. \& Cartwright, R. (2014). ``NOOP: A Domain-Theoretic
  Model of Nominally-Typed OOP.'' ENTCS.
\item
  Abdelgawad, M. (2016). ``Why Nominal-Typing Matters in OOP.''
  arXiv:1606.03809.
\item
  Gil, J. \& Maman, I. (2008). ``Whiteoak: Introducing Structural Typing
  into Java.'' OOPSLA.
\item
  Veldhuizen, T. (2006). ``Tradeoffs in Metaprogramming.'' ACM Computing
  Surveys.
\item
  Damaševičius, R. \& Štuikys, V. (2010). ``Complexity Metrics for
  Metaprograms.'' Information Technology and Control.
\item
  Liskov, B. \& Wing, J. (1994). ``A Behavioral Notion of Subtyping.''
  ACM TOPLAS.
\item
  Blum, M. (1967). ``On the Size of Machines.'' Information and Control.
\item
  Cook, W., Hill, W. \& Canning, P. (1990). ``Inheritance is not
  Subtyping.'' POPL.
\item
  de Moura, L. \& Ullrich, S. (2021). ``The Lean 4 Theorem Prover and
  Programming Language.'' CADE.
\item
  Leroy, X. (2009). ``Formal verification of a realistic compiler.''
  Communications of the ACM.
\item
  Klein, G., et al.~(2009). ``seL4: Formal verification of an OS
  kernel.'' SOSP.
\item
  Siek, J. \& Taha, W. (2006). ``Gradual Typing for Functional
  Languages.'' Scheme and Functional Programming Workshop.
\item
  Wadler, P. \& Findler, R. (2009). ``Well-Typed Programs Can't Be
  Blamed.'' ESOP.
\item
  Gamma, E., Helm, R., Johnson, R. \& Vlissides, J. (1994). ``Design
  Patterns: Elements of Reusable Object-Oriented Software.''
  Addison-Wesley.
\item
  Peters, T. (2004). ``PEP 20 -- The Zen of Python.'' Python Enhancement
  Proposals.
\item
  TypeScript GitHub Issue \#202 (2014). ``Nominal types.''
  https://github.com/microsoft/TypeScript/issues/202
\item
  TypeScript GitHub Issue \#33038 (2019). ``Proposal: Nominal Type
  Tags.'' https://github.com/microsoft/TypeScript/issues/33038
\end{enumerate}

