\section{Conclusion}\label{conclusion}

We have formalized and proved that minimal theories are unique and computable from 
query spaces.

\textbf{Main Results:}
\begin{enumerate}
\item \textbf{Unique Minimal Theory (Theorem~\ref{thm:unique-minimal-theory}):} For any 
finite domain, exactly one minimal theory exists up to isomorphism.

\item \textbf{Compression Necessity (Theorem~\ref{thm:compression-necessity}):} Infinite 
query spaces require theories smaller than implementations.

\item \textbf{Computability from Queries (Theorem~\ref{thm:computability-from-queries}):} 
The minimal theory is algorithmic function of query space.

\item \textbf{Incoherence of Pluralism (Theorem~\ref{thm:anti-pluralism}):} Multiple 
non-isomorphic minimal theories cannot coexist.
\end{enumerate}

\textbf{Implications:}
\begin{itemize}
\item Scientific theories converge because minimal structure is unique, not by convention
\item Theory choice is mechanical extraction, not creative interpretation
\item Occam's razor is computational requirement, not aesthetic preference
\item Papers 1--3 are instances of general uniqueness pattern
\end{itemize}

\subsection{Broader Context}\label{broader-context}

This paper provides epistemological foundation for the pentalogy:

\textbf{Paper 1:} Proves axis orthogonality is unique minimal theory for attribute
independence queries.

\textbf{Paper 2:} Proves DOF = 1 is unique minimal theory for multi-scale coherence queries.

\textbf{Paper 3:} Proves leverage is unique minimal theory for statistical influence queries.

\textbf{Paper 7 (this paper):} Proves all minimal theories are unique---explains why
Papers 1--3 found uniqueness.

\textbf{Paper 6 (Formality and Universality):} Follows logically from this result.
If minimal theories are unique (Paper 7), and formality guarantees universality, then
formal minimal theories ARE the structure of reality---not descriptions of it. Paper 6
makes the metaphysical move that Paper 7 enables mathematically.

The pattern is not coincidence but mathematical necessity.

\subsection{Future Work}\label{future-work}

\begin{itemize}
\item Extend to infinite domains with suitable convergence criteria
\item Formalize complexity classes for theory discovery algorithms
\item Connect to category theory: theories as initial objects in query-answering categories
\item Apply to other domains: physics, biology, economics
\end{itemize}

\textbf{The Central Insight:} Reality has minimal structure answering all queries. 
Science is the algorithm extracting this structure. Theories converge because 
the minimal compression is unique.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

