\section{Introduction}\label{introduction}

\textbf{Theorem (Main Result).} There exists a computable function $f: \text{Requirements} \to \text{Architecture}$ such that $f(R)$ minimizes expected error probability among all architectures satisfying $R$.

\textbf{Proof sketch.} Define leverage $L(A) = |\text{Capabilities}(A)|/\text{DOF}(A)$. We prove:
\begin{enumerate}
\item Architecture with $n$ DOF is isomorphic to series system with $n$ components (Theorem~\ref{thm:dof-reliability})
\item Series system error probability: $P_{\text{error}}(n) = 1-(1-p)^n$ (standard reliability theory)
\item For $\text{Capabilities}(A_1) = \text{Capabilities}(A_2)$: $L(A_1) > L(A_2) \iff P_{\text{error}}(A_1) < P_{\text{error}}(A_2)$ (Theorem~\ref{thm:leverage-error})
\item Therefore: $f(R) = \arg\max_{A: \text{Cap}(A) \supseteq R} L(A)$ (Theorem~\ref{thm:optimal})
\end{enumerate}

This establishes decidability of architectural optimization for the error minimization objective.

\subsection{Definitions}

\textbf{Definition (Informal):} \emph{Leverage} is the ratio of capabilities to degrees of freedom:
\[
L = \frac{|\text{Capabilities}|}{\text{DOF}}
\]

\textbf{Degrees of Freedom (DOF):} Independent state variables in the architecture. Each DOF represents a location that can be modified independently:
\begin{itemize}
\item $n$ microservices $\to$ DOF $= n$ (each service is independently modifiable)
\item Code copied to $n$ locations $\to$ DOF $= n$ (each copy is independent)
\item Single source with $n$ derivations $\to$ DOF $= 1$ (only source is independent)
\item $k$ API endpoints $\to$ DOF $= k$ (each endpoint independently defined)
\end{itemize}

\textbf{Capabilities:} Requirements the architecture satisfies (e.g., ``support horizontal scaling,'' ``provide type provenance,'' ``enable independent deployment'').

\textbf{Interpretation:} High leverage means gaining many capabilities from few DOF. Low leverage means paying many DOF for few capabilities.

\subsection{DOF-Reliability Isomorphism}

\textbf{Preview of Section~\ref{probability-model}.}
The formal results establish:
\begin{enumerate}
\item DOF-reliability isomorphism and round-trip identity (Theorem~\ref{thm:dof-reliability})
\item Exact ordering preservation for linear vs. series error models (Theorem~\ref{thm:approx-bound})
\item Quantitative leverage-gap identities for expected modification/error ratios (Theorem~\ref{thm:leverage-gap})
\item Physical edit-energy floor with explicit positive per-edit conversion constant (Theorem~\ref{thm:physical-energy-floor})
\item Budget feasibility boundary and explicit assumption-necessity witnesses (Theorem~\ref{thm:physical-budget-boundary}, Corollary~\ref{cor:physical-assumption-necessity})
\end{enumerate}

\subsection{Leverage Gap}

\textbf{Interpretation.}
Given equal capabilities, DOF reduction simultaneously improves expected error and expected modification cost by the same ratio; Section~\ref{probability-model} formalizes this identity.

\subsection{Building on Prior Results}

This paper does not \emph{subsume} Papers 1 and 2---it \emph{builds on} their epistemic foundations to provide an optimization criterion.

\textbf{Dependency chain:}
\begin{enumerate}
\item \textbf{Paper 1} proves axis orthogonality $\to$ enables error independence (Theorem~\ref{thm:error-independence})
\item \textbf{Paper 2} proves DOF = 1 guarantees coherence $\to$ establishes the coherence constraint
\item \textbf{This paper} provides the optimization criterion \emph{within} the coherence constraint
\end{enumerate}

Paper 2 supplies the coherence constraint (DOF = 1 per structural fact); this paper supplies the optimization rule inside that feasible set (maximize leverage). Paper 1 supplies axis orthogonality; this paper transports that structure into an explicit probabilistic ordering over architectures.

\subsection{Organization}

Section~\ref{foundations} defines Architecture, DOF, Capabilities, and Leverage. Section~\ref{probability-model} derives the error model from Paper 1's orthogonality and Paper 2's coherence theorems. Section~\ref{main-theorems} proves decidability and optimality. Section~\ref{instances} demonstrates integration with Papers 1 and 2. Section~\ref{appendix-lean} describes Lean mechanization.

\subsection{Scope and Limitations}

\textbf{What this paper provides:}
\begin{itemize}
\item Formal framework for comparing architectural alternatives
\item Provable connection between leverage and error probability
\item Decision procedure: maximize leverage subject to coherence constraint
\item Demonstration via before/after examples from production code
\end{itemize}

\textbf{Scope:}
Leverage characterizes the capability-to-DOF ratio. Performance, security, and other dimensions remain orthogonal concerns. The framework applies when requirements permit multiple architectural choices with different DOF. Error independence is \emph{derived} from Paper 1's axis orthogonality theorem, not assumed. Physical edit-energy claims use a single explicit calibration constant (`joulesPerEdit > 0`) and no additional architectural axioms.

\subsection{Roadmap}

Section 2 provides formal foundations (definitions). Section 3 derives the probability model from Papers 1 and 2. Section 4 proves main theorems. Section 5 presents instances (SSOT, typing, microservices, APIs, configuration, databases). Section 6 demonstrates practical application via before/after examples. Section 7 surveys related work. Section 8 concludes.

\subsection{Anticipated Objections}\label{sec:objection-summary}

Before proceeding, we address objections readers are likely forming. Each is refuted in detail in Appendix~\ref{appendix-rebuttals}; here we summarize the key points.

\paragraph{``Leverage is just a heuristic, not rigorous.''}
Leverage is \emph{formally defined} (Definition 1.4) and \emph{machine-checked} in Lean 4. Theorem 3.1 proves that maximizing leverage minimizes error probability. Current mechanization size is \LeanTotalLines{} lines with \LeanTotalTheorems{} theorems/lemmas and \LeanTotalSorry{} \texttt{sorry}.

\paragraph{``Different domains need different metrics.''}
The framework is domain-agnostic. We demonstrate instances across programming languages (SSOT, nominal typing), system architecture (microservices), API design (REST), configuration (convention vs explicit), and database design (normalization). The same principle ($L = |\text{Cap}|/\text{DOF}$) applies universally.

\paragraph{``Capabilities can't be quantified.''}
We don't need absolute quantification. Theorem 3.1 requires only \emph{relative ordering}: if capabilities are equal and DOF differs, the lower-DOF architecture has higher leverage. For architectures with different capabilities, we count cardinality.

\paragraph{``The independence assumption is unrealistic.''}
The model treats DOF-level errors as independent under the orthogonality assumptions used in the mechanization. Even with correlated errors (coefficient $\rho$), error probability remains monotone in DOF under standard positive-correlation models. Extension to explicit correlation models is future work.

\paragraph{``Performance matters more than error probability.''}
We agree. The framework addresses \emph{one dimension}: error probability. Multi-objective optimization (Pareto frontier over leverage, performance, security) is recommended for full architectural evaluation. For domains where correctness dominates (safety-critical, financial), leverage should be primary.

\paragraph{``The proofs are trivial.''}
The value is not in difficulty but in \emph{existence}. Machine-checked proofs provide precision (every step explicit), verification (computer-checked), and reproducibility (anyone can run them). ``Trivial'' proofs that compile are more valuable than ``deep'' proofs with errors.

\medskip
\noindent\textbf{If you have an objection not listed above,} check Appendix~\ref{appendix-rebuttals} (8 objections addressed) before concluding it has not been considered.
