\section{Instances}\label{instances}

We demonstrate that the leverage framework unifies prior results and applies to diverse architectural decisions.

\subsection{Instance 1: Single Source of Truth (SSOT)}

We previously formalized the DRY principle, proving that Python uniquely provides SSOT for structural facts via definition-time hooks and introspection. Here we show SSOT is an instance of leverage maximization.

\subsubsection{Prior Result}

\textbf{Published Theorem:} A language enables SSOT for structural facts if and only if it provides (1) definition-time hooks AND (2) introspectable derivation results. Python is the only mainstream language satisfying both requirements.

\textbf{Modification Complexity:} For structural fact $F$ with $n$ use sites:
\begin{itemize}
\item SSOT: $M(\text{change } F) = 1$ (modify source, derivations update automatically)
\item Non-SSOT: $M(\text{change } F) = n$ (modify each use site independently)
\end{itemize}

\subsubsection{Leverage Perspective}

\begin{definition}[SSOT Architecture]
Architecture $A_{\text{SSOT}}$ for structural fact $F$ has:
\begin{itemize}
\item Single source $S$ defining $F$
\item Derived use sites updated automatically from $S$
\item DOF $= 1$ (only $S$ is independently modifiable)
\end{itemize}
\end{definition}

\begin{definition}[Non-SSOT Architecture]
Architecture $A_{\text{non-SSOT}}$ for structural fact $F$ with $n$ use sites has:
\begin{itemize}
\item $n$ independent definitions (copied or manually synchronized)
\item DOF $= n$ (each definition independently modifiable)
\end{itemize}
\end{definition}

\begin{theorem}[SSOT Leverage Dominance]\label{thm:ssot-leverage}
For structural fact with $n$ use sites:
\[
\frac{L(A_{\text{SSOT}})}{L(A_{\text{non-SSOT}})} = n
\]
\end{theorem}
\leanmeta{\LH{L_48108f64},
\LH{L_3a9932b8},
\LH{L_2027e0f3}}

\begin{proof}
Both architectures provide same capabilities: $|\text{Cap}(A_{\text{SSOT}})| = |\text{Cap}(A_{\text{non-SSOT}})| = c$.

DOF:
\begin{align*}
\text{DOF}(A_{\text{SSOT}}) &= 1 \\
\text{DOF}(A_{\text{non-SSOT}}) &= n
\end{align*}

Leverage:
\begin{align*}
L(A_{\text{SSOT}}) &= c/1 = c \\
L(A_{\text{non-SSOT}}) &= c/n
\end{align*}

Ratio:
\[
\frac{L(A_{\text{SSOT}})}{L(A_{\text{non-SSOT}})} = \frac{c}{c/n} = n
\]
\end{proof}

\begin{corollary}[Unbounded Advantage]
As use sites grow ($n \to \infty$), leverage advantage grows unbounded.
\end{corollary}

\begin{corollary}[Error Probability]
For small $p$:
\[
\frac{P_{\text{error}}(A_{\text{non-SSOT}})}{P_{\text{error}}(A_{\text{SSOT}})} \approx n
\]
\end{corollary}

\textbf{Connection to Prior Work:} Our published Theorem 6.3 (Unbounded Complexity Gap) showed $M(\text{SSOT}) = O(1)$ vs $M(\text{non-SSOT}) = \Omega(n)$. Theorem \ref{thm:ssot-leverage} provides the leverage perspective: SSOT achieves $n$-times better leverage.

\subsection{Instance 2: Nominal Typing Dominance}

We previously proved nominal typing strictly dominates structural and duck typing for OO systems with inheritance. Here we show this is an instance of leverage maximization.

\subsubsection{Prior Result}

\textbf{Published Theorems:}
\begin{enumerate}
\item Theorem 3.13 (Provenance Impossibility): No shape discipline can compute provenance
\item Theorem 3.19 (Capability Gap): Gap = B-dependent queries = \{provenance, identity, enumeration, conflict resolution\}
\item Theorem 3.5 (Strict Dominance): Nominal strictly dominates duck typing
\end{enumerate}

\subsubsection{Leverage Perspective}

\begin{definition}[Typing Discipline as Architecture]
A typing discipline $D$ is an architecture where:
\begin{itemize}
\item Components = type checker, runtime dispatch, introspection APIs
\item Capabilities = queries answerable by the discipline
\end{itemize}
\end{definition}

\textbf{Duck Typing:} Uses only Shape axis ($S$: methods, attributes)
\begin{itemize}
\item Capabilities: Shape checking (``Does object have method $m$?'')
\item Cannot answer: provenance, identity, enumeration, conflict resolution
\end{itemize}

\textbf{Nominal Typing:} Uses Name + Bases + Shape axes ($N + B + S$)
\begin{itemize}
\item Capabilities: All duck capabilities PLUS 4 B-dependent capabilities
\item Can answer: ``Which type provided method $m$?'' (provenance), ``Is this exactly type $T$?'' (identity), ``List all subtypes of $T$'' (enumeration), ``Which method wins in diamond?'' (conflict)
\end{itemize}

\begin{theorem}[Nominal Leverage Dominance]\label{thm:nominal-leverage}
Assuming $\mathrm{DOF}(\mathrm{Nominal}) = \mathrm{DOF}(\mathrm{Duck}) = d$ (equal implementation cost hypothesis):
\[
L(\text{Nominal}) > L(\text{Duck})
\]
\end{theorem}
\leanmeta{\LH{L_4372537d},
\LH{L_d6782c96},
\LH{L_dbf653bd}}

\begin{proof}
Let $c_{\text{duck}} = |\text{Cap}(\text{Duck})|$ and $c_{\text{nominal}} = |\text{Cap}(\text{Nominal})|$.

By Theorem 3.19 (published): $c_{\text{nominal}} = c_{\text{duck}} + 4$.

By hypothesis: $\text{DOF}(\text{Nominal}) = \text{DOF}(\text{Duck}) = d$.

Therefore:
\[
L(\text{Nominal}) = \frac{c_{\text{duck}} + 4}{d} > \frac{c_{\text{duck}}}{d} = L(\text{Duck})
\]

\textbf{Status of hypothesis.} The equal-DOF assumption is formalized as an explicit premise in the Lean mechanization (\LH{L_4372537d}). Whether it holds in practice depends on the implementation. Abstractly: the typing discipline is one component; DOF is the number of independent change points in that component. The hypothesis says adding name-based dispatch (N + B axes) does not introduce more independent change points than shape-based dispatch (S axis alone).
\end{proof}

\textbf{Connection to Prior Work:} Our published Theorem 3.5 (Strict Dominance) showed nominal typing provides strictly more capabilities for same DOF cost. Theorem \ref{thm:nominal-leverage} provides the leverage formulation.

\subsection{Instance 3: Microservices Architecture}

Should a system use microservices or a monolith? How many services are optimal? The leverage framework provides answers. This architectural style, popularized by Fowler and Lewis \cite{fowler2014microservices}, traces its roots to the Unix philosophy of ``doing one thing well'' \cite{pike1984program}.

\subsubsection{Architecture Comparison}

\textbf{Monolith:}
\begin{itemize}
\item Components: Single deployment unit
\item DOF $= 1$
\item Capabilities: Basic functionality, simple deployment
\end{itemize}

\textbf{$n$ Microservices:}
\begin{itemize}
\item Components: $n$ independent services
\item DOF $= n$ (each service independently deployable/modifiable)
\item Additional Capabilities: Independent scaling, independent deployment, fault isolation, team autonomy, polyglot persistence \cite{fowler2014microservices}
\end{itemize}

\subsubsection{Leverage Analysis}

Let $c_0$ = capabilities provided by monolith.

Let $\Delta c = 5$ denote the additional capabilities from microservices:
independent scaling, independent deployment, fault isolation, team autonomy,
and polyglot persistence.

\textbf{Leverage:}
\begin{align*}
L(\text{Monolith}) &= c_0 / 1 = c_0 \\
L(n \text{ Microservices}) &= (c_0 + \Delta c) / n = (c_0 + 5) / n
\end{align*}

\textbf{Break-even Point:}
\[
L(\text{Microservices}) \geq L(\text{Monolith}) \iff \frac{c_0 + 5}{n} \geq c_0 \iff n \leq 1 + \frac{5}{c_0}
\]

\textbf{Interpretation:} If base capabilities $c_0 = 5$, then $n \leq 2$ services is optimal. For $c_0 = 20$, up to $n = 1.25$ (i.e., monolith still better). Microservices justified only when additional capabilities significantly outweigh DOF cost.

\subsection{Instance 4: REST API Design}

Generic endpoints vs specific endpoints: a leverage tradeoff.

\subsubsection{Architecture Comparison}

\textbf{Specific Endpoints:} One endpoint per use case
\begin{itemize}
\item Example: \texttt{GET /users}, \texttt{GET /posts}, \texttt{GET /comments}, ...
\item For $n$ use cases: DOF $= n$
\item Capabilities: Serve $n$ use cases
\end{itemize}

\textbf{Generic Endpoint:} Single parameterized endpoint
\begin{itemize}
\item Example: \texttt{GET /resources/:type/:id}
\item DOF $= 1$
\item Capabilities: Serve $n$ use cases (same as specific)
\end{itemize}

\subsubsection{Leverage Analysis}

\begin{align*}
L(\text{Generic}) &= n / 1 = n \\
L(\text{Specific}) &= n / n = 1
\end{align*}

\textbf{Advantage:} $L(\text{Generic}) / L(\text{Specific}) = n$

\textbf{Tradeoff:} Generic endpoint has higher leverage but may sacrifice:
\begin{itemize}
\item Type safety (dynamic routing)
\item Specific validation per resource
\item Tailored response formats
\end{itemize}

\textbf{Decision Rule:} Use generic if $n > k$ where $k$ is complexity threshold (typically $k \approx 3$--$5$).

\subsection{Instance 5: Configuration Systems}

Convention over configuration (CoC) reduces developer decisions and preserves
flexibility \cite{hansson2005rails}. In this framework it is leverage
maximization via defaults.

\subsubsection{Architecture Comparison}

\textbf{Explicit Configuration:} Must set all $m$ parameters
\begin{itemize}
\item DOF $= m$ (each parameter independently set)
\item Capabilities: Configure $m$ aspects
\end{itemize}

\textbf{Convention over Configuration:} Provide defaults, override only $k$ parameters
\begin{itemize}
\item DOF $= k$ where $k \ll m$
\item Capabilities: Configure same $m$ aspects (defaults handle rest)
\end{itemize}

\textbf{Example (Rails vs Java EE):}
\begin{itemize}
\item Rails: 5 config parameters (convention for rest)
\item Java EE: 50 config parameters (explicit for all)
\end{itemize}

\subsubsection{Leverage Analysis}

\begin{align*}
L(\text{Convention}) &= m / k \\
L(\text{Explicit}) &= m / m = 1
\end{align*}

\textbf{Advantage:} $L(\text{Convention}) / L(\text{Explicit}) = m/k$

For Rails example: $m/k = 50/5 = 10$ (10$\times$ leverage improvement).

\subsection{Instance 6: Database Schema Normalization}

Normalization eliminates redundancy, maximizing leverage. This concept, introduced by Codd \cite{codd1970relational}, is the foundation of relational database design \cite{date2003introduction}.

\subsubsection{Architecture Comparison}

Consider customer address stored in database:

\textbf{Denormalized (Address in 3 tables):}
\begin{itemize}
\item \texttt{Users} table: address columns
\item \texttt{Orders} table: shipping address columns
\item \texttt{Invoices} table: billing address columns
\item DOF $= 3$ (address stored 3 times)
\end{itemize}

\textbf{Normalized (Address in 1 table):}
\begin{itemize}
\item \texttt{Addresses} table: single source
\item Foreign keys from \texttt{Users}, \texttt{Orders}, \texttt{Invoices}
\item DOF $= 1$ (address stored once)
\end{itemize}

\subsubsection{Leverage Analysis}

Both provide same capability: store/retrieve addresses.

\begin{align*}
L(\text{Normalized}) &= c / 1 = c \\
L(\text{Denormalized}) &= c / 3
\end{align*}

\textbf{Advantage:} $L(\text{Normalized}) / L(\text{Denormalized}) = 3$

\textbf{Modification Complexity:}
\begin{itemize}
\item Change address format: Normalized $M = 1$, Denormalized $M = 3$
\item Error probability: $P_{\text{denorm}} = 3p$ vs $P_{\text{norm}} = p$
\end{itemize}

\textbf{Tradeoff:} Normalization increases leverage but may sacrifice query performance (joins required).

\subsection{Summary of Instances}

\begin{center}
\begin{tabular}{lccc}
\toprule
\textbf{Instance} & \textbf{High Leverage} & \textbf{Low Leverage} & \textbf{Ratio} \\
\midrule
SSOT & DOF = 1 & DOF = $n$ & $n$ \\
Nominal Typing & $c+4$ caps, DOF $d$ & $c$ caps, DOF $d$ & $(c+4)/c$ \\
Microservices & Monolith (DOF = 1) & $n$ services (DOF = $n$) & $n/(c_0+5)$ \\
REST API & Generic (DOF = 1) & Specific (DOF = $n$) & $n$ \\
Configuration & Convention (DOF = $k$) & Explicit (DOF = $m$) & $m/k$ \\
Database & Normalized (DOF = 1) & Denormalized (DOF = $n$) & $n$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Pattern:} High leverage architectures achieve $n$-fold improvement where $n$ is the consolidation factor (use sites, services, endpoints, parameters, or redundant storage).
