\section{Preemptive Rebuttals}\label{appendix-rebuttals}

We address anticipated objections.

\subsection{Objection 1: ``Leverage is just a heuristic, not rigorous''}

\textbf{Response:} Leverage is \emph{formally defined} (Definition 1.4) and \emph{machine-checked} in Lean 4. Theorem 3.1 \emph{proves} (not assumes) that maximizing leverage minimizes error probability. This is mathematics, not heuristics.

\textbf{Evidence:} \LeanTotalLines{} lines, \LeanTotalTheorems{} theorems/lemmas, \LeanTotalSorry{} \texttt{sorry}. Claim-to-proof links are generated in the auto claim coverage matrix.

\subsection{Objection 2: ``Different domains need different metrics''}

\textbf{Response:} The framework is \emph{domain-agnostic}. We prove this by demonstrating instances across:
\begin{itemize}
\item Programming languages (SSOT, nominal typing)
\item System architecture (microservices)
\item API design (REST endpoints)
\item Configuration (convention vs explicit)
\item Database design (normalization)
\end{itemize}

The same principle (maximize $L = |\text{Cap}|/\text{DOF}$) applies universally.

\subsection{Objection 3: ``Capabilities can't be quantified''}

\textbf{Response:} We \emph{don't need absolute quantification}. Theorem 3.1 requires only \emph{relative ordering}: if $\text{Cap}(A_1) = \text{Cap}(A_2)$ and $\text{DOF}(A_1) < \text{DOF}(A_2)$, then $L(A_1) > L(A_2)$.

For architectures with \emph{different} capabilities, we count cardinality. This suffices for comparing alternatives (e.g., nominal vs duck: nominal has 4 additional capabilities).

\subsection{Objection 4: ``SSOT is only relevant for Python''}

\textbf{Response:} SSOT is \emph{implementable} in any language with definition-time hooks and introspection. Our prior work proved Python uniquely provides \emph{both} among mainstream languages, but:
\begin{itemize}
\item Common Lisp (CLOS) provides SSOT
\item Smalltalk provides SSOT
\item Future languages could provide SSOT
\end{itemize}

The \emph{principle} (leverage via SSOT) is universal. The \emph{implementation} depends on language features.

\subsection{Objection 5: ``Independence assumption is unrealistic''}

\textbf{Response:} In the mechanization, independence is derived from orthogonality assumptions on axis structure (not inserted as an untracked theorem-level axiom). Real systems may still exhibit additional correlated failure modes outside this model.

\textbf{Mitigation:} Even with correlation, DOF remains relevant. If correlation coefficient is $\rho$, then:
\[
P_{\text{error}}(n) \approx n \cdot p \cdot (1 + (n-1)\rho)
\]

Still monotonically increasing in $n$. High-leverage architectures still preferable.

\textbf{Future work:} Extend framework to correlated errors (Section 8.3).

\subsection{Objection 6: ``Performance matters more than error probability''}

\textbf{Response:} We \emph{agree}. Performance, security, and other quality attributes matter. Our framework addresses \emph{one dimension}: error probability.

\textbf{Recommended approach:} Multi-objective optimization (Future Work, Section 8.3). Compute Pareto frontier over (leverage, performance, security).

For domains where correctness dominates (safety-critical systems, financial software), leverage should be primary criterion.

\subsection{Objection 7: ``Case studies are cherry-picked''}

\textbf{Response:} The instances presented (SSOT, nominal typing, microservices, APIs, configuration, databases) demonstrate the framework's domain-agnostic applicability. Each instance is derived mathematically from the leverage definition, not selected based on favorable results.

PR \#44 in OpenHCS provides a publicly verifiable example of DOF collapse in practice. The theoretical framework stands independently of any specific codebase.

\subsection{Objection 8: ``The Lean proofs are trivial''}

\textbf{Objection:} ``The Lean proofs just formalize obvious definitions. There's no deep mathematics here.''

\textbf{Response:} The value is not in the difficulty of the proofs but in their \emph{existence}. Machine-checked proofs provide:

\begin{enumerate}
\item \textbf{Precision:} Informal arguments can be vague. Lean requires every step to be explicit.
\item \textbf{Verification:} The proofs are checked by a computer. Human error is eliminated.
\item \textbf{Reproducibility:} Anyone can run the proofs and verify the results.
\end{enumerate}

``Trivial'' proofs that compile are infinitely more valuable than ``deep'' proofs that contain errors. Every theorem in this paper has been validated by the Lean type checker.
