\begin{abstract}
We study zero-error class identification under constrained observations with three resources: tag rate $L$ (bits per entity), identification cost $W$ (attribute queries), and distortion $D$ (misidentification probability). If the attribute-profile map $\pi$ is not injective on classes, then attribute-only observation cannot identify class identity with zero error. Let $A_\pi := \max_u |\{c : \pi(c)=u\}|$ be collision multiplicity. Any $D=0$ scheme must satisfy $L \ge \log_2 A_\pi$, and this bound is tight. In maximal-barrier domains ($A_\pi=k$), the nominal point $(L,W,D)=(\lceil\log_2 k\rceil,O(1),0)$ is the unique Pareto-optimal zero-error point, so zero-error identification forces explicit naming information. Without tags ($L=0$), zero-error identification requires $W=\Omega(d)$ queries, where $d$ is the distinguishing dimension. Minimal sufficient query sets form the bases of a matroid, making $d$ well-defined as an invariant of the observation structure rather than of a particular query strategy. We also prove exact finite-block confusability laws: at blocklength $t$, zero-error feasibility is equivalent to a tag alphabet threshold $A_\pi^t$, yielding exact linear log-bit scaling. These results turn a familiar design heuristic into an explicit resource tradeoff: once classes collide under the available observations, zero-error identification requires either additional naming information or additional query cost. We illustrate the framework on databases, biology, typed software systems, and model registries, and machine-check the main claims in Lean 4 (\LeanTotalLines\ lines, \LeanTotalTheorems\ theorem/lemma statements, \LeanTotalSorry\ \texttt{sorry}).

\textbf{Keywords:} rate-distortion theory, zero-error identification, query complexity, matroid structure, classification systems
\end{abstract}
