\section{Static Regime Complexity}\label{sec:hardness}

This section studies the static regime, where the input is a finite coordinate-structured decision problem under the succinct encoding of Section~\ref{sec:encoding}. The core question is whether a candidate coordinate set preserves the optimizer map.

\subsection{Problem Definitions}

\begin{problem}[SUFFICIENCY-CHECK]
\textbf{Input:} Decision problem $\mathcal{D}=(A,X_1,\ldots,X_n,U)$ and coordinate set $I \subseteq \{1,\ldots,n\}$. \\
\textbf{Question:} Is $I$ sufficient for $\mathcal{D}$?
\end{problem}

\begin{problem}[MINIMUM-SUFFICIENT-SET]
\textbf{Input:} Decision problem $\mathcal{D}=(A,X_1,\ldots,X_n,U)$ and integer $k$. \\
\textbf{Question:} Does there exist a sufficient set $I$ with $|I| \le k$?
\end{problem}

\begin{problem}[ANCHOR-SUFFICIENCY]
\textbf{Input:} Decision problem $\mathcal{D}=(A,X_1,\ldots,X_n,U)$ and fixed coordinate set $I$. \\
\textbf{Question:} Does there exist an assignment $\alpha$ to $I$ such that $\Opt(s)$ is constant on the fiber $\{s : s_I=\alpha\}$?
\end{problem}

\subsection{coNP-Completeness Results}

\begin{theorem}[SUFFICIENCY-CHECK is coNP-complete]\label{thm:sufficiency-conp}
SUFFICIENCY-CHECK is coNP-complete.
\end{theorem}

\begin{proof}
Membership: by Proposition~\ref{prop:insufficiency-counterexample}, insufficiency has a polynomial witness $(s,s')$ with $s_I=s'_I$ and $\Opt(s)\ne\Opt(s')$, so the complement is in NP.

Hardness: reduce TAUTOLOGY to the empty-set instance. Given formula $\varphi(x_1,\ldots,x_n)$, build a two-action decision problem where one action tracks $\varphi$ and the other is a baseline action. Then $I=\emptyset$ is sufficient iff $\varphi$ is a tautology, because empty-set sufficiency is exactly constancy of $\Opt$ (Proposition~\ref{prop:empty-sufficient-constant}). The reduction is polynomial in formula size.
\end{proof}

\begin{theorem}[MINIMUM-SUFFICIENT-SET is coNP-complete]\label{thm:minsuff-conp}
MINIMUM-SUFFICIENT-SET is coNP-complete.
\end{theorem}

\begin{proof}
By Proposition~\ref{prop:minimal-relevant-equiv}, the minimum sufficient set is exactly the relevant-coordinate set. Hence the question ``$|I^*|\le k$?'' is equivalent to a coNP check over relevance witnesses. CoNP-hardness follows from the $k=0$ slice, which is exactly SUFFICIENCY-CHECK with $I=\emptyset$.
\end{proof}

\subsection{$\Sigma_2^P$-Completeness}

\begin{theorem}[ANCHOR-SUFFICIENCY is $\Sigma_2^P$-complete]\label{thm:anchor-sigma2p}
ANCHOR-SUFFICIENCY is $\Sigma_2^P$-complete.
\end{theorem}

\begin{proof}
Membership is immediate from the quantifier pattern:
\[
\exists\alpha\;\forall s\in S:\ (s_I=\alpha)\Rightarrow \Opt(s)=\Opt(s_\alpha).
\]
Hardness follows by reduction from $\exists\forall$-SAT: existential variables encode the anchor assignment, and universal variables range over the residual coordinates. Utility values are set so that fiberwise constancy holds exactly when the source formula is true.
\end{proof}

\subsection{Certificate-Size Lower Bound}

\begin{theorem}[Witness-Checking Duality]\label{thm:witness-checking-duality}
For Boolean state spaces $S=\{0,1\}^n$, any sound checker for the empty-set sufficiency core must inspect at least $2^{n-1}$ witness pairs in the worst case.
\end{theorem}

\begin{proof}
Empty-set sufficiency asks whether $\Opt$ is constant on all $2^n$ states. A sound refutation-complete checker must be able to separate constant maps from maps that differ on some hidden antipodal partition. An adversary can place the first disagreement in any of $2^{n-1}$ independent pair slots; if fewer than $2^{n-1}$ slots are inspected, two instances remain indistinguishable to the checker but have opposite truth values. Therefore at least $2^{n-1}$ pair checks are necessary.
\end{proof}

\paragraph{Mechanization note.}
The TAUTOLOGY reduction stack, the coNP and $\Sigma_2^P$ classifications, and the witness-budget lower-bound core are formalized in Lean 4 and indexed in Appendix~\ref{app:lean}.
