\section{Introduction}\label{sec:introduction}

Consider any decision problem $\mathcal{D}$ with actions $A$ and factored state space $S = X_1 \times \cdots \times X_n$. A coordinate set $I \subseteq \{1, \ldots, n\}$ is \emph{sufficient} if knowing only coordinates in $I$ determines the optimal action:
\[
s_I = s'_I \implies \Opt(s) = \Opt(s')
\]
This is \emph{coordinate sufficiency}---the formalization of \emph{information sufficiency} (whether available information determines the optimal action) over a factored state space.

Paper 4 established that SUFFICIENCY-CHECK is \coNP-complete for static decision problems $(A, S, U)$. This paper extends that landscape to two natural extensions: stochastic and sequential regimes.

\subsection{Stochastic Regime}

When the state is drawn from a distribution $P : \Delta(S)$, sufficiency must account for randomness. We use the distribution-conditional definition: $I$ is stochastically sufficient if $P(s \mid s_I) \otimes U$ determines the same optimal action as $P(s) \otimes U$. This subsumes expected-value sufficiency (where only the first moment matters) and connects to classical sufficient statistics.

\begin{center}
\begin{tabularx}{\linewidth}{@{}llX@{}}
\toprule
\textbf{Regime} & \textbf{Tuple} & \textbf{Complexity} \\
\midrule
Static & $(A, S, U)$ & \coNP-complete (Paper 4) \\
Stochastic & $(A, S, U, P)$ & \PP-complete \\
Sequential & $(A, S, U, T, O)$ & \PSPACE-complete \\
\bottomrule
\end{tabularx}
\end{center}

The \PP upper bound comes from counting satisfying assignments weighted by $P$. The lower bound uses a MAJSAT reduction: given $\varphi$, construct uniform $P$ over $\{0,1\}^n$ and let $U(\text{accept}, s) = \varphi(s)$, $U(\text{reject}, s) = 1/2$. Then $\emptyset$ is stochastically sufficient iff $\mathbb{E}_P[\varphi(s)] \geq 1/2$ or $< 1/2$---i.e., majority of assignments satisfy $\varphi$.

Tractable subcases: product distributions (expected utility decomposes), bounded support ($|\text{supp}(P)| \leq k$), log-concave distributions.

\subsection{Sequential Regime}

When decisions unfold over time with transitions $T : A \times S \to \Delta(S)$ and observations $O : S \to \Delta(\Omega)$, the problem becomes a POMDP. Sequential sufficiency asks whether observation history restricted to $I$-coordinates determines the optimal policy.

The \PSPACE upper bound follows from Papadimitriou \& Tsitsiklis (1987). The lower bound uses TQBF reduction: given $\forall x_1 \exists x_2 \forall x_3 \ldots \varphi$, construct $n$ time steps, adversarial transitions at universal quantifiers, agent choices at existential quantifiers, terminal utility $\varphi$. Then full observation history is sequentially sufficient iff the QBF is true.

Tractable subcases: fully observable (MDP), bounded horizon, tree-structured transitions, deterministic transitions.

\subsection{Substrate Cost}

We extend Paper 4's integrity-competence matrix with substrate cost $\kappa(\text{cell}, \text{agentType})$. The matrix verdict is substrate-independent: $\text{verdict}(c)$ is the same for all agent types. But trajectories diverge: $\text{trajectory}(\sigma, \tau_1) \neq \text{trajectory}(\sigma, \tau_2)$ in general. This is the minimal formal hook for substrate in decision theory.

\subsection{Temporal Learning}

An agent facing sequential problem instances drawn from unknown distribution $D$ can learn the structure class via Bayesian updates. As $k \to \infty$, posterior $P(C \mid I_1, \ldots, I_k) \to \delta_{C*}$ if structure classes are identifiable. This reduces the abstention frontier: $A_1 \supseteq A_2 \supseteq \ldots$.

\subsection{Temporal Integrity}

Claims across sequences must satisfy evidence-monotone retractions: an agent retracts a claim only when new evidence contradicts it. Retraction without evidence violates temporal integrity; refinement ($I' \subset I$) is allowed.

\subsection{Main Theorems}

\begin{enumerate}
\item \textbf{Theorem~\ref{th:stochastic-sufficiency-pp}:} STOCHASTIC-SUFFICIENCY-CHECK is \PP-complete via MAJSAT reduction.

\item \textbf{Theorem~\ref{th:sequential-sufficiency-pspace}:} SEQUENTIAL-SUFFICIENCY-CHECK is \PSPACE-complete via TQBF reduction.

\item \textbf{Propositions~\ref{prop:static-stochastic-strict} and \ref{prop:stochastic-sequential-strict} (Regime Hierarchy):} Static $\subset$ Stochastic $\subset$ Sequential; \coNP $\subset$ \PP $\subset$ \PSPACE.

\item \textbf{Theorem~\ref{th:substrate-independence-verdict} (Substrate Independence of Verdict):} For all cells $c$, $\text{verdict}(c)$ is $\kappa$-independent.

\item \textbf{Theorem~\ref{th:substrate-dependence-trajectory} (Substrate Dependence of Trajectory):} There exist $\tau_1, \tau_2, \sigma$ with $\text{trajectory}(\sigma, \tau_1) \neq \text{trajectory}(\sigma, \tau_2)$.
\end{enumerate}

\leanmeta{\LH{CC58},
\LH{CC48},
\LH{CC64},
\LH{CC63}}

\subsection{Claim-Typing Bridges (Inherited Static Core)}

\begin{enumerate}
\item \textbf{Typed claim admissibility (inherited static core):} report admissibility is typed by certificate class (\(\mathsf{ABSTAIN}\), \(\mathsf{EXACT}\), \(\mathsf{EPSILON}(\varepsilon)\)).
\item \textbf{Evidence-admissibility equivalence (inherited static core):} first-class evidence-object existence is equivalent to typed report admissibility.
\item \textbf{Empty-sufficient characterization (inherited static core):} the \(I=\emptyset\) query is exactly universal constancy of the decision boundary, and non-sufficiency is exactly existence of a counterexample pair.
\item \textbf{Exact requires evidence (inherited static core):} exact admissibility is equivalent to exact evidence existence.
\item \textbf{No evidence implies zero certified confidence (inherited static core):} under signal consistency, no evidence implies zero certified confidence.
\end{enumerate}
\claimstamp{P4-core typed-claim chain}{DC,AR}
\leanmeta{\LH{CC75},\LH{IC7},\LH{DP2},\LH{DP5},\LH{IC9},\LH{IC17}}

\subsection{Paper Structure}

Section~\ref{sec:stochastic}: Stochastic decision problems, \PP-completeness, tractable cases.
Section~\ref{sec:sequential}: Sequential (POMDP) problems, \PSPACE-completeness, tractable cases.
Section~\ref{sec:regime-hierarchy}: Strict hierarchy of regimes.
Section~\ref{sec:substrate-cost}: Substrate cost formalization and independence results.
Section~\ref{sec:temporal-learning}: Bayesian structure detection, abstention reduction.
Section~\ref{sec:temporal-integrity}: Temporal integrity constraints.
Section~\ref{sec:cross-regime-transfer}: Transfer conditions between regimes.
Section~\ref{sec:related}: Related work.
