\section{Conclusion}

This paper studies the convergence structure of decision-relevant information. Starting from shared definitions of sufficiency and relevance, it identifies the optimizer quotient as the canonical decision-preserving abstraction and shows that multiple mathematical frameworks recover the same structural core.

\subsection*{What Converges}

The common invariant is structural rank: the size of the relevant-coordinate support. Quotient entropy, Fisher-style support counting, and lossless zero-distortion summaries recover that same support-size quantity once the optimizer quotient is fixed. The significance is not that these frameworks become identical, but that they agree on what counts as the dimension of decision relevance.

\subsection*{Bayesian Consequence}

The Bayesian argument is equally direct. Counting on finite sets yields probability after normalization; $\log x \le x - 1$ yields Gibbs' inequality; Gibbs' inequality yields cross-entropy minimization; and Bayes emerges as the unique optimizer under log loss. This places the Bayesian conclusions on the same foundational footing as the quotient and entropy results.

\subsection*{Contribution}

The contribution is to place quotient structure, structural-rank recovery, and Bayesian optimality in a single mathematical framework.

\subsection*{Outlook}

Natural next questions are refinement questions rather than boundary questions: which additional frameworks recover the same invariant, how much of the quotient story can be internalized categorically, and which convergence statements admit the cleanest mechanized forms in Lean.
