\section{Temporal Integrity}\label{sec:temporal-integrity}

We formalize consistency of claims across sequences of problem instances.

\subsection{Definition}

\begin{definition}[Temporal Integrity]\label{def:temporal-integrity}
Temporal integrity = consistency of claims across a sequence of instances.

A physical controller has temporal integrity if:
\begin{itemize}
\item Each per-instance claim satisfies Paper 4's integrity constraint
\item Retractions are evidence-monotone: the controller retracts a claim only when new evidence contradicts it, never without cause
\item Refinements are monotone: if the controller claims $I$ is sufficient at time $t$, and later claims $I' \subset I$ is sufficient at time $t' > t$, this is a refinement (allowed). If it claims $I' \supset I$, this is a weakening that requires justification.
\end{itemize}
\end{definition}

\begin{decision}\label{dec:temporal-integrity-check}
\textbf{TEMPORAL-INTEGRITY-CHECK}: Given a sequence of claims and evidence, is the sequence temporally integrity-preserving?
\end{decision}

\begin{claim}{Retraction with evidence preserves integrity}{prop:retraction-evidence-integrity}
If a controller claimed $I$ sufficient, observed a counterexample, then retracted, this preserves temporal integrity.
\claimstamp{P\ref{prop:retraction-evidence-integrity}}{RG}
\leanmeta{\LH{CC70}.}
\end{claim}

\leanmeta{\LH{CC70}}

\begin{claim}{Retraction without evidence violates integrity}{prop:retraction-no-evidence-violates}
If a controller claimed $I$ sufficient, observed no new information, then retracted, this violates temporal integrity.
\claimstamp{P\ref{prop:retraction-no-evidence-violates}}{RG}
\leanmeta{\LH{CC71}.}
\end{claim}

\leanmeta{\LH{CC71}}

\begin{claim}{Refinement strengthens integrity}{prop:refinement-strengthens}
If a controller claimed $I$ sufficient, discovers $I' \subset I$ also sufficient, updates claim to $I'$. This preserves and strengthens temporal integrity.
\claimstamp{P\ref{prop:refinement-strengthens}}{RG}
\leanmeta{\LH{CC69}.}
\end{claim}

\leanmeta{\LH{CC69}}

\subsection{Complexity of Temporal Integrity}

\begin{claim}{Complexity lower bound}{th:temporal-integrity-complexity}
TEMPORAL-INTEGRITY-CHECK is at least as hard as the base regime's SUFFICIENCY-CHECK.
\claimstamp{T\ref{th:temporal-integrity-complexity}}{RG}
\end{claim}

Each claim in the sequence requires verification. The temporal constraint (monotonicity) adds structure but doesn't reduce worst-case complexity.

Paper 4 proves integrity forces abstention in hardness-blocked regimes. Paper 5 proves that credibility requires costly signals. Together: abstention is structurally required when certification is unavailable, and credibility is recovered only through verifiable costly evidence traces.
