\section{Conservation Law for Decision-Relevant Coordinates}\label{sec:simplicity-tax}

Formally: this section defines expressive-gap tax and proves redistribution, growth, and amortization identities under fixed required-coordinate sets.

The load-bearing fact in this section is not the set identity itself; it is the difficulty of shrinking the required set $R(P)$ in the first place. By Theorem~\ref{thm:sufficiency-conp} (and Theorem~\ref{thm:minsuff-conp} for minimization), exact relevance identification is intractable in the worst case under succinct encoding. The identities below therefore quantify how unresolved relevance is redistributed between central and per-site work.\claimstamp{T\ref{thm:sufficiency-conp},\ref{thm:minsuff-conp}}{S}
For encoded physical systems, the same redistribution laws apply under physical-to-core transport: unresolved coordinates remain per-site external work, while resolved coordinates are paid once in centralized structure.\claimstamp{P\ref{prop:physical-claim-transport};T\ref{thm:tax-conservation},\ref{thm:tax-grows},\ref{thm:amortization}}{AR,LC}
\leanmeta{\LH{CT6},\LH{HD5},\LH{HD26},\LH{HD21},\LH{HD4}.}

\begin{definition}
Let $R(P)$ be the required dimensions (those affecting $\Opt$) and $A(M)$ the dimensions model $M$ handles natively. The \emph{expressive gap} is $\text{Gap}(M,P) = R(P) \setminus A(M)$.
\end{definition}

\begin{definition}[Simplicity Tax]\label{def:simplicity-tax}
The \emph{simplicity tax} is the size of the expressive gap:
\[
\text{SimplicityTax}(M,P) := |\text{Gap}(M,P)|.
\]
\end{definition}

\begin{theorem}[Redistribution Identity]\label{thm:tax-conservation}
$|\text{Gap}(M, P)| + |R(P) \cap A(M)| = |R(P)|$. The total cannot be reduced, only redistributed between ``handled natively'' and ``handled externally.''
\leanmeta{\LH{HD5}}
\end{theorem}

\begin{proof}
In the finite-coordinate model this is the exact set-cardinality identity
\[
|R\setminus A| + |R\cap A| = |R|,
\]
formalized as \LH{HD5}.
\end{proof}

\begin{remark}[Why this is nontrivial in context]
The algebraic identity in Theorem~\ref{thm:tax-conservation} is elementary. Its force comes from upstream hardness: reducing $|R(P)|$ by exact relevance minimization is worst-case intractable under the succinct encoding, so redistribution is often the only tractable lever available.
\end{remark}

\begin{theorem}[Linear Growth]\label{thm:tax-grows}
For $n$ decision sites:
\[
\text{TotalExternalWork} = n \times \text{SimplicityTax}(M, P).
\]
\leanmeta{\LH{HD26}}
\end{theorem}

\begin{proof}
This is by definition of per-site externalization and is mechanized as
\LH{HD26}.
\end{proof}

\begin{theorem}[Amortization]\label{thm:amortization}
Let $H_{\text{central}}$ be the one-time cost of using a complete model. There exists
\[
n^* = \left\lfloor \frac{H_{\text{central}}}{\text{SimplicityTax}(M,P)} \right\rfloor
\]
such that for $n > n^*$, the complete model has lower total cost.
\leanmeta{\LH{HD4}}
\end{theorem}

\begin{proof}
For positive per-site tax, the threshold inequality
\[
n > \left\lfloor \frac{H_{\text{central}}}{\text{SimplicityTax}} \right\rfloor
\implies
H_{\text{central}} < n\cdot \text{SimplicityTax}
\]
is mechanized as
\LH{HD4}.
\end{proof}

\begin{corollary}[Gap Externalization]\label{cor:gap-externalization}
\claimstamp{C\ref{cor:gap-externalization}}{H=cost-growth}
If $\text{Gap}(M,P)\neq\emptyset$, then external handling cost scales linearly with the number of decision sites.
\leanmeta{\LH{HD26},
\LH{HD21}}
\end{corollary}

\begin{proof}
The exact linear form is
\LH{HD26}.
When the gap is nonempty (positive tax), monotone growth with $n$ is
\LH{HD21}.
\end{proof}

\begin{corollary}[Exact Minimization Criterion]\label{cor:gap-minimization-hard}
\claimstamp{C\ref{cor:gap-minimization-hard}}{H=succinct-hard}
For mechanized Boolean-coordinate instances, ``there exists a sufficient set of size at most $k$'' is equivalent to ``the relevant-coordinate set has cardinality at most $k$.''
\leanmeta{\LH{S2P2}}
\end{corollary}

\begin{proof}
This is \LH{S2P2}.
\end{proof}

Informally: when exact minimization is hard, unresolved relevance shifts between centralized structure and repeated local work.

Appendix~\ref{app:lean} provides theorem statements and module paths for the corresponding Lean formalization.
