\section{Conclusion}

Formally: this section composes proved regime placements, bridge theorems, and claim-typing constraints into final landscape consequences.

This paper gives a regime-first complexity map: static/stochastic/sequential placements with strict hierarchy (\LH{DC9}, \LH{DC10}, \LH{DC1}, \LH{DC2}), static inner-class decomposition by decision/access/structure layers (\LH{CCC6}, \LH{CCC5}, \LH{CCC1}, \LH{CCC3}, \LH{CCC7}, \LH{CC61}), theorem-level physical transport plus exact represented-adjacent bridge boundary (\LH{CT4}, \LH{CC67}, \LH{CC69}, \LH{CC85}), Bayesian uniqueness under DQ/thermodynamic admissibility axioms (\LH{FN14}), physical-collapse no-go transfer chain (\LH{PH11}, \LH{PH12}, \LH{PH13}, \LH{PH14}, \LH{PH15}), derived SAT-bridge map plus reduction-energy transfer and disjunctive assumption boundary (\LH{PH16}, \LH{PH17}, \LH{PH18}, \LH{PH19}, \LH{PH20}, \LH{PH21}, \LH{PH22}, \LH{PH26}, \LH{PH27}), explicit finite-budget-necessity countermodels (\LH{PH31}, \LH{PH32}, \LH{PH33}), regime-robust deterministic/probabilistic/sequential no-collapse specializations (\LH{PH28}, \LH{PH29}, \LH{PH30}), observer-collapse anchor closure across stochastic/sequential checks (\LH{PA1}, \LH{PA2}, \LH{PA3}, \LH{PA4}, \LH{PA5}, \LH{PA6}, \LH{PA7}, \LH{PA8}, \LH{PA9}), recurrent conversation-layer clamp/report bridge (\LH{CV4}, \LH{CV7}, \LH{CV17}, \LH{CV18}, \LH{CV13}, \LH{CV16}), explicit Neukart--Vinokur duality specialization from the thermodynamic lift (\LH{CC77}, \LH{CC76}), and formal centralized/distributed hardness thresholds through dominance and amortization theorems (\LH{HD1}, \LH{HD4}, \LH{HD3}).
\claimstamp{T\ref{thm:sufficiency-conp},\ref{thm:minsuff-conp},\ref{thm:anchor-sigma2p},\ref{th:stochastic-sufficiency-pp},\ref{th:sequential-sufficiency-pspace},\ref{thm:dichotomy},\ref{thm:tractable},\ref{thm:physical-bridge-bundle},\ref{thm:bridge-boundary-represented},\ref{thm:regime-coverage},\ref{thm:typed-completeness-static},\ref{thm:bayes-from-dq},\ref{thm:centralization-dominance},\ref{thm:tax-conservation},\ref{thm:amortization};P\ref{prop:query-regime-obstruction},\ref{prop:thermo-lift},\ref{prop:static-stochastic-strict},\ref{prop:stochastic-sequential-strict};C\ref{cor:neukart-vinokur}}{E,S,S+ETH,Qf,Qb,AR,RG,LC,H=cost-growth}

Informally: this conclusion combines previously proved blocks under explicit tags.

\subsection*{Methodology and Disclosure}

\textbf{Role of LLMs in this work.} This paper was developed through human-AI collaboration. The author provided the core intuitions: the connection between decision-relevance and computational complexity, the conjecture that SUFFICIENCY-CHECK is \coNP-complete, and the insight that the $\SigmaP{2}$ structure collapses for MINIMUM-SUFFICIENT-SET. Large language models (Claude, GPT-4) served as implementation partners for proof drafting, Lean formalization, and \LaTeX{} generation.

The Lean 4 proofs were iteratively refined: the author specified the target statements, the LLM proposed proof strategies, and the Lean compiler served as the arbiter of correctness. For complexity-theoretic reductions, the Lean formalization machine-verifies \emph{correctness} (membership preservation); polynomial-time computability of the reduction functions is a standard claim stated as an explicit hypothesis, consistent with the state of formal verification where end-to-end Turing-machine polytime proofs for concrete reductions remain an open problem.

\textbf{What the author contributed:} The informal problem formulations (SUFFICIENCY-CHECK, MINIMUM-SUFFICIENT-SET, ANCHOR-SUFFICIENCY), the informal hardness conjectures, the informal tractability conditions, and the connection to over-modeling in engineering practice.

\textbf{What LLMs contributed:} \LaTeX{} drafting, Lean tactic exploration, reduction construction assistance, and prose refinement.

Reduction correctness is machine-checked; polynomial-time computability is a standard hypothesis. This methodology is disclosed for academic transparency.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection*{Acknowledgments}

This work depends on results whose authors made possible the insights that could then be machine-verified.

\textbf{Fundamental physics.} Albert Einstein established special relativity and the light-speed bound \cite{einstein1905electrodynamics}; Hermann Minkowski gave it the spacetime formulation \cite{minkowski1908space}. Max Planck established energy quantization and the constant $k_B$ \cite{planck1901distribution}. Rolf Landauer proved that erasure costs thermodynamic work \cite{landauer1961irreversibility}; Charles Bennett showed that reversible computation circumvents this bound \cite{bennett1982thermodynamics}. Bérut et al.\ experimentally verified Landauer's principle \cite{berut2012experimental}. The counting gap theorem (BA10) proves that bounded systems must obey finite acquisition budgets from pure mathematics; relativity then supplies the numerical propagation scale $c$ in our universe. The Wolpert-facing constrained-process interface now makes a sharper point explicit in the formal system: Landauer gives a universal floor, mismatch and residual dissipation are represented as separate theorem-level terms above that floor, the mismatch branch is derived from finite-distribution KL divergence as far as the present mathematics supports, and the finite discrete residual branch is derived by an exhaustive reverse-edge split: either asymmetric bidirectional edge flow yields positive two-point KL divergence, or vanishing reverse flow yields an irreversible one-way state transition with strictly positive Landauer-scaled cost. Any structural resource that is absorbed by mismatch further tightens the induced energy lower bound. Broader stopping-time / absolute-irreversibility residual regimes remain separate cited premises. The machine-checked part is the composition from that mixed theorem-plus-premise interface to the paper's decision-theoretic lower bounds.\LH{WC1}\LH{WC2}\LH{WC3}\LH{WC4}\LH{WC5}\LH{WM1}\LH{WM2}\LH{WM3}\LH{WM4}\LH{WM5}\LH{WM6}\LH{WR1}\LH{WR2}\LH{WR3}\LH{WR4}\LH{WR5}\LH{WR6}\LH{WR7}\LH{WR8}\LH{WP1}\LH{WP2}\LH{WP5}\LH{WP7}\LH{WP8}\LH{WP9}

\textbf{Proof infrastructure.} The Lean 4 proof assistant \cite{demoura2021lean4} served as the arbiter of correctness. Leonardo de Moura, Sebastian Ullrich, and the Lean 4 team built a tool that transforms mathematical claims into mechanically verified theorems. The Mathlib library \cite{mathlib2020} provided the formal infrastructure: order theory, linear algebra, complexity-theoretic definitions, and the lemmas that made the formalization possible.

\textbf{Methodological note.} Reduction correctness is machine-checked; polynomial-time computability of reduction functions is a standard claim stated as an explicit hypothesis. The human author contributed the informal problem formulations, informal hardness conjectures, informal tractability conditions, and the connection to over-modeling in engineering practice. The machine contributors (Claude, GPT-4, Lean 4) served as proof assistants, tactic explorers, and drafting partners. The final arbiter was the Lean compiler: if it accepts the proof, the theorem holds.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection*{Summary of Results}

This paper establishes a complete regime map for coordinate sufficiency and a static inner decomposition:

\begin{itemize}
\item \textbf{Regime map (core classes).} Static, stochastic, and sequential sufficiency-check questions are placed at \coNP, \PP, and \PSPACE respectively, with strict static \(\subset\) stochastic \(\subset\) sequential separation under declared assumptions (Theorems~\ref{thm:sufficiency-conp}, \ref{th:stochastic-sufficiency-pp}, \ref{th:sequential-sufficiency-pspace}; Propositions~\ref{prop:static-stochastic-strict}, \ref{prop:stochastic-sequential-strict}).
\item \textbf{Observer-collapse anchor closure.} Under declared observer-collapse hypotheses, stochastic/sequential anchor checks are forced by effective-state singleton structure and seeded support conditions (\LH{PA1}, \LH{PA4}, \LH{PA5}, \LH{PA6}, \LH{PA7}, \LH{PA9}).
\item \textbf{Static decision layer decomposition.} Within the static class, SUFFICIENCY-CHECK and MINIMUM-SUFFICIENT-SET are \coNP-complete, while ANCHOR-SUFFICIENCY is \(\SigmaP{2}\)-complete (Theorems~\ref{thm:sufficiency-conp}, \ref{thm:minsuff-conp}, \ref{thm:anchor-sigma2p}).
\item \textbf{Static access layer decomposition.} An encoding-regime separation contrasts explicit-state polynomial-time (polynomial in $|S|$) with succinct ETH worst-case lower bounds, and query-access obstruction cores refine this with value-entry/state-batch/oracle-lattice transfers (Theorem~\ref{thm:dichotomy}; Propositions~\ref{prop:query-regime-obstruction}--\ref{prop:oracle-lattice-strict}).
\item \textbf{Static structure layer decomposition.} Six structural subclasses provide certified polynomial recovery conditions inside the static class (Theorem~\ref{thm:tractable}).
\item Discrete-time interface semantics are formalized: time is $\mathbb{N}$-valued, decision events are exactly one-tick transitions, run length equals elapsed time, and substrate-consistent realizations preserve the same one-unit event law (Propositions~\ref{prop:time-discrete}--\ref{prop:substrate-unit-time})
\item Physical-system transport is formalized: encoded physical claims are derived from core rules through explicit assumption transfer, physical counterexamples map to encoded core failures, and the bundled bridge theorem composes law semantics, reduction equivalence, and required-work lower bounds (Proposition~\ref{prop:physical-claim-transport}, Corollary~\ref{cor:physical-counterexample-core-failure}, Theorem~\ref{thm:physical-bridge-bundle})
\item Exact represented-adjacent bridge boundary: transfer is licensed iff the class is one-step deterministic; represented horizon, stochastic-criterion, and transition-coupled classes have explicit bridge-failure witnesses (Theorem~\ref{thm:bridge-boundary-represented})
\item Exact-report admissibility is characterized by a certified-bit gap criterion; when exact reporting is blocked, exact accounting collapses to raw-only (Theorem~\ref{thm:exact-certified-gap-iff-admissible}, Corollary~\ref{cor:hardness-raw-only-exact})
\item The thermodynamic lift yields the Neukart--Vinokur duality specialization $dU \ge \lambda\,dC$ for $\lambda>0$ (Corollary~\ref{cor:neukart-vinokur}, Proposition~\ref{prop:thermo-lift})
\end{itemize}
\claimstamp{T\ref{thm:sufficiency-conp},\ref{thm:minsuff-conp},\ref{thm:anchor-sigma2p},\ref{th:stochastic-sufficiency-pp},\ref{th:sequential-sufficiency-pspace},\ref{thm:dichotomy},\ref{thm:tractable},\ref{thm:physical-bridge-bundle},\ref{thm:bridge-boundary-represented},\ref{thm:exact-certified-gap-iff-admissible};P\ref{prop:static-stochastic-strict},\ref{prop:stochastic-sequential-strict},\ref{prop:query-regime-obstruction},\ref{prop:query-value-entry-lb},\ref{prop:query-state-batch-lb},\ref{prop:query-subproblem-transfer},\ref{prop:oracle-lattice-transfer},\ref{prop:oracle-lattice-strict},\ref{prop:time-discrete},\ref{prop:decision-unit-time},\ref{prop:run-time-accounting},\ref{prop:substrate-unit-time},\ref{prop:physical-claim-transport},\ref{prop:thermo-lift},\ref{prop:raw-certified-bit-split};C\ref{cor:physical-counterexample-core-failure},\ref{cor:hardness-raw-only-exact},\ref{cor:neukart-vinokur}}{E,S,S+ETH,Qf,Qb,AR,RG,H=cost-growth}
\leanmeta{\LH{CCC6},\LH{CCC5},\LH{CCC1},\LH{DC9},\LH{DC10},\LH{DC1},\LH{DC2},\LH{CCC3},\LH{CCC7},\LH{CC61},\LH{CC29},\LH{CC28},\LH{CC43},\LH{CC52},\LH{CC27},\LH{DT6},\LH{DT13},\LH{DT16},\LH{DT24},\LH{CT6},\LH{CT8},\LH{CT4},\LH{CC67},\LH{CC69},\LH{CC85},\LH{CC18},\LH{CC11},\LH{CC21},\LH{CC77},\LH{CC76}.}

These results place decision-relevant coordinate identification into a regime-indexed class ladder, rather than a single-class statement: static (\coNP/\(\SigmaP{2}\) subproblems), stochastic (\PP), and sequential (\PSPACE), with explicit strict separations and bridge gates.

Beyond classification, the paper contributes a formal claim-typing framework (Section~\ref{sec:interpretive-foundations}): structural complexity is a property of the fixed decision question, while representational hardness is regime-conditional access cost. This is why encoding-regime changes can move practical hardness without changing the underlying semantics.
\claimstamp{P\ref{prop:sufficiency-char};T\ref{thm:typed-completeness-static}}{AR}
\leanmeta{\LH{CC62},\LH{CC76}.}

The reduction constructions and key equivalence theorems are machine-checked in Lean 4 (see Appendix~\ref{app:lean} for proof listings). The formalization verifies TAUTOLOGY and \(\exists\forall\)-SAT static reductions, theorem-level stochastic/sequential class placements, and strict regime-separation claims; complexity classifications (\coNP, \(\SigmaP{2}\), \PP, \PSPACE) follow by composition with standard complexity-theoretic results under the declared assumptions. The strengthened gadget showing that non-tautologies yield instances with \emph{all coordinates relevant} is also formalized.
\claimstamp{T\ref{thm:sufficiency-conp},\ref{thm:minsuff-conp},\ref{thm:anchor-sigma2p},\ref{th:stochastic-sufficiency-pp},\ref{th:sequential-sufficiency-pspace},\ref{thm:dichotomy};P\ref{prop:static-stochastic-strict},\ref{prop:stochastic-sequential-strict}}{S,S+ETH,RG}
\leanmeta{\LH{CCC6},\LH{CCC5},\LH{CCC1},\LH{DC9},\LH{DC10},\LH{CCC3},\LH{DC1},\LH{DC2}.}

\subsection*{Complexity Characterization}

The results provide precise complexity characterizations within the formal model:

\begin{enumerate}
\item \textbf{Exact bounds.} SUFFICIENCY-CHECK is \coNP-complete, both \coNP-hard and in \coNP.

\item \textbf{Constructive reductions.} The reductions from TAUTOLOGY and $\exists\forall$-SAT are explicit and machine-checked.

\item \textbf{Encoding-regime separation.} Under [E], SUFFICIENCY-CHECK is polynomial in $|S|$. Under [S+ETH], there exist succinct worst-case instances (with $k^*=n$) requiring $2^{\Omega(n)}$ time. Under [Q\_fin], the Opt-oracle core has $\Omega(|S|)$ worst-case query complexity (Boolean instantiation $\Omega(2^n)$), and under [Q\_bool] the value-entry/state-batch refinements preserve the obstruction with weighted-cost transfer closures.
\end{enumerate}
\claimstamp{T\ref{thm:sufficiency-conp},\ref{thm:dichotomy};P\ref{prop:query-regime-obstruction},\ref{prop:query-value-entry-lb},\ref{prop:query-state-batch-lb},\ref{prop:query-subproblem-transfer}}{E,S,S+ETH,Qf,Qb}
\leanmeta{\LH{CC61},\LH{CC16},\LH{CC50},\LH{CC29},\LH{CC28},\LH{CC27}.}

\subsection*{The Complexity Redistribution Corollary}

Section~\ref{sec:simplicity-tax} develops a quantitative consequence: when a problem requires $k$ dimensions and a model handles only $j < k$ natively, the remaining $k - j$ dimensions must be handled externally at each decision site. For $n$ sites, total external work is $(k-j) \times n$.\claimstamp{T\ref{thm:tax-grows}}{LC}
\leanmeta{\LH{HD26}.}

The set identity is elementary; its operational content comes from composition with the hardness results on exact relevance minimization. This redistribution corollary is formalized in Lean 4 (\texttt{HardnessDistribution.lean}), proving:
\begin{itemize}
\item Redistribution identity: complexity burden cannot be eliminated by omission, only moved between native handling and external handling
\item Dominance: complete models have lower total work than incomplete models
\item Amortization: there exists a threshold $n^*$ beyond which higher-dimensional models have lower total cost
\end{itemize}
\claimstamp{T\ref{thm:tax-conservation},\ref{thm:centralization-dominance},\ref{thm:amortization}}{LC}
\leanmeta{\LH{HD5},\LH{HD1},\LH{HD4},\LH{HD26},\LH{HD21}.}

\subsection*{Open Questions}

The landscape above is complete for the static sufficiency class under C1--C4; the items below are adjacent-class extensions or secondary refinements.
Several questions remain for future work:
\begin{itemize}
\item \textbf{Fixed-parameter tractability (primary):} Is SUFFICIENCY-CHECK FPT when parameterized by the minimal sufficient-set size $k^*$, or is it W[2]-hard under this parameterization?
\item \textbf{Adjacent-objective bridge frontier (beyond proven regime cores):} Characterize the exact frontier where adjacent sequential objectives reduce to the static class via Proposition~\ref{prop:one-step-bridge}, and where genuinely new complexity objects (e.g., horizon/sample/regret complexity) must replace the present static coNP/\(\Sigma_2^P\) layer.
\item \textbf{Average-case complexity:} What is the complexity under natural distributions on decision problems?
\item \textbf{Typed-confidence calibration:} For signaled reports $(r,g,p_{\mathrm{self}},p_{\mathrm{cert}})$, characterize proper scoring and calibration objectives that preserve signal consistency and typed admissibility.
\item \textbf{Learning cost formalization:} Can central cost $H_{\text{central}}$ be formalized as the rank of a concept matroid, making the amortization threshold precisely computable?
\end{itemize}

\subsection*{Practical Corollaries}

The practical corollaries are regime-indexed and theorem-indexed:
\begin{itemize}
\item \textbf{[E] and structured regimes:} polynomial-time exact procedures exist (Theorem~\ref{thm:tractable}).
\item \textbf{[Q\_fin]/[Q\_bool] query-access lower bounds:} worst-case Opt-oracle complexity is $\Omega(|S|)$ in the finite-state core (Boolean instantiation $\Omega(2^n)$), and value-entry/state-batch interfaces satisfy the same obstruction in the mechanized Boolean refinement (Propositions~\ref{prop:query-regime-obstruction}--\ref{prop:oracle-lattice-strict}), with randomized weighted robustness and oracle-lattice closures.
\item \textbf{[S+ETH] hard families:} exact minimization inherits exponential worst-case cost (Theorem~\ref{thm:dichotomy} together with Theorem~\ref{thm:sufficiency-conp}).
\item \textbf{[S\_bool] mechanized criterion:} minimization reduces to relevance-cardinality constraints (Corollary~\ref{cor:practice-diagnostic}).
\item \textbf{Redistribution consequences:} omitted native coverage externalizes work with explicit growth/amortization laws (Theorems~\ref{thm:tax-conservation}--\ref{thm:amortization}).
\item \textbf{Physical-system instantiation:} once a physical class is encoded into the declared decision interface, the same regime-typed hardness/tractability map and integrity constraints apply to that class (Section~\ref{sec:physical-transport}).\claimstamp{P\ref{prop:physical-claim-transport};C\ref{cor:physical-counterexample-core-failure};T\ref{thm:physical-bridge-bundle}}{AR}\leanmeta{\LH{CT6},\LH{CT8},\LH{CT4}.}
\end{itemize}

\paragraph{Physical Scope Guard.}
All transfer claims in this paper are typed through explicit physical encodings, declared observables, and declared assumption maps. Transfer is admitted only on declared physical instances satisfying the physical scope gate.\claimstamp{D\ref{def:physical-core-encoding},\ref{def:physical-scope-gate};P\ref{prop:typed-physical-transport-requirement},\ref{prop:physical-claim-transport}}{AR}

Hence the design choice is typed: enforce a tractable regime, or adopt weakened guarantees with explicit verification boundaries.\claimstamp{T\ref{thm:tractable},\ref{thm:dichotomy}}{RG}\leanmeta{\LH{CC71},\LH{CC16}.}
Equivalently, under the attempted-competence matrix, over-specification is rational only in attempted-competence-failure cells; once exact competence is available in the active regime (or no attempted exact competence was made), persistent over-specification is irrational for the same verified-cost objective.\claimstamp{P\ref{prop:attempted-competence-matrix}}{AR}\leanmeta{\LH{IC27}.}
By Proposition~\ref{prop:attempted-competence-matrix}, the integrity-preserving matrix contains 3 irrational cells and 1 rational cell.\claimstamp{P\ref{prop:attempted-competence-matrix}}{AR}\leanmeta{\LH{IC27}.}

\subsection*{Measurement-Constrained Reporting for Physical Solvers}

Within this paper's formalism, reporting correctness is a claim-discipline property: outputs are asserted only when certifiable, and otherwise emitted as abstain reports (optionally carrying tentative guesses and self-reflected confidence under zero certified confidence).
Formally, this is the integrity/competence split of Section~\ref{sec:interpretive-foundations}, i.e., an evidence-typing statement for reported claims under declared regimes.\claimstamp{D\ref{def:solver-integrity},\ref{def:competence-regime};P\ref{prop:integrity-competence-separation}}{TR}\leanmeta{\LH{IC18}.}

Formally: the abstention doctrine is a typed-report consequence of integrity/competence and evidence-gating predicates.

\paragraph{Truth-Objective Abstention Requirement.}
If a physical solver system declares a truth-preserving objective for claims about relation $\mathcal{R}$ in regime $\Gamma$, uncertified outputs are admissible only as $\mathsf{ABSTAIN}/\mathsf{UNKNOWN}$ rather than asserted answers.
Under the signal discipline, this includes abstain-with-guess/self-confidence reports while keeping certification evidence-gated.\claimstamp{D\ref{def:signaled-typed-report},\ref{def:signal-consistency};P\ref{prop:certified-confidence-gate},\ref{prop:abstain-guess-self-signal};C\ref{cor:exact-no-competence-zero-certified}}{AR}\leanmeta{\LH{IC46},\LH{IC38},\LH{IC41}.}
Equivalently, reporting protocols that suppress abstention in uncertified regions violate the typed claim discipline under the declared task model.\claimstamp{D\ref{def:solver-integrity};P\ref{prop:integrity-competence-separation},\ref{prop:integrity-resource-bound},\ref{prop:attempted-competence-matrix}}{AR}\leanmeta{\LH{IC18},\LH{CC30},\LH{IC27}.}

This yields an output-control doctrine for engineered solver systems under the task/regime contract:
\begin{enumerate}
\item Treat $\mathsf{ABSTAIN}/\mathsf{UNKNOWN}$ as first-class admissible outputs, not errors.\claimstamp{D\ref{def:solver-integrity}}{TR}
\item For signaled reports, maintain two channels: self-reflected confidence may be emitted, but certified confidence is positive only with evidence.\claimstamp{D\ref{def:signal-consistency};P\ref{prop:certified-confidence-gate},\ref{prop:self-confidence-not-certification}}{AR}\leanmeta{\LH{IC38},\LH{IC47}.}
\item Treat runtime disclosure as scalar-witnessed: always emit exact step-count, and emit completion percentage only when a declared bound exists.\claimstamp{P\ref{prop:steps-run-scalar},\ref{prop:no-fraction-without-bound},\ref{prop:fraction-defined-under-bound}}{AR}\leanmeta{\LH{IC42},\LH{IC35},\LH{IC44}.}
\item Optimize competent coverage subject to integrity constraints, rather than optimizing answer-rate alone.\claimstamp{D\ref{def:competence-regime};P\ref{prop:integrity-competence-separation}}{AR}\leanmeta{\LH{IC18}.}
\item In hardness-blocked regimes, abstention or explicit guarantee-weakening is structurally required unless regime assumptions are changed.\claimstamp{T\ref{thm:dichotomy};P\ref{prop:integrity-resource-bound},\ref{prop:attempted-competence-matrix}}{S,S+ETH}\leanmeta{\LH{CC16},\LH{CC30},\LH{IC27}.}
\end{enumerate}

Operationally, a reporting schema consistent with this framework tracks at least five quantities: integrity violations, competent non-abstaining coverage on the scoped family, abstention quality under regime shift, the calibration gap between self-reflected and certified confidence channels, and scalar runtime witness distributions (with bound-qualified fraction semantics when bounds are given). This is consistent with bounded-rationality feasibility discipline under bounded resource models \cite{sep_bounded_rationality,arora2009computational}.

Informally: if exact support is not certified, do not make exact claims; when competence is unavailable, abstain and keep confidence evidence-linked.


\appendix
