\section{Lean 4 Proof Listings}\label{app:lean}

The complete Lean 4 formalization is available in the companion artifact (Zenodo DOI listed on the title page). The mechanization consists of \LeanTotalLines\ lines across \LeanTotalFiles\ files, with \LeanTotalTheorems\ theorem/lemma statements.

\subsection{What Is Machine-Checked}

The Lean formalization establishes:

\begin{enumerate}
\item \textbf{Correctness of the TAUTOLOGY reduction:} The theorem \texttt{tautology\_iff\_sufficient} proves that the mapping from Boolean formulas to decision problems preserves the decision structure (accept iff the formula is a tautology).

\item \textbf{Decision problem definitions:} Formal definitions of sufficiency, optimality, and the decision quotient.

\item \textbf{Economic theorems:} Simplicity Tax redistribution identities and hardness distribution results.

\item \textbf{Query-access lower-bound core:} Formalized Boolean query-model indistinguishability theorem for the full problem via the $I=\emptyset$ subproblem (\texttt{emptySufficiency\_query\_indistinguishable\_pair}), plus obstruction-scale identities (\texttt{queryComplexityLowerBound}, \texttt{exponential\_query\_complexity}).
\end{enumerate}

\textbf{Complexity classifications} (coNP-completeness, $\Sigma_2^P$-completeness) follow by conditional composition with standard results (e.g., TAUTOLOGY coNP-completeness and $\exists\forall$-SAT $\Sigma_2^P$-completeness), represented explicitly as hypotheses in the conditional transfer theorems listed below. The Lean proofs verify the reduction constructions and the transfer closures under those hypotheses.
The assumptions themselves are unpacked by an explicit ledger projection theorem (\nolinkurl{ClaimClosure.standard_assumption_ledger_unpack}) so dependency tracking is machine-auditable.
\IfFileExists{content/assumption_ledger_auto.tex}{\input{content/assumption_ledger_auto.tex}}{}

\subsection{Polynomial-Time Reduction Definition}

We use Mathlib's Turing machine framework to define polynomial-time computability:

\begin{verbatim}
/-- Polynomial-time computable function using Turing machines -/
def PolyTime {α β : Type} (ea : FinEncoding α) (eb : FinEncoding β) 
    (f : α → β) : Prop :=
  Nonempty (Turing.TM2ComputableInPolyTime ea eb f)

/-- Polynomial-time many-one (Karp) reduction -/
def ManyOneReducesPoly {α β : Type} (ea : FinEncoding α) (eb : FinEncoding β)
    (A : Set α) (B : Set β) : Prop :=
  ∃ f : α → β, PolyTime ea eb f ∧ ∀ x, x ∈ A ↔ f x ∈ B
\end{verbatim}

This uses the standard definition: a reduction is polynomial-time computable via Turing machines and preserves membership.

\subsection{The Main Reduction Theorem}

\begin{theorem}[TAUTOLOGY Reduction Correctness, Lean]
The reduction from TAUTOLOGY to SUFFICIENCY-CHECK is correct:
\end{theorem}

\begin{verbatim}
theorem tautology_iff_sufficient (φ : Formula n) :
    φ.isTautology ↔ (reductionProblem φ).isSufficient Finset.empty
\end{verbatim}

This theorem is proven by showing both directions:
\begin{itemize}
\item If $\varphi$ is a tautology, then the empty coordinate set is sufficient
\item If the empty coordinate set is sufficient, then $\varphi$ is a tautology
\end{itemize}

The proof verifies that the utility construction in \texttt{reductionProblem} creates the appropriate decision structure where:
\begin{itemize}
\item At reference states, \texttt{accept} is optimal with utility 1
\item At assignment states, \texttt{accept} is optimal iff $\varphi(a) = \text{true}$
\end{itemize}

\subsection{Economic Results}

The hardness distribution theorems (Section~\ref{sec:simplicity-tax}) are fully formalized:

\begin{verbatim}
theorem simplicityTax_conservation (P : SpecificationProblem)
    (S : SolutionArchitecture P) :
    S.centralDOF + simplicityTax P S ≥ P.intrinsicDOF

theorem simplicityTax_grows (P : SpecificationProblem)
    (S : SolutionArchitecture P) (n₁ n₂ : ℕ)
    (hn : n₁ < n₂) (htax : simplicityTax P S > 0) :
    totalDOF S n₁ < totalDOF S n₂

theorem native_dominates_manual (P : SpecificationProblem) (n : Nat)
    (hn : n > P.intrinsicDOF) :
    totalDOF (nativeTypeSystem P) n < totalDOF (manualApproach P) n

theorem totalDOF_eventually_constant_iff_zero_distributed
    (S : SolutionArchitecture P) :
    IsEventuallyConstant (fun n => totalDOF S n) ↔ S.distributedDOF = 0

theorem no_positive_slope_linear_represents_saturating
    (c d K : ℕ) (hd : d > 0) :
    ¬ (∀ n, c + n * d = generalizedTotalDOF c (saturatingSiteCost K) n)
\end{verbatim}

\noindent\textbf{Identifier note.}
Lean identifiers retain internal naming (\texttt{intrinsicDOF}, \texttt{simplicityTax\_conservation}); in paper terminology these correspond to \emph{baseline hardness} and the \emph{redistribution lower-bound identity}, respectively.

\subsection{Complete Claim Coverage Matrix}

\begingroup
\scriptsize
\sloppy
\setlength{\tabcolsep}{4pt}
\begin{longtable}{@{}>{\raggedright\arraybackslash}p{0.20\linewidth}>{\raggedright\arraybackslash}p{0.11\linewidth}>{\raggedright\arraybackslash}p{0.42\linewidth}>{\raggedright\arraybackslash}p{0.23\linewidth}@{}}
\toprule
\textbf{Paper handle} & \textbf{Status} & \textbf{Lean support} & \textbf{Notes} \\
\midrule
\nolinkurl{cor:exact-identifiability} & Full &
\nolinkurl{Sigma2PHardness.exactlyIdentifiesRelevant_iff_sufficient_and_subset_relevantFinset} & Direct theorem mapping. \\
\nolinkurl{cor:gap-externalization} & Full &
\nolinkurl{HardnessDistribution.totalExternalWork_eq_n_mul_gapCard}, \nolinkurl{HardnessDistribution.simplicityTax_grows} & Composite from two mechanized theorems. \\
\nolinkurl{cor:gap-minimization-hard} & Full &
\nolinkurl{Sigma2PHardness.min_sufficient_set_iff_relevant_card} & Direct theorem mapping. \\
\nolinkurl{prop:representation-gap} & Full &
\nolinkurl{Sigma2PHardness.representationGap}, \nolinkurl{Sigma2PHardness.representationGap_eq_zero_iff}, \nolinkurl{Sigma2PHardness.representationGap_zero_iff_minimalSufficient}, \nolinkurl{Sigma2PHardness.representationGap_missing_eq_gapCard}, \nolinkurl{Sigma2PHardness.min_representationGap_zero_iff_relevant_card} &
Formal $\varepsilon$ object plus zero-gap, minimality, simplicity-tax component, and bounded-search collapse mapping. \\
\nolinkurl{cor:generalized-eventual-dominance} & Full &
\nolinkurl{HardnessDistribution.generalized_right_eventually_dominates_wrong} & Direct theorem mapping. \\
\nolinkurl{cor:linear-positive-no-saturation} & Full &
\nolinkurl{HardnessDistribution.no_positive_slope_linear_represents_saturating} & Direct theorem mapping. \\
\nolinkurl{cor:no-auto-minimize} & Full (conditional) &
\nolinkurl{ClaimClosure.no_auto_minimize_of_p_neq_conp} &
Mechanized conditional closure: non-collapse plus collapse-implication yields impossibility. \\
\nolinkurl{cor:overmodel-diagnostic-implication} & Full &
\nolinkurl{Sigma2PHardness.sufficient_iff_relevant_subset} & Contrapositive of mechanized theorem. \\
\nolinkurl{cor:practice-bounded} & Full &
\nolinkurl{sufficiency_poly_bounded_actions} & Direct theorem mapping. \\
\nolinkurl{cor:practice-diagnostic} & Full &
\nolinkurl{Sigma2PHardness.min_sufficient_set_iff_relevant_card} & Direct theorem mapping. \\
\nolinkurl{cor:practice-structured} & Full &
\nolinkurl{sufficiency_poly_separable} & Direct theorem mapping. \\
\nolinkurl{cor:practice-tree} & Full &
\nolinkurl{sufficiency_poly_tree_structured} & Direct theorem mapping. \\
\nolinkurl{cor:practice-unstructured} & Full &
\nolinkurl{all_coords_relevant_of_not_tautology} & Direct theorem mapping. \\
\nolinkurl{cor:right-wrong-hardness} & Full &
\nolinkurl{HardnessDistribution.right_dominates_wrong} & Direct theorem mapping. \\
\nolinkurl{cor:type-system-threshold} & Full &
\nolinkurl{HardnessDistribution.native_dominates_manual} & Direct theorem mapping. \\
\nolinkurl{prop:adq-ordering} & Full &
\nolinkurl{ClaimClosure.adq_ordering} &
Strict ordering theorem for shared nonempty behavior denominator. \\
\nolinkurl{prop:dominance-modes} & Full &
\nolinkurl{HardnessDistribution.right_dominates_wrong}, \nolinkurl{HardnessDistribution.centralized_higher_leverage}, \nolinkurl{HardnessDistribution.generalized_right_dominates_wrong_of_bounded_vs_identity_lower}, \nolinkurl{HardnessDistribution.generalized_dominance_can_fail_without_wrong_growth}, \nolinkurl{HardnessDistribution.generalized_dominance_can_fail_without_right_boundedness} & Compositional index proposition. \\
\nolinkurl{prop:generalized-assumption-boundary} & Full &
\nolinkurl{HardnessDistribution.generalized_dominance_can_fail_without_wrong_growth}, \nolinkurl{HardnessDistribution.generalized_dominance_can_fail_without_right_boundedness} & Direct pair mapping. \\
\nolinkurl{prop:hardness-conservation} & Full &
\nolinkurl{HardnessDistribution.totalDOF_ge_intrinsic} & Direct structural core theorem. \\
\nolinkurl{prop:hardness-efficiency-interpretation} & Full &
\nolinkurl{HardnessDistribution.hardnessEfficiency_eq_central_share} & Direct definitional equivalence. \\
\nolinkurl{prop:integrity-competence-separation} & Full &
\nolinkurl{IntegrityCompetence.competence_implies_integrity}, \nolinkurl{IntegrityCompetence.integrity_not_competent_of_nonempty_scope} & Direct pair mapping. \\
\nolinkurl{prop:bridge-failure-horizon} & Full &
\nolinkurl{ClaimClosure.horizon_gt_one_bridge_can_fail_on_sufficiency}, \nolinkurl{ClaimClosure.horizonTwoWitness_immediate_empty_sufficient} &
Mechanized counterexample: dropping one-step horizon can break bridge transfer on sufficiency claims. \\
\nolinkurl{prop:bridge-failure-stochastic} & Full &
\nolinkurl{ClaimClosure.stochastic_objective_bridge_can_fail_on_sufficiency} &
Mechanized counterexample: stochastic-criterion optimization can diverge from expected-utility static transfer. \\
\nolinkurl{prop:bridge-failure-transition} & Full &
\nolinkurl{ClaimClosure.transition_coupled_bridge_can_fail_on_sufficiency} &
Mechanized counterexample: transition-coupled objective can break static transfer. \\
\nolinkurl{thm:bridge-boundary-represented} & Full &
\nolinkurl{ClaimClosure.bridge_transfer_iff_one_step_class}, \nolinkurl{ClaimClosure.bridge_failure_witness_non_one_step}, \nolinkurl{ClaimClosure.bridge_boundary_represented_family} &
Represented-family boundary closure: transfer licensed iff one-step deterministic, with explicit non-one-step witnesses. \\
\nolinkurl{prop:minimal-relevant-equiv} & Full &
\nolinkurl{DecisionProblem.minimalSufficient_iff_relevant}, \nolinkurl{DecisionProblem.relevantSet_is_minimal} &
Product-space bridge: minimal sufficient sets coincide with the witness-defined relevance set. \\
\nolinkurl{prop:query-regime-obstruction} & Full &
\nolinkurl{ClaimClosure.query_obstruction_finite_state_core}, \nolinkurl{emptySufficiency_query_indistinguishable_pair_finite}, \nolinkurl{spikeFinite_empty_not_sufficient} &
Finite-state query-regime lower-bound core (arbitrary finite state type, $|S|\ge 2$) via $I=\emptyset$ indistinguishability. \\
\nolinkurl{cor:query-obstruction-bool} & Full &
\nolinkurl{ClaimClosure.query_obstruction_boolean_corollary}, \nolinkurl{emptySufficiency_query_indistinguishable_pair} &
Boolean-coordinate corollary of the finite-state core (recovers $\Omega(2^n)$ bound). \\
\nolinkurl{prop:query-value-entry-lb} & Full &
\nolinkurl{emptySufficiency_valueEntry_indistinguishable_pair}, \nolinkurl{valueEntryView_eq_if_hidden_untouched}, \nolinkurl{touchedStates_card_le_query_card} &
Mechanized Boolean value-entry query lower bound (full problem via $I=\emptyset$ indistinguishability). \\
\nolinkurl{prop:query-subproblem-transfer} & Full &
\nolinkurl{ClaimClosure.subproblem_hardness_lifts_to_full} &
Generic mechanized closure from subproblem hardness to full-problem hardness under restriction maps. \\
\nolinkurl{prop:query-randomized-robustness} & Full &
\nolinkurl{indistinguishable_pair_forces_one_error}, \nolinkurl{indistinguishable_pair_forces_one_error_per_seed}, \nolinkurl{decode_error_sum_two_labels} &
Seedwise mechanized randomized-robustness closure for indistinguishable yes/no pairs. \\
\nolinkurl{prop:query-randomized-weighted} & Full &
\nolinkurl{weighted_seed_error_identity}, \nolinkurl{weighted_seed_half_floor} &
Weighted finite-support randomized closure: aggregate error mass identity on the hard pair. \\
\nolinkurl{prop:query-state-batch-lb} & Full &
\nolinkurl{emptySufficiency_stateBatch_indistinguishable_pair}, \nolinkurl{stateBatchView_eq_if_hidden_untouched} &
Mechanized state-batch query lower bound via hidden-state indistinguishability. \\
\nolinkurl{prop:query-finite-state-generalization} & Full &
\nolinkurl{emptySufficiency_query_indistinguishable_pair_finite}, \nolinkurl{spikeFinite_empty_not_sufficient} &
Finite-state generalization of the empty-subproblem obstruction core (cardinality $\ge 2$). \\
\nolinkurl{prop:query-tightness-full-scan} & Full &
\nolinkurl{full_query_distinguishes_const_spike_finite} &
Mechanized adversary-family tightness witness under full-state scan. \\
\nolinkurl{prop:query-weighted-transfer} & Full &
\nolinkurl{weightedQueryCost_ge_min_mul_card}, \nolinkurl{weightedQueryCost_ge_min_mul_of_card_lb} &
Weighted-query transfer: cardinality lower bounds lift to weighted-cost lower bounds. \\
\nolinkurl{prop:oracle-lattice-transfer} & Full &
\nolinkurl{valueEntryView_eq_of_stateBatchView_eq_on_touched} &
Mechanized oracle-lattice transfer: state-batch agreement implies entry-view agreement on touched states. \\
\nolinkurl{prop:oracle-lattice-strict} & Full &
\nolinkurl{const_vs_scaled_opt_view_equal}, \nolinkurl{const_vs_scaled_value_entry_diff_at_true} &
Mechanized strictness witness: `Opt`-oracle view can hide value-level differences visible to value-entry queries. \\
\nolinkurl{prop:set-to-selector} & Full &
\nolinkurl{DecisionProblem.sufficient_implies_selectorSufficient} &
Set-valued sufficiency implies selector-level sufficiency for any deterministic selector. \\
\nolinkurl{prop:zero-epsilon-reduction} & Full &
\nolinkurl{DecisionProblem.epsOpt_zero_eq_opt}, \nolinkurl{DecisionProblem.sufficient_iff_zeroEpsilonSufficient} &
Mechanized zero-$\varepsilon$ reduction: exact sufficiency is exactly the $\varepsilon=0$ case. \\
\nolinkurl{prop:zero-epsilon-competence} & Full &
\nolinkurl{IntegrityCompetence.zero_epsilon_competence_iff_exact}, \nolinkurl{IntegrityCompetence.epsilon_competence_implies_integrity} &
Mechanized reduction from exact competence to the $\varepsilon=0$ competence instance, with integrity preservation for each $\varepsilon$-typed relation. \\
\nolinkurl{prop:selector-separation} & Full &
\nolinkurl{ClaimClosure.selectorSufficient_not_implies_setSufficient} &
Mechanized counterexample: selector-sufficiency does not imply set-valued sufficiency. \\
\nolinkurl{prop:attempted-competence-matrix} & Full &
\nolinkurl{IntegrityCompetence.overModelVerdict_rational_iff}, \nolinkurl{IntegrityCompetence.admissible_matrix_counts}, \nolinkurl{IntegrityCompetence.admissible_irrational_strictly_more_than_rational} &
Integrity-preserving matrix is mechanized: one rational admissible cell and three irrational admissible cells. \\
\nolinkurl{prop:bridge-transfer-scope} & Full &
\nolinkurl{ClaimClosure.one_step_bridge} &
Transfer rule follows from the one-step deterministic bridge equivalence. \\
\nolinkurl{prop:snapshot-process-typing} & Full &
\nolinkurl{ClaimClosure.agent_transfer_licensed_iff_snapshot}, \nolinkurl{ClaimClosure.process_bridge_failure_witness}, \nolinkurl{ClaimClosure.snapshot_vs_process_typed_boundary} &
Agent snapshot/process typing closure over represented adjacent classes: transfer license iff snapshot, with mechanized process-side bridge-failure witnesses. \\
\nolinkurl{prop:one-step-bridge} & Full &
\nolinkurl{ClaimClosure.one_step_bridge} &
One-step deterministic bridge is mechanized as an equivalence of sufficiency predicates. \\
\nolinkurl{prop:sufficiency-char} & Full &
\nolinkurl{ClaimClosure.sufficiency_iff_dq_ratio}, \nolinkurl{ClaimClosure.sufficiency_iff_projectedOptCover_eq_opt} &
Finite-model quotient characterization mechanized in both ratio form and projected-cover form. \\
\nolinkurl{thm:amortization} & Full &
\nolinkurl{HardnessDistribution.complete_model_dominates_after_threshold} & Direct theorem mapping. \\
\nolinkurl{thm:anchor-sigma2p} & Full (conditional) &
\nolinkurl{ClaimClosure.anchor_sigma2p_reduction_core}, \nolinkurl{ClaimClosure.anchor_sigma2p_complete_conditional}, \nolinkurl{anchor_sufficiency_sigma2p} &
Reduction core is mechanized; completeness lift is mechanized as a conditional transfer on standard source-class facts. \\
\nolinkurl{thm:centralization-dominance} & Full &
\nolinkurl{HardnessDistribution.centralization_dominance_bundle}, \nolinkurl{HardnessDistribution.centralization_step_saves_n_minus_one} & Direct pair mapping. \\
\nolinkurl{thm:config-reduction} & Full &
\nolinkurl{ConfigReduction.config_sufficiency_iff_behavior_preserving} & Direct theorem mapping. \\
\nolinkurl{thm:cost-asymmetry-eth} & Full (conditional) &
\nolinkurl{ClaimClosure.cost_asymmetry_eth_conditional}, \nolinkurl{HardnessDistribution.linear_lt_exponential_plus_constant_eventually} &
Asymptotic dominance and ETH-conditioned lift are both mechanized. \\
\nolinkurl{thm:regime-coverage} & Full &
\nolinkurl{ClaimClosure.declaredRegimeFamily_complete}, \nolinkurl{ClaimClosure.regime_core_claim_proved} &
Declared static-class regime family is finite and exhaustive, with one mechanized core claim per regime constructor. \\
\nolinkurl{thm:typed-completeness-static} & Full (conditional) &
\nolinkurl{ClaimClosure.typed_static_class_completeness} &
Typed static-class completeness package: class-label closures + dichotomy closure + regime coverage + family exhaustiveness. \\
\nolinkurl{thm:dichotomy} & Full (conditional) &
\nolinkurl{ClaimClosure.dichotomy_conditional}, \nolinkurl{ClaimClosure.hard_family_all_coords_core}, \nolinkurl{ClaimClosure.explicit_state_upper_core} &
Upper-branch core and hard-family core are mechanized; ETH lower branch is mechanized as conditional transfer. \\
\nolinkurl{thm:generalized-dominance} & Full &
\nolinkurl{HardnessDistribution.generalized_right_dominates_wrong_of_bounded_vs_identity_lower} & Direct theorem mapping. \\
\nolinkurl{thm:generalized-saturation-possible} & Full &
\nolinkurl{HardnessDistribution.saturatingSiteCost_eventually_constant}, \nolinkurl{HardnessDistribution.generalizedTotal_with_saturation_eventually_constant} & Direct construction mapping. \\
\nolinkurl{thm:linear-saturation-iff-zero} & Full &
\nolinkurl{HardnessDistribution.totalDOF_eventually_constant_iff_zero_distributed} & Direct theorem mapping. \\
\nolinkurl{thm:minsuff-collapse} & Full (conditional) &
\nolinkurl{ClaimClosure.minsuff_collapse_core}, \nolinkurl{ClaimClosure.minsuff_collapse_to_conp_conditional}, \nolinkurl{Sigma2PHardness.min_sufficient_set_iff_relevant_card} &
Quantifier collapse core and coNP-reading lift are mechanized. \\
\nolinkurl{thm:minsuff-conp} & Full (conditional) &
\nolinkurl{ClaimClosure.minsuff_collapse_core}, \nolinkurl{ClaimClosure.minsuff_conp_complete_conditional} &
coNP-completeness transfer is mechanized conditionally from the collapse core. \\
\nolinkurl{thm:overmodel-diagnostic} & Full &
\nolinkurl{ClaimClosure.no_exact_identifier_implies_not_boundary_characterized}, \nolinkurl{ClaimClosure.boundaryCharacterized_iff_exists_sufficient_subset}, \nolinkurl{ConfigReduction.config_sufficiency_iff_behavior_preserving} &
Contrapositive diagnostic step mechanized under explicit boundary-characterization definition. \\
\nolinkurl{thm:sufficiency-conp} & Full (conditional) &
\nolinkurl{ClaimClosure.sufficiency_conp_reduction_core}, \nolinkurl{ClaimClosure.sufficiency_conp_complete_conditional}, \nolinkurl{tautology_iff_sufficient} &
Reduction core and coNP-completeness transfer are mechanized conditionally. \\
\nolinkurl{thm:tax-conservation} & Full &
\nolinkurl{HardnessDistribution.gap_conservation_card} & Direct theorem mapping. \\
\nolinkurl{thm:tax-grows} & Full &
\nolinkurl{HardnessDistribution.totalExternalWork_eq_n_mul_gapCard} & Direct theorem mapping. \\
\nolinkurl{thm:tractable} & Full (conditional) &
\nolinkurl{ClaimClosure.tractable_bounded_core}, \nolinkurl{ClaimClosure.tractable_separable_core}, \nolinkurl{ClaimClosure.tractable_tree_core}, \nolinkurl{ClaimClosure.tractable_subcases_conditional} &
All three tractable branch cores and assembly closure are mechanized. \\
\bottomrule
\end{longtable}
\endgroup

\subsection{Claims Not Fully Mechanized}

\textbf{Status:} all theorem/proposition/corollary handles in this paper now have Lean backing.
Entries marked \textbf{Full (conditional)} are explicitly mechanized transfer theorems that depend on standard external complexity facts (e.g., source-class completeness or ETH assumptions), with those dependencies represented as hypotheses in Lean.

\subsection{Module Structure}

\begin{itemize}
\item \texttt{Basic.lean} -- Core definitions (DecisionProblem, sufficiency, optimality)
\item \texttt{Sufficiency.lean} -- Sufficiency checking algorithms and properties
\item \texttt{Reduction.lean} -- TAUTOLOGY reduction construction and correctness
\item \texttt{Hardness/ConfigReduction.lean} -- Sentinel-action configuration reduction theorem
\item \texttt{Complexity.lean} -- Polynomial-time reduction definitions using mathlib
\item \texttt{HardnessDistribution.lean} -- Simplicity Tax redistribution and amortization theorems
\item \texttt{IntegrityCompetence.lean} -- Solver integrity vs regime competence separation
\item \texttt{ClaimClosure.lean} -- Mechanized closure of paper-level bridge and diagnostic claims
\item \texttt{Tractability/} -- Bounded actions, separable utilities, tree structure
\end{itemize}

\subsection{Verification}

The proofs compile with Lean 4 and contain no \texttt{sorry} placeholders. Run \texttt{lake build} in the proof directory to verify.
