We study physical information sufficiency as a decision-theoretic meta-problem. Given a decision problem $\mathcal{D}=(A,S,U)$ with factored state space $S=X_1\times\cdots\times X_n$, does a coordinate set $I$ determine the optimal action? Formally: $s_I=s'_I\implies\Opt(s)=\Opt(s')$. This asks whether projected information preserves optimal-action correspondence.

The paper proves static/stochastic/sequential complexity placement and strict regime separation (\LH{CC61}, \LH{DC9}, \LH{DC10}, \LH{DC1}, \LH{DC2}), static-class coNP/\(\Sigma_2^P\) placement with encoding/query separation (\LH{CC36}, \LH{CC5}, \LH{CC16}, \LH{CC50}), theorem-level physical transport plus an exact represented-adjacent bridge boundary (\LH{CT4}, \LH{CC10}, \LH{CC9}, \LH{CC8}), static-class completeness boundaries (\LH{CC14}, \LH{CC76}), Bayesian uniqueness from DQ/thermodynamic admissibility axioms (\LH{FN14}), layered graph nontriviality/belief/physical-lift composition (\LH{GN2}, \LH{GN5}, \LH{GN7}, \LH{GN8}, \LH{GN9}, \LH{GN10}, \LH{GN11}, \LH{GN12}, \LH{GN13}), conditional physical-collapse no-go transfer and canonical \(P{=}NP\) physical-impossibility specialization under declared collapse-map assumptions (\LH{PH11}, \LH{PH12}, \LH{PH13}, \LH{PH14}, \LH{PH15}), derived SAT-bridge collapse map with reduction-level energy transfer (\LH{PH16}, \LH{PH17}, \LH{PH18}, \LH{PH19}, \LH{PH20}, \LH{PH21}, \LH{PH22}), sharp assumption-failure disjunction and regime-robust specializations (\LH{PH26}, \LH{PH27}, \LH{PH28}, \LH{PH29}, \LH{PH30}), conversation-layer shared-channel clamp semantics with typed YES/NO/ABSTAIN projection (\LH{CV4}, \LH{CV7}, \LH{CV8}, \LH{CV9}, \LH{CV12}, \LH{CV13}, \LH{CV16}), and mechanized hardness-placement laws with explicit dominance and amortization thresholds (\LH{HD1}, \LH{HD4}, \LH{HD3}).\claimstamp{T\ref{thm:sufficiency-conp},\ref{thm:minsuff-conp},\ref{thm:anchor-sigma2p},\ref{th:stochastic-sufficiency-pp},\ref{th:sequential-sufficiency-pspace},\ref{thm:dichotomy},\ref{thm:typed-completeness-static},\ref{thm:physical-bridge-bundle},\ref{thm:bridge-boundary-represented},\ref{thm:bayes-from-dq},\ref{thm:centralization-dominance},\ref{thm:tax-conservation},\ref{thm:amortization};P\ref{prop:query-regime-obstruction},\ref{prop:static-stochastic-strict},\ref{prop:stochastic-sequential-strict}}{S,S+ETH,Qf,Qb,AR,LC}

\textbf{Complexity landscape.} Static: SUFFICIENCY-CHECK is \coNP-complete; MINIMUM-SUFFICIENT-SET is \coNP-complete; ANCHOR-SUFFICIENCY is $\Sigma_2^P$-complete. Stochastic: STOCHASTIC-SUFFICIENCY-CHECK is PP-complete. Sequential: SEQUENTIAL-SUFFICIENCY-CHECK is PSPACE-complete. Under standard assumptions, static $\subset$ stochastic $\subset$ sequential is strict. Under explicit-state encoding, polynomial-time algorithms hold for bounded actions, separable utility, low tensor rank, tree structure, bounded treewidth, and coordinate symmetry. Under succinct encoding with ETH, worst-case requires $2^{\Omega(n)}$ time. Under query access, $\Omega(|S|)$ lower bound.

\textbf{Bayesian reasoning.} Two mechanized paths coincide: counting $\to$ conditional probability $\to$ decision-quotient certainty, and thermodynamic admissibility (no free information, energy cost $\ge kT\ln2$ per bit, integrity preservation) $\to$ unique update rule. Independently, provable uncertainty with forced action implies nondegenerate belief, and under a strictly informative evidence channel a Bayes update exists \LH{BF2}, \LH{BF3}, \LH{BF4}. The observer-path bridge also proves forced-belief and Bayes-update existence \LH{GN8}, \LH{GN9}. Deadline-forced action yields positive decision-bit and thermodynamic lower bounds \LH{CF5}, \LH{CF8}, \LH{CF9}, and with uncertainty gives nondegenerate belief \LH{CF7}. Lean proves crossEntropy = entropy + KL \LH{FN12}, KL >= 0 (Gibbs) \LH{FN7}, entropy <= crossEntropy \LH{FN8}, Bayes optimality \LH{FN14}.

\textbf{Physical grounding.} A physical-to-core encoding $E:\mathcal{P}\to\mathcal{D}$ transfers assumptions and lifts core claims to encoded physical instances; discrete-time event semantics and bit-to-energy/carbon lifts are formalized. Logic/time scaffolds alone do not determine unique physical law/objective under separating parameterizations \LH{CF2}, \LH{CF4}. A coupled recurrent conversation layer fixes shared-node updates, clamp-observed state, one-bit decision events, and positive Landauer lower bound under positive per-bit conversion \LH{CV4}, \LH{CV5}, \LH{CV7}, \LH{CV8}, \LH{CV9}. Typed report projection preserves YES/NO as exact-claim channel and ABSTAIN as abstain-claim channel \LH{CV11}, \LH{CV12}, \LH{CV13}, \LH{CV15}, \LH{CV16}. The represented-adjacent boundary is exact: transfer is licensed iff the class is one-step deterministic, with explicit witnesses for horizon, stochastic-criterion, and transition-coupled failures (Theorem~\ref{thm:bridge-boundary-represented}). The thermodynamic lift yields the Neukart--Vinokur specialization $dU \ge \lambda\,dC$ for $\lambda>0$ (Corollary~\ref{cor:neukart-vinokur}, Proposition~\ref{prop:thermo-lift}). Conditional physical-collapse no-go closure is explicit at requirement and canonical levels \LH{PH12}, \LH{PH13}, \LH{PH14}, \LH{PH15}; with bounded budget, positive per-bit cost, and exponential required-op lower bound, collapse implies assumption failure (unbounded budget or zero bit-cost or failed exponential lower bound) \LH{PH26}, \LH{PH27}. The SAT-side map is derived through an explicit SAT witness plus reduction bridge and runtime-to-energy transfer \LH{PH16}, \LH{PH17}, \LH{PH18}, \LH{PH19}, \LH{PH20}, \LH{PH21}, \LH{PH22}. Countermodels isolate necessity of finite-budget, positive bit-cost, exponential lower bounds, and map premises \LH{PH23}, \LH{PH24}, \LH{PH25}, \LH{PH31}, \LH{PH32}, \LH{PH33}. Deterministic/probabilistic/sequential regime specializations preserve the same no-collapse form \LH{PH28}, \LH{PH29}, \LH{PH30}. Cycle/material witnesses give positive Landauer/NV floors and room-temperature stability witnesses \LH{GN10}, \LH{GN11}, \LH{GN12}, \LH{GN13}. Atoms with orbital transitions are agents; ground-state atoms are passive.\claimstamp{P\ref{prop:physical-claim-transport},\ref{prop:thermo-lift};C\ref{cor:neukart-vinokur};T\ref{thm:physical-bridge-bundle},\ref{thm:bridge-boundary-represented}}{AR,H=cost-growth}\leanmeta{\LH{CT6},\LH{CT4},\LH{CC10},\LH{CC9},\LH{CC8},\LH{CC64},\LH{CC65},\LH{CF2},\LH{CF4},\LH{CV4},\LH{CV7},\LH{CV9},\LH{CV13},\LH{CV16},\LH{PH12},\LH{PH13},\LH{PH14},\LH{PH15},\LH{PH16},\LH{PH18},\LH{PH20},\LH{PH22},\LH{PH26},\LH{PH27},\LH{PH28},\LH{PH29},\LH{PH30},\LH{PH31},\LH{PH32},\LH{PH33},\LH{GN10},\LH{GN11},\LH{GN12},\LH{GN13}.}

\textbf{Machine-checked.} Lean 4 (21045 lines, 901 theorems, 0 sorries). All theorem-level claims are regime-typed with explicit encoding assumptions.
