\section{Conclusion}\label{conclusion}

\subsection{Methodology and Disclosure}

\textbf{Role of LLMs in this work.} This paper was developed through
human-AI collaboration. The author provided the core insight---that
leverage ($L = \text{Capabilities}/\text{DOF}$) unifies architectural
decision-making---while large language models (Claude, GPT-4) served as
implementation partners for formalization, proof drafting, and LaTeX
generation.

The Lean 4 proofs (\LeanTotalLines{} lines, \LeanTotalSorry{} \texttt{sorry} placeholders) were iteratively
developed: the author specified theorems, the LLM proposed proof
strategies, and the Lean compiler verified correctness. This
methodology is epistemically sound: machine-checked proofs are correct
regardless of generation method.

\textbf{What the author contributed:} The leverage framework itself,
the metatheorem that SSOT and nominal typing are instances of leverage
maximization, the connection to error probability, the case study
selection from OpenHCS, and the weighted leverage extension.

\textbf{What LLMs contributed:} LaTeX drafting, Lean tactic suggestions,
prose refinement, and exploration of proof strategies.

This disclosure reflects our commitment to transparency. The contribution
is the unifying insight; the proofs stand on their machine-checked
validity.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Summary}

We provided the first formal connection between software architecture and reliability theory. Key results:

\textbf{1. DOF-Reliability Isomorphism (Theorem \ref{thm:dof-reliability}):} An architecture with $n$ DOF is isomorphic to a series system with $n$ components. Each DOF is a failure point. This is the core theoretical contribution---it grounds architectural decisions in reliability-theoretic foundations.

\textbf{2. Leverage Gap Theorem (Theorem \ref{thm:leverage-gap}):} For equal-capability architectures, modification cost ratio equals DOF ratio. This is a \emph{testable prediction}: $k\times$ leverage yields $k\times$ fewer modifications.

\textbf{3. Ordering Equivalence (Theorem \ref{thm:approx-bound}):} The linear expected-error model and exact series model induce the same architecture ordering in the discrete formalization.

\textbf{4. Leverage-Error Tradeoff (Theorem \ref{thm:leverage-error}):} Maximizing leverage minimizes error probability.

\textbf{5. Physical Edit-Energy Floor (Theorem \ref{thm:physical-energy-floor}, Corollary \ref{cor:leverage-energy}):} Under one explicit physical calibration constant, minimum edit-energy is proportional to DOF, and higher leverage in a fixed capability class yields strictly lower energy floor.

\textbf{6. Budget Boundary and Necessity (Theorem \ref{thm:physical-budget-boundary}, Corollary \ref{cor:physical-assumption-necessity}):} infeasibility is exactly budget-below-floor, and necessity witnesses isolate the two physical premises used in this layer.

\textbf{7. Unifying Framework:} SSOT and nominal typing are instances of leverage maximization.

\textbf{8. Practical Demonstration:} Before/after examples from OpenHCS demonstrating DOF collapse. PR \#44 provides a publicly verifiable instance.

\subsection{Decision Procedure}

Given requirements $R$, choose optimal architecture via:

\begin{enumerate}
\item \textbf{Enumerate:} List candidate architectures $\{A_1, \ldots, A_n\}$
\item \textbf{Filter:} Keep only $A_i$ with $\text{Cap}(A_i) \supseteq R$ (feasible architectures)
\item \textbf{Compute:} Calculate $L(A_i) = |\text{Cap}(A_i)|/\text{DOF}(A_i)$ for each
\item \textbf{Optimize:} Choose $A^* = \arg\max_i L(A_i)$
\end{enumerate}

\textbf{Justification:} By Theorem 3.4, this minimizes error probability among feasible architectures.

\subsection{Limitations}

\textbf{1. Independence Assumption:} The mechanized probability model treats DOF-level errors as independent under the orthogonality assumptions. Real systems may include additional correlated failure modes.

\textbf{2. Constant Error Rate:} Assumes $p$ is constant across components. Reality: some components are more error-prone than others.

\textbf{3. Single Codebase Examples:} Demonstrations drawn from OpenHCS. The mechanism is general; specific patterns may vary across domains.

\textbf{4. Capability Quantification:} We count capabilities qualitatively (unweighted). Some capabilities may be more valuable than others.

\textbf{5. Static Analysis:} Framework evaluates architecture statically. Dynamic factors (runtime performance, scalability) require separate analysis.

\subsection{Future Work}

\textbf{1. Weighted Capabilities:} Extend framework to assign weights to capabilities based on business value: $L = \sum w_i c_i / \text{DOF}$.

\textbf{2. Correlated Errors:} Relax independence assumption. Model error correlation via covariance matrix.

\textbf{3. Multi-Objective Optimization:} Combine leverage with performance, security, and other quality attributes. Pareto frontier analysis.

\textbf{4. Tool Support:} Develop automated leverage calculator. Static analysis to compute DOF, capability inference from specifications.

\textbf{5. Language Extensions:} Design languages that make high-leverage patterns easier (e.g., first-class support for SSOT).

\textbf{6. Broader Validation:} Replicate case studies across diverse domains (web services, embedded systems, data pipelines).

\subsection{Impact}

This work provides:

\textbf{For Practitioners:} Principled method for architectural decisions. When choosing between alternatives, compute leverage and select maximum (subject to requirements).

\textbf{For Researchers:} Unifying framework connecting SSOT, nominal typing, microservices, API design, configuration, and database normalization. This integrates cleanly with Paper 4's decision-sufficiency viewpoint: Paper 4 characterizes when information is sufficient for an optimal decision, while Paper 3 provides a computable architectural objective within that sufficiency envelope.

\textbf{For Educators:} Formal foundation for teaching software architecture. Explains \emph{why} design patterns work (leverage maximization).

\subsection{Final Remarks}

Software architecture has long relied on heuristics and experience. This paper provides formal foundations: \emph{architectural quality is fundamentally about leverage}. By maximizing capabilities per degree of freedom, we minimize error probability and modification cost.

The framework unifies diverse prior results (SSOT, nominal typing) and applies to new domains (microservices, APIs, configuration, databases).

We invite the community to apply the leverage framework to additional domains, develop tool support, and extend the theory to weighted capabilities and correlated errors.


\appendix
