Five independent scientific frameworks---engineering optimization, epistemic coherence, information geometry, computational complexity, and statistical physics---each characterize the same architectural property: having exactly one degree of freedom.

\textbf{Central Result (Five-Way Equivalence, Theorem~\ref{thm:five-way}).} For any architecture $A$ with $\mathrm{Cap}(A) > 0$:
\[
\mathrm{DOF}(A) = 1
\;\iff\; \text{max leverage}
\;\iff\; \text{SSOT}
\;\iff\; \mathrm{srank} = 1
\;\iff\; \text{tractable sufficiency}
\;\iff\; \text{zero thermodynamic cost}.
\]
This convergence is not a coincidence. The five frameworks are logically independent---each was developed for a distinct purpose---yet all select the single-source condition as their ground state. All equivalences are machine-checked in Lean 4 via live cross-paper imports (Paper~3 $\to$ Paper~4 $\to$ Mathlib).

Building on axis orthogonality (Paper~1) and Single Source of Truth (Paper~2), the engineering consequences follow: define leverage
\[L(A) = \frac{|\mathrm{Capabilities}(A)|}{\mathrm{DOF}(A)}.\]
Maximizing $L$ subject to feasibility minimizes expected error probability (DOF--Reliability Isomorphism), minimizes modification cost (Leverage Gap), and minimizes physical edit-energy under Landauer calibration with explicit per-edit constant $j_{\mathcal{M}} > 0$. These are \emph{corollaries} of the convergence theorem, not independent contributions.

\textbf{Open Conjectures.} The thermodynamic selection theorem currently assumes P~$\neq$~coNP. We conjecture that thermodynamic cost scales directly with structural rank \emph{unconditionally}---removing the complexity-theoretic hypothesis would reduce the five-way equivalence to a purely physical statement. A second open problem connects DOF~$=1$ architectures to England's replication inequality: single-source architectures that generate derived instances may satisfy a formal entropy-production bound that multi-DOF architectures cannot match.

All core theorems are machine-checked in Lean~4. An assumption ledger and proof listing accompany the paper.

Keywords: software architecture, degrees of freedom, five-way equivalence, Landauer principle, structural rank, formal methods
